### 2020-01-09 manuscript.Rmd
Version send to co-authors

---
title: "Automated filtering of species occurrence data can identify problems, but needs to be customized"
author: 
- Alexander Zizka^1,2^,
- Alice Calvente^3^,
- Fernanda Antunes Carvalho^4,5^,
- Ana Paula Alves da Silva,
- Andressa Cabral,
- Arthur de Souza Soares,
- Eduardo Calisto Tomaz,
- Fernanda Gondim Lambert Moreira,
- Filipe Serrano,
- Jéssica Fernanda Ramos Coelho,
- Mabel Rocio Baez-Lizarazo,
- Mariana Ramos Fantinati,
- Matheus Colli-Silva,
- Moabe Ferreira Fernandes,
- Nathália Michelly da Cunha Santos,
- Paolla G. C. Souza,
- Renata Clícia Santos da Costa,
- Thais Ferreira Pinto de Araújo,
- Tiago Andrade Borges Santos,
- Tiago Luiz Vieira,
- Valéria Fonseca Vale,
- Alexandre Antonelli^6^
#bibliography: course_natal.bib
csl: jbi-3.csl
output:
  word_document:
  #bookdown::pdf_document2:
    toc: false
    toc_depth: 1
    fig_caption: true
   # number_sections: false
   # keep_tex: yes
header-includes: |
    \usepackage{caption}
    \usepackage{fancyhdr}
    \usepackage{booktabs}
    \pagestyle{fancy}
    \fancyhead[L]{Automated filter effect}
    \fancyhead[LE,RO]{\thepage}
    \fancyfoot{}
    \usepackage{float}
    \floatplacement{figure}{H}
    \usepackage{lineno}
    \linenumbers
linestretch: 1.3
fontsize: 12pt
always_allow_html: yes
---

1. sDiv,German Center for Integrative Biodiversity Research (iDiv), Leipzig, Germany
2. Naturalis Biodiversity Center, Leiden, The Netherlands
3. Laboratório de Botânica Sistemática, Departamento de Botânica e Zoologia, Centro de Biociências, Universidade Federal do Rio Grande do Norte.
4. Departamento de Genética, Ecologia e Evolução, Instituto de Ciências Biológicas, Universidade Federal de Minas Gerais, Av. Antônio Carlos 6627 - Pampulha, 31270-901 - Belo Horizonte - MG - Brasil

\newpage{}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
library(knitr)
library(kableExtra)
library(readxl)
library(dplyr)
library(readr)
library(tidyr)

knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark=",")
})
```

```{r}
# load and prepare data
dat <- read_csv("output/all_records.csv") %>% 
  filter(!is.na(decimalLongitude)) %>% 
  filter(!is.na(decimalLatitude))

dat_cl <-  dat %>% 
  filter(summary)
```

# Abstract
Automatic flagging and subsequent filtering are a scalable and reproducible means to identify potentially problematic records in 'big data' datasets of species geographic occurrences from public databases (e.g., www.gbif.org). However, it is unclear how big the data loss by filtering is, if similar test should be applied across taxonomic groups, and what the effect of filtering is for downstream analyses. Here, we obtained occurrence records for 16 taxa including plants and animals as well as terrestrial and marine groups from www.gbif.org and documented the effect of 13 recently proposed filters in each group. We evaluated the effect of filtering on the inference of species richness patterns and the accuracy of automated conservation assessments as examples for downstream analyses. We found 29-67% of the records potentially problematic across taxonomic groups. The tests for duplicated information, collection year, basis of record as well as urban areas and political centroids were most important. While many of the flagged records might not be erroneous per se, they might be overly imprecise, increase uncertainty of downstream analyses and generate unnecessary computational burden. Overall, our results stress the importance of (meta-)data exploration for biogeographic research and suggest that automated flagging can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the taxonomic group and geographic area of interest.

# Keywords
Automated cleaning, automated conservation assessment, data quality, GBIF, Neotropics

\newpage{}

# Introduction
Publicly available species distribution data have become a crucial resource in biogeographic research. In particular, the availability of digitized collections from museums and herbaria, and citizen science observations has increased in the last years. As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.3 billion geo-referenced occurrence records for species from across the globe and the tree of life.

A central challenge to the use of these publicly available species occurrence data in research are erroneous geographic coordinates [@Anderson2016]. Errors mostly arise, because public databases integrate records collected with different methodology in different places at different times, often with little centralized curation and meta-data. For instance, erroneous coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common [@Maldonado2015; @Yesson2007] and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records from species in captivity [@Zizka2019].

Manual data cleaning based on expert knowledge can detect these issues, but is only feasible on small taxonomic or geographic scales and is difficult to reproduce. Automated flagging methods to identify potentially problematic records have been proposed as a scalable alternative for datasets with millions of records from many different taxa. Those methods are usually based on geographic gazetteers [e.g., @Chamberlain2016; @Zizka2019; @Jin2020] or on additional data, such as environmental variables [@Robertson2016]. Additionally, filtering procedures based on record meta-data, for instance collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biogeographic research [@Zizka2019]. While these methods can identify common errors, it is unclear if errors are equally common across taxonomic groups, and which errors are most severe for downstream analyses.

Erroneous records are especially problematic in conservation, where stakes are high. Recently, automated conservation assessment methods have been developed to use publicly available occurrence records for conservation, in particular to support the assessment procedures for the global Red List of the International Union for the Conservation of Nature (IUCN) [@Dauby2017; @Bachman2011; @Pelletier2018]. These methods approximate species' range size, namely the Extent of Occurrence (EOO, which is the area of a convex hull polygon comprising all records of a species), the Area of Occupancy (AOO, which is the sum of the area actually occupied by a species, calculated based on a small-scale regular grid), and the number of locations for a preliminary conservation assessment following IUCN Criterion B. Automated assessments using these methods have been used to speed up global [@Stevart2019; @Zizka2019a] and regional [@Schmidt2017; @Cosiaux2018] Red List assessments. Since especially EOO is sensitive to individual records with erroneous coordinates automated conservation assessments might be prone to bias by erroneous records, especially, if the number of records is low, as is the case for many species from the tropical regions of Earth.

Here, we test the effect of automated filtering of species geographic occurrence records on the number of records available in different plant and animal groups. Furthermore, we test the impact of automated filtering procedures on the accuracy of automated conservation assessments compared to full IUCN assessments. Specifically, we test a pipeline of 13 tests to automatically flag possibly erroneous records 1) using record meta-data and 2) using geographic gazetteers. We address thee questions:

1. Which is the most relevant filter, i.e. which type of error is most common?
2. Does the importance of individual filters differ among taxonomic groups?
3. Does automated filtering improve the accuracy of automated conservation assessment compared to the global IUCN Red List?

# Material and Methods
We did all analysis in the R computing environment [@rcoreteam2019], using standard libraries for data handling and visualization [@Wickham2018; @Garnier2018; @Ooms2014; @Ooms2019; @Hijmans2019]. All analysis scripts are available from https://github.com/idiv-biodiversity/effects_of_automated_cleaning.

## Choice of study taxa
This study is the outcome of a workshop at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering taxonomists and biogeographers working with different taxonomic groups of plants and animals across the Neotropics. Each participant analysed geographic occurrence data from their taxonomic group of interest and provided an assessment of the results for this group. Hence, we chose the study groups based on the expertise of the participants which enabled us to compare the results across the tree of life (at the coast of an arbitrary choice of taxa and taxonomic ranks). We obtained public occurrence records for 16 taxa, including one plant family, eight plant genera, one genus of fungi, two families and one genus of terrestrial arthropods, one family of snakes, one family of skates and one genus of fish (Table 1).

## Species occurrence data
We downloaded occurrence information for all study groups from www.gbif.org using the `rgbif` v1.4.0 package [@Chamberlain2017] in R [@GBIForg2019; @GBIForg2019a; @GBIForg2019b; @GBIForg2019c; @GBIForg2019d; @GBIForg2019e; @GBIForg2019f; @GBIForg2019g; @GBIForg2019h; @GBIForg2019i; @GBIForg2019j; @GBIForg2019k; @GBIForg2019l; @GBIForg2019m; @GBIForg2019n; @GBIForg2019o]. We chose www.gbif.org for this study, since it is the largest and most widely used provider of species occurrence data and covers the entire tree of life. We downloaded only records with geographic coordinates, and limited the study area to latitudes smaller than 33$^\circ$ N and longitudes smaller than 35$^\circ$ W and larger than 120$^\circ$ W reflecting our area of expertise [the Neotropics, @Morrone2014].

##  Automated cleaning
We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the raw data downloaded from GBIF ("raw", hereafter) using meta-data for those records for which it was available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise and often reflect different land-use types); and 4) records with less than 1 and more than 99 individual records. Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location. In a second step, we used the `clean_coordinates` function of the `CoordinateCleaner v2.0-11` package [@Zizka2019] with default options to flag errors that are common to biological collections, including: sea coordinates, zero coordinates, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions ("filtered", hereafter). See Table 2 for a summary of all flagging tests we used.

## Downstream analyses
We first generated species richness maps using 100x100 km grid cells for the raw and filtered dataset respectively, using the `speciesgeocodeR v2.0-10` package [@Topel2016] and then performed a preliminary conservation assessment for all study groups based on both datasets using the `ConR v1.2.4` package [@Dauby2017]. `ConR` estimates the EOO, AOO, and the number of locations, and then suggests a preliminary conservation status based on Criterion B of the global IUCN Red List. While these assessments are preliminary [see @IUCN2017], they can be a proxy for full assessments and are used by the IUCN to speed up full assessments. We then benchmarked the preliminary conservation assessments with the global IUCN Red List assessments, which we obtained from www.iucn.org via the `rredlist v.0.5.0` package [@Chamberlain2018].

## Evaluation of the results
Each specialist evaluated the performance of the raw and cleaned datasets, respectively concerning: 1) the number of flagged records; 2) the importance of specific tests; 3) the accuracy of the overall species richness maps; 4) the accuracy of the preliminary conservation assessment.

# Results
```{r}
# Table 1
tab1 <- read_excel("figures_tables/table_study_taxa.xlsx")

tab1 <- tab1 %>%
  mutate_all(linebreak) %>%
  select(-researcher) %>% 
  filter(!taxon %in% c("Entomobryidae", "Opisthonema", "Eugenia", "Harpalyce"))

# Table 2
tab2 <- read_csv("output/summary_table_flags.csv") %>% 
  dplyr::select(-coordinate_missing, -invalid_coords, -sea_coords, -gbif_coords,-inst_coords, -equal_coords) %>% 
  mutate(fraction = round(fraction * 100,1))

tab2 <- tab2[,c(1:3,14,5,7,4,8,6,10,13,11,12,9)]

tab2[,-c(1:4)] <- round((tab2[,-c(1:4)] / unlist(tab2[,2]) * 100), 1)

tab2 <- tab1 %>% 
  select(taxon) %>%  
  left_join(tab2, by = "taxon") %>% 
  replace_na(list(total_records = 0))

tab3 <- read_csv("output/table_results_conservation_assessment.csv")

tab3 <- tab1 %>% 
  select(taxon) %>%  
  left_join(tab3, by = "taxon") %>% 
  replace_na(list(iucn_spnum = 0, raw_spnum = 0, flag_spnum = 0)) %>% 
  filter(iucn_spnum > 1)

aa <- read_csv("output/supplement_full_results_conservation_assessment.csv") %>% 
  mutate(iucn_category = ifelse(iucn_category %in% c("LC", "NT"), "LC or NT", iucn_category)) %>% 
  mutate(iucn_status = ifelse(iucn_category == "LC or NT", "not_threatened", "possibly_threatened"))
```

We found a total of `r nrow(dat)` occurrence records, with a median of `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per study group and `r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per species (Table 3). We found most records for `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(taxon) %>%  unlist()` (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs) %>%  unlist()`) and fewest for *`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(taxon) %>%  unlist()`* (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(nrecs) %>%  unlist()`). The species with most records was *`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(species)`* (`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs)`).

Our automated tests flagged a total of `r sum(!dat$summary)` records (Fig. \ref{fig:total}), with a median of `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per group . Overall, the most important test was for duplicated records (on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!.dpl)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per taxonomic group). The most important filtering steps based on record meta-data were for the basis of records (`r dat %>% group_by(taxon) %>% summarize(num = mean(!coordinate_base)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the collection year (`r dat %>% group_by(taxon) %>% summarize(num = mean(!record_age)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%). The most important automated tests were for urban area (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.urb)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%), the capital coordinates (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.cap)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the centroids of political units (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.cen)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%, Table 3, see Supplementary material 1 for the absolute numbers).  Only few records were flagged by the zero coordinates and biodiversity institutions tests (Fig. \ref{fig:split}), and no records were flagged by the sea test.

*`r tab2 %>% arrange(fraction) %>% top_n(1) %>% select(taxon) %>% unlist()`*, *`r tab2 %>% arrange(fraction) %>% top_n(3) %>% select(taxon) %>% slice(2) %>% unlist()`*, and *`r tab2 %>% arrange(fraction) %>% top_n(3) %>% select(taxon) %>% slice(3) %>% unlist()`* had the highest fraction of flagged records (Table 3). In general, the different filters we tested were of similar importance for different study groups. There were few outstanding exceptions, including the particularly high proportions of records flagged by the basis of record test for *Tityus* (7.0%), Dipsadidae (5.6%), *Prosthechea* (5.0%) and *Tillandsia* (4.9%), by the collection year for Dipsadidae (11.3%), by the taxonomic identification level for  *Tityus* (1.6%), by the capital coordinates for *Oocephalus* (6.1%) and *Gaylussacia* (3.2%), as well as by the urban areas test for *Oocephalus* (13.3%) and Iridaceae (12.3%).

Geographically, the records flagged by the basis of record and individuals collected tests were concentrated in Central America and southern North America, and a relative high number of records flagged due to their proximity to the centroids of political entities were located on Caribbean islands (Fig. \ref{fig:split}). See Supplementary material 2 for species richness maps using the raw and cleaned data for all study groups.

<!-- For the discussion check those individual groups and see what exactly was flagged here. -->
We found IUCN assessments for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>% nrow()` species that were also included in our distribution data from 10 of our study groups (Table 4, Supplementary material 3). The fraction of species evaluated varied among the study group, with a maximum of `r tab3 %>% select(iucn_frac_eval) %>% unlist() %>%  max()`% for *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>% slice(1L) %>% select(taxon) %>%  unlist()`* and *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>%  select(taxon) %>% slice(2) %>% unlist()`* and a minimum of `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(iucn_frac_eval) %>%  unlist()`% for `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(taxon) %>%  unlist()` (note that the number of species varied considerably among the groups). The median percentage of species per study group with an IUCN assessment was `r tab3 %>% select(iucn_frac_eval) %>% unlist() %>%  median()`%. A total of `r aa %>% filter(!is.na(iucn_category)) %>%  filter(iucn_category != "DD") %>%filter(iucn_category != "EX") %>%  mutate(test = iucn_status == "possibly_threatened") %>% select(test) %>% unlist() %>%  sum()` species were listed as Threatened by the IUCN global Red List (CR = `r aa %>% filter(iucn_category == "CR") %>% nrow()`, EN = `r aa %>% filter(iucn_category == "EN") %>% nrow()`, VU = `r aa %>% filter(iucn_category == "VU") %>% nrow()`) and `r  aa %>% filter(!is.na(iucn_category)) %>%  filter(iucn_category != "DD") %>%filter(iucn_category != "EX") %>%  mutate(test = iucn_status == "not_threatened") %>% select(test) %>% unlist() %>%  sum()` as Not Threatened.

We obtained automated conservation assessments for `r aa %>% select(flag_Category_CriteriaB) %>% filter(!is.na(flag_Category_CriteriaB)) %>% nrow()` species in the filtered dataset. Based on the filtered data the automated conservation assessment evaluated `r sum(aa$flag_status == "possibly_threatened", na.rm = TRUE)` species as possibly threatened (`r round(sum(aa$flag_status == "possibly_threatened", na.rm=TRUE) / aa %>% filter(!is.na(flag_status)) %>% nrow() * 100, 1)`%, CR = `r aa %>% filter(flag_Category_CriteriaB == "CR") %>% nrow()`, EN = `r aa %>% filter(flag_Category_CriteriaB == "EN") %>% nrow()`, VU = `r aa %>% filter(flag_Category_CriteriaB == "VU") %>% nrow()`). The automated assessment based on the filtered dataset agreed with the IUCN assessment for identifying species as possibly threatened (CR, EN, VU) for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status)) %>% mutate(agreement = iucn_status == flag_status) %>% select(agreement) %>% unlist() %>% sum() ` species (`r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status))  %>% mutate(agreement = iucn_status == flag_status) %>% select(agreement) %>% unlist() %>% mean() %>%  round(3) * 100`%; Table 4). Filtering reduced the EOO by `r tab3 %>%  select(flag_median_EOO_change) %>%  unlist() %>%  median()`% and the AOO by `r tab3 %>%  select(flag_median_AOO_change) %>%  unlist() %>%  median()`% on median per group. For the raw dataset the agreement with IUCN was higher at `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% mutate(agreement = iucn_status == raw_status) %>% select(agreement) %>% unlist() %>% sum() ` species (`r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% mutate(agreement = iucn_status == raw_status) %>% select(agreement) %>% unlist() %>% mean() %>%  round(3) * 100`%).

# Discussion
We found, that automated flagging based on meta-data and automatic tests flagged on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% of the records per taxonomic group. The filters for basis of record, duplicates, collection year, and urban areas flag the highest fraction of records (question 1). The importance of different tests was similar among taxonomic groups, with few exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (question 2). The results for species richness were similar between the raw and filtered data with some improvements by the filters and we found little impact of filtering on the accuracy of the automated conservation assessments (question 3).

<!-- 1. Which is the most relevant flagging mechanism? -->
The high proportion of records flagged as duplicates, may have several reasons. First, the deposition of duplicates from the same specimen hosted at different collections is common practice, especially for plants, where a duplication is easily feasible. Second, independent collections at similar localities may occur, in particular for local endemics, and third low coordinate precision, for instance based on automated geo-referencing from locality descriptions, may lump records from nearby localities. Of course, duplicated records do not represent errors per se, but they represent an unnecessary computational burden, which is often a bottleneck for analyses of datasets with millions of records.

The aim of automated filtering is to identify possibly problematic records that are unsuitable for particular downstream analyses. Hence our tests include errors in the narrow sense (e.g., zero coordinates, equal lat/lon and seas tests) as well as records that are not necessary erroneous, but have a low precision (e.g., collection year, coordinate precision, centroids of political units) or are unnecessary (duplicates). The importance of each test and the linked thresholds must be judged based on the downstream analyses. For instance, records assigned to country centroids might be acceptable for inter-continental comparisons, but can be consider erroneous for species distribution modelling on a local scale. As our results show, adapting automated tests to the geographic study area or the taxonomic study group can be warranted. For instance, the high number of records flagged for centroids on the Lesser Antilles (Fig. \ref{fig:split}) might be overly strict, although we chose a conservative distance for the Political centroid test (1 km).

<!-- 2. Does the importance of individual errors differ among taxonomic groups? -->
The number of records flagged by individual tests was similar across the study groups, suggesting that similar error types might be relevant for collections of plants and animals and partly also terrestrial and marine organisms. Therefore, the same tests can be used across taxonomic groups. Some notable exceptions stress the need to adapt each test to the taxonomic study group to balance data quality and data availability. The high fraction of records flagged by the basis of record test for *Tityus*, Dipsadidae, *Prosthechea* and *Tillandsia*, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups. The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection activity already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700. The latter records likely represent a data entry error. They are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700. The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae, *Oocephalus*, and *Gaylussacia*, might be related to horticulture since at least some species of this group are commonly cultivated as ornamentals.

The general agreement between the species richness maps based on raw and filtered data was encouraging, for the use of this data for large-scale biogeographic research (Fig. \ref{fig:speciesrichness}, Supplementary material 1). The filter based on political centroids had an important impact on species richness patterns, which is congruent with the results from previous studied [@Maldonado2015]. Records assigned to country or province centroids are often old records, which are geo-referenced based on vague locality descriptions. These records are at the same time more likely to represent dubious species names, since they might be old synonyms or type specimen of species that have only been collected and described once, which are erroneously increasing species numbers.

The accuracy of the automated conservation assessment was in the same range as found by previous studies [@NicLughadha2019; @Zizka2019a]. The similar accuracy of the raw and filtered dataset for the automated conservation assessment was surprising, in particular given the EOO and AOO reduction observed in the filtered dataset (Table 4) and the impact of errors on spatial analyses observed in previous studies [@Gueta2016]. The robustness of the automated assessment was likely due to the fact that the EOO for most species was large, even after the considerable reduction caused by filtering. This might be caused by the structure of our comparison, which only included species that were evaluated by the IUCN Red List (and not considered as Data Deficient) and at the same time have occurrences recorded in GBIF, which biases the dataset towards species with larger ranges, since generally more data for them is available. The robustness of automated conservation assessments to data quality is encouraging, although these methods are only an approximation, especially for species with few collection records [@Rivers2011].

# Conclusions
Our results suggest that between 1/3rd and 2/3rds of the occurrence records obtained from public databases might be unsuitable for downstream biogeographic analyses, dependent on the taxonomic group. While the majority of these records might not be erroneous per se, they might be overly imprecise and thereby increase uncertainty of downstream results or put additional computational burden on the analyses.

Overall, our results stress the importance of (meta-)data exploration before analyses and suggest that automated flagging can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the specific taxonomic groups and geographic area of interest.

# Acknowledgements
This study enrolled participants of the workshop "Biodiversity data: from field to yield" lead by Alexandre Antonelli and Alexander Zizka through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI. We thank the Pró-reitoria de Pesquisa and the Pró-reitoria de Pós-graduação of UFRN for the financial support (edital 02/2016 - internacionalização). AZ is thankful for funding by iDiv via the German Research Foundation (DFG FZT 118), specifically through sDiv, the Synthesis Centre of iDiv.

# Supplementary material
Appendix 1 - Table S1, the absolute number of flagged records per taxonomic group and test

Appendix 2 - Taxon specific richness maps and comments.

Appendix 3 - The full results of the conservation assessment

\newpage{}

# Tables
<!-- A table briefly describing the tests we ran -->
```{r tabletests, warning=FALSE}
tab0 <- read_xlsx("figures_tables/table_test_performed.xlsx")

kable(tab0, booktabs = T, escape = T, format="latex",
        caption = "The automated filters used in this study.")%>%
  kable_styling(font_size = 9, latex_options =c("striped", "hold_position")) %>%
   column_spec(1, width = "2cm", bold = FALSE, italic = FALSE)%>%
   column_spec(2, width = "2.5cm", bold = FALSE, italic = FALSE)%>%
   column_spec(3, width = "10cm", bold = FALSE, italic = FALSE)
```

<!-- A table presenting the study taxa -->
```{r tabletaxa, warning=FALSE}
kable(tab1, booktabs = T, escape = T, format="latex",
      col.names = linebreak(c("Taxon",
                              "Taxon rank",
                              "Realm",
                              "Common name",
                              "'Phylum'",
                              "Order",
                              "Family")),
        caption = "The study groups and their taxonomy. This study includes three marine and 13 terrestrial taxa, 6 of them animals, one group of fungi and 9 plants, belonging to 16 different orders.")%>%
  landscape()%>%
  kable_styling(font_size = 11) %>%
   row_spec(6, hline_after = TRUE) %>%
   row_spec(7, hline_after = TRUE) %>%
   column_spec(1, width = "2.5cm", bold = FALSE, italic = FALSE)%>%
   column_spec(2, width = "2cm", bold = FALSE, italic = FALSE)%>%
   column_spec(3, width = "2cm", bold = FALSE, italic = FALSE)%>%
   column_spec(4:5, width = "2.5cm", bold = FALSE, italic = FALSE)
```

<!-- A table presenting the overall results, the decrease in records by filtering, the change in conservation status -->
```{r tablecoords, warning=FALSE}
tab2 <- tab2 %>%
  mutate_all(linebreak)

kable(tab2, booktabs = T, escape = T, format="latex",
      col.names = linebreak(c("Taxon",
                              "Total records",
                              "Records flagged",
                              "Fraction flagged [%]",
                              "Basis of record [%]",
                              "Collection year [%]",
                              "Coordinate precision [%]",
                              "Id-level [%]",
                              "Individual count [%]",
                              "Capitals [%]",
                              "Duplicates [%]",
                              "Political centroids [%]",
                              "Urban areas [%]",
                              "Zeros [%]")),
        caption = "The impact of automated flagging on occurrence records for 16 Neotropical taxa downloaded from www.gbif.org. From column four onwards the numbers show the percentage of records flagged by the respective test. Only tests that flagged at least 0.1\\% of the records in any group are shown.", format.args = list(big.mark = ","))%>%
  landscape()%>%
  kable_styling(font_size = 9) %>%
   row_spec(6, hline_after = TRUE) %>%
   row_spec(7, hline_after = TRUE) %>%
   column_spec(1, width = "2cm", bold = FALSE, italic = FALSE)%>%
   column_spec(2:6, width = "1.1cm", bold = FALSE, italic = FALSE)%>%
   column_spec(7:12, width = "1.3cm", bold = FALSE, italic = FALSE)%>%
  column_spec(13:14, width = "1cm", bold = FALSE, italic = FALSE)%>%
   add_header_above(c(" " = 1,
                      "Summary" = 3,
                     "Meta data filters" = 5,
                     "Automated filters" = 5))
```

```{r}
tab3 <- tab3 %>%
  mutate_all(linebreak)

kable(tab3, booktabs = T, escape = T, format="latex",
      col.names = linebreak(c("Taxon",
                              "n taxa",
                              "Evaluated [%]",
                              "Threatened [%]",
                              "n taxa",
                              "Threatened [%]",
                              "Match with IUCN [%]",
                              "n taxa",
                              "Threatened [%]",
                              "Match with IUCN [%]",
                              "EOO change compared to raw [%]",
                              "AOO change compared to raw [%]")),
        caption = "Conservation assessment for 16 Neotropical taxa of plants and animals based on three datasets. IUCN: global red list assessment obtained from www.iucn.org; GBIF Raw: Preliminary conservation assessment based on IUCN Criterion B using ConR and the raw dataset from GBIF; GBIF filtered: Preliminary conservation assessment based on IUCN Criterion B using ConR and the filtered dataset. Only taxa with at least on species evaluated by IUCN shown.", 
      format.args = list(big.mark = ","))%>%
  landscape()%>%
  kable_styling(font_size = 9) %>%
   column_spec(2:10, width = "1.2cm", bold = FALSE, italic = FALSE)%>%
   column_spec(11:12, width = "1.5cm", bold = FALSE, italic = FALSE)%>%
   add_header_above(c(" " = 1,
                      "IUCN" = 3,
                     "GBIF Raw" = 3,
                     "GBIF Filtered" = 5))
```

\newpage{}

# Figures
<!-- A map with number of species and where they have been removed -->
```{r total, out.width="\\linewidth", fig.cap=c('The difference of records available between the raw and filtered dataset, in a 100x100 km grid across the Neotropics'), echo=FALSE}
knitr::include_graphics(path = "./output/figure_number_of_records.jpg")
```

```{r split, out.width="\\linewidth", fig.cap=c('The geographic pattern of the occurrence records flagged by the automated tests of this study. The concentration of records flagged by the basis of record test and individual count test in Central A,erica and southern North America, suggest some specific problem with a contributing dataset from the region. Only filters that flagged at least one record are shown.'), echo=FALSE}
knitr::include_graphics(path = "./output/figure_number_of_records_split.jpg")
```

```{r speciesrichness, out.width="\\linewidth", fig.cap=c('The difference in species richness derived from the raw and filtered datasets (raw - filtered), for 16 taxa of plants and animals.'), echo=FALSE}
knitr::include_graphics(path = "./output/figure_number_species_richness_difference.jpg")
```

\newpage{}

# References

### 2020-01-09 manuscript.tex
Version send to co-authors

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{setspace}
\setstretch{1.3}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Automated filtering of species occurrence data can identify problems, but needs to be customized},
  pdfauthor={Alexander Zizka1,2,; Alice Calvente3,; Fernanda Antunes Carvalho4,5,; Ana Paula Alves da Silva,; Andressa Cabral,; Arthur de Souza Soares,; Eduardo Calisto Tomaz,; Fernanda Gondim Lambert Moreira,; Filipe Serrano,; Jéssica Fernanda Ramos Coelho,; Mabel Rocio Baez-Lizarazo,; Mariana Ramos Fantinati,; Matheus Colli-Silva,; Moabe Ferreira Fernandes,; Nathália Michelly da Cunha Santos,; Paolla G. C. Souza,; Renata Clícia Santos da Costa,; Thais Ferreira Pinto de Araújo,; Tiago Andrade Borges Santos,; Tiago Luiz Vieira,; Valéria Fonseca Vale,; Alexandre Antonelli6},
  pdfborder={0 0 0},
  breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-2}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{caption}
\usepackage{fancyhdr}
\usepackage{booktabs}
\pagestyle{fancy}
\fancyhead[L]{Automated filter effect}
\fancyhead[LE,RO]{\thepage}
\fancyfoot{}
\usepackage{float}
\floatplacement{figure}{H}
\usepackage{lineno}
\linenumbers
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{Automated filtering of species occurrence data can identify problems, but needs to be customized}
\author{Alexander Zizka\textsuperscript{1,2}, \and Alice Calvente\textsuperscript{3}, \and Fernanda Antunes Carvalho\textsuperscript{4,5}, \and Ana Paula Alves da Silva, \and Andressa Cabral, \and Arthur de Souza Soares, \and Eduardo Calisto Tomaz, \and Fernanda Gondim Lambert Moreira, \and Filipe Serrano, \and Jéssica Fernanda Ramos Coelho, \and Mabel Rocio Baez-Lizarazo, \and Mariana Ramos Fantinati, \and Matheus Colli-Silva, \and Moabe Ferreira Fernandes, \and Nathália Michelly da Cunha Santos, \and Paolla G. C. Souza, \and Renata Clícia Santos da Costa, \and Thais Ferreira Pinto de Araújo, \and Tiago Andrade Borges Santos, \and Tiago Luiz Vieira, \and Valéria Fonseca Vale, \and Alexandre Antonelli\textsuperscript{6}}
\date{}

\begin{document}
\maketitle

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  sDiv,German Center for Integrative Biodiversity Research (iDiv), Leipzig, Germany
\item
  Naturalis Biodiversity Center, Leiden, The Netherlands
\item
  Laboratório de Botânica Sistemática, Departamento de Botânica e Zoologia, Centro de Biociências, Universidade Federal do Rio Grande do Norte.
\item
  Departamento de Genética, Ecologia e Evolução, Instituto de Ciências Biológicas, Universidade Federal de Minas Gerais, Av. Antônio Carlos 6627 - Pampulha, 31270-901 - Belo Horizonte - MG - Brasil
\end{enumerate}

\newpage{}

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

Automatic flagging and subsequent filtering are a scalable and reproducible means to identify potentially problematic records in `big data' datasets of species geographic occurrences from public databases (e.g., www.gbif.org). However, it is unclear how big the data loss by filtering is, if similar test should be applied across taxonomic groups, and what the effect of filtering is for downstream analyses. Here, we obtained occurrence records for 16 taxa including plants and animals as well as terrestrial and marine groups from www.gbif.org and documented the effect of 13 recently proposed filters in each group. We evaluated the effect of filtering on the inference of species richness patterns and the accuracy of automated conservation assessments as examples for downstream analyses. We found 29-67\% of the records potentially problematic across taxonomic groups. The tests for duplicated information, collection year, basis of record as well as urban areas and political centroids were most important. While many of the flagged records might not be erroneous per se, they might be overly imprecise, increase uncertainty of downstream analyses and generate unnecessary computational burden. Overall, our results stress the importance of (meta-)data exploration for biogeographic research and suggest that automated flagging can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the taxonomic group and geographic area of interest.

\hypertarget{keywords}{%
\section{Keywords}\label{keywords}}

Automated cleaning, automated conservation assessment, data quality, GBIF, Neotropics

\newpage{}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Publicly available species distribution data have become a crucial resource in biogeographic research. In particular, the availability of digitized collections from museums and herbaria, and citizen science observations has increased in the last years. As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.3 billion geo-referenced occurrence records for species from across the globe and the tree of life.

A central challenge to the use of these publicly available species occurrence data in research are erroneous geographic coordinates (Anderson et al., 2016). Errors mostly arise, because public databases integrate records collected with different methodology in different places at different times, often with little centralized curation and meta-data. For instance, erroneous coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common (Maldonado et al., 2015; Yesson et al., 2007) and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records from species in captivity (Zizka, Silvestro, et al., 2019).

Manual data cleaning based on expert knowledge can detect these issues, but is only feasible on small taxonomic or geographic scales and is difficult to reproduce. Automated flagging methods to identify potentially problematic records have been proposed as a scalable alternative for datasets with millions of records from many different taxa. Those methods are usually based on geographic gazetteers (e.g., Chamberlain, 2016; Jin \& Yang, 2020; Zizka, Silvestro, et al., 2019) or on additional data, such as environmental variables (Robertson, Visser, \& Hui, 2016). Additionally, filtering procedures based on record meta-data, for instance collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biogeographic research (Zizka, Silvestro, et al., 2019). While these methods can identify common errors, it is unclear if errors are equally common across taxonomic groups, and which errors are most severe for downstream analyses.

Erroneous records are especially problematic in conservation, where stakes are high. Recently, automated conservation assessment methods have been developed to use publicly available occurrence records for conservation, in particular to support the assessment procedures for the global Red List of the International Union for the Conservation of Nature (IUCN) (Bachman et al., 2011; Dauby et al., 2017; Pelletier et al., 2018). These methods approximate species' range size, namely the Extent of Occurrence (EOO, which is the area of a convex hull polygon comprising all records of a species), the Area of Occupancy (AOO, which is the sum of the area actually occupied by a species, calculated based on a small-scale regular grid), and the number of locations for a preliminary conservation assessment following IUCN Criterion B. Automated assessments using these methods have been used to speed up global (Stévart et al., 2019; Zizka, Azevedo, et al., 2019) and regional (Cosiaux et al., 2018; Schmidt et al., 2017) Red List assessments. Since especially EOO is sensitive to individual records with erroneous coordinates automated conservation assessments might be prone to bias by erroneous records, especially, if the number of records is low, as is the case for many species from the tropical regions of Earth.

Here, we test the effect of automated filtering of species geographic occurrence records on the number of records available in different plant and animal groups. Furthermore, we test the impact of automated filtering procedures on the accuracy of automated conservation assessments compared to full IUCN assessments. Specifically, we test a pipeline of 13 tests to automatically flag possibly erroneous records 1) using record meta-data and 2) using geographic gazetteers. We address thee questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Which is the most relevant filter, i.e.~which type of error is most common?
\item
  Does the importance of individual filters differ among taxonomic groups?
\item
  Does automated filtering improve the accuracy of automated conservation assessment compared to the global IUCN Red List?
\end{enumerate}

\hypertarget{material-and-methods}{%
\section{Material and Methods}\label{material-and-methods}}

We did all analysis in the R computing environment (R Core Team, 2019), using standard libraries for data handling and visualization (Garnier, 2018; Hijmans, 2019; Ooms, 2014, 2019; Wickham, 2018). All analysis scripts are available from \url{https://github.com/idiv-biodiversity/effects_of_automated_cleaning}.

\hypertarget{choice-of-study-taxa}{%
\subsection{Choice of study taxa}\label{choice-of-study-taxa}}

This study is the outcome of a workshop at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering taxonomists and biogeographers working with different taxonomic groups of plants and animals across the Neotropics. Each participant analysed geographic occurrence data from their taxonomic group of interest and provided an assessment of the results for this group. Hence, we chose the study groups based on the expertise of the participants which enabled us to compare the results across the tree of life (at the coast of an arbitrary choice of taxa and taxonomic ranks). We obtained public occurrence records for 16 taxa, including one plant family, eight plant genera, one genus of fungi, two families and one genus of terrestrial arthropods, one family of snakes, one family of skates and one genus of fish (Table 1).

\hypertarget{species-occurrence-data}{%
\subsection{Species occurrence data}\label{species-occurrence-data}}

We downloaded occurrence information for all study groups from www.gbif.org using the \texttt{rgbif} v1.4.0 package (Chamberlain, 2017) in R (GBIF.org, 2019a, 2019b, 2019c, 2019d, 2019e, 2019f, 2019g, 2019h, 2019i, 2019j, 2019k, 2019l, 2019m, 2019n, 2019o, 2019p). We chose www.gbif.org for this study, since it is the largest and most widely used provider of species occurrence data and covers the entire tree of life. We downloaded only records with geographic coordinates, and limited the study area to latitudes smaller than 33\(^\circ\) N and longitudes smaller than 35\(^\circ\) W and larger than 120\(^\circ\) W reflecting our area of expertise (the Neotropics, Morrone, 2014).

\hypertarget{automated-cleaning}{%
\subsection{Automated cleaning}\label{automated-cleaning}}

We followed the cleaning pipeline outlined by Zizka, Silvestro, et al. (2019) and first filtered the raw data downloaded from GBIF (``raw'', hereafter) using meta-data for those records for which it was available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise and often reflect different land-use types); and 4) records with less than 1 and more than 99 individual records. Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location. In a second step, we used the \texttt{clean\_coordinates} function of the \texttt{CoordinateCleaner\ v2.0-11} package (Zizka, Silvestro, et al., 2019) with default options to flag errors that are common to biological collections, including: sea coordinates, zero coordinates, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions (``filtered'', hereafter). See Table 2 for a summary of all flagging tests we used.

\hypertarget{downstream-analyses}{%
\subsection{Downstream analyses}\label{downstream-analyses}}

We first generated species richness maps using 100x100 km grid cells for the raw and filtered dataset respectively, using the \texttt{speciesgeocodeR\ v2.0-10} package (Töpel et al., 2016) and then performed a preliminary conservation assessment for all study groups based on both datasets using the \texttt{ConR\ v1.2.4} package (Dauby et al., 2017). \texttt{ConR} estimates the EOO, AOO, and the number of locations, and then suggests a preliminary conservation status based on Criterion B of the global IUCN Red List. While these assessments are preliminary (see IUCN Standards and Petitions Subcommittee, 2017), they can be a proxy for full assessments and are used by the IUCN to speed up full assessments. We then benchmarked the preliminary conservation assessments with the global IUCN Red List assessments, which we obtained from www.iucn.org via the \texttt{rredlist\ v.0.5.0} package (Chamberlain, 2018).

\hypertarget{evaluation-of-the-results}{%
\subsection{Evaluation of the results}\label{evaluation-of-the-results}}

Each specialist evaluated the performance of the raw and cleaned datasets, respectively concerning: 1) the number of flagged records; 2) the importance of specific tests; 3) the accuracy of the overall species richness maps; 4) the accuracy of the preliminary conservation assessment.

\hypertarget{results}{%
\section{Results}\label{results}}

We found a total of 215,262 occurrence records, with a median of 3,460 records per study group and 11 records per species (Table 3). We found most records for Dipsadidae (64,249) and fewest for \emph{Thozetella} (51). The species with most records was \emph{Harengula jaguana} (19,878).

Our automated tests flagged a total of 88,648 records (Fig. \ref{fig:total}), with a median of 40.3\% per group . Overall, the most important test was for duplicated records (on average 32.9\% per taxonomic group). The most important filtering steps based on record meta-data were for the basis of records (2.1\%) and the collection year (3.3\%). The most important automated tests were for urban area (5\%), the capital coordinates (1.1\%) and the centroids of political units (0.5\%, Table 3, see Supplementary material 1 for the absolute numbers). Only few records were flagged by the zero coordinates and biodiversity institutions tests (Fig. \ref{fig:split}), and no records were flagged by the sea test.

\emph{Prosthechea}, \emph{Harengula}, and \emph{Thozetella} had the highest fraction of flagged records (Table 3). In general, the different filters we tested were of similar importance for different study groups. There were few outstanding exceptions, including the particularly high proportions of records flagged by the basis of record test for \emph{Tityus} (7.0\%), Dipsadidae (5.6\%), \emph{Prosthechea} (5.0\%) and \emph{Tillandsia} (4.9\%), by the collection year for Dipsadidae (11.3\%), by the taxonomic identification level for \emph{Tityus} (1.6\%), by the capital coordinates for \emph{Oocephalus} (6.1\%) and \emph{Gaylussacia} (3.2\%), as well as by the urban areas test for \emph{Oocephalus} (13.3\%) and Iridaceae (12.3\%).

Geographically, the records flagged by the basis of record and individuals collected tests were concentrated in Central America and southern North America, and a relative high number of records flagged due to their proximity to the centroids of political entities were located on Caribbean islands (Fig. \ref{fig:split}). See Supplementary material 2 for species richness maps using the raw and cleaned data for all study groups.

We found IUCN assessments for 575 species that were also included in our distribution data from 10 of our study groups (Table 4, Supplementary material 3). The fraction of species evaluated varied among the study group, with a maximum of 100\% for \emph{Harengula} and \emph{Lepismium} and a minimum of 2.3\% for Iridaceae (note that the number of species varied considerably among the groups). The median percentage of species per study group with an IUCN assessment was 32.45\%. A total of 101 species were listed as Threatened by the IUCN global Red List (CR = 19, EN = 40, VU = 42) and 474 as Not Threatened.

We obtained automated conservation assessments for 2,092 species in the filtered dataset. Based on the filtered data the automated conservation assessment evaluated 1,299 species as possibly threatened (62.1\%, CR = 464, EN = 543, VU = 292). The automated assessment based on the filtered dataset agreed with the IUCN assessment for identifying species as possibly threatened (CR, EN, VU) for 356 species (64\%; Table 4). Filtering reduced the EOO by -16.1\% and the AOO by -9.3\% on median per group. For the raw dataset the agreement with IUCN was higher at 378 species (65.6\%).

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

We found, that automated flagging based on meta-data and automatic tests flagged on average 40.3\% of the records per taxonomic group. The filters for basis of record, duplicates, collection year, and urban areas flag the highest fraction of records (question 1). The importance of different tests was similar among taxonomic groups, with few exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (question 2). The results for species richness were similar between the raw and filtered data with some improvements by the filters and we found little impact of filtering on the accuracy of the automated conservation assessments (question 3).

The high proportion of records flagged as duplicates, may have several reasons. First, the deposition of duplicates from the same specimen hosted at different collections is common practice, especially for plants, where a duplication is easily feasible. Second, independent collections at similar localities may occur, in particular for local endemics, and third low coordinate precision, for instance based on automated geo-referencing from locality descriptions, may lump records from nearby localities. Of course, duplicated records do not represent errors per se, but they represent an unnecessary computational burden, which is often a bottleneck for analyses of datasets with millions of records.

The aim of automated filtering is to identify possibly problematic records that are unsuitable for particular downstream analyses. Hence our tests include errors in the narrow sense (e.g., zero coordinates, equal lat/lon and seas tests) as well as records that are not necessary erroneous, but have a low precision (e.g., collection year, coordinate precision, centroids of political units) or are unnecessary (duplicates). The importance of each test and the linked thresholds must be judged based on the downstream analyses. For instance, records assigned to country centroids might be acceptable for inter-continental comparisons, but can be consider erroneous for species distribution modelling on a local scale. As our results show, adapting automated tests to the geographic study area or the taxonomic study group can be warranted. For instance, the high number of records flagged for centroids on the Lesser Antilles (Fig. \ref{fig:split}) might be overly strict, although we chose a conservative distance for the Political centroid test (1 km).

The number of records flagged by individual tests was similar across the study groups, suggesting that similar error types might be relevant for collections of plants and animals and partly also terrestrial and marine organisms. Therefore, the same tests can be used across taxonomic groups. Some notable exceptions stress the need to adapt each test to the taxonomic study group to balance data quality and data availability. The high fraction of records flagged by the basis of record test for \emph{Tityus}, Dipsadidae, \emph{Prosthechea} and \emph{Tillandsia}, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups. The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection activity already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700. The latter records likely represent a data entry error. They are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700. The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae, \emph{Oocephalus}, and \emph{Gaylussacia}, might be related to horticulture since at least some species of this group are commonly cultivated as ornamentals.

The general agreement between the species richness maps based on raw and filtered data was encouraging, for the use of this data for large-scale biogeographic research (Fig. \ref{fig:speciesrichness}, Supplementary material 1). The filter based on political centroids had an important impact on species richness patterns, which is congruent with the results from previous studied (Maldonado et al., 2015). Records assigned to country or province centroids are often old records, which are geo-referenced based on vague locality descriptions. These records are at the same time more likely to represent dubious species names, since they might be old synonyms or type specimen of species that have only been collected and described once, which are erroneously increasing species numbers.

The accuracy of the automated conservation assessment was in the same range as found by previous studies (Nic Lughadha et al., 2019; Zizka, Azevedo, et al., 2019). The similar accuracy of the raw and filtered dataset for the automated conservation assessment was surprising, in particular given the EOO and AOO reduction observed in the filtered dataset (Table 4) and the impact of errors on spatial analyses observed in previous studies (Gueta \& Carmel, 2016). The robustness of the automated assessment was likely due to the fact that the EOO for most species was large, even after the considerable reduction caused by filtering. This might be caused by the structure of our comparison, which only included species that were evaluated by the IUCN Red List (and not considered as Data Deficient) and at the same time have occurrences recorded in GBIF, which biases the dataset towards species with larger ranges, since generally more data for them is available. The robustness of automated conservation assessments to data quality is encouraging, although these methods are only an approximation, especially for species with few collection records (Rivers et al., 2011).

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

Our results suggest that between 1/3rd and 2/3rds of the occurrence records obtained from public databases might be unsuitable for downstream biogeographic analyses, dependent on the taxonomic group. While the majority of these records might not be erroneous per se, they might be overly imprecise and thereby increase uncertainty of downstream results or put additional computational burden on the analyses.

Overall, our results stress the importance of (meta-)data exploration before analyses and suggest that automated flagging can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the specific taxonomic groups and geographic area of interest.

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

This study enrolled participants of the workshop ``Biodiversity data: from field to yield'' lead by Alexandre Antonelli and Alexander Zizka through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI. We thank the Pró-reitoria de Pesquisa and the Pró-reitoria de Pós-graduação of UFRN for the financial support (edital 02/2016 - internacionalização). AZ is thankful for funding by iDiv via the German Research Foundation (DFG FZT 118), specifically through sDiv, the Synthesis Centre of iDiv.

\hypertarget{supplementary-material}{%
\section{Supplementary material}\label{supplementary-material}}

Appendix 1 - Table S1, the absolute number of flagged records per taxonomic group and test

Appendix 2 - Taxon specific richness maps and comments.

Appendix 3 - Table S2, the full results of the conservation assessment

\newpage{}

\hypertarget{tables}{%
\section{Tables}\label{tables}}

\begin{table}[!h]

\caption{\label{tab:tabletests}The automated filters used in this study.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{10cm}}
\toprule
Test & Type & Rationale\\
\midrule
\rowcolor{gray!6}  Basis of record & Meta-data & Records might be unsuitable or unreliable for certain analysis dependent on their source, e.g. "fossil" or "unknown"\\
Collection year & Meta-data & Coordinates from old records are more likely to be imprecise or erroneous coordinates since they are derived from  geo-referencing based on the locality description. Since is more problematic for older records, since names or borders of places may change\\
\rowcolor{gray!6}  Coordinate precision & Meta-data & Records may be unsuitable for a study if their precision is lower than the studies analysis scale\\
Identification level & Meta-data & Records may be unsuitable if they are not identified to species level.\\
\rowcolor{gray!6}  Individual count & Meta-data & Records may be unsuitable if the number of recorded individuals is 0 (record of absence) or if the count is to high, as this is often related to records from barcoding or indicative of data entry problems.\\
\addlinespace
Biodiversity institutions & Gazetteer-based & Records may have coordinatesat the location of biodiversity institutions, e.g. because they were erroneously entered with the physical location of the specimen or because they represent individuals from captivity or horticulture\\
\rowcolor{gray!6}  Duplicates & Gazetteer-based & Duplicated records may add unnecessary computational burden to analyses\\
Equal lat/lon & Gazetteer-based & Coordinates with equal latitude and longitude are usually indicative of data entry errors\\
\rowcolor{gray!6}  Capitals & Gazetteer-based & Records may be assigned to the coordinates of countries capitals based on a vague locality description\\
Political centroids & Gazetteer-based & Records may be assigned to the coordinates of the centroids of political entities based on a vague locality description\\
\addlinespace
\rowcolor{gray!6}  Sea & Gazetteer-based & Coordinates from terrestrial organisms in the sea are usually indicative of data entry errors, e.g. swapped latitude and longitude\\
Urban areas & Gazetteer-based & Records from urban areas are not necessarily errors, but often represent imprecise records automatically geo-referenced from vague locality descriptions or old records from different land-use types\\
\rowcolor{gray!6}  Zeros & Gazetteer-based & Coordinates with plain zeros are often indicative of data entry errors\\
\bottomrule
\end{tabular}
\end{table}

\begin{landscape}\begin{table}

\caption{\label{tab:tabletaxa}The study groups and their taxonomy. This study includes three marine and 13 terrestrial taxa, 6 of them animals, one group of fungi and 9 plants, belonging to 16 different orders.}
\centering
\fontsize{11}{13}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}ll}
\toprule
Taxon & Taxon rank & Realm & Common name & 'Phylum' & Order & Family\\
\midrule
Diogenidae & Family & Marine & Hermit crabs & Arthropoda & Decapoda & Diogenidae\\
Neanuridae & Family & Terrestrial & Springtails & Arthropoda & Poduromorpha & Neanuridae\\
Tityus & Genus & Terrestrial & Scorpions & Arthropoda & Scorpiones & Buthidae\\
Arhynchobatidae & Family & Marine & Skates & Chordata & Rajiformes & Arhynchobatidae\\
Dipsadidae & Family & Terrestrial & Snakes & Chordata & Squamata & Colubridae\\
\addlinespace
Harengula & Genus & Marine & Herings & Chordata & Clupeiformes & Clupeidae\\
\hline
Thozetella & Genus & Terrestrial & Sac fungi & Ascomycota & Chaetosphaeriales & Chaetosphaeriaceae\\
\hline
Conchocarpus & Genus & Terrestrial & NA & Angiosperms & Sapindales & Rutaceae\\
Gaylussacia & Genus & Terrestrial & Huckleberries & Angiosperms & Ericales & Ericaceae\\
Iridaceae & Family & Terrestrial & NA & Angiosperms & Asparagales & Iridaceae\\
\addlinespace
Lepismium & Genus & Terrestrial & Cacti & Angiosperms & Caryophyllales & Cactaceae\\
Oocephalus & Genus & Terrestrial & NA & Angiosperms & Lamiales & Lamiaceae\\
Pilosocereus & Genus & Terrestrial & NA & Angiosperms & Caryophyllales & Cactaceae\\
Prosthechea & Genus & Terrestrial & Orchids & Angiosperms & Asparagales & Orchidaceae\\
Tillandsia & Genus & Terrestrial & Bromeliads & Angiosperms & Poales & Bromeliaceae\\
\addlinespace
Tocoyena & Genus & Terrestrial & NA & Angiosperms & Gentianales & Rubiaceae\\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

\begin{landscape}\begin{table}

\caption{\label{tab:tablecoords}The impact of automated flagging on occurrence records for 16 Neotropical taxa downloaded from www.gbif.org. From column four onwards the numbers show the percentage of records flagged by the respective test. Only tests that flagged at least 0.1\% of the records in any group are shown.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{1.1cm}>{\raggedleft\arraybackslash}p{1.1cm}>{\raggedleft\arraybackslash}p{1.1cm}>{\raggedleft\arraybackslash}p{1.1cm}>{\raggedleft\arraybackslash}p{1.1cm}>{\raggedleft\arraybackslash}p{1.3cm}>{\raggedleft\arraybackslash}p{1.3cm}>{\raggedleft\arraybackslash}p{1.3cm}>{\raggedleft\arraybackslash}p{1.3cm}>{\raggedleft\arraybackslash}p{1.3cm}>{\raggedleft\arraybackslash}p{1.3cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Summary} & \multicolumn{5}{c}{Meta data filters} & \multicolumn{5}{c}{Automated filters} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-9} \cmidrule(l{3pt}r{3pt}){10-14}
Taxon & Total records & Records flagged & Fraction flagged [\%] & Basis of record [\%] & Collection year [\%] & Coordinate precision [\%] & Id-level [\%] & Individual count [\%] & Capitals [\%] & Duplicates [\%] & Political centroids [\%] & Urban areas [\%] & Zeros [\%]\\
\midrule
Diogenidae & 13,840 & 5,294 & 38.3 & 1.7 & 2.5 & 0.0 & 0.0 & 0.0 & 0.7 & 33.8 & 0.2 & 1.3 & 0.0\\
Neanuridae & 689 & 461 & 66.9 & 2.9 & 1.3 & 0.0 & 0.0 & 0.0 & 0.0 & 62.4 & 0.0 & 2.0 & 0.0\\
Tityus & 1,018 & 562 & 55.2 & 7.0 & 0.4 & 1.8 & 1.6 & 0.0 & 1.2 & 43.5 & 0.1 & 6.9 & 0.0\\
Arhynchobatidae & 14,633 & 5,469 & 37.4 & 1.7 & 1.3 & 0.0 & 0.9 & 0.0 & 0.0 & 35.4 & 0.0 & 1.9 & 0.0\\
Dipsadidae & 64,249 & 37,054 & 57.7 & 5.6 & 11.3 & 0.8 & 0.0 & 0.1 & 1.8 & 46.3 & 0.4 & 8.5 & 0.0\\
\addlinespace
Harengula & 36,697 & 10,230 & 27.9 & 1.0 & 0.4 & 0.0 & 0.3 & 0.0 & 0.2 & 27.0 & 0.1 & 0.2 & 0.0\\
\hline
Thozetella & 51 & 15 & 29.4 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 27.5 & 0.0 & 2.0 & 0.0\\
\hline
Conchocarpus & 1,551 & 665 & 42.9 & 0.5 & 1.9 & 0.1 & 0.0 & 0.0 & 0.0 & 39.6 & 0.9 & 2.3 & 0.0\\
Gaylussacia & 3,998 & 1,884 & 47.1 & 0.7 & 4.4 & 0.6 & 0.0 & 0.0 & 3.2 & 41.8 & 1.1 & 5.2 & 0.0\\
Iridaceae & 23,127 & 7,760 & 33.6 & 0.9 & 4.7 & 0.1 & 0.0 & 1.3 & 1.0 & 17.1 & 0.4 & 12.3 & 0.0\\
\addlinespace
Lepismium & 825 & 245 & 29.7 & 0.0 & 2.1 & 0.0 & 0.0 & 0.0 & 0.1 & 21.9 & 0.1 & 7.8 & 0.0\\
Oocephalus & 883 & 435 & 49.3 & 0.0 & 0.7 & 0.3 & 0.0 & 0.1 & 6.1 & 41.9 & 0.8 & 13.3 & 0.0\\
Pilosocereus & 1,940 & 501 & 25.8 & 1.8 & 7.0 & 0.0 & 0.0 & 0.9 & 0.5 & 16.8 & 0.5 & 2.1 & 0.0\\
Prosthechea & 6,617 & 2,088 & 31.6 & 5.0 & 8.3 & 0.1 & 0.0 & 0.2 & 0.4 & 19.6 & 1.7 & 0.9 & 0.1\\
Tillandsia & 42,222 & 14,893 & 35.3 & 4.9 & 5.1 & 0.1 & 0.0 & 1.0 & 0.7 & 19.8 & 0.7 & 9.2 & 0.0\\
\addlinespace
Tocoyena & 2,922 & 1,092 & 37.4 & 0.1 & 1.9 & 0.2 & 0.0 & 0.5 & 0.8 & 32.3 & 0.8 & 5.0 & 0.0\\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

\begin{landscape}\begin{table}

\caption{\label{tab:unnamed-chunk-3}Conservation assessment for 16 Neotropical taxa of plants and animals based on three datasets. IUCN: global red list assessment obtained from www.iucn.org; GBIF Raw: Preliminary conservation assessment based on IUCN Criterion B using ConR and the raw dataset from GBIF; GBIF filtered: Preliminary conservation assessment based on IUCN Criterion B using ConR and the filtered dataset. Only taxa with at least on species evaluated by IUCN shown.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{l>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{IUCN} & \multicolumn{3}{c}{GBIF Raw} & \multicolumn{5}{c}{GBIF Filtered} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7} \cmidrule(l{3pt}r{3pt}){8-12}
Taxon & n taxa & Evaluated [\%] & Threatened [\%] & n taxa & Threatened [\%] & Match with IUCN [\%] & n taxa & Threatened [\%] & Match with IUCN [\%] & EOO change compared to raw [\%] & AOO change compared to raw [\%]\\
\midrule
Arhynchobatidae & 37 & 51.3 & 17.9 & 39 & 35.9 & 45.0 & 39 & 41.0 & 40.0 & -21.0 & -13.6\\
Dipsadidae & 519 & 68.0 & 8.8 & 637 & 58.4 & 63.0 & 597 & 60.0 & 61.2 & -2.1 & -15.6\\
Harengula & 4 & 100.0 & 0.0 & 4 & 0.0 & 100.0 & 4 & 0.0 & 100.0 & -20.3 & -13.2\\
Conchocarpus & 4 & 8.7 & 0.0 & 46 & 63.0 & 100.0 & 45 & 62.2 & 100.0 & -12.0 & -6.6\\
Gaylussacia & 2 & 3.3 & 0.0 & 61 & 59.0 & 50.0 & 58 & 60.3 & 50.0 & -20.7 & -8.2\\
\addlinespace
Iridaceae & 13 & 2.3 & 0.2 & 531 & 64.4 & 50.0 & 466 & 62.9 & 62.5 & -17.5 & -12.3\\
Lepismium & 6 & 100.0 & 0.0 & 6 & 16.7 & 83.3 & 6 & 16.7 & 83.3 & -33.9 & -7.9\\
Pilosocereus & 41 & 80.9 & 19.1 & 47 & 55.3 & 73.7 & 46 & 56.5 & 71.1 & -8.5 & -5.8\\
Tillandsia & 54 & 11.6 & 6.0 & 464 & 61.4 & 85.2 & 453 & 62.7 & 83.3 & -12.8 & -9.8\\
Tocoyena & 3 & 13.6 & 4.5 & 22 & 31.8 & 66.7 & 21 & 38.1 & 66.7 & -14.7 & -8.8\\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

\newpage{}

\hypertarget{figures}{%
\section{Figures}\label{figures}}

\begin{figure}
\includegraphics[width=\linewidth]{./output/figure_number_of_records} \caption{The difference of records available between the raw and filtered dataset, in a 100x100 km grid across the Neotropics}\label{fig:total}
\end{figure}

\begin{figure}
\includegraphics[width=\linewidth]{./output/figure_number_of_records_split} \caption{The geographic pattern of the occurrence records flagged by the automated tests of this study. The concentration of records flagged by the basis of record test and individual count test in Central A,erica and southern North America, suggest some specific problem with a contributing dataset from the region. Only filters that flagged at least one record are shown.}\label{fig:split}
\end{figure}

\begin{figure}
\includegraphics[width=\linewidth]{./output/figure_number_species_richness_difference} \caption{The difference in species richness derived from the raw and filtered datasets (raw - filtered), for 16 taxa of plants and animals.}\label{fig:speciesrichness}
\end{figure}

\newpage{}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Anderson2016}{}%
Anderson, R. P., Araújo, M., Guisan, A., Lobo, J. M., Martínez-Meyer, E., Peterson, T., \& Soberón, J. (2016). \emph{Final Report of the Task Group on GBIF Data Fitness for Use in Distribution Modelling - Are species occurrence data in global online repositories fit for modeling species distributions? The case of the Global Biodiversity Information Facility (GBIF)} (p. 27pp). GBIF.

\leavevmode\hypertarget{ref-Bachman2011}{}%
Bachman, S. P., Moat, J., Hill, A., Torre, J. de la, \& Scott, B. (2011). Supporting Red List threat assessments with GeoCAT: geospatial conservation assessment tool. \emph{ZooKeys}, \emph{150}, 117--126. \url{https://doi.org/10.3897/zookeys.150.2109}

\leavevmode\hypertarget{ref-Chamberlain2016}{}%
Chamberlain, S. (2016). scrubr: Clean Biological Occurrence Records. Retrieved from \url{https://cran.r-project.org/package=scrubr}

\leavevmode\hypertarget{ref-Chamberlain2018}{}%
Chamberlain, S. (2018). \emph{rredlist: 'IUCN' Red List Client}. Retrieved from \url{https://cran.r-project.org/package=rredlist}

\leavevmode\hypertarget{ref-Chamberlain2017}{}%
Chamberlain, S. A. (2017). rgbif: Interface to the Global Biodiversity Information Facility API. R package version 0.9.9. Retrieved from \url{https://github.com/ropensci/rgbif}

\leavevmode\hypertarget{ref-Cosiaux2018}{}%
Cosiaux, A., Gardiner, L. M., Stauffer, F. W., Bachman, S. P., Sonké, B., Baker, W. J., \& Couvreur, T. L. (2018). Low extinction risk for an important plant resource: Conservation assessments of continental African palms (Arecaceae/Palmae). \emph{Biological Conservation}, \emph{221}, 323--333. \url{https://doi.org/10.1016/j.biocon.2018.02.025}

\leavevmode\hypertarget{ref-Dauby2017}{}%
Dauby, G., Stévart, T., Droissart, V., Cosiaux, A., Deblauwe, V., Simo-Droissart, M., \ldots{} Couvreur, T. L. P. (2017). ConR: An R package to assist large-scale multispecies preliminary conservation assessments using distribution data. \emph{Ecology and Evolution}, \emph{7}(24), 11292--11303. \url{https://doi.org/10.1002/ece3.3704}

\leavevmode\hypertarget{ref-Garnier2018}{}%
Garnier, S. (2018). \emph{viridis: Default color maps from 'matplotlib'}. Retrieved from \url{https://cran.r-project.org/package=viridis}

\leavevmode\hypertarget{ref-GBIForg2019}{}%
GBIF.org. (2019a). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.sojrfp}

\leavevmode\hypertarget{ref-GBIForg2019a}{}%
GBIF.org. (2019b). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.bx0jjw}

\leavevmode\hypertarget{ref-GBIForg2019b}{}%
GBIF.org. (2019c). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.zv6kuq}

\leavevmode\hypertarget{ref-GBIForg2019c}{}%
GBIF.org. (2019d). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.uutyb6}

\leavevmode\hypertarget{ref-GBIForg2019d}{}%
GBIF.org. (2019e). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.8hnzfo}

\leavevmode\hypertarget{ref-GBIForg2019e}{}%
GBIF.org. (2019f). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.zznjbv}

\leavevmode\hypertarget{ref-GBIForg2019f}{}%
GBIF.org. (2019g). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.zjjpmh}

\leavevmode\hypertarget{ref-GBIForg2019g}{}%
GBIF.org. (2019h). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.4srw8a}

\leavevmode\hypertarget{ref-GBIForg2019h}{}%
GBIF.org. (2019i). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.rpkjsh}

\leavevmode\hypertarget{ref-GBIForg2019i}{}%
GBIF.org. (2019j). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.762543}

\leavevmode\hypertarget{ref-GBIForg2019j}{}%
GBIF.org. (2019k). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.nmzgi9}

\leavevmode\hypertarget{ref-GBIForg2019k}{}%
GBIF.org. (2019l). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.scmkx5}

\leavevmode\hypertarget{ref-GBIForg2019l}{}%
GBIF.org. (2019m). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.d34gos}

\leavevmode\hypertarget{ref-GBIForg2019m}{}%
GBIF.org. (2019n). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.6bzfz4}

\leavevmode\hypertarget{ref-GBIForg2019n}{}%
GBIF.org. (2019o). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.wkwque}

\leavevmode\hypertarget{ref-GBIForg2019o}{}%
GBIF.org. (2019p). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.zj2cyj}

\leavevmode\hypertarget{ref-Gueta2016}{}%
Gueta, T., \& Carmel, Y. (2016). Quantifying the value of user-level data cleaning for big data: A case study using mammal distribution models. \emph{Ecological Informatics}, \emph{34}, 139--145. \url{https://doi.org/10.1016/j.ecoinf.2016.06.001}

\leavevmode\hypertarget{ref-Hijmans2019}{}%
Hijmans, R. J. (2019). raster: Geographic data analysis and modeling. Retrieved from \url{https://cran.r-project.org/package=raster}

\leavevmode\hypertarget{ref-IUCN2017}{}%
IUCN Standards and Petitions Subcommittee. (2017). \emph{Guidelines for Using the IUCN Red List - Categories and Criteria. Version 13. Prepared by the Standards and Petitions Subcommittee. Downloadable from http://www.iucnredlist.org/documents/RedListGuidelines.pdf} (pp. 1--60).

\leavevmode\hypertarget{ref-Jin2020}{}%
Jin, J., \& Yang, J. (2020). BDcleaner: A workflow for cleaning taxonomic and geographic errors in occurrence data archived in biodiversity databases. \emph{Global Ecology and Conservation}, \emph{21}, e00852. \url{https://doi.org/10.1016/j.gecco.2019.e00852}

\leavevmode\hypertarget{ref-Maldonado2015}{}%
Maldonado, C., Molina, C. I., Zizka, A., Persson, C., Taylor, C. M., Albán, J., \ldots{} Antonelli, A. (2015). Estimating species diversity and distribution in the era of Big Data: to what extent can we trust public databases? \emph{Global Ecology and Biogeography}, \emph{24}(8), 973--984. \url{https://doi.org/10.1111/geb.12326}

\leavevmode\hypertarget{ref-Morrone2014}{}%
Morrone, J. J. (2014). Biogeographical regionalisation of the Neotropical region. \emph{Zootaxa}, \emph{3782}(1), 1. \url{https://doi.org/10.11646/zootaxa.3782.1.1}

\leavevmode\hypertarget{ref-NicLughadha2019}{}%
Nic Lughadha, E., Walker, B. E., Canteiro, C., Chadburn, H., Davis, A. P., Hargreaves, S., \ldots{} Rivers, M. C. (2019). The use and misuse of herbarium specimens in evaluating plant extinction risks. \emph{Philosophical Transactions of the Royal Society B: Biological Sciences}, \emph{374}(1763), 20170402. \url{https://doi.org/10.1098/rstb.2017.0402}

\leavevmode\hypertarget{ref-Ooms2014}{}%
Ooms, J. (2014). The jsonlite Package: A Practical and Consistent Mapping Between JSON Data and R Objects. \emph{arXiv}. Retrieved from \url{https://arxiv.org/abs/1403.2805}

\leavevmode\hypertarget{ref-Ooms2019}{}%
Ooms, J. (2019). \emph{writexl: Export Data Frames to Excel 'xlsx' Format}. Retrieved from \url{https://cran.r-project.org/package=writexl}

\leavevmode\hypertarget{ref-Pelletier2018}{}%
Pelletier, T. A., Carstens, B. C., Tank, D. C., Sullivan, J., \& Espíndola, A. (2018). Predicting plant conservation priorities on a global scale. \emph{Proceedings of the National Academy of Sciences}, \emph{115}(51), 13027--13032. \url{https://doi.org/10.1073/pnas.1804098115}

\leavevmode\hypertarget{ref-rcoreteam2019}{}%
R Core Team. (2019). R: A language and environment for statistical computing. Austria, Vienna: R Foundation for Statistical Computing. Retrieved from \url{https://www.r-project.org/}

\leavevmode\hypertarget{ref-Rivers2011}{}%
Rivers, M. C., Taylor, L., Brummitt, N. a., Meagher, T. R., Roberts, D. L., \& Lughadha, E. N. (2011). How many herbarium specimens are needed to detect threatened species? \emph{Biological Conservation}, \emph{144}(10), 2541--2547. \url{https://doi.org/10.1016/j.biocon.2011.07.014}

\leavevmode\hypertarget{ref-Robertson2016}{}%
Robertson, M. P., Visser, V., \& Hui, C. (2016). Biogeo: an R package for assessing and improving data quality of occurrence record datasets. \emph{Ecography}, \emph{39}, 394--401. \url{https://doi.org/10.1111/ecog.02118}

\leavevmode\hypertarget{ref-Schmidt2017}{}%
Schmidt, M., Zizka, A., Traoré, S., Ataholo, M., Chatelain, C., Daget, P., \ldots{} Thiombiano, A. (2017). Diversity, distribution and preliminary conservation status of the flora of Burkina Faso. \emph{Phytotaxa Monographs}, \emph{304}(1), 1--215.

\leavevmode\hypertarget{ref-Stevart2019}{}%
Stévart, T., Dauby, G., Lowry, P. P., Blach-Overgaard, A., Droissart, V., Harris, D. J., \ldots{} Couvreur, T. L. P. (2019). A third of the tropical African flora is potentially threatened with extinction. \emph{Science Advances}, \emph{5}(11), eaax9444. \url{https://doi.org/10.1126/sciadv.aax9444}

\leavevmode\hypertarget{ref-Topel2016}{}%
Töpel, M., Zizka, A., Calió, M. F., Scharn, R., Silvestro, D., \& Antonelli, A. (2016). SpeciesGeoCoder: Fast Categorization of Species Occurrences for Analyses of Biodiversity, Biogeography, Ecology, and Evolution. \emph{Systematic Biology}, \emph{66}(2), syw064. \url{https://doi.org/10.1093/sysbio/syw064}

\leavevmode\hypertarget{ref-Wickham2018}{}%
Wickham, H. (2018). tidyverse: Easily install and load the 'Tidyverse'. Retrieved from \url{https://cran.r-project.org/package=tidyverse}

\leavevmode\hypertarget{ref-Yesson2007}{}%
Yesson, C., Brewer, P. W., Sutton, T., Caithness, N., Pahwa, J. S., Burgess, M., \ldots{} Culham, A. (2007). How Global Is the Global Biodiversity Information Facility? \emph{PLoS ONE}, \emph{2}(11), e1124. \url{https://doi.org/10.1371/journal.pone.0001124}

\leavevmode\hypertarget{ref-Zizka2019a}{}%
Zizka, A., Azevedo, J., Leme, E., Neves, B., Ferreira, A., Caceres, D., \& Zizka, G. (2019). Biogeography and conservation status of the pineapple family ( Bromeliaceae ), (June), 1--13. \url{https://doi.org/10.1111/ddi.13004}

\leavevmode\hypertarget{ref-Zizka2019}{}%
Zizka, A., Silvestro, D., Andermann, T., Azevedo, J., Duarte Ritter, C., Edler, D., \ldots{} Antonelli, A. (2019). \textless{}scp\textgreater{}CoordinateCleaner\textless{}/scp\textgreater{} : Standardized cleaning of occurrence records from biological collection databases. \emph{Methods in Ecology and Evolution}, \emph{10}(5), 744--751. \url{https://doi.org/10.1111/2041-210X.13152}

\end{document}

### 2020-01-09 01_download_data_from_gbif.R

@@ -1,0 +1,84 @@
+# A script to download data from gbif
+
+library(tidyverse)
+library(rgbif)
+library(writexl)
+
+taxon <- c("Diogenidae", 
+           "Entomobrydiae",
+           "Neanuridae",
+           "Tityus",
+           "Arhynchobatidae",
+           "Dipsadidae",
+           "Harengula",
+           "Opisthonema",
+           "Thozetella",
+           "Conchocarpus",
+           "Eugenia",
+           "Gaylussacia",
+           "Harpalyce",
+           "Iridaceae",
+           "Lepismium",
+           "Oocephalus",
+           "Pilosocereus",
+           "Prosthechea",
+           "Tillandsia",
+           "Tocoyena")
+
+taxon_rank <- c("family",
+                "family",
+                "family",
+                "genus",
+                "family",
+                "family",
+                "genus",
+                "genus",
+                "genus",
+                "genus",
+                "genus",
+                "genus",
+                "genus",
+                "family",
+                "genus",
+                "genus",
+                "genus",
+                "genus",
+                "genus",
+                "genus")
+
+
+for(i in 1:length(taxon)){
+  print(i)
+  #Use the name_suggest function to get the gbif taxon key
+  tax_key <- name_suggest(q = taxon[i], rank = taxon_rank[i])
+  
+  if(length(tax_key) > 0){
+    #Sometimes groups have multiple taxon keys, in this case three, so we will check how many records are available for them
+    lapply(tax_key$key, "occ_count")
+    
+    #Here the firsrt one is relevant, check for your group!
+    tax_key <- tax_key$key[1]
+    # 
+    # # Download data for Neotropics
+    # ## The extent of Morrone's 2014 bioregionalization
+    # study_a <-"POLYGON((-34.7 32.8, -117.2 32.8, -117.2 -55.8, -34.7 -55.8, -34.7 32.8))"
+    # 
+    # dat  <- occ_search(taxonKey = tax_key, return = "data", hasCoordinate = T,
+    #                    geometry = study_a, limit = 250000)
+    # write_xlsx(dat, paste("input/", "redownload_", taxon[i], ".xlsx", sep = ""))
+    
+    # THis is a similar download to get the API
+    occ_download(sprintf("taxonKey = %s", tax_key),
+                 "hasCoordinate = TRUE",
+                 "decimalLatitude <= 35",
+                 "decimalLongitude <= -33",
[…]

### 2020-01-09 02_automated_cleaning.R

@@ -1,0 +1,116 @@
+# A script to combine the data from gbif and run the metadata and automated filtering
+
+# libraries
+library(tidyverse)
+library(countrycode)
+library(CoordinateCleaner)
+
+# load and combine data
+li <- list.files("input/gbif/", pattern = "_")
+
+dat <- data.frame()
+
+for(i in 1:length(li)){
+  print(i)
+  sub <- read_delim(paste("input/gbif/", li[i], "/occurrence.txt", sep = ""), delim = "\t") %>% 
+    mutate_all(as.character)
+  if(nrow(sub) > 0){
+    sub$taxon <- str_split(li[i], pattern = "_")[[1]][1]
+    sub <- select(sub, taxon, decimalLongitude, decimalLatitude,
+                  coordinateUncertaintyInMeters, 
+                  countryCode, gbifID, species, class, family, genus, taxonRank,
+                  year, basisOfRecord, individualCount)
+    dat <- bind_rows(dat,sub)
+  }
+}
+
+
+
+# meta data filtering
+dat <- dat %>% 
+  mutate(coordinateUncertaintyInMeters = parse_number(coordinateUncertaintyInMeters)) %>% 
+  mutate(decimalLongitude = round(parse_number(decimalLongitude), 4)) %>% 
+  mutate(decimalLatitude = round(parse_number(decimalLatitude), 4)) %>%  
+  filter(!is.na(species)) %>% # remove entires without species information
+  filter(decimalLongitude >= -120)%>% 
+  mutate(individualCount = ifelse(individualCount == TRUE, NA, individualCount)) %>%
+  mutate(individualCount =  ifelse(individualCount == FALSE, NA, individualCount)) %>% 
+  mutate(coordinate_missing =!is.na(decimalLongitude) | !is.na(decimalLatitude)) %>% 
+  mutate(coordinate_precision = coordinateUncertaintyInMeters/1000 <= 100 | is.na(coordinateUncertaintyInMeters)) %>% 
+  mutate(coordinate_base = basisOfRecord == "HUMAN_OBSERVATION" | 
+           basisOfRecord == "OBSERVATION" |
+           basisOfRecord == "PRESERVED_SPECIMEN" | 
+           is.na(basisOfRecord)) %>% 
+  mutate(individual_count = (individualCount > 0 | is.na(individualCount)) &
+           (individualCount <= 99 | is.na(individualCount))) %>% 
[…]

### 2020-01-09 03_automated_conservation_Assessment.R

@@ -1,0 +1,189 @@
+# Automated conservation assessment
+
+library(ConR)
+library(tidyverse)
+library(rredlist)
+library(jsonlite)
+
+# load and prepare data
+dat <- read_csv("output/all_records.csv") %>% 
+  filter(!is.na(decimalLongitude)) %>% 
+  filter(!is.na(decimalLatitude))
+
+dat_cl <-  dat %>% 
+  filter(summary)
+
+# Automated assessment
+## raw
+rw_in <- dat%>%
+  dplyr::select(ddlat = decimalLatitude,
+                ddlon = decimalLongitude, 
+                tax = species) %>% 
+  distinct() %>% 
+  filter(!is.na(tax))
+
+rw_in <- rw_in %>% filter(ddlat >-80) # there is one erroneous record that makes the automated assessment crash
+
+## filtered
+cl_in <- dat_cl%>%
+  dplyr::select(ddlat = decimalLatitude,
+                ddlon = decimalLongitude, 
+                tax = species) %>% 
+  distinct()
+
+
+# AUtomated assessment
+
+## raw
+splist <- unique(rw_in$tax)
+
+for(i in 1:length(splist)){
+  print(paste(i, length(splist), sep = "/"))
+  sub <- rw_in %>% filter(tax == splist[i])
+  
+  out <- IUCN.eval(sub)
+  
+  if(i == 1){
+    write_csv(out, "output/automated_assessment_raw.csv")
+  }else{
+    write_csv(out, "output/automated_assessment_raw.csv", append = TRUE)
+  }
+}
+
+## filtered
+splist <- unique(cl_in$tax)
+
+for(i in 1:length(splist)){
+  print(paste(i, length(splist), sep = "/"))
+  sub <- cl_in %>% filter(tax == splist[i])
+  
+  out <- IUCN.eval(sub)
+  
+  if(i == 1){
+    write_csv(out, "output/automated_assessment_clean.csv")
+  }else{
+    write_csv(out, "output/automated_assessment_clean.csv", append = TRUE)
+  }
+}
+
+
+# IUCN assessment for the species
+iucn.key <- "01524b67f4972521acd1ded2d8b3858e7fedc7da5fd75b8bb2c5456ea18b01ba"
+
+sp.list <- rw_in$tax %>% 
+  as.character() %>% 
+  unique()
+
+#get conservation status from IUCN
+for(i in 1:length(sp.list)){
+  print(i)
[…]

### 2020-01-09 04_figure_number_of_occurrences.R

@@ -1,0 +1,160 @@
+# a script to visualize the raw and cleaned records
+library(tidyverse)
+library(raster)
+library(viridis)
+library(ggthemes)
+library(speciesgeocodeR)
+
+# load data
+dat <- read_csv("output/all_records.csv") %>% 
+  filter(!is.na(decimalLongitude)) %>% 
+  filter(!is.na(decimalLatitude))
+
+dat_cl <-  dat %>% 
+  filter(summary)
+
+be <- raster("input/ABROCOMIDAE_ABROCOMIDAE.tif")
+
+# define projections
+behr <- '+proj=cea +lon_0=0 +lat_ts=30 +x_0=0 +y_0=0 +datum=WGS84 +ellps=WGS84 +units=m +no_defs'
+wgs1984 <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
+
+
+# Figure one - total
+# convert data
+pts_rw <- dat[, c("decimalLongitude", "decimalLatitude")]%>%
+  SpatialPoints(proj4string = CRS(wgs1984))%>%
+  spTransform(behr)
+pts_rw <- data.frame(accepted_name_species = dat$species,
+                  coordinates(pts_rw))
+
+pts_cl <- dat_cl[, c("decimalLongitude", "decimalLatitude")]%>%
+  SpatialPoints(proj4string = CRS(wgs1984))%>%
+  spTransform(behr)
+pts_cl <- data.frame(accepted_name_species = dat_cl$species,
+                     coordinates(pts_cl))
+
+# generate raster with species occurrence
+abu_rw <- RichnessGrid(x = pts_rw, ras = be, type = "abu")
+abu_cl <- RichnessGrid(x = pts_cl, ras = be, type = "abu")
+
+plo <- abu_rw - abu_cl
+
+plo <-data.frame(rasterToPoints(plo))
+# 
+# plo_rw <-data.frame(rasterToPoints(abu_rw))%>%
+#   # filter(layer > 0 ) %>% 
+#   mutate(dataset = "Raw")
+# 
+# plo_cl <-data.frame(rasterToPoints(abu_cl))%>%
+#   # filter(layer > 0 ) %>% 
+#   mutate(dataset = "Filtered")
+# 
+# plo <- bind_rows(plo_rw, plo_cl)
+
+# background map
+world.inp  <- suppressWarnings(rnaturalearth::ne_download(scale = 50, 
+                                                          type = 'land', 
+                                                          category = 'physical',
[…]

### 2020-01-09 05_supplementary_table_S1.R

@@ -1,0 +1,30 @@
+# A script to generate supplementary table 1, with the absolute numbers for each flag
+library(tidyverse)
+
+tab2 <- read_csv("output/summary_table_flags.csv") %>%  
+  mutate(fraction = round(fraction * 100,1))
+
+tab2 <- tab2[,c(1:3,ncol(tab2),4:(ncol(tab2) -1))]
+
+names(tab2) <- c("Taxon",
+                 "Total number of records",
+                 "Total number of flags",
+                 "Fraction of records flagged",
+                 "Missing coordinates",
+                 "Coordinate precision",
+                 "Basis of record",
+                 "Number of individuals collected",
+                 "Collection year",
+                 "Identification level",
+                 "Invalid records",
+                 "Equal lat/lon",
+                 "Zero coordinates",
+                 "Capital coordinates",
+                 "Political centroids",
+                 "Sea coordinates",
+                 "Urban areas",
+                 "GBIF headquaters",
+                 "Biodiversity instiutions",
+                 "Duplicates")
+
+write_csv(tab2, "output/supplement_absolute_numbers_flags.csv")

### 2020-01-09 06_species_richness_maps.R

@@ -1,0 +1,133 @@
+# a script to visualize the raw and cleaned records
+library(tidyverse)
+library(raster)
+library(viridis)
+library(ggthemes)
+library(speciesgeocodeR)
+
+# load data
+dat <- read_csv("output/all_records.csv") %>% 
+  filter(!is.na(decimalLongitude)) %>% 
+  filter(!is.na(decimalLatitude))
+
+dat_cl <-  dat %>% 
+  filter(summary)
+
+be <- raster("input/ABROCOMIDAE_ABROCOMIDAE.tif")
+
+# define projections
+behr <- '+proj=cea +lon_0=0 +lat_ts=30 +x_0=0 +y_0=0 +datum=WGS84 +ellps=WGS84 +units=m +no_defs'
+wgs1984 <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
+
+# background map
+world.inp  <- suppressWarnings(rnaturalearth::ne_download(scale = 50, 
+                                                          type = 'land', 
+                                                          category = 'physical',
+                                                          load = TRUE))
+world.behr <- spTransform(world.inp, CRS(behr)) %>% fortify()
+
+#cleaned data for main mansucript
+plo <- dat %>%
+  filter(summary) %>% 
+  select(species, taxon, decimalLongitude, decimalLatitude)
+
+plo <- split(plo, f = plo$taxon)
+
+plo <-  lapply(plo, 
+               function(k){
+                 pts <- k[, c("decimalLongitude", "decimalLatitude")]%>%
+                   SpatialPoints(proj4string = CRS(wgs1984))%>%
+                   spTransform(behr) %>% 
+                   coordinates()
+                 pts <-  data.frame(species = k$species,
+                                    pts)
+                 out <- pts %>% 
+                   RichnessGrid(ras = be, type = "spnum") %>% 
+                   rasterToPoints() %>% 
+                   data.frame()})
+
+plo_cl <- bind_rows(plo, .id = "taxon")
+names(plo_cl)[4] <- "layer_cl"
+
+# unfiltered for the supplement
+plo <- dat %>%
+  select(species, taxon, decimalLongitude, decimalLatitude)
+
+plo <- split(plo, f = plo$taxon)
+
+plo <-  lapply(plo, 
+               function(k){
+                 pts <- k[, c("decimalLongitude", "decimalLatitude")]%>%
+                   SpatialPoints(proj4string = CRS(wgs1984))%>%
[…]

### 2020-01-09 07_supplement_per_group_richness.R

@@ -1,0 +1,124 @@
+# a script to generate the suppemebnt for each group
+library(tidyverse)
+library(raster)
+library(viridis)
+library(ggthemes)
+library(speciesgeocodeR)
+
+# load data
+dat <- read_csv("output/all_records.csv") %>% 
+  filter(!is.na(decimalLongitude)) %>% 
+  filter(!is.na(decimalLatitude))
+
+dat_cl <-  dat %>% 
+  filter(summary)
+
+# explore what is going on with the high filter fraction taxa
+# dat %>%  filter(taxon == "Tityus") %>%  group_by(basisOfRecord, coordinate_base) %>% summarize(num = n()) %>%
+#   group_by(coordinate_base) %>% mutate(frac = num / sum(num))
+# dat %>%  filter(taxon == "Dipsadidae") %>%  group_by(basisOfRecord, coordinate_base)  %>% summarize(num = n()) %>%
+#   group_by(coordinate_base) %>% mutate(frac = num / sum(num))
+# dat %>%  filter(taxon == "Prosthechea") %>%  group_by(basisOfRecord, coordinate_base)  %>% summarize(num = n()) %>%
+#   group_by(coordinate_base) %>% mutate(frac = num / sum(num))
+# dat %>%  filter(taxon == "Tillandsia") %>%  group_by(basisOfRecord, coordinate_base)  %>% summarize(num = n()) %>%
+#   group_by(coordinate_base) %>% mutate(frac = num / sum(num))
+# 
+# dat %>%  filter(taxon == "Dipsadidae") %>%  group_by(year, record_age)  %>% summarize(num = n()) %>%
+#   group_by(record_age) %>% mutate(frac = num / sum(num)) %>%  View()
+# 
+# dat %>%  filter(taxon == "Diogenidae") %>%  group_by(individualCount, individual_count)  %>% summarize(num = n()) %>%
+#   group_by(individual_count) %>% mutate(frac = num / sum(num)) %>%  View()
+# dat %>%  filter(taxon == "Harengula") %>%  group_by(individualCount, individual_count)  %>% summarize(num = n()) %>%
+#   group_by(individual_count) %>% mutate(frac = num / sum(num)) %>%  View()
+# 
+# dat %>%  filter(taxon == "Tityus") %>%  group_by(taxonRank, record_id) %>% summarize(num = n()) %>%
+#   group_by(record_id) %>% mutate(frac = num / sum(num))
[…]

### 2020-01-27 manuscript.Rmd
version for co-authors

@@
 - Alexander Zizka^1,2^,
-- Alice Calvente^3^,
 - Fernanda Antunes Carvalho^4,5^,
-- Ana Paula Alves da Silva,
+- Alice Calvente^4,5^,
+- Ana Paula Alves da Silva^4,5^,
 - Andressa Cabral,
-- Arthur de Souza Soares,
-- Eduardo Calisto Tomaz,
-- Fernanda Gondim Lambert Moreira,
+- Arthur de Souza Soares^4,5^,
+- Eduardo Calisto Tomaz^4^,
+- Fernanda Gondim Lambert Moreira^4,5^,
 - Filipe Serrano,
@@
 - Mabel Rocio Baez-Lizarazo,
-- Mariana Ramos Fantinati,
+- Mariana Ramos Fantinati^4,5^,
 - Matheus Colli-Silva,
@@
 - Alexandre Antonelli^6^
-#bibliography: course_natal.bib
+bibliography: course_natal.bib
 csl: jbi-3.csl
 output:
-  word_document:
-  #bookdown::pdf_document2:
+  #word_document:
+  bookdown::pdf_document2:
     toc: false
@@
     fig_caption: true
-   # number_sections: false
-   # keep_tex: yes
+    number_sections: false
+    keep_tex: yes
 header-includes: |
@@
 2. Naturalis Biodiversity Center, Leiden, The Netherlands
-3. Laboratório de Botânica Sistemática, Departamento de Botânica e Zoologia, Centro de Biociências, Universidade Federal do Rio Grande do Norte.
-4. Departamento de Genética, Ecologia e Evolução, Instituto de Ciências Biológicas, Universidade Federal de Minas Gerais, Av. Antônio Carlos 6627 - Pampulha, 31270-901 - Belo Horizonte - MG - Brasil
+3. Departamento de Genética, Ecologia e Evolução, Instituto de Ciências Biológicas, Universidade Federal de Minas Gerais, Av. Antônio Carlos 6627 - Pampulha, 31270-901 - Belo Horizonte - MG - Brasil
+4. Laboratório de Botânica Sistemática, Departamento de Botânica e Zoologia, Centro de Biociências, Universidade Federal do Rio Grande do Norte.
+5. Programa de pós graduação em Sistemática e Evolução, Universidade Federal do Rio Grande do Norte, Natal, Brasil
+6. Gothenburg Global Biodiversity Centre, University of Gothenburg, Box 461, 405 30 Gothenburg, Sweden
+7. Department for Biological and Environmental Sciences, University of Gothenburg, Box 461, 405 30 Gothenburg, Sweden
+8. Royal Botanic Gardens Kew, TW9 3AE, Richmond, Surrey, United Kingdom
 
@@
 Automatic flagging and subsequent filtering are a scalable and reproducible means to identify potentially problematic records in 'big data' datasets of species geographic occurrences from public databases (e.g., www.gbif.org).
-However, it is unclear how big the data loss by filtering is, if similar test should be applied across taxonomic groups, and what the effect of filtering is for downstream analyses.
-Here, we obtained occurrence records for 16 taxa including plants and animals as well as terrestrial and marine groups from www.gbif.org and documented the effect of 13 recently proposed filters in each group.
+However, it is unclear how much data is lost by filtering is, if similar test should be applied across taxonomic groups, and what is the effect of filtering for downstream analyses.
+Here, we obtained occurrence records for 16 taxa including plants and animals (terrestrial and marine groups) from www.gbif.org and documented the effect of 13 recently proposed filters in each group.
 We evaluated the effect of filtering on the inference of species richness patterns and the accuracy of automated conservation assessments as examples for downstream analyses.
@@
 
-Manual data cleaning based on expert knowledge can detect these issues, but is only feasible on small taxonomic or geographic scales and is difficult to reproduce.
+Manual data cleaning based on expert knowledge can detect these issues, but is only feasible on small taxonomic or geographic scales and is time-consuming and difficult to reproduce.
 Automated flagging methods to identify potentially problematic records have been proposed as a scalable alternative for datasets with millions of records from many different taxa.
 Those methods are usually based on geographic gazetteers [e.g., @Chamberlain2016; @Zizka2019; @Jin2020] or on additional data, such as environmental variables [@Robertson2016].
-Additionally, filtering procedures based on record meta-data, for instance collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biogeographic research [@Zizka2019].
+Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biogeographic research [@Zizka2019].
 While these methods can identify common errors, it is unclear if errors are equally common across taxonomic groups, and which errors are most severe for downstream analyses.
@@
 Automated assessments using these methods have been used to speed up global [@Stevart2019; @Zizka2019a] and regional [@Schmidt2017; @Cosiaux2018] Red List assessments.
-Since especially EOO is sensitive to individual records with erroneous coordinates automated conservation assessments might be prone to bias by erroneous records, especially, if the number of records is low, as is the case for many species from the tropical regions of Earth.
+Since especially EOO is sensitive to individual records with erroneous coordinates automated conservation assessments might be prone to bias by erroneous records, especially, if the number of records is low, as is the case for many species from the tropical regions on Earth.
 
@@
 Furthermore, we test the impact of automated filtering procedures on the accuracy of automated conservation assessments compared to full IUCN assessments.
-Specifically, we test a pipeline of 13 tests to automatically flag possibly erroneous records 1) using record meta-data and 2) using geographic gazetteers.
+Specifically, we evaluate a pipeline of 13 automated filters to flag possibly erroneous records 1) using record meta-data and 2) using geographic gazetteers.
 We address thee questions:
@@
 We did all analysis in the R computing environment [@rcoreteam2019], using standard libraries for data handling and visualization [@Wickham2018; @Garnier2018; @Ooms2014; @Ooms2019; @Hijmans2019].
-All analysis scripts are available from https://github.com/idiv-biodiversity/effects_of_automated_cleaning.
+All analysis scripts are available from https://github.com/idiv-biodiversity/effects_of_automated_cleaning, or from the Appendix 1 in the electronic supplement of this study.
 
 ## Choice of study taxa
-This study is the outcome of a workshop at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering taxonomists and biogeographers working with different taxonomic groups of plants and animals across the Neotropics.
+This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering taxonomists and biogeographers working with different taxonomic groups of plants and animals across the Neotropics.
 Each participant analysed geographic occurrence data from their taxonomic group of interest and provided an assessment of the results for this group.
-Hence, we chose the study groups based on the expertise of the participants which enabled us to compare the results across the tree of life (at the coast of an arbitrary choice of taxa and taxonomic ranks).
+Hence, we chose the study groups based on the expertise of the participants which enabled us to compare the results across the tree of life (at the cost of an arbitrary choice of taxa and taxonomic ranks).
 We obtained public occurrence records for 16 taxa, including one plant family, eight plant genera, one genus of fungi, two families and one genus of terrestrial arthropods, one family of snakes, one family of skates and one genus of fish (Table 1).
@@
 We chose www.gbif.org for this study, since it is the largest and most widely used provider of species occurrence data and covers the entire tree of life.
-We downloaded only records with geographic coordinates, and limited the study area to latitudes smaller than 33$^\circ$ N and longitudes smaller than 35$^\circ$ W and larger than 120$^\circ$ W reflecting our area of expertise [the Neotropics, @Morrone2014].
+We downloaded only records with geographic coordinates, and limited the study area to latitudes smaller than 33$^\circ$ N and longitudes smaller than 35$^\circ$ W and larger than 120$^\circ$ W reflecting the Neotropics, our area of expertise [@Morrone2014].All included taxa are restricted within the Neotropics (natural populations) except Arhynchobatidae, Iridaceae and Neanuridae for which only the Neotropical occurrence was explored.
 
@@
 We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the raw data downloaded from GBIF ("raw", hereafter) using meta-data for those records for which it was available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise and often reflect different land-use types); and 4) records with less than 1 and more than 99 individual records.
-Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location.
+Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e.
+test for duplicated records.
 In a second step, we used the `clean_coordinates` function of the `CoordinateCleaner v2.0-11` package [@Zizka2019] with default options to flag errors that are common to biological collections, including: sea coordinates, zero coordinates, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions ("filtered", hereafter).
@@
 `ConR` estimates the EOO, AOO, and the number of locations, and then suggests a preliminary conservation status based on Criterion B of the global IUCN Red List.
-While these assessments are preliminary [see @IUCN2017], they can be a proxy for full assessments and are used by the IUCN to speed up full assessments.
+While these assessments are preliminary [see @IUCN2017], they can be a proxy used by the IUCN to speed up full assessments.
 We then benchmarked the preliminary conservation assessments with the global IUCN Red List assessments, which we obtained from www.iucn.org via the `rredlist v.0.5.0` package [@Chamberlain2018].
@@
 
-We found a total of `r nrow(dat)` occurrence records, with a median of `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per study group and `r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per species (Table 3).
+We found a total of `r nrow(dat)` occurrence records, with a median of `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per study group and `r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per species (Table 3, Appendix 2).
 We found most records for `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(taxon) %>%  unlist()` (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs) %>%  unlist()`) and fewest for *`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(taxon) %>%  unlist()`* (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(nrecs) %>%  unlist()`).
@@
 Our automated tests flagged a total of `r sum(!dat$summary)` records (Fig.
-\ref{fig:total}), with a median of `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per group .
+\ref{fig:total}), with a median of `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per group.
 Overall, the most important test was for duplicated records (on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!.dpl)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per taxonomic group).
@@
 
-Geographically, the records flagged by the basis of record and individuals collected tests were concentrated in Central America and southern North America, and a relative high number of records flagged due to their proximity to the centroids of political entities were located on Caribbean islands (Fig.
+Geographically, the records flagged by the basis of record and individuals collected tests were concentrated in Central America and southern North America, and a relatively high number of records flagged due to their proximity to the centroids of political entities were located on Caribbean islands (Fig.
 \ref{fig:split}).
@@
 <!-- For the discussion check those individual groups and see what exactly was flagged here. -->
-We found IUCN assessments for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>% nrow()` species that were also included in our distribution data from 10 of our study groups (Table 4, Supplementary material 3).
+We found IUCN assessments for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>% nrow()` species that were also included in our distribution data from 10 of our study groups (Table 4, Appendix 3).
 The fraction of species evaluated varied among the study group, with a maximum of `r tab3 %>% select(iucn_frac_eval) %>% unlist() %>%  max()`% for *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>% slice(1L) %>% select(taxon) %>%  unlist()`* and *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>%  select(taxon) %>% slice(2) %>% unlist()`* and a minimum of `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(iucn_frac_eval) %>%  unlist()`% for `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(taxon) %>%  unlist()` (note that the number of species varied considerably among the groups).
@@
 We found, that automated flagging based on meta-data and automatic tests flagged on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% of the records per taxonomic group.
-The filters for basis of record, duplicates, collection year, and urban areas flag the highest fraction of records (question 1).
+The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (question 1).
 The importance of different tests was similar among taxonomic groups, with few exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (question 2).
@@
 <!-- 1. Which is the most relevant flagging mechanism? -->
-The high proportion of records flagged as duplicates, may have several reasons.
+Several reasons may explain the high proportion of records flagged as duplicates.
 First, the deposition of duplicates from the same specimen hosted at different collections is common practice, especially for plants, where a duplication is easily feasible.
@@
 The importance of each test and the linked thresholds must be judged based on the downstream analyses.
-For instance, records assigned to country centroids might be acceptable for inter-continental comparisons, but can be consider erroneous for species distribution modelling on a local scale.
+For instance, records assigned to country centroids might be acceptable for inter-continental comparisons, but can be considered erroneous for species distribution modelling on a local scale.
 As our results show, adapting automated tests to the geographic study area or the taxonomic study group can be warranted.
@@
 Some notable exceptions stress the need to adapt each test to the taxonomic study group to balance data quality and data availability.
-The high fraction of records flagged by the basis of record test for *Tityus*, Dipsadidae, *Prosthechea* and *Tillandsia*, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups.
+The high fraction of records flagged by the basis of record test for *Tityus*, Dipsadidae, *Prosthechea* and *Tillandsia*, were mostly caused by a high number of records in these groups based on unknown collection methods (See Appendix 4 for f the individual groups), which might be caused by the contribution of specific datasets lacking this information for these groups.
 The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection activity already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700.
@@
 They are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700.
+This emphasizes the importance to better train researchers and students to manage biodiversity databases.
 The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae, *Oocephalus*, and *Gaylussacia*, might be related to horticulture since at least some species of this group are commonly cultivated as ornamentals.
@@
 # Supplementary material
-Appendix 1 - Table S1, the absolute number of flagged records per taxonomic group and test
+Appendix 1 - Analyses scripts
 
-Appendix 2 - Taxon specific richness maps and comments.
+Appendix 2 - Table S1, the absolute number of flagged records per taxonomic group and test
 
 Appendix 3 - The full results of the conservation assessment
+
+Appendix 4 - Taxon specific richness maps and comments.
 
@@
 # Tables
-<!-- A table briefly describing the tests we ran -->
-```{r tabletests, warning=FALSE}
-tab0 <- read_xlsx("figures_tables/table_test_performed.xlsx")
-
-kable(tab0, booktabs = T, escape = T, format="latex",
-        caption = "The automated filters used in this study.")%>%
-  kable_styling(font_size = 9, latex_options =c("striped", "hold_position")) %>%
-   column_spec(1, width = "2cm", bold = FALSE, italic = FALSE)%>%
-   column_spec(2, width = "2.5cm", bold = FALSE, italic = FALSE)%>%
-   column_spec(3, width = "10cm", bold = FALSE, italic = FALSE)
-```
 
@@
    column_spec(4:5, width = "2.5cm", bold = FALSE, italic = FALSE)
+```
+
+<!-- A table briefly describing the tests we ran -->
+```{r tabletests, warning=FALSE}
+tab0 <- read_xlsx("figures_tables/table_test_performed.xlsx")
+
+kable(tab0, booktabs = T, escape = T, format="latex",
+        caption = "The automated filters used in this study.")%>%
+  kable_styling(font_size = 9, latex_options =c("striped", "hold_position")) %>%
+   column_spec(1, width = "2cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(2, width = "2.5cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(3, width = "10cm", bold = FALSE, italic = FALSE)
 ```

### 2020-01-27 manuscript.tex
version for co-authors

@@
   pdftitle={Automated filtering of species occurrence data can identify problems, but needs to be customized},
-  pdfauthor={Alexander Zizka1,2,; Alice Calvente3,; Fernanda Antunes Carvalho4,5,; Ana Paula Alves da Silva,; Andressa Cabral,; Arthur de Souza Soares,; Eduardo Calisto Tomaz,; Fernanda Gondim Lambert Moreira,; Filipe Serrano,; Jéssica Fernanda Ramos Coelho,; Mabel Rocio Baez-Lizarazo,; Mariana Ramos Fantinati,; Matheus Colli-Silva,; Moabe Ferreira Fernandes,; Nathália Michelly da Cunha Santos,; Paolla G.
+  pdfauthor={Alexander Zizka1,2,; Fernanda Antunes Carvalho4,5,; Alice Calvente4,5,; Ana Paula Alves da Silva4,5,; Andressa Cabral,; Arthur de Souza Soares4,5,; Eduardo Calisto Tomaz4,; Fernanda Gondim Lambert Moreira4,5,; Filipe Serrano,; Jéssica Fernanda Ramos Coelho,; Mabel Rocio Baez-Lizarazo,; Mariana Ramos Fantinati4,5,; Matheus Colli-Silva,; Moabe Ferreira Fernandes,; Nathália Michelly da Cunha Santos,; Paolla G.
 C.
@@
 \title{Automated filtering of species occurrence data can identify problems, but needs to be customized}
-\author{Alexander Zizka\textsuperscript{1,2}, \and Alice Calvente\textsuperscript{3}, \and Fernanda Antunes Carvalho\textsuperscript{4,5}, \and Ana Paula Alves da Silva, \and Andressa Cabral, \and Arthur de Souza Soares, \and Eduardo Calisto Tomaz, \and Fernanda Gondim Lambert Moreira, \and Filipe Serrano, \and Jéssica Fernanda Ramos Coelho, \and Mabel Rocio Baez-Lizarazo, \and Mariana Ramos Fantinati, \and Matheus Colli-Silva, \and Moabe Ferreira Fernandes, \and Nathália Michelly da Cunha Santos, \and Paolla G.
+\author{Alexander Zizka\textsuperscript{1,2}, \and Fernanda Antunes Carvalho\textsuperscript{4,5}, \and Alice Calvente\textsuperscript{4,5}, \and Ana Paula Alves da Silva\textsuperscript{4,5}, \and Andressa Cabral, \and Arthur de Souza Soares\textsuperscript{4,5}, \and Eduardo Calisto Tomaz\textsuperscript{4}, \and Fernanda Gondim Lambert Moreira\textsuperscript{4,5}, \and Filipe Serrano, \and Jéssica Fernanda Ramos Coelho, \and Mabel Rocio Baez-Lizarazo, \and Mariana Ramos Fantinati\textsuperscript{4,5}, \and Matheus Colli-Silva, \and Moabe Ferreira Fernandes, \and Nathália Michelly da Cunha Santos, \and Paolla G.
 C.
@@
 \item
+  Departamento de Genética, Ecologia e Evolução, Instituto de Ciências Biológicas, Universidade Federal de Minas Gerais, Av. Antônio Carlos 6627 - Pampulha, 31270-901 - Belo Horizonte - MG - Brasil
+\item
   Laboratório de Botânica Sistemática, Departamento de Botânica e Zoologia, Centro de Biociências, Universidade Federal do Rio Grande do Norte.
 \item
-  Departamento de Genética, Ecologia e Evolução, Instituto de Ciências Biológicas, Universidade Federal de Minas Gerais, Av. Antônio Carlos 6627 - Pampulha, 31270-901 - Belo Horizonte - MG - Brasil
+  Programa de pós graduação em Sistemática e Evolução, Universidade Federal do Rio Grande do Norte, Natal, Brasil
+\item
+  Gothenburg Global Biodiversity Centre, University of Gothenburg, Box 461, 405 30 Gothenburg, Sweden
+\item
+  Department for Biological and Environmental Sciences, University of Gothenburg, Box 461, 405 30 Gothenburg, Sweden
+\item
+  Royal Botanic Gardens Kew, TW9 3AE, Richmond, Surrey, United Kingdom
 \end{enumerate}
@@
 Automatic flagging and subsequent filtering are a scalable and reproducible means to identify potentially problematic records in `big data' datasets of species geographic occurrences from public databases (e.g., www.gbif.org).
-However, it is unclear how big the data loss by filtering is, if similar test should be applied across taxonomic groups, and what the effect of filtering is for downstream analyses.
-Here, we obtained occurrence records for 16 taxa including plants and animals as well as terrestrial and marine groups from www.gbif.org and documented the effect of 13 recently proposed filters in each group.
+However, it is unclear how much data is lost by filtering is, if similar test should be applied across taxonomic groups, and what is the effect of filtering for downstream analyses.
+Here, we obtained occurrence records for 16 taxa including plants and animals (terrestrial and marine groups) from www.gbif.org and documented the effect of 13 recently proposed filters in each group.
 We evaluated the effect of filtering on the inference of species richness patterns and the accuracy of automated conservation assessments as examples for downstream analyses.
@@
 
-Manual data cleaning based on expert knowledge can detect these issues, but is only feasible on small taxonomic or geographic scales and is difficult to reproduce.
+Manual data cleaning based on expert knowledge can detect these issues, but is only feasible on small taxonomic or geographic scales and is time-consuming and difficult to reproduce.
 Automated flagging methods to identify potentially problematic records have been proposed as a scalable alternative for datasets with millions of records from many different taxa.
 Those methods are usually based on geographic gazetteers (e.g., Chamberlain, 2016; Jin \& Yang, 2020; Zizka, Silvestro, et al., 2019) or on additional data, such as environmental variables (Robertson, Visser, \& Hui, 2016).
-Additionally, filtering procedures based on record meta-data, for instance collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biogeographic research (Zizka, Silvestro, et al., 2019).
+Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biogeographic research (Zizka, Silvestro, et al., 2019).
 While these methods can identify common errors, it is unclear if errors are equally common across taxonomic groups, and which errors are most severe for downstream analyses.
@@
 Automated assessments using these methods have been used to speed up global (Stévart et al., 2019; Zizka, Azevedo, et al., 2019) and regional (Cosiaux et al., 2018; Schmidt et al., 2017) Red List assessments.
-Since especially EOO is sensitive to individual records with erroneous coordinates automated conservation assessments might be prone to bias by erroneous records, especially, if the number of records is low, as is the case for many species from the tropical regions of Earth.
+Since especially EOO is sensitive to individual records with erroneous coordinates automated conservation assessments might be prone to bias by erroneous records, especially, if the number of records is low, as is the case for many species from the tropical regions on Earth.
 
@@
 We did all analysis in the R computing environment (R Core Team, 2019), using standard libraries for data handling and visualization (Garnier, 2018; Hijmans, 2019; Ooms, 2014, 2019; Wickham, 2018).
-All analysis scripts are available from \url{https://github.com/idiv-biodiversity/effects_of_automated_cleaning}.
+All analysis scripts are available from \url{https://github.com/idiv-biodiversity/effects_of_automated_cleaning}, or from the Appendix 1 in the electronic supplement of this study.
 
@@
 
-This study is the outcome of a workshop at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering taxonomists and biogeographers working with different taxonomic groups of plants and animals across the Neotropics.
+This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering taxonomists and biogeographers working with different taxonomic groups of plants and animals across the Neotropics.
 Each participant analysed geographic occurrence data from their taxonomic group of interest and provided an assessment of the results for this group.
-Hence, we chose the study groups based on the expertise of the participants which enabled us to compare the results across the tree of life (at the coast of an arbitrary choice of taxa and taxonomic ranks).
+Hence, we chose the study groups based on the expertise of the participants which enabled us to compare the results across the tree of life (at the cost of an arbitrary choice of taxa and taxonomic ranks).
 We obtained public occurrence records for 16 taxa, including one plant family, eight plant genera, one genus of fungi, two families and one genus of terrestrial arthropods, one family of snakes, one family of skates and one genus of fish (Table 1).
@@
 We chose www.gbif.org for this study, since it is the largest and most widely used provider of species occurrence data and covers the entire tree of life.
-We downloaded only records with geographic coordinates, and limited the study area to latitudes smaller than 33\(^\circ\) N and longitudes smaller than 35\(^\circ\) W and larger than 120\(^\circ\) W reflecting our area of expertise (the Neotropics, Morrone, 2014).
+We downloaded only records with geographic coordinates, and limited the study area to latitudes smaller than 33\(^\circ\) N and longitudes smaller than 35\(^\circ\) W and larger than 120\(^\circ\) W reflecting the Neotropics, our area of expertise (Morrone, 2014).All included taxa are restricted within the Neotropics (natural populations) except Arhynchobatidae, Iridaceae and Neanuridae for which only the Neotropical occurrence was explored.
 
@@
 (2019) and first filtered the raw data downloaded from GBIF (``raw'', hereafter) using meta-data for those records for which it was available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise and often reflect different land-use types); and 4) records with less than 1 and more than 99 individual records.
-Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location.
+Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e.~test for duplicated records.
 In a second step, we used the \texttt{clean\_coordinates} function of the \texttt{CoordinateCleaner\ v2.0-11} package (Zizka, Silvestro, et al., 2019) with default options to flag errors that are common to biological collections, including: sea coordinates, zero coordinates, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions (``filtered'', hereafter).
@@
 \texttt{ConR} estimates the EOO, AOO, and the number of locations, and then suggests a preliminary conservation status based on Criterion B of the global IUCN Red List.
-While these assessments are preliminary (see IUCN Standards and Petitions Subcommittee, 2017), they can be a proxy for full assessments and are used by the IUCN to speed up full assessments.
+While these assessments are preliminary (see IUCN Standards and Petitions Subcommittee, 2017), they can be a proxy used by the IUCN to speed up full assessments.
 We then benchmarked the preliminary conservation assessments with the global IUCN Red List assessments, which we obtained from www.iucn.org via the \texttt{rredlist\ v.0.5.0} package (Chamberlain, 2018).
@@
 
-We found a total of 215,262 occurrence records, with a median of 3,460 records per study group and 11 records per species (Table 3).
+We found a total of 215,262 occurrence records, with a median of 3,460 records per study group and 11 records per species (Table 3, Appendix 2).
 We found most records for Dipsadidae (64,249) and fewest for \emph{Thozetella} (51).
@@
 Our automated tests flagged a total of 88,648 records (Fig.
-\ref{fig:total}), with a median of 40.3\% per group .
+\ref{fig:total}), with a median of 40.3\% per group.
 Overall, the most important test was for duplicated records (on average 32.9\% per taxonomic group).
@@
 
-We found IUCN assessments for 575 species that were also included in our distribution data from 10 of our study groups (Table 4, Supplementary material 3).
+We found IUCN assessments for 575 species that were also included in our distribution data from 10 of our study groups (Table 4, Appendix 3).
 The fraction of species evaluated varied among the study group, with a maximum of 100\% for \emph{Harengula} and \emph{Lepismium} and a minimum of 2.3\% for Iridaceae (note that the number of species varied considerably among the groups).
@@
 We found, that automated flagging based on meta-data and automatic tests flagged on average 40.3\% of the records per taxonomic group.
-The filters for basis of record, duplicates, collection year, and urban areas flag the highest fraction of records (question 1).
+The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (question 1).
 The importance of different tests was similar among taxonomic groups, with few exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (question 2).
@@
 
-The high proportion of records flagged as duplicates, may have several reasons.
+Several reasons may explain the high proportion of records flagged as duplicates.
 First, the deposition of duplicates from the same specimen hosted at different collections is common practice, especially for plants, where a duplication is easily feasible.
@@
 Some notable exceptions stress the need to adapt each test to the taxonomic study group to balance data quality and data availability.
-The high fraction of records flagged by the basis of record test for \emph{Tityus}, Dipsadidae, \emph{Prosthechea} and \emph{Tillandsia}, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups.
+The high fraction of records flagged by the basis of record test for \emph{Tityus}, Dipsadidae, \emph{Prosthechea} and \emph{Tillandsia}, were mostly caused by a high number of records in these groups based on unknown collection methods (See Appendix 4 for f the individual groups), which might be caused by the contribution of specific datasets lacking this information for these groups.
 The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection activity already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700.
@@
 They are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700.
+This emphasizes the importance to better train researchers and students to manage biodiversity databases.
 The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae, \emph{Oocephalus}, and \emph{Gaylussacia}, might be related to horticulture since at least some species of this group are commonly cultivated as ornamentals.
@@
 
-Appendix 1 - Table S1, the absolute number of flagged records per taxonomic group and test
+Appendix 1 - Analyses scripts
 
-Appendix 2 - Taxon specific richness maps and comments.
+Appendix 2 - Table S1, the absolute number of flagged records per taxonomic group and test
 
-Appendix 3 - Table S2, the full results of the conservation assessment
+Appendix 3 - The full results of the conservation assessment
+
+Appendix 4 - Taxon specific richness maps and comments.
 
@@
 \section{Tables}\label{tables}}
-
-\begin{table}[!h]
-
-\caption{\label{tab:tabletests}The automated filters used in this study.}
-\centering
-\fontsize{9}{11}\selectfont
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{10cm}}
-\toprule
-Test & Type & Rationale\\
-\midrule
-\rowcolor{gray!6}  Basis of record & Meta-data & Records might be unsuitable or unreliable for certain analysis dependent on their source, e.g. "fossil" or "unknown"\\
-Collection year & Meta-data & Coordinates from old records are more likely to be imprecise or erroneous coordinates since they are derived from  geo-referencing based on the locality description.
-Since is more problematic for older records, since names or borders of places may change\\
-\rowcolor{gray!6}  Coordinate precision & Meta-data & Records may be unsuitable for a study if their precision is lower than the studies analysis scale\\
-Identification level & Meta-data & Records may be unsuitable if they are not identified to species level.\\
-\rowcolor{gray!6}  Individual count & Meta-data & Records may be unsuitable if the number of recorded individuals is 0 (record of absence) or if the count is to high, as this is often related to records from barcoding or indicative of data entry problems.\\
-\addlinespace
-Biodiversity institutions & Gazetteer-based & Records may have coordinatesat the location of biodiversity institutions, e.g.
-because they were erroneously entered with the physical location of the specimen or because they represent individuals from captivity or horticulture\\
-\rowcolor{gray!6}  Duplicates & Gazetteer-based & Duplicated records may add unnecessary computational burden to analyses\\
-Equal lat/lon & Gazetteer-based & Coordinates with equal latitude and longitude are usually indicative of data entry errors\\
-\rowcolor{gray!6}  Capitals & Gazetteer-based & Records may be assigned to the coordinates of countries capitals based on a vague locality description\\
-Political centroids & Gazetteer-based & Records may be assigned to the coordinates of the centroids of political entities based on a vague locality description\\
-\addlinespace
-\rowcolor{gray!6}  Sea & Gazetteer-based & Coordinates from terrestrial organisms in the sea are usually indicative of data entry errors, e.g. swapped latitude and longitude\\
-Urban areas & Gazetteer-based & Records from urban areas are not necessarily errors, but often represent imprecise records automatically geo-referenced from vague locality descriptions or old records from different land-use types\\
-\rowcolor{gray!6}  Zeros & Gazetteer-based & Coordinates with plain zeros are often indicative of data entry errors\\
-\bottomrule
-\end{tabular}
-\end{table}
 
@@
 \end{landscape}
+
+\begin{table}[!h]
+
+\caption{\label{tab:tabletests}The automated filters used in this study.}
+\centering
+\fontsize{9}{11}\selectfont
+\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{10cm}}
+\toprule
+Test & Type & Rationale\\
+\midrule
+\rowcolor{gray!6}  Basis of record & Meta-data & Records might be unsuitable or unreliable for certain analysis dependent on their source, e.g. "fossil" or "unknown"\\
+Collection year & Meta-data & Coordinates from old records are more likely to be imprecise or erroneous coordinates since they are derived from  geo-referencing based on the locality description.
+Since is more problematic for older records, since names or borders of places may change\\
+\rowcolor{gray!6}  Coordinate precision & Meta-data & Records may be unsuitable for a study if their precision is lower than the studies analysis scale\\
+Identification level & Meta-data & Records may be unsuitable if they are not identified to species level.\\
+\rowcolor{gray!6}  Individual count & Meta-data & Records may be unsuitable if the number of recorded individuals is 0 (record of absence) or if the count is to high, as this is often related to records from barcoding or indicative of data entry problems.\\
+\addlinespace
+Biodiversity institutions & Gazetteer-based & Records may have coordinates at the location of biodiversity institutions, e.g.
+because they were erroneously entered with the physical location of the specimen or because they represent individuals from captivity or horticulture\\
+\rowcolor{gray!6}  Duplicates & Gazetteer-based & Duplicated records may add unnecessary computational burden to analyses\\
+Equal lat/lon & Gazetteer-based & Coordinates with equal latitude and longitude are usually indicative of data entry errors\\
+\rowcolor{gray!6}  Capitals & Gazetteer-based & Records may be assigned to the coordinates of countries capitals based on a vague locality description\\
+Political centroids & Gazetteer-based & Records may be assigned to the coordinates of the centroids of political entities based on a vague locality description\\
+\addlinespace
+\rowcolor{gray!6}  Sea & Gazetteer-based & Coordinates from terrestrial organisms in the sea are usually indicative of data entry errors, e.g. swapped latitude and longitude\\
+Urban areas & Gazetteer-based & Records from urban areas are not necessarily errors, but often represent imprecise records automatically geo-referenced from vague locality descriptions or old records from different land-use types\\
+\rowcolor{gray!6}  Zeros & Gazetteer-based & Coordinates with plain zeros are often indicative of data entry errors\\
+\bottomrule
+\end{tabular}
+\end{table}
 

### 2020-02-25 manuscript.Rmd
include antonellis feedback

@@
 ---
-title: "Automated filtering of species occurrence data can identify problems, but needs to be customized"
+title: "No one-size-fits-all solution to clean GBIF"
 author: 
 - Alexander Zizka^1,2^,
-- Fernanda Antunes Carvalho^4,5^,
-- Alice Calvente^4,5^,
-- Ana Paula Alves da Silva^4,5^,
-- Andressa Cabral,
-- Arthur de Souza Soares^4,5^,
+- Fernanda Antunes Carvalho^4^,
+- Alice Calvente^4^,
+- Ana Paula Alves da Silva^4^,
+- Andressa Cabral^5^,
+- Arthur de Souza Soares^4^,
 - Eduardo Calisto Tomaz^4^,
-- Fernanda Gondim Lambert Moreira^4,5^,
-- Filipe Serrano,
-- Jéssica Fernanda Ramos Coelho,
-- Mabel Rocio Baez-Lizarazo,
-- Mariana Ramos Fantinati^4,5^,
-- Matheus Colli-Silva,
-- Moabe Ferreira Fernandes,
-- Nathália Michelly da Cunha Santos,
-- Paolla G. C. Souza,
-- Renata Clícia Santos da Costa,
-- Thais Ferreira Pinto de Araújo,
-- Tiago Andrade Borges Santos,
-- Tiago Luiz Vieira,
-- Valéria Fonseca Vale,
-- Alexandre Antonelli^6^
+- Fernanda Gondim Lambert Moreira^4^,
+- Filipe Cabreirinha Serrano^6^,
+- Jéssica Fernanda Ramos Coelho^4^,
+- Mabel Rocio Baez-Lizarazo,^7^
+- Mariana Ramos Fantinati^4^,
+- Matheus Colli-Silva^5^,
+- Moabe Ferreira Fernandes^8^,
+- Nathália Michelly da Cunha Santos^4^,
+- Paolla Gabryelle Cavalcante de Souza^4^,
+- Renata Clicia dos Santos‐Costa^4^,
+- Thais Ferreira-Araújo^4^,
+- Tiago Andrade Borges Santos^9^,
+- Tiago Luiz Vieira^9^,
+- Valéria Fonseca Vale^4^,
+- Alexandre Antonelli^10,11,12^
 bibliography: course_natal.bib
@@
     keep_tex: yes
+nocite: |
+    @GBIForg2019; @GBIForg2019a; @GBIForg2019b; @GBIForg2019c; @GBIForg2019d; @GBIForg2019e; @GBIForg2019f; @GBIForg2019g; @GBIForg2019h; @GBIForg2019i; @GBIForg2019j; @GBIForg2019k; @GBIForg2019l; @GBIForg2019m; @GBIForg2019n; @GBIForg2019o;@GBIForg2019p;@GBIForg2019q
 header-includes: |
@@
 
-1. sDiv,German Center for Integrative Biodiversity Research (iDiv), Leipzig, Germany
+1. sDiv, German Center for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Leipzig, Germany
 2. Naturalis Biodiversity Center, Leiden, The Netherlands
-3. Departamento de Genética, Ecologia e Evolução, Instituto de Ciências Biológicas, Universidade Federal de Minas Gerais, Av. Antônio Carlos 6627 - Pampulha, 31270-901 - Belo Horizonte - MG - Brasil
-4. Laboratório de Botânica Sistemática, Departamento de Botânica e Zoologia, Centro de Biociências, Universidade Federal do Rio Grande do Norte.
-5. Programa de pós graduação em Sistemática e Evolução, Universidade Federal do Rio Grande do Norte, Natal, Brasil
-6. Gothenburg Global Biodiversity Centre, University of Gothenburg, Box 461, 405 30 Gothenburg, Sweden
-7. Department for Biological and Environmental Sciences, University of Gothenburg, Box 461, 405 30 Gothenburg, Sweden
-8. Royal Botanic Gardens Kew, TW9 3AE, Richmond, Surrey, United Kingdom
+3. Departamento de Genética, Ecologia e Evolução, Universidade Federal de Minas Gerais, Belo Horizonte, Brazil
+4. Departamento de Botânica e Zoologia, Universidade Federal do Rio Grande do Norte, Natal, Brazil
+5. Departamento de Botânica, Universidade de São Paulo, São Paulo, Brazil
+6. Departamento de Ecologia, Universidade de São Paulo, São Paulo, Brazil
+7. Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil
+8. Departamento de Ciências Biológicas, Universidade Estadual de Feira de Santana, Feira de Santana, Brazil
+9. Universidade Estadual de Feira de Santana, Bahia, Brazil
+10. Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden
+11. Department for Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden
+12. Royal Botanic Gardens Kew, Richmond, United Kingdom
 
@@
 # Abstract
-Automatic flagging and subsequent filtering are a scalable and reproducible means to identify potentially problematic records in 'big data' datasets of species geographic occurrences from public databases (e.g., www.gbif.org).
-However, it is unclear how much data is lost by filtering is, if similar test should be applied across taxonomic groups, and what is the effect of filtering for downstream analyses.
-Here, we obtained occurrence records for 16 taxa including plants and animals (terrestrial and marine groups) from www.gbif.org and documented the effect of 13 recently proposed filters in each group.
-We evaluated the effect of filtering on the inference of species richness patterns and the accuracy of automated conservation assessments as examples for downstream analyses.
-We found 29-67% of the records potentially problematic across taxonomic groups.
-The tests for duplicated information, collection year, basis of record as well as urban areas and political centroids were most important.
-While many of the flagged records might not be erroneous per se, they might be overly imprecise, increase uncertainty of downstream analyses and generate unnecessary computational burden.
-Overall, our results stress the importance of (meta-)data exploration for biogeographic research and suggest that automated flagging can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the taxonomic group and geographic area of interest.
+Species occurrence records provide the basis for many biodiversity studies.
+They derive from geo-referenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications.
+Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern.
+Automatic flagging and filtering are a scalable and reproducible means to identify potentially problematic records in datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org).
+However, it is unclear how much data may be lost by filtering, whether the same tests should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses.
+Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 16 Neotropical taxa including plants and animals, terrestrial and marine, downloaded from GBIF.
+We find that 29-67% of the records are potentially erroneous, with large variation across taxonomic groups.
+Tests for duplicated information, collection year, basis of record as well as urban areas and political centroids have the greatest effect.
+While many flagged records might not be *de facto* erroneous, they could be overly imprecise and increase uncertainty in downstream analyses.
+Automated flagging can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus.
+Our results stress the importance of thorough exploration of the meta-data associated with species records for biodiversity research.
 
 # Keywords
-Automated cleaning, automated conservation assessment, data quality, GBIF, Neotropics
+Automated cleaning, automated conservation assessment, data quality, GBIF, Neotropics.
 
@@
 # Introduction
-Publicly available species distribution data have become a crucial resource in biogeographic research.
-In particular, the availability of digitized collections from museums and herbaria, and citizen science observations has increased in the last years.
+Publicly available species distribution data have become a crucial resource in biodiversity research, including studies in ecology, biogeography, systematics and conservation biology.
+In particular, the availability of digitized collections from museums and herbaria, and citizen science observations has increased drastically over the last few years.
 As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.3 billion geo-referenced occurrence records for species from across the globe and the tree of life.
@@
 A central challenge to the use of these publicly available species occurrence data in research are erroneous geographic coordinates [@Anderson2016].
-Errors mostly arise, because public databases integrate records collected with different methodology in different places at different times, often with little centralized curation and meta-data.
-For instance, erroneous coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common [@Maldonado2015; @Yesson2007] and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records from species in captivity [@Zizka2019].
+Errors mostly arise because public databases integrate records collected with different methodologies in different places, at different times; often without centralized curation and only rudimentary meta-data.
+For instance, erroneous coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common [@Maldonado2015; @Yesson2007] and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in captivity [@Zizka2019].
 
-Manual data cleaning based on expert knowledge can detect these issues, but is only feasible on small taxonomic or geographic scales and is time-consuming and difficult to reproduce.
-Automated flagging methods to identify potentially problematic records have been proposed as a scalable alternative for datasets with millions of records from many different taxa.
+Manual data cleaning based on expert knowledge can detect these issues, but it is only feasible on small taxonomic or geographic scales, and it is time-consuming and difficult to reproduce.
+As an alternative, automated flagging methods to identify potentially problematic records have been proposed as a scalable option, as they are able to deal with datasets containing up to millions of records and many different taxa.
 Those methods are usually based on geographic gazetteers [e.g., @Chamberlain2016; @Zizka2019; @Jin2020] or on additional data, such as environmental variables [@Robertson2016].
-Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biogeographic research [@Zizka2019].
-While these methods can identify common errors, it is unclear if errors are equally common across taxonomic groups, and which errors are most severe for downstream analyses.
+Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biodiveristy research [@Zizka2019].
 
 Erroneous records are especially problematic in conservation, where stakes are high.
-Recently, automated conservation assessment methods have been developed to use publicly available occurrence records for conservation, in particular to support the assessment procedures for the global Red List of the International Union for the Conservation of Nature (IUCN) [@Dauby2017; @Bachman2011; @Pelletier2018].
-These methods approximate species' range size, namely the Extent of Occurrence (EOO, which is the area of a convex hull polygon comprising all records of a species), the Area of Occupancy (AOO, which is the sum of the area actually occupied by a species, calculated based on a small-scale regular grid), and the number of locations for a preliminary conservation assessment following IUCN Criterion B.
-Automated assessments using these methods have been used to speed up global [@Stevart2019; @Zizka2019a] and regional [@Schmidt2017; @Cosiaux2018] Red List assessments.
-Since especially EOO is sensitive to individual records with erroneous coordinates automated conservation assessments might be prone to bias by erroneous records, especially, if the number of records is low, as is the case for many species from the tropical regions on Earth.
+Recently proposed methods for automated conservation assessments could support the formal assessment procedures for the global Red List of the International Union for the Conservation of Nature (IUCN) [@Dauby2017; @Bachman2011; @Pelletier2018].
+These methods approximate species' range size, namely the Extent of Occurrence (EOO, which is the area of a convex hull polygon comprising all records of a species), the Area of Occupancy (AOO, which is the sum of the area actually occupied by a species, calculated based on a small-scale regular grid), and the number of locations for a preliminary conservation assessment following IUCN Criterion B ("Geographic range").
+These methods have been used topropose preliminary global [@Stevart2019; @Zizka2019a] and regional [@Schmidt2017; @Cosiaux2018] Red List assessments.
+However, all metrics but especially EOO, are sensitive to individual records with erroneous coordinates **(REFERENCE)**.
+Automated conservation assessments may therefore be biased, particularly if the number of records are few, as it is the case for many tropical species.
 
-Here, we test the effect of automated filtering of species geographic occurrence records on the number of records available in different plant and animal groups.
-Furthermore, we test the impact of automated filtering procedures on the accuracy of automated conservation assessments compared to full IUCN assessments.
-Specifically, we evaluate a pipeline of 13 automated filters to flag possibly erroneous records 1) using record meta-data and 2) using geographic gazetteers.
-We address thee questions:
+While newly proposed methods for automated cleaning of species records hold great promise for biodiversity research, their use across taxonomic groups and datasets remains poorly explored.
+Here, we test the effect of automated filtering of species geographic occurrence records on the number of records available in different groups of animals, funghi, and plants.
+Furthermore, we test the impact of automated filtering procedures for the accuracy of automated conservation assessments compared to full IUCN assessments.
+Specifically, we evaluate a pipeline of 13 automated filters to flag possibly erroneous records by using 1) record meta-data and 2) geographic gazetteers.
+We address three questions:
 
-1. Which is the most relevant filter, i.e. which type of error is most common?
+1. Which filters lead to the biggest loss of data when applied?
 2. Does the importance of individual filters differ among taxonomic groups?
-3. Does automated filtering improve the accuracy of automated conservation assessment compared to the global IUCN Red List?
+3. Does automated filtering improve the accuracy of automated conservation assessments?
 
 # Material and Methods
-We did all analysis in the R computing environment [@rcoreteam2019], using standard libraries for data handling and visualization [@Wickham2018; @Garnier2018; @Ooms2014; @Ooms2019; @Hijmans2019].
-All analysis scripts are available from https://github.com/idiv-biodiversity/effects_of_automated_cleaning, or from the Appendix 1 in the electronic supplement of this study.
-
 ## Choice of study taxa
-This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering taxonomists and biogeographers working with different taxonomic groups of plants and animals across the Neotropics.
-Each participant analysed geographic occurrence data from their taxonomic group of interest and provided an assessment of the results for this group.
-Hence, we chose the study groups based on the expertise of the participants which enabled us to compare the results across the tree of life (at the cost of an arbitrary choice of taxa and taxonomic ranks).
-We obtained public occurrence records for 16 taxa, including one plant family, eight plant genera, one genus of fungi, two families and one genus of terrestrial arthropods, one family of snakes, one family of skates and one genus of fish (Table 1).
+This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering graduate students, taxonomists and biogeographers working with different taxonomic groups of animals, funghi, and plants across the Neotropics.
+Each participant analysed geographic occurrence data from their taxonomic group of interest and provided an informed assessment of the results for their group.
+Hence, we include groups based on the expertise of the participants wrather than following some arbitrary choice of taxa and taxonomic ranks.
+We acknowledge a varying degree in documented expertise and number of years working on each group.
+We obtained public occurrence records for 16 taxa, including one plant family, eight plant genera, one genus of fungi, two families and one genus of terrestrial arthropods, one family of snakes, one family of skates, and one genus of bony fish (Table 1).
 
 ## Species occurrence data
-We downloaded occurrence information for all study groups from www.gbif.org using the `rgbif` v1.4.0 package [@Chamberlain2017] in R [@GBIForg2019; @GBIForg2019a; @GBIForg2019b; @GBIForg2019c; @GBIForg2019d; @GBIForg2019e; @GBIForg2019f; @GBIForg2019g; @GBIForg2019h; @GBIForg2019i; @GBIForg2019j; @GBIForg2019k; @GBIForg2019l; @GBIForg2019m; @GBIForg2019n; @GBIForg2019o].
-We chose www.gbif.org for this study, since it is the largest and most widely used provider of species occurrence data and covers the entire tree of life.
-We downloaded only records with geographic coordinates, and limited the study area to latitudes smaller than 33$^\circ$ N and longitudes smaller than 35$^\circ$ W and larger than 120$^\circ$ W reflecting the Neotropics, our area of expertise [@Morrone2014].All included taxa are restricted within the Neotropics (natural populations) except Arhynchobatidae, Iridaceae and Neanuridae for which only the Neotropical occurrence was explored.
+<!-- @GBIForg2019; @GBIForg2019a; @GBIForg2019b; @GBIForg2019c; @GBIForg2019d; @GBIForg2019e; @GBIForg2019f; @GBIForg2019g; @GBIForg2019h; @GBIForg2019i; @GBIForg2019j; @GBIForg2019k; @GBIForg2019l; @GBIForg2019m; @GBIForg2019n; @GBIForg2019o -->
+
+We downloaded occurrence information for all study groups from www.gbif.org using the `rgbif` v1.4.0 package [@Chamberlain2017] in R (GBIF.org, 2019a-p).
+We downloaded only records with geographic coordinates and limited the study area to latitudes smaller than 33$^\circ$ N and longitudes smaller than 35$^\circ$ W and larger than 120$^\circ$ W reflecting the Neotropics, our main area of expertise [@Morrone2014].
+The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, *Gaylussacia*, Iridaceae, Neanuridae, and *Tillandsia* for which we only obtained the Neotropical occurrences.
 
 ##  Automated cleaning
-We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the raw data downloaded from GBIF ("raw", hereafter) using meta-data for those records for which it was available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise and often reflect different land-use types); and 4) records with less than 1 and more than 99 individual records.
+We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the raw data downloaded from GBIF ("raw", hereafter) using meta-data for those records for which they were available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99.
 Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e.
-test for duplicated records.
-In a second step, we used the `clean_coordinates` function of the `CoordinateCleaner v2.0-11` package [@Zizka2019] with default options to flag errors that are common to biological collections, including: sea coordinates, zero coordinates, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions ("filtered", hereafter).
+test for duplicated records).
+In a second step, we used the `clean_coordinates` function of the `CoordinateCleaner v2.0-11` package [@Zizka2019] with default options to flag errors that are common to biological data sets (“filtered”, hereafter).
+These including: coordinates in the sea for terrestrial taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions.
 See Table 2 for a summary of all flagging tests we used.
@@
 ## Downstream analyses
-We first generated species richness maps using 100x100 km grid cells for the raw and filtered dataset respectively, using the `speciesgeocodeR v2.0-10` package [@Topel2016] and then performed a preliminary conservation assessment for all study groups based on both datasets using the `ConR v1.2.4` package [@Dauby2017].
+We first generated species richness maps using 100x100 km grid cells for the raw and filtered datasets respectively, using the package `speciesgeocodeR v2.0-10` [@Topel2016].
+We then performed an automated conservation assessment for all study groups based on both datasets using the `ConR v1.2.4` package [@Dauby2017].
 `ConR` estimates the EOO, AOO, and the number of locations, and then suggests a preliminary conservation status based on Criterion B of the global IUCN Red List.
 While these assessments are preliminary [see @IUCN2017], they can be a proxy used by the IUCN to speed up full assessments.
-We then benchmarked the preliminary conservation assessments with the global IUCN Red List assessments, which we obtained from www.iucn.org via the `rredlist v.0.5.0` package [@Chamberlain2018].
+We then benchmarked the preliminary conservation assessments against the global IUCN Red List assessments as far as available for the same taxa, which we obtained from www.iucn.org via the `rredlist v.0.5.0` package [@Chamberlain2018].
 
-## Evaluation of the results
-Each specialist evaluated the performance of the raw and cleaned datasets, respectively concerning: 1) the number of flagged records; 2) the importance of specific tests; 3) the accuracy of the overall species richness maps; 4) the accuracy of the preliminary conservation assessment.
+## Evaluation of results
+<!-- This session is exciting and makes the study more reliable and interesting than one just based on data mining, but how this was done is very unclear.
+Did people provide a specific score for those, or how were their assessments recorded?
+I think we need to clarify this, as it’s the weakest part of the paper at the moment.
+We don’t report the results for this evaluation nor what they mean.
+-->
+Each specialist provided an informed assessment on the performance of the raw and cleaned datasets, respectively concerning: 1) the number of flagged records; 2) the importance of specific tests; 3) the accuracy of the overall species richness maps; 4) the accuracy of the preliminary conservation assessments.
+
+We did all analyses in the R computing environment [@rcoreteam2019], using standard libraries for data handling and visualization [@Wickham2018; @Garnier2018; @Ooms2014; @Ooms2019; @Hijmans2019].
+All scripts are available from https://github.com/idiv-biodiversity/effects_of_automated_cleaning, or from the Appendix 1 in the electronic supplement of this study.
 
@@
 
-We found a total of `r nrow(dat)` occurrence records, with a median of `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per study group and `r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per species (Table 3, Appendix 2).
-We found most records for `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(taxon) %>%  unlist()` (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs) %>%  unlist()`) and fewest for *`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(taxon) %>%  unlist()`* (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(nrecs) %>%  unlist()`).
+We retrieved a total of `r nrow(dat)` species occurrence records, with a median of `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per study group and `r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per species (Table 3, Appendix 2).
+We obtained most records for `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(taxon) %>%  unlist()` (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs) %>%  unlist()`) and fewest for *`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(taxon) %>%  unlist()`* (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(nrecs) %>%  unlist()`).
 The species with most records was *`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(species)`* (`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs)`).
@@
 Overall, the most important test was for duplicated records (on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!.dpl)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per taxonomic group).
-The most important filtering steps based on record meta-data were for the basis of records (`r dat %>% group_by(taxon) %>% summarize(num = mean(!coordinate_base)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the collection year (`r dat %>% group_by(taxon) %>% summarize(num = mean(!record_age)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%).
-The most important automated tests were for urban area (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.urb)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%), the capital coordinates (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.cap)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the centroids of political units (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.cen)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%, Table 3, see Supplementary material 1 for the absolute numbers).
-Only few records were flagged by the zero coordinates and biodiversity institutions tests (Fig.
+The filtering steps based on record meta-data that flagged the largest number of records were the basis of records (`r dat %>% group_by(taxon) %>% summarize(num = mean(!coordinate_base)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the collection year (`r dat %>% group_by(taxon) %>% summarize(num = mean(!record_age)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%).
+The most important automated tests were for urban area (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.urb)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%), the capital coordinates (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.cap)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the centroids of political units (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.cen)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%, see Table 3 and Supplementary material 1 for further details and the absolute numbers).
+Only few records were flagged by the zero coordinates and biodiversity institution tests (Fig.
 \ref{fig:split}), and no records were flagged by the sea test.
@@
 In general, the different filters we tested were of similar importance for different study groups.
-There were few outstanding exceptions, including the particularly high proportions of records flagged by the basis of record test for *Tityus* (7.0%), Dipsadidae (5.6%), *Prosthechea* (5.0%) and *Tillandsia* (4.9%), by the collection year for Dipsadidae (11.3%), by the taxonomic identification level for  *Tityus* (1.6%), by the capital coordinates for *Oocephalus* (6.1%) and *Gaylussacia* (3.2%), as well as by the urban areas test for *Oocephalus* (13.3%) and Iridaceae (12.3%).
+There were few outstanding exceptions, including the particularly high proportions of records flagged on the "basis of record test" for *Tityus* (7.0%), Dipsadidae (5.6%), *Prosthechea* (5.0%) and *Tillandsia* (4.9%), by the collection year for Dipsadidae (11.3%), by the taxonomic identification level for  *Tityus* (1.6%), by the capital coordinates for *Oocephalus* (6.1%) and *Gaylussacia* (3.2%), and by the urban areas test for *Oocephalus* (13.3%) and Iridaceae (12.3%).
 
-Geographically, the records flagged by the basis of record and individuals collected tests were concentrated in Central America and southern North America, and a relatively high number of records flagged due to their proximity to the centroids of political entities were located on Caribbean islands (Fig.
+Geographically, the records flagged by the "basis of record" and "individual count" tests were concentrated in Central America and southern North America, and a relatively high number of records flagged due to their proximity to the centroids of political entities were located on Caribbean islands (Fig.
 \ref{fig:split}).
@@
 
-<!-- For the discussion check those individual groups and see what exactly was flagged here. -->
 We found IUCN assessments for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>% nrow()` species that were also included in our distribution data from 10 of our study groups (Table 4, Appendix 3).
-The fraction of species evaluated varied among the study group, with a maximum of `r tab3 %>% select(iucn_frac_eval) %>% unlist() %>%  max()`% for *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>% slice(1L) %>% select(taxon) %>%  unlist()`* and *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>%  select(taxon) %>% slice(2) %>% unlist()`* and a minimum of `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(iucn_frac_eval) %>%  unlist()`% for `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(taxon) %>%  unlist()` (note that the number of species varied considerably among the groups).
+The fraction of species evaluated varied among the study group, with a maximum of `r tab3 %>% select(iucn_frac_eval) %>% unlist() %>%  max()`% for *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>% slice(1L) %>% select(taxon) %>%  unlist()`* and *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>%  select(taxon) %>% slice(2) %>% unlist()`* and a minimum of `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(iucn_frac_eval) %>%  unlist()`% for `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(taxon) %>%  unlist()` (note that the number of species varied considerably among groups).
 The median percentage of species per study group with an IUCN assessment was `r tab3 %>% select(iucn_frac_eval) %>% unlist() %>%  median()`%.
@@
 We obtained automated conservation assessments for `r aa %>% select(flag_Category_CriteriaB) %>% filter(!is.na(flag_Category_CriteriaB)) %>% nrow()` species in the filtered dataset.
-Based on the filtered data the automated conservation assessment evaluated `r sum(aa$flag_status == "possibly_threatened", na.rm = TRUE)` species as possibly threatened (`r round(sum(aa$flag_status == "possibly_threatened", na.rm=TRUE) / aa %>% filter(!is.na(flag_status)) %>% nrow() * 100, 1)`%, CR = `r aa %>% filter(flag_Category_CriteriaB == "CR") %>% nrow()`, EN = `r aa %>% filter(flag_Category_CriteriaB == "EN") %>% nrow()`, VU = `r aa %>% filter(flag_Category_CriteriaB == "VU") %>% nrow()`).
+Based on the filtered data, the automated conservation assessment evaluated `r sum(aa$flag_status == "possibly_threatened", na.rm = TRUE)` species as possibly threatened (`r round(sum(aa$flag_status == "possibly_threatened", na.rm=TRUE) / aa %>% filter(!is.na(flag_status)) %>% nrow() * 100, 1)`%, CR = `r aa %>% filter(flag_Category_CriteriaB == "CR") %>% nrow()`, EN = `r aa %>% filter(flag_Category_CriteriaB == "EN") %>% nrow()`, VU = `r aa %>% filter(flag_Category_CriteriaB == "VU") %>% nrow()`).
 The automated assessment based on the filtered dataset agreed with the IUCN assessment for identifying species as possibly threatened (CR, EN, VU) for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status)) %>% mutate(agreement = iucn_status == flag_status) %>% select(agreement) %>% unlist() %>% sum() ` species (`r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status))  %>% mutate(agreement = iucn_status == flag_status) %>% select(agreement) %>% unlist() %>% mean() %>%  round(3) * 100`%; Table 4).
@@
 # Discussion
-We found, that automated flagging based on meta-data and automatic tests flagged on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% of the records per taxonomic group.
-The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (question 1).
-The importance of different tests was similar among taxonomic groups, with few exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (question 2).
-The results for species richness were similar between the raw and filtered data with some improvements by the filters and we found little impact of filtering on the accuracy of the automated conservation assessments (question 3).
+<!-- add some subheadings to discussion -->
+Automated flagging based on meta-data and automatic tests flagged on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% of the records per taxonomic group.
+The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (Question 1).
+The importance of different tests was similar across taxonomic groups, with few exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (Question 2).
+The results for species richness were similar between the raw and filtered data with some improvements by the filters.
+We found little impact of filtering on the accuracy of the automated conservation assessments (Question 3).
 
@@
 Several reasons may explain the high proportion of records flagged as duplicates.
-First, the deposition of duplicates from the same specimen hosted at different collections is common practice, especially for plants, where a duplication is easily feasible.
-Second, independent collections at similar localities may occur, in particular for local endemics, and third low coordinate precision, for instance based on automated geo-referencing from locality descriptions, may lump records from nearby localities.
-Of course, duplicated records do not represent errors per se, but they represent an unnecessary computational burden, which is often a bottleneck for analyses of datasets with millions of records.
+First, the deposition of duplicates from the same specimen hosted at different collections is common practice, especially for plants, where a specimen duplication is easily feasible.
+Second, independent collections at similar localities may occur, in particular for local endemics.
+Third, low coordinate precision, for instance based on automated geo-referencing from locality descriptions, may lump records from nearby localities.
+Duplicated records do not represent errors per se, but an unnecessary computational burden, which is often a bottleneck for analyses of datasets with millions of records.
 
 The aim of automated filtering is to identify possibly problematic records that are unsuitable for particular downstream analyses.
-Hence our tests include errors in the narrow sense (e.g., zero coordinates, equal lat/lon and seas tests) as well as records that are not necessary erroneous, but have a low precision (e.g., collection year, coordinate precision, centroids of political units) or are unnecessary (duplicates).
-The importance of each test and the linked thresholds must be judged based on the downstream analyses.
+Hence, our tests may identify true errors  (e.g., wrong coordinates such as those with exclusively zeros, equal latitude and longitude and records of terestrial species in the sea) as well as records that are not necessarily erroneous, but have a low precision (e.g., collection year, coordinate precision, centroids of political units) or are redundant for certain analyses (duplicates).
+The importance of each test and the linked thresholds must be judged based on the specific downstream analyses.
 For instance, records assigned to country centroids might be acceptable for inter-continental comparisons, but can be considered erroneous for species distribution modelling on a local scale.
@@
 <!-- 2. Does the importance of individual errors differ among taxonomic groups? -->
-The number of records flagged by individual tests was similar across the study groups, suggesting that similar error types might be relevant for collections of plants and animals and partly also terrestrial and marine organisms.
+The number of records flagged by individual tests was similar across study groups, suggesting that similar error types might be relevant for collections of plants and animals.
 Therefore, the same tests can be used across taxonomic groups.
 Some notable exceptions stress the need to adapt each test to the taxonomic study group to balance data quality and data availability.
-The high fraction of records flagged by the basis of record test for *Tityus*, Dipsadidae, *Prosthechea* and *Tillandsia*, were mostly caused by a high number of records in these groups based on unknown collection methods (See Appendix 4 for f the individual groups), which might be caused by the contribution of specific datasets lacking this information for these groups.
-The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection activity already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700.
-The latter records likely represent a data entry error.
-They are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700.
-This emphasizes the importance to better train researchers and students to manage biodiversity databases.
-The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae, *Oocephalus*, and *Gaylussacia*, might be related to horticulture since at least some species of this group are commonly cultivated as ornamentals.
+The high fraction of records flagged by the "basis of record test" for *Tityus*, Dipsadidae, *Prosthechea* and *Tillandsia*, were mostly caused by a high number of records in these groups based on unknown collection methods (See Appendix 4 for the individual groups), which might be caused by the contribution of specific datasets lacking this information for these groups.
+The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection effort already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700.
+The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700.
+The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae, and *Oocephalus* might be related to horticulture since at least some species in those groups are commonly cultivated as ornamentals.
 
@@
 \ref{fig:speciesrichness}, Supplementary material 1).
-The filter based on political centroids had an important impact on species richness patterns, which is congruent with the results from previous studied [@Maldonado2015].
-Records assigned to country or province centroids are often old records, which are geo-referenced based on vague locality descriptions.
-These records are at the same time more likely to represent dubious species names, since they might be old synonyms or type specimen of species that have only been collected and described once, which are erroneously increasing species numbers.
+The filter based on political centroids had an important impact on species richness patterns, which is congruent with the results from a previous study in the coffee family [@Maldonado2015].
+Records assigned to country or province centroids are often old records, which are geo-referenced at a later point based on vague locality descriptions.
+These records are at the same time more likely to represent dubious species names, since they might be old synonyms or type specimens of species that have only been collected and described once, which are erroneously increasing species numbers.
 
@@
 The robustness of the automated assessment was likely due to the fact that the EOO for most species was large, even after the considerable reduction caused by filtering.
-This might be caused by the structure of our comparison, which only included species that were evaluated by the IUCN Red List (and not considered as Data Deficient) and at the same time have occurrences recorded in GBIF, which biases the dataset towards species with larger ranges, since generally more data for them is available.
+This might be caused by the structure of our comparison, which only included species that were evaluated by the IUCN Red List (and not considered as Data Deficient) and at the same time had occurrences recorded in GBIF.
+Those inclusion criteria are likely to have, biased the datasets towards species with large ranges, since generally more data for them are available.
 The robustness of automated conservation assessments to data quality is encouraging, although these methods are only an approximation, especially for species with few collection records [@Rivers2011].
@@
 # Conclusions
-Our results suggest that between 1/3rd and 2/3rds of the occurrence records obtained from public databases might be unsuitable for downstream biogeographic analyses, dependent on the taxonomic group.
-While the majority of these records might not be erroneous per se, they might be overly imprecise and thereby increase uncertainty of downstream results or put additional computational burden on the analyses.
+Our results suggest that between one to two thirds of the occurrence records obtained from public databases might be unsuitable for downstream biodiversity analyses.
+While the majority of these records might not be erroneous per se, they might be overly imprecise and thereby increase uncertainty of downstream results or add computational burden on big data analyses.
 
-Overall, our results stress the importance of (meta-)data exploration before analyses and suggest that automated flagging can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the specific taxonomic groups and geographic area of interest.
+Our results stress the importance of (meta-)data exploration prior to most biodiversity analyses.
+Automated flagging can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the specific taxonomic groups and geographic area of interest.
+The putative errors we encountered point to the importance to train researchers and students to curate species occurrence datasets and to allow users to provide feedback for particular records, thus contributing to overall accuracy and precision of public biodiversity databases.
 
 # Acknowledgements
-This study enrolled participants of the workshop "Biodiversity data: from field to yield" lead by Alexandre Antonelli and Alexander Zizka through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI.
-We thank the Pró-reitoria de Pesquisa and the Pró-reitoria de Pós-graduação of UFRN for the financial support (edital 02/2016 - internacionalização).
-AZ is thankful for funding by iDiv via the German Research Foundation (DFG FZT 118), specifically through sDiv, the Synthesis Centre of iDiv.
+This study enrolled participants of the workshop "Biodiversity data: from field to yield" led by Alexandre Antonelli and Alexander Zizka through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI.
+We thank the Pró-reitoria de Pesquisa and the Pró-reitoria de Pós-graduação of UFRN for financial support (edital 02/2016 - internacionalização).
+AZ is funded by iDiv via the German Research Foundation (DFG FZT 118), specifically through sDiv, the Synthesis Centre of iDiv.
+AA is supported by the Swedish Research Council, the Knut and Alice Wallenberg Foundation, the Swedish Foundation for Strategic Research and the Royal Botanic Gardens, Kew.
 
@@
 
-Appendix 2 - Table S1, the absolute number of flagged records per taxonomic group and test
+Appendix 2 - Table S1, reporting the absolute number of flagged records per taxonomic group and test
 
@@
         caption = "The study groups and their taxonomy.
-This study includes three marine and 13 terrestrial taxa, 6 of them animals, one group of fungi and 9 plants, belonging to 16 different orders.")%>%
+This study includes three marine and 13 terrestrial taxa, six of them animals, one group of fungi and nine plants, belonging to 16 different orders.")%>%
   landscape()%>%
@@
 From column four onwards the numbers show the percentage of records flagged by the respective test.
-Only tests that flagged at least 0.1\\% of the records in any group are shown.", format.args = list(big.mark = ","))%>%
+Only tests that flagged at least 0.1\\% of the records in any group are shown.
+Individual records can be flagged by multiple tests, therefore the sum of percentge from all tests can supersede the total percentage.", format.args = list(big.mark = ","))%>%
   landscape()%>%
@@
                               "AOO change compared to raw [%]")),
-        caption = "Conservation assessment for 16 Neotropical taxa of plants and animals based on three datasets.
+        caption = "Conservation assessment for 10 Neotropical taxa of plants and animals based on three datasets.
 IUCN: global red list assessment obtained from www.iucn.org; GBIF Raw: Preliminary conservation assessment based on IUCN Criterion B using ConR and the raw dataset from GBIF; GBIF filtered: Preliminary conservation assessment based on IUCN Criterion B using ConR and the filtered dataset.
@@
 <!-- A map with number of species and where they have been removed -->
-```{r total, out.width="\\linewidth", fig.cap=c('The difference of records available between the raw and filtered dataset, in a 100x100 km grid across the Neotropics'), echo=FALSE}
+```{r total, out.width="\\linewidth", fig.cap=c('The difference in records available between the raw and filtered dataset of 16 Neotropical taxa including plants and animals, plotted in a 100x100 km grid across the Neotropics.'), echo=FALSE}
 knitr::include_graphics(path = "./output/figure_number_of_records.jpg")
@@
 
-```{r split, out.width="\\linewidth", fig.cap=c('The geographic pattern of the occurrence records flagged by the automated tests of this study.
-The concentration of records flagged by the basis of record test and individual count test in Central A,erica and southern North America, suggest some specific problem with a contributing dataset from the region.
+```{r split, out.width="\\linewidth", fig.cap=c('Geographic location of the occurrence records flagged by the automated tests  applied in this study.
+The concentration of records flagged on the basis of record test and individual count test in Central America and southern North America.
 Only filters that flagged at least one record are shown.'), echo=FALSE}
@@
 
-```{r speciesrichness, out.width="\\linewidth", fig.cap=c('The difference in species richness derived from the raw and filtered datasets (raw - filtered), for 16 taxa of plants and animals.'), echo=FALSE}
+```{r speciesrichness, out.width="\\linewidth", fig.cap=c('The difference in species richness derived from the raw and filtered datasets (raw - filtered), for 16 taxa of animals, funghi, and plants.'), echo=FALSE}
 knitr::include_graphics(path = "./output/figure_number_species_richness_difference.jpg")

### 2020-02-25 manuscript.tex
include antonellis feedback

@@
-\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
 \PassOptionsToPackage{hyphens}{url}
@@
 \usepackage{setspace}
-\setstretch{1.3}
 \usepackage{amssymb,amsmath}
@@
   \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provides euro and other symbols
-\else % if luatex or xelatex
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
   \usepackage{unicode-math}
@@
 \fi
-% use upquote if available, for straight quotes in verbatim environments
+% Use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 \hypersetup{
-  pdftitle={Automated filtering of species occurrence data can identify problems, but needs to be customized},
-  pdfauthor={Alexander Zizka1,2,; Fernanda Antunes Carvalho4,5,; Alice Calvente4,5,; Ana Paula Alves da Silva4,5,; Andressa Cabral,; Arthur de Souza Soares4,5,; Eduardo Calisto Tomaz4,; Fernanda Gondim Lambert Moreira4,5,; Filipe Serrano,; Jéssica Fernanda Ramos Coelho,; Mabel Rocio Baez-Lizarazo,; Mariana Ramos Fantinati4,5,; Matheus Colli-Silva,; Moabe Ferreira Fernandes,; Nathália Michelly da Cunha Santos,; Paolla G.
-C.
-Souza,; Renata Clícia Santos da Costa,; Thais Ferreira Pinto de Araújo,; Tiago Andrade Borges Santos,; Tiago Luiz Vieira,; Valéria Fonseca Vale,; Alexandre Antonelli6},
-  pdfborder={0 0 0},
-  breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
+  pdftitle={No one-size-fits-all solution to clean GBIF},
+  pdfauthor={Alexander Zizka1,2,; Fernanda Antunes Carvalho4,; Alice Calvente4,; Ana Paula Alves da Silva4,; Andressa Cabral5,; Arthur de Souza Soares4,; Eduardo Calisto Tomaz4,; Fernanda Gondim Lambert Moreira4,; Filipe Cabreirinha Serrano6,; Jéssica Fernanda Ramos Coelho4,; Mabel Rocio Baez-Lizarazo,7; Mariana Ramos Fantinati4,; Matheus Colli-Silva5,; Moabe Ferreira Fernandes8,; Nathália Michelly da Cunha Santos4,; Paolla Gabryelle Cavalcante de Souza4,; Renata Clicia dos Santos‐Costa4,; Thais Ferreira-Araújo4,; Tiago Andrade Borges Santos9,; Tiago Luiz Vieira9,; Valéria Fonseca Vale4,; Alexandre Antonelli10,11,12},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
 \usepackage[margin=1in]{geometry}
 \usepackage{longtable,booktabs}
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
 % Allow footnotes in longtable head/foot
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-2}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-% set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 \usepackage{caption}
@@
 
-\title{Automated filtering of species occurrence data can identify problems, but needs to be customized}
-\author{Alexander Zizka\textsuperscript{1,2}, \and Fernanda Antunes Carvalho\textsuperscript{4,5}, \and Alice Calvente\textsuperscript{4,5}, \and Ana Paula Alves da Silva\textsuperscript{4,5}, \and Andressa Cabral, \and Arthur de Souza Soares\textsuperscript{4,5}, \and Eduardo Calisto Tomaz\textsuperscript{4}, \and Fernanda Gondim Lambert Moreira\textsuperscript{4,5}, \and Filipe Serrano, \and Jéssica Fernanda Ramos Coelho, \and Mabel Rocio Baez-Lizarazo, \and Mariana Ramos Fantinati\textsuperscript{4,5}, \and Matheus Colli-Silva, \and Moabe Ferreira Fernandes, \and Nathália Michelly da Cunha Santos, \and Paolla G.
-C.
-Souza, \and Renata Clícia Santos da Costa, \and Thais Ferreira Pinto de Araújo, \and Tiago Andrade Borges Santos, \and Tiago Luiz Vieira, \and Valéria Fonseca Vale, \and Alexandre Antonelli\textsuperscript{6}}
+\title{No one-size-fits-all solution to clean GBIF}
+\author{Alexander Zizka\textsuperscript{1,2}, \and Fernanda Antunes Carvalho\textsuperscript{4}, \and Alice Calvente\textsuperscript{4}, \and Ana Paula Alves da Silva\textsuperscript{4}, \and Andressa Cabral\textsuperscript{5}, \and Arthur de Souza Soares\textsuperscript{4}, \and Eduardo Calisto Tomaz\textsuperscript{4}, \and Fernanda Gondim Lambert Moreira\textsuperscript{4}, \and Filipe Cabreirinha Serrano\textsuperscript{6}, \and Jéssica Fernanda Ramos Coelho\textsuperscript{4}, \and Mabel Rocio Baez-Lizarazo,\textsuperscript{7} \and Mariana Ramos Fantinati\textsuperscript{4}, \and Matheus Colli-Silva\textsuperscript{5}, \and Moabe Ferreira Fernandes\textsuperscript{8}, \and Nathália Michelly da Cunha Santos\textsuperscript{4}, \and Paolla Gabryelle Cavalcante de Souza\textsuperscript{4}, \and Renata Clicia dos Santos‐Costa\textsuperscript{4}, \and Thais Ferreira-Araújo\textsuperscript{4}, \and Tiago Andrade Borges Santos\textsuperscript{9}, \and Tiago Luiz Vieira\textsuperscript{9}, \and Valéria Fonseca Vale\textsuperscript{4}, \and Alexandre Antonelli\textsuperscript{10,11,12}}
 \date{}
@@
 
+\setstretch{1.3}
 \begin{enumerate}
@@
 \item
-  sDiv,German Center for Integrative Biodiversity Research (iDiv), Leipzig, Germany
+  sDiv, German Center for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Leipzig, Germany
 \item
@@
 \item
-  Departamento de Genética, Ecologia e Evolução, Instituto de Ciências Biológicas, Universidade Federal de Minas Gerais, Av. Antônio Carlos 6627 - Pampulha, 31270-901 - Belo Horizonte - MG - Brasil
+  Departamento de Genética, Ecologia e Evolução, Universidade Federal de Minas Gerais, Belo Horizonte, Brazil
 \item
-  Laboratório de Botânica Sistemática, Departamento de Botânica e Zoologia, Centro de Biociências, Universidade Federal do Rio Grande do Norte.
+  Departamento de Botânica e Zoologia, Universidade Federal do Rio Grande do Norte, Natal, Brazil
 \item
-  Programa de pós graduação em Sistemática e Evolução, Universidade Federal do Rio Grande do Norte, Natal, Brasil
+  Departamento de Botânica, Universidade de São Paulo, São Paulo, Brazil
 \item
-  Gothenburg Global Biodiversity Centre, University of Gothenburg, Box 461, 405 30 Gothenburg, Sweden
+  Departamento de Ecologia, Universidade de São Paulo, São Paulo, Brazil
 \item
-  Department for Biological and Environmental Sciences, University of Gothenburg, Box 461, 405 30 Gothenburg, Sweden
+  Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil
 \item
-  Royal Botanic Gardens Kew, TW9 3AE, Richmond, Surrey, United Kingdom
+  Departamento de Ciências Biológicas, Universidade Estadual de Feira de Santana, Feira de Santana, Brazil
+\item
+  Universidade Estadual de Feira de Santana, Bahia, Brazil
+\item
+  Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden
+\item
+  Department for Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden
+\item
+  Royal Botanic Gardens Kew, Richmond, United Kingdom
 \end{enumerate}
@@
 
-Automatic flagging and subsequent filtering are a scalable and reproducible means to identify potentially problematic records in `big data' datasets of species geographic occurrences from public databases (e.g., www.gbif.org).
-However, it is unclear how much data is lost by filtering is, if similar test should be applied across taxonomic groups, and what is the effect of filtering for downstream analyses.
-Here, we obtained occurrence records for 16 taxa including plants and animals (terrestrial and marine groups) from www.gbif.org and documented the effect of 13 recently proposed filters in each group.
-We evaluated the effect of filtering on the inference of species richness patterns and the accuracy of automated conservation assessments as examples for downstream analyses.
-We found 29-67\% of the records potentially problematic across taxonomic groups.
-The tests for duplicated information, collection year, basis of record as well as urban areas and political centroids were most important.
-While many of the flagged records might not be erroneous per se, they might be overly imprecise, increase uncertainty of downstream analyses and generate unnecessary computational burden.
-Overall, our results stress the importance of (meta-)data exploration for biogeographic research and suggest that automated flagging can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the taxonomic group and geographic area of interest.
+Species occurrence records provide the basis for many biodiversity studies.
+They derive from geo-referenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications.
+Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern.
+Automatic flagging and filtering are a scalable and reproducible means to identify potentially problematic records in datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org).
+However, it is unclear how much data may be lost by filtering, whether the same tests should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses.
+Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 16 Neotropical taxa including plants and animals, terrestrial and marine, downloaded from GBIF.
+We find that 29-67\% of the records are potentially erroneous, with large variation across taxonomic groups.
+Tests for duplicated information, collection year, basis of record as well as urban areas and political centroids have the greatest effect.
+While many flagged records might not be \emph{de facto} erroneous, they could be overly imprecise and increase uncertainty in downstream analyses.
+Automated flagging can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus.
+Our results stress the importance of thorough exploration of the meta-data associated with species records for biodiversity research.
 
@@
 
-Automated cleaning, automated conservation assessment, data quality, GBIF, Neotropics
+Automated cleaning, automated conservation assessment, data quality, GBIF, Neotropics.
 
@@
 
-Publicly available species distribution data have become a crucial resource in biogeographic research.
-In particular, the availability of digitized collections from museums and herbaria, and citizen science observations has increased in the last years.
+Publicly available species distribution data have become a crucial resource in biodiversity research, including studies in ecology, biogeography, systematics and conservation biology.
+In particular, the availability of digitized collections from museums and herbaria, and citizen science observations has increased drastically over the last few years.
 As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.3 billion geo-referenced occurrence records for species from across the globe and the tree of life.
@@
 A central challenge to the use of these publicly available species occurrence data in research are erroneous geographic coordinates (Anderson et al., 2016).
-Errors mostly arise, because public databases integrate records collected with different methodology in different places at different times, often with little centralized curation and meta-data.
-For instance, erroneous coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common (Maldonado et al., 2015; Yesson et al., 2007) and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records from species in captivity (Zizka, Silvestro, et al., 2019).
+Errors mostly arise because public databases integrate records collected with different methodologies in different places, at different times; often without centralized curation and only rudimentary meta-data.
+For instance, erroneous coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common (Maldonado et al., 2015; Yesson et al., 2007) and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in captivity (Zizka, Silvestro, et al., 2019).
 
-Manual data cleaning based on expert knowledge can detect these issues, but is only feasible on small taxonomic or geographic scales and is time-consuming and difficult to reproduce.
-Automated flagging methods to identify potentially problematic records have been proposed as a scalable alternative for datasets with millions of records from many different taxa.
+Manual data cleaning based on expert knowledge can detect these issues, but it is only feasible on small taxonomic or geographic scales, and it is time-consuming and difficult to reproduce.
+As an alternative, automated flagging methods to identify potentially problematic records have been proposed as a scalable option, as they are able to deal with datasets containing up to millions of records and many different taxa.
 Those methods are usually based on geographic gazetteers (e.g., Chamberlain, 2016; Jin \& Yang, 2020; Zizka, Silvestro, et al., 2019) or on additional data, such as environmental variables (Robertson, Visser, \& Hui, 2016).
-Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biogeographic research (Zizka, Silvestro, et al., 2019).
-While these methods can identify common errors, it is unclear if errors are equally common across taxonomic groups, and which errors are most severe for downstream analyses.
+Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biodiveristy research (Zizka, Silvestro, et al., 2019).
 
 Erroneous records are especially problematic in conservation, where stakes are high.
-Recently, automated conservation assessment methods have been developed to use publicly available occurrence records for conservation, in particular to support the assessment procedures for the global Red List of the International Union for the Conservation of Nature (IUCN) (Bachman et al., 2011; Dauby et al., 2017; Pelletier et al., 2018).
-These methods approximate species' range size, namely the Extent of Occurrence (EOO, which is the area of a convex hull polygon comprising all records of a species), the Area of Occupancy (AOO, which is the sum of the area actually occupied by a species, calculated based on a small-scale regular grid), and the number of locations for a preliminary conservation assessment following IUCN Criterion B.
-Automated assessments using these methods have been used to speed up global (Stévart et al., 2019; Zizka, Azevedo, et al., 2019) and regional (Cosiaux et al., 2018; Schmidt et al., 2017) Red List assessments.
-Since especially EOO is sensitive to individual records with erroneous coordinates automated conservation assessments might be prone to bias by erroneous records, especially, if the number of records is low, as is the case for many species from the tropical regions on Earth.
+Recently proposed methods for automated conservation assessments could support the formal assessment procedures for the global Red List of the International Union for the Conservation of Nature (IUCN) (Bachman et al., 2011; Dauby et al., 2017; Pelletier et al., 2018).
+These methods approximate species' range size, namely the Extent of Occurrence (EOO, which is the area of a convex hull polygon comprising all records of a species), the Area of Occupancy (AOO, which is the sum of the area actually occupied by a species, calculated based on a small-scale regular grid), and the number of locations for a preliminary conservation assessment following IUCN Criterion B (``Geographic range'').
+These methods have been used topropose preliminary global (Stévart et al., 2019; Zizka, Azevedo, et al., 2019) and regional (Cosiaux et al., 2018; Schmidt et al., 2017) Red List assessments.
+However, all metrics but especially EOO, are sensitive to individual records with erroneous coordinates \textbf{(REFERENCE)}.
+Automated conservation assessments may therefore be biased, particularly if the number of records are few, as it is the case for many tropical species.
 
-Here, we test the effect of automated filtering of species geographic occurrence records on the number of records available in different plant and animal groups.
-Furthermore, we test the impact of automated filtering procedures on the accuracy of automated conservation assessments compared to full IUCN assessments.
-Specifically, we evaluate a pipeline of 13 automated filters to flag possibly erroneous records 1) using record meta-data and 2) using geographic gazetteers.
-We address thee questions:
+While newly proposed methods for automated cleaning of species records hold great promise for biodiversity research, their use across taxonomic groups and datasets remains poorly explored.
+Here, we test the effect of automated filtering of species geographic occurrence records on the number of records available in different groups of animals, funghi, and plants.
+Furthermore, we test the impact of automated filtering procedures for the accuracy of automated conservation assessments compared to full IUCN assessments.
+Specifically, we evaluate a pipeline of 13 automated filters to flag possibly erroneous records by using 1) record meta-data and 2) geographic gazetteers.
+We address three questions:
 
@@
 \item
-  Which is the most relevant filter, i.e.~which type of error is most common?
+  Which filters lead to the biggest loss of data when applied?
 \item
@@
 \item
-  Does automated filtering improve the accuracy of automated conservation assessment compared to the global IUCN Red List?
+  Does automated filtering improve the accuracy of automated conservation assessments?
 \end{enumerate}
@@
 
-We did all analysis in the R computing environment (R Core Team, 2019), using standard libraries for data handling and visualization (Garnier, 2018; Hijmans, 2019; Ooms, 2014, 2019; Wickham, 2018).
-All analysis scripts are available from \url{https://github.com/idiv-biodiversity/effects_of_automated_cleaning}, or from the Appendix 1 in the electronic supplement of this study.
-
 \hypertarget{choice-of-study-taxa}{%
@@
 
-This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering taxonomists and biogeographers working with different taxonomic groups of plants and animals across the Neotropics.
-Each participant analysed geographic occurrence data from their taxonomic group of interest and provided an assessment of the results for this group.
-Hence, we chose the study groups based on the expertise of the participants which enabled us to compare the results across the tree of life (at the cost of an arbitrary choice of taxa and taxonomic ranks).
-We obtained public occurrence records for 16 taxa, including one plant family, eight plant genera, one genus of fungi, two families and one genus of terrestrial arthropods, one family of snakes, one family of skates and one genus of fish (Table 1).
+This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering graduate students, taxonomists and biogeographers working with different taxonomic groups of animals, funghi, and plants across the Neotropics.
+Each participant analysed geographic occurrence data from their taxonomic group of interest and provided an informed assessment of the results for their group.
+Hence, we include groups based on the expertise of the participants wrather than following some arbitrary choice of taxa and taxonomic ranks.
+We acknowledge a varying degree in documented expertise and number of years working on each group.
+We obtained public occurrence records for 16 taxa, including one plant family, eight plant genera, one genus of fungi, two families and one genus of terrestrial arthropods, one family of snakes, one family of skates, and one genus of bony fish (Table 1).
 
@@
 
-We downloaded occurrence information for all study groups from www.gbif.org using the \texttt{rgbif} v1.4.0 package (Chamberlain, 2017) in R (GBIF.org, 2019a, 2019b, 2019c, 2019d, 2019e, 2019f, 2019g, 2019h, 2019i, 2019j, 2019k, 2019l, 2019m, 2019n, 2019o, 2019p).
-We chose www.gbif.org for this study, since it is the largest and most widely used provider of species occurrence data and covers the entire tree of life.
-We downloaded only records with geographic coordinates, and limited the study area to latitudes smaller than 33\(^\circ\) N and longitudes smaller than 35\(^\circ\) W and larger than 120\(^\circ\) W reflecting the Neotropics, our area of expertise (Morrone, 2014).All included taxa are restricted within the Neotropics (natural populations) except Arhynchobatidae, Iridaceae and Neanuridae for which only the Neotropical occurrence was explored.
+We downloaded occurrence information for all study groups from www.gbif.org using the \texttt{rgbif} v1.4.0 package (Chamberlain, 2017) in R (GBIF.org, 2019a-p).
+We downloaded only records with geographic coordinates and limited the study area to latitudes smaller than 33\(^\circ\) N and longitudes smaller than 35\(^\circ\) W and larger than 120\(^\circ\) W reflecting the Neotropics, our main area of expertise (Morrone, 2014).
+The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, \emph{Gaylussacia}, Iridaceae, Neanuridae, and \emph{Tillandsia} for which we only obtained the Neotropical occurrences.
 
@@
 We followed the cleaning pipeline outlined by Zizka, Silvestro, et al.
-(2019) and first filtered the raw data downloaded from GBIF (``raw'', hereafter) using meta-data for those records for which it was available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise and often reflect different land-use types); and 4) records with less than 1 and more than 99 individual records.
-Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e.~test for duplicated records.
-In a second step, we used the \texttt{clean\_coordinates} function of the \texttt{CoordinateCleaner\ v2.0-11} package (Zizka, Silvestro, et al., 2019) with default options to flag errors that are common to biological collections, including: sea coordinates, zero coordinates, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions (``filtered'', hereafter).
+(2019) and first filtered the raw data downloaded from GBIF (``raw'', hereafter) using meta-data for those records for which they were available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99.
+Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e.~test for duplicated records).
+In a second step, we used the \texttt{clean\_coordinates} function of the \texttt{CoordinateCleaner\ v2.0-11} package (Zizka, Silvestro, et al., 2019) with default options to flag errors that are common to biological data sets (``filtered'', hereafter).
+These including: coordinates in the sea for terrestrial taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions.
 See Table 2 for a summary of all flagging tests we used.
@@
 
-We first generated species richness maps using 100x100 km grid cells for the raw and filtered dataset respectively, using the \texttt{speciesgeocodeR\ v2.0-10} package (Töpel et al., 2016) and then performed a preliminary conservation assessment for all study groups based on both datasets using the \texttt{ConR\ v1.2.4} package (Dauby et al., 2017).
+We first generated species richness maps using 100x100 km grid cells for the raw and filtered datasets respectively, using the package \texttt{speciesgeocodeR\ v2.0-10} (Töpel et al., 2016).
+We then performed an automated conservation assessment for all study groups based on both datasets using the \texttt{ConR\ v1.2.4} package (Dauby et al., 2017).
 \texttt{ConR} estimates the EOO, AOO, and the number of locations, and then suggests a preliminary conservation status based on Criterion B of the global IUCN Red List.
 While these assessments are preliminary (see IUCN Standards and Petitions Subcommittee, 2017), they can be a proxy used by the IUCN to speed up full assessments.
-We then benchmarked the preliminary conservation assessments with the global IUCN Red List assessments, which we obtained from www.iucn.org via the \texttt{rredlist\ v.0.5.0} package (Chamberlain, 2018).
+We then benchmarked the preliminary conservation assessments against the global IUCN Red List assessments as far as available for the same taxa, which we obtained from www.iucn.org via the \texttt{rredlist\ v.0.5.0} package (Chamberlain, 2018).
 
-\hypertarget{evaluation-of-the-results}{%
-\subsection{Evaluation of the results}\label{evaluation-of-the-results}}
+\hypertarget{evaluation-of-results}{%
+\subsection{Evaluation of results}\label{evaluation-of-results}}
 
-Each specialist evaluated the performance of the raw and cleaned datasets, respectively concerning: 1) the number of flagged records; 2) the importance of specific tests; 3) the accuracy of the overall species richness maps; 4) the accuracy of the preliminary conservation assessment.
+Each specialist provided an informed assessment on the performance of the raw and cleaned datasets, respectively concerning: 1) the number of flagged records; 2) the importance of specific tests; 3) the accuracy of the overall species richness maps; 4) the accuracy of the preliminary conservation assessments.
+
+We did all analyses in the R computing environment (R Core Team, 2019), using standard libraries for data handling and visualization (Garnier, 2018; Hijmans, 2019; Ooms, 2014, 2019; Wickham, 2018).
+All scripts are available from \url{https://github.com/idiv-biodiversity/effects_of_automated_cleaning}, or from the Appendix 1 in the electronic supplement of this study.
 
@@
 
-We found a total of 215,262 occurrence records, with a median of 3,460 records per study group and 11 records per species (Table 3, Appendix 2).
-We found most records for Dipsadidae (64,249) and fewest for \emph{Thozetella} (51).
+We retrieved a total of 215,262 species occurrence records, with a median of 3,460 records per study group and 11 records per species (Table 3, Appendix 2).
+We obtained most records for Dipsadidae (64,249) and fewest for \emph{Thozetella} (51).
 The species with most records was \emph{Harengula jaguana} (19,878).
@@
 Overall, the most important test was for duplicated records (on average 32.9\% per taxonomic group).
-The most important filtering steps based on record meta-data were for the basis of records (2.1\%) and the collection year (3.3\%).
-The most important automated tests were for urban area (5\%), the capital coordinates (1.1\%) and the centroids of political units (0.5\%, Table 3, see Supplementary material 1 for the absolute numbers).
-Only few records were flagged by the zero coordinates and biodiversity institutions tests (Fig.
+The filtering steps based on record meta-data that flagged the largest number of records were the basis of records (2.1\%) and the collection year (3.3\%).
+The most important automated tests were for urban area (5\%), the capital coordinates (1.1\%) and the centroids of political units (0.5\%, see Table 3 and Supplementary material 1 for further details and the absolute numbers).
+Only few records were flagged by the zero coordinates and biodiversity institution tests (Fig.
 \ref{fig:split}), and no records were flagged by the sea test.
@@
 In general, the different filters we tested were of similar importance for different study groups.
-There were few outstanding exceptions, including the particularly high proportions of records flagged by the basis of record test for \emph{Tityus} (7.0\%), Dipsadidae (5.6\%), \emph{Prosthechea} (5.0\%) and \emph{Tillandsia} (4.9\%), by the collection year for Dipsadidae (11.3\%), by the taxonomic identification level for \emph{Tityus} (1.6\%), by the capital coordinates for \emph{Oocephalus} (6.1\%) and \emph{Gaylussacia} (3.2\%), as well as by the urban areas test for \emph{Oocephalus} (13.3\%) and Iridaceae (12.3\%).
+There were few outstanding exceptions, including the particularly high proportions of records flagged on the ``basis of record test'' for \emph{Tityus} (7.0\%), Dipsadidae (5.6\%), \emph{Prosthechea} (5.0\%) and \emph{Tillandsia} (4.9\%), by the collection year for Dipsadidae (11.3\%), by the taxonomic identification level for \emph{Tityus} (1.6\%), by the capital coordinates for \emph{Oocephalus} (6.1\%) and \emph{Gaylussacia} (3.2\%), and by the urban areas test for \emph{Oocephalus} (13.3\%) and Iridaceae (12.3\%).
 
-Geographically, the records flagged by the basis of record and individuals collected tests were concentrated in Central America and southern North America, and a relatively high number of records flagged due to their proximity to the centroids of political entities were located on Caribbean islands (Fig.
+Geographically, the records flagged by the ``basis of record'' and ``individual count'' tests were concentrated in Central America and southern North America, and a relatively high number of records flagged due to their proximity to the centroids of political entities were located on Caribbean islands (Fig.
 \ref{fig:split}).
@@
 We found IUCN assessments for 575 species that were also included in our distribution data from 10 of our study groups (Table 4, Appendix 3).
-The fraction of species evaluated varied among the study group, with a maximum of 100\% for \emph{Harengula} and \emph{Lepismium} and a minimum of 2.3\% for Iridaceae (note that the number of species varied considerably among the groups).
+The fraction of species evaluated varied among the study group, with a maximum of 100\% for \emph{Harengula} and \emph{Lepismium} and a minimum of 2.3\% for Iridaceae (note that the number of species varied considerably among groups).
 The median percentage of species per study group with an IUCN assessment was 32.45\%.
@@
 We obtained automated conservation assessments for 2,092 species in the filtered dataset.
-Based on the filtered data the automated conservation assessment evaluated 1,299 species as possibly threatened (62.1\%, CR = 464, EN = 543, VU = 292).
+Based on the filtered data, the automated conservation assessment evaluated 1,299 species as possibly threatened (62.1\%, CR = 464, EN = 543, VU = 292).
 The automated assessment based on the filtered dataset agreed with the IUCN assessment for identifying species as possibly threatened (CR, EN, VU) for 356 species (64\%; Table 4).
@@
 
-We found, that automated flagging based on meta-data and automatic tests flagged on average 40.3\% of the records per taxonomic group.
-The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (question 1).
-The importance of different tests was similar among taxonomic groups, with few exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (question 2).
-The results for species richness were similar between the raw and filtered data with some improvements by the filters and we found little impact of filtering on the accuracy of the automated conservation assessments (question 3).
+Automated flagging based on meta-data and automatic tests flagged on average 40.3\% of the records per taxonomic group.
+The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (Question 1).
+The importance of different tests was similar across taxonomic groups, with few exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (Question 2).
+The results for species richness were similar between the raw and filtered data with some improvements by the filters.
+We found little impact of filtering on the accuracy of the automated conservation assessments (Question 3).
 
 Several reasons may explain the high proportion of records flagged as duplicates.
-First, the deposition of duplicates from the same specimen hosted at different collections is common practice, especially for plants, where a duplication is easily feasible.
-Second, independent collections at similar localities may occur, in particular for local endemics, and third low coordinate precision, for instance based on automated geo-referencing from locality descriptions, may lump records from nearby localities.
-Of course, duplicated records do not represent errors per se, but they represent an unnecessary computational burden, which is often a bottleneck for analyses of datasets with millions of records.
+First, the deposition of duplicates from the same specimen hosted at different collections is common practice, especially for plants, where a specimen duplication is easily feasible.
+Second, independent collections at similar localities may occur, in particular for local endemics.
+Third, low coordinate precision, for instance based on automated geo-referencing from locality descriptions, may lump records from nearby localities.
+Duplicated records do not represent errors per se, but an unnecessary computational burden, which is often a bottleneck for analyses of datasets with millions of records.
 
 The aim of automated filtering is to identify possibly problematic records that are unsuitable for particular downstream analyses.
-Hence our tests include errors in the narrow sense (e.g., zero coordinates, equal lat/lon and seas tests) as well as records that are not necessary erroneous, but have a low precision (e.g., collection year, coordinate precision, centroids of political units) or are unnecessary (duplicates).
-The importance of each test and the linked thresholds must be judged based on the downstream analyses.
+Hence, our tests may identify true errors (e.g., wrong coordinates such as those with exclusively zeros, equal latitude and longitude and records of terestrial species in the sea) as well as records that are not necessarily erroneous, but have a low precision (e.g., collection year, coordinate precision, centroids of political units) or are redundant for certain analyses (duplicates).
+The importance of each test and the linked thresholds must be judged based on the specific downstream analyses.
 For instance, records assigned to country centroids might be acceptable for inter-continental comparisons, but can be considered erroneous for species distribution modelling on a local scale.
@@
 
-The number of records flagged by individual tests was similar across the study groups, suggesting that similar error types might be relevant for collections of plants and animals and partly also terrestrial and marine organisms.
+The number of records flagged by individual tests was similar across study groups, suggesting that similar error types might be relevant for collections of plants and animals.
 Therefore, the same tests can be used across taxonomic groups.
 Some notable exceptions stress the need to adapt each test to the taxonomic study group to balance data quality and data availability.
-The high fraction of records flagged by the basis of record test for \emph{Tityus}, Dipsadidae, \emph{Prosthechea} and \emph{Tillandsia}, were mostly caused by a high number of records in these groups based on unknown collection methods (See Appendix 4 for f the individual groups), which might be caused by the contribution of specific datasets lacking this information for these groups.
-The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection activity already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700.
-The latter records likely represent a data entry error.
-They are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700.
-This emphasizes the importance to better train researchers and students to manage biodiversity databases.
-The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae, \emph{Oocephalus}, and \emph{Gaylussacia}, might be related to horticulture since at least some species of this group are commonly cultivated as ornamentals.
+The high fraction of records flagged by the ``basis of record test'' for \emph{Tityus}, Dipsadidae, \emph{Prosthechea} and \emph{Tillandsia}, were mostly caused by a high number of records in these groups based on unknown collection methods (See Appendix 4 for the individual groups), which might be caused by the contribution of specific datasets lacking this information for these groups.
+The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection effort already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700.
+The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700.
+The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae, and \emph{Oocephalus} might be related to horticulture since at least some species in those groups are commonly cultivated as ornamentals.
 
@@
 \ref{fig:speciesrichness}, Supplementary material 1).
-The filter based on political centroids had an important impact on species richness patterns, which is congruent with the results from previous studied (Maldonado et al., 2015).
-Records assigned to country or province centroids are often old records, which are geo-referenced based on vague locality descriptions.
-These records are at the same time more likely to represent dubious species names, since they might be old synonyms or type specimen of species that have only been collected and described once, which are erroneously increasing species numbers.
+The filter based on political centroids had an important impact on species richness patterns, which is congruent with the results from a previous study in the coffee family (Maldonado et al., 2015).
+Records assigned to country or province centroids are often old records, which are geo-referenced at a later point based on vague locality descriptions.
+These records are at the same time more likely to represent dubious species names, since they might be old synonyms or type specimens of species that have only been collected and described once, which are erroneously increasing species numbers.
 
@@
 The robustness of the automated assessment was likely due to the fact that the EOO for most species was large, even after the considerable reduction caused by filtering.
-This might be caused by the structure of our comparison, which only included species that were evaluated by the IUCN Red List (and not considered as Data Deficient) and at the same time have occurrences recorded in GBIF, which biases the dataset towards species with larger ranges, since generally more data for them is available.
+This might be caused by the structure of our comparison, which only included species that were evaluated by the IUCN Red List (and not considered as Data Deficient) and at the same time had occurrences recorded in GBIF.
+Those inclusion criteria are likely to have, biased the datasets towards species with large ranges, since generally more data for them are available.
 The robustness of automated conservation assessments to data quality is encouraging, although these methods are only an approximation, especially for species with few collection records (Rivers et al., 2011).
@@
 
-Our results suggest that between 1/3rd and 2/3rds of the occurrence records obtained from public databases might be unsuitable for downstream biogeographic analyses, dependent on the taxonomic group.
-While the majority of these records might not be erroneous per se, they might be overly imprecise and thereby increase uncertainty of downstream results or put additional computational burden on the analyses.
+Our results suggest that between one to two thirds of the occurrence records obtained from public databases might be unsuitable for downstream biodiversity analyses.
+While the majority of these records might not be erroneous per se, they might be overly imprecise and thereby increase uncertainty of downstream results or add computational burden on big data analyses.
 
-Overall, our results stress the importance of (meta-)data exploration before analyses and suggest that automated flagging can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the specific taxonomic groups and geographic area of interest.
+Our results stress the importance of (meta-)data exploration prior to most biodiversity analyses.
+Automated flagging can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the specific taxonomic groups and geographic area of interest.
+The putative errors we encountered point to the importance to train researchers and students to curate species occurrence datasets and to allow users to provide feedback for particular records, thus contributing to overall accuracy and precision of public biodiversity databases.
 
@@
 
-This study enrolled participants of the workshop ``Biodiversity data: from field to yield'' lead by Alexandre Antonelli and Alexander Zizka through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI.
-We thank the Pró-reitoria de Pesquisa and the Pró-reitoria de Pós-graduação of UFRN for the financial support (edital 02/2016 - internacionalização).
-AZ is thankful for funding by iDiv via the German Research Foundation (DFG FZT 118), specifically through sDiv, the Synthesis Centre of iDiv.
+This study enrolled participants of the workshop ``Biodiversity data: from field to yield'' led by Alexandre Antonelli and Alexander Zizka through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI.
+We thank the Pró-reitoria de Pesquisa and the Pró-reitoria de Pós-graduação of UFRN for financial support (edital 02/2016 - internacionalização).
+AZ is funded by iDiv via the German Research Foundation (DFG FZT 118), specifically through sDiv, the Synthesis Centre of iDiv.
+AA is supported by the Swedish Research Council, the Knut and Alice Wallenberg Foundation, the Swedish Foundation for Strategic Research and the Royal Botanic Gardens, Kew.
 
@@
 \caption{\label{tab:tabletaxa}The study groups and their taxonomy.
-This study includes three marine and 13 terrestrial taxa, 6 of them animals, one group of fungi and 9 plants, belonging to 16 different orders.}
+This study includes three marine and 13 terrestrial taxa, six of them animals, one group of fungi and nine plants, belonging to 16 different orders.}
 \centering
@@
 Arhynchobatidae & Family & Marine & Skates & Chordata & Rajiformes & Arhynchobatidae\\
-Dipsadidae & Family & Terrestrial & Snakes & Chordata & Squamata & Colubridae\\
+Dipsadidae & Family & Terrestrial & Snakes & Chordata & Squamata & Dipsadidae\\
 \addlinespace
-Harengula & Genus & Marine & Herings & Chordata & Clupeiformes & Clupeidae\\
+Harengula & Genus & Marine & Herrings & Chordata & Clupeiformes & Clupeidae\\
 \hline
@@
 \midrule
-\rowcolor{gray!6}  Basis of record & Meta-data & Records might be unsuitable or unreliable for certain analysis dependent on their source, e.g. "fossil" or "unknown"\\
+\rowcolor{gray!6}  Basis of record & Meta-data & Records might be unsuitable or unreliable for certain analyses dependent on their source, e.g. "fossil" or "unknown"\\
 Collection year & Meta-data & Coordinates from old records are more likely to be imprecise or erroneous coordinates since they are derived from  geo-referencing based on the locality description.
-Since is more problematic for older records, since names or borders of places may change\\
-\rowcolor{gray!6}  Coordinate precision & Meta-data & Records may be unsuitable for a study if their precision is lower than the studies analysis scale\\
+This is more problematic for older records, since names or borders of places may change\\
+\rowcolor{gray!6}  Coordinate precision & Meta-data & Records may be unsuitable for a study if their precision is lower than the study analysis scale\\
 Identification level & Meta-data & Records may be unsuitable if they are not identified to species level.\\
@@
 Equal lat/lon & Gazetteer-based & Coordinates with equal latitude and longitude are usually indicative of data entry errors\\
-\rowcolor{gray!6}  Capitals & Gazetteer-based & Records may be assigned to the coordinates of countries capitals based on a vague locality description\\
+\rowcolor{gray!6}  Capitals & Gazetteer-based & Records may be assigned to the coordinates of country capitals based on a vague locality description\\
 Political centroids & Gazetteer-based & Records may be assigned to the coordinates of the centroids of political entities based on a vague locality description\\
@@
 
-\caption{\label{tab:unnamed-chunk-3}Conservation assessment for 16 Neotropical taxa of plants and animals based on three datasets.
+\caption{\label{tab:unnamed-chunk-3}Conservation assessment for 10 Neotropical taxa of plants and animals based on three datasets.
 IUCN: global red list assessment obtained from www.iucn.org; GBIF Raw: Preliminary conservation assessment based on IUCN Criterion B using ConR and the raw dataset from GBIF; GBIF filtered: Preliminary conservation assessment based on IUCN Criterion B using ConR and the filtered dataset.
@@
 \begin{figure}
-\includegraphics[width=\linewidth]{./output/figure_number_of_records} \caption{The difference of records available between the raw and filtered dataset, in a 100x100 km grid across the Neotropics}\label{fig:total}
+\includegraphics[width=\linewidth]{./output/figure_number_of_records} \caption{The difference in records available between the raw and filtered dataset of 16 Neotropical taxa including plants and animals, plotted in a 100x100 km grid across the Neotropics.}\label{fig:total}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\linewidth]{./output/figure_number_of_records_split} \caption{The geographic pattern of the occurrence records flagged by the automated tests of this study.
-The concentration of records flagged by the basis of record test and individual count test in Central A,erica and southern North America, suggest some specific problem with a contributing dataset from the region.
+\includegraphics[width=\linewidth]{./output/figure_number_of_records_split} \caption{Geographic location of the occurrence records flagged by the automated tests  applied in this study.
+The concentration of records flagged on the basis of record test and individual count test in Central America and southern North America.
 Only filters that flagged at least one record are shown.}\label{fig:split}
@@
 \begin{figure}
-\includegraphics[width=\linewidth]{./output/figure_number_species_richness_difference} \caption{The difference in species richness derived from the raw and filtered datasets (raw - filtered), for 16 taxa of plants and animals.}\label{fig:speciesrichness}
+\includegraphics[width=\linewidth]{./output/figure_number_species_richness_difference} \caption{The difference in species richness derived from the raw and filtered datasets (raw - filtered), for 16 taxa of animals, funghi, and plants.}\label{fig:speciesrichness}
 \end{figure}
@@
 L.
+P.
 (2018).
@@
 (2019).
-\textless{}scp\textgreater{}CoordinateCleaner\textless{}/scp\textgreater{} : Standardized cleaning of occurrence records from biological collection databases.
+CoordinateCleaner: Standardized cleaning of occurrence records from biological collection databases.
 \emph{Methods in Ecology and Evolution}, \emph{10}(5), 744--751.

### 2020-02-25 02_automated_cleaning.R

@@ -12,7 +12,7 @@ dat <- data.frame()
 
 for(i in 1:length(li)){
   print(i)
-  sub <- read_delim(paste("input/gbif/", li[i], "/occurrence.txt", sep = ""), delim = "\t") %>% 
+  sub <- read_delim(paste("input/gbif/", li[i], "/occurrence.txt", sep = ""), delim = "\t", quote = "") %>% 
     mutate_all(as.character)
   if(nrow(sub) > 0){
     sub$taxon <- str_split(li[i], pattern = "_")[[1]][1]

### 2020-02-29 manuscript_peerj.Rmd
peerj format

--- 2020-02-25 manuscript.Rmd
@@
 ---
-title: "No one-size-fits-all solution to clean GBIF"
+title: No one-size-fits-all solution to clean GBIF
+preprint: false
 author: 
-- Alexander Zizka^1,2^,
-- Fernanda Antunes Carvalho^4^,
-- Alice Calvente^4^,
-- Ana Paula Alves da Silva^4^,
-- Andressa Cabral^5^,
-- Arthur de Souza Soares^4^,
-- Eduardo Calisto Tomaz^4^,
-- Fernanda Gondim Lambert Moreira^4^,
-- Filipe Cabreirinha Serrano^6^,
-- Jéssica Fernanda Ramos Coelho^4^,
-- Mabel Rocio Baez-Lizarazo,^7^
-- Mariana Ramos Fantinati^4^,
-- Matheus Colli-Silva^5^,
-- Moabe Ferreira Fernandes^8^,
-- Nathália Michelly da Cunha Santos^4^,
-- Paolla Gabryelle Cavalcante de Souza^4^,
-- Renata Clicia dos Santos‐Costa^4^,
-- Thais Ferreira-Araújo^4^,
-- Tiago Andrade Borges Santos^9^,
-- Tiago Luiz Vieira^9^,
-- Valéria Fonseca Vale^4^,
-- Alexandre Antonelli^10,11,12^
+  - name: Alexander Zizka
+    affiliation: 1,2
+    corresponding: true
+    email: alexander.zizka@idiv.de
+  - name: Fernanda Antunes Carvalho
+    affiliation: 3
+  - name: Alice Calvente
+    affiliation: 4
+  - name: Ana Paula Alves da Silva
+    affiliation: 4
+  - name: Andressa Cabral
+    affiliation: 5
+  - name: Arthur de Souza Soares
+    affiliation: 4
+  - name: Eduardo Calisto Tomaz
+    affiliation: 4
+  - name: Fernanda Gondim Lambert Moreira
+    affiliation: 4
+  - name: Filipe Cabreirinha Serrano
+    affiliation: 6
+  - name: Jéssica Fernanda Ramos Coelho
+    affiliation: 4
+  - name: Mabel Rocio Baez-Lizarazo
+    affiliation: 7
+  - name: Mariana Ramos Fantinati
+    affiliation: 4
+  - name: Matheus Colli-Silva
+    affiliation: 5
+  - name: Moabe Ferreira Fernandes
+    affiliation: 8
+  - name: Nathália Michelly da Cunha Santos
+    affiliation: 4
+  - name: Paolla Gabryelle Cavalcante de Souza
+    affiliation: 4
+  - name: Renata Clicia dos Santos‐Costa
+    affiliation: 4
+  - name: Thais Ferreira-Araújo
+    affiliation: 4
+  - name: Tiago Andrade Borges Santos
+    affiliation: 9
+  - name: Tiago Luiz Vieira
+    affiliation: 9
+  - name: Valéria Fonseca Vale
+    affiliation: 4
+  - name: Alexandre Antonelli
+    affiliation: 10,11,12
+affiliation:
+  - code: 1
+    address: sDiv, German Center for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Leipzig, Germany
+  - code: 2
+    address: Naturalis Biodiversity Center, Leiden, The Netherlands
+  - code: 3
+    address: Departamento de Genética, Ecologia e Evolução, Universidade Federal de Minas Gerais, Belo Horizonte, Brazil
+  - code: 4
+    address: Departamento de Botânica e Zoologia, Universidade Federal do Rio Grande do Norte, Natal, Brazil
+  - code: 5
+    address: Departamento de Botânica, Universidade de São Paulo, São Paulo, Brazil
+  - code: 6
+    address: Departamento de Ecologia, Universidade de São Paulo, São Paulo, Brazil
+  - code: 7
+    address: Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil
+  - code: 8
+    address: Departamento de Ciências Biológicas, Universidade Estadual de Feira de Santana, Feira de Santana, Brazil
+  - code: 9
+    address: Universidade Estadual de Feira de Santana, Bahia, Brazil
+  - code: 10
+    address: Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden
+  - code: 11
+    address: Department for Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden
+  - code: 12
+    address: Royal Botanic Gardens Kew, Richmond, United Kingdom
+abstract: >
+  Species occurrence records provide the basis for many biodiversity studies.
+They derive from geo-referenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications.
+Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern.
+Automatic flagging and filtering are a scalable and reproducible means to identify potentially problematic records in datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org).
+However, it is unclear how much data may be lost by filtering, whether the same tests should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses.
+Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 16 Neotropical taxa including plants and animals, terrestrial and marine, downloaded from GBIF.
+We find that 29-67% of the records are potentially erroneous, with large variation across taxonomic groups.
+Tests for duplicated information, collection year, basis of record as well as urban areas and political centroids have the greatest effect.
+While many flagged records might not be *de facto* erroneous, they could be overly imprecise and increase uncertainty in downstream analyses.
+Automated flagging can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus.
+Our results stress the importance of thorough exploration of the meta-data associated with species records for biodiversity research.
 bibliography: course_natal.bib
-csl: jbi-3.csl
 output:
-  #word_document:
-  bookdown::pdf_document2:
-    toc: false
-    toc_depth: 1
-    fig_caption: true
-    number_sections: false
-    keep_tex: yes
+  bookdown::pdf_book:
+    base_format: rticles::peerj_article # for using bookdown features like \@ref()
+  rticles::peerj_article: default
 nocite: |
     @GBIForg2019; @GBIForg2019a; @GBIForg2019b; @GBIForg2019c; @GBIForg2019d; @GBIForg2019e; @GBIForg2019f; @GBIForg2019g; @GBIForg2019h; @GBIForg2019i; @GBIForg2019j; @GBIForg2019k; @GBIForg2019l; @GBIForg2019m; @GBIForg2019n; @GBIForg2019o;@GBIForg2019p;@GBIForg2019q
-header-includes: |
-    \usepackage{caption}
-    \usepackage{fancyhdr}
-    \usepackage{booktabs}
-    \pagestyle{fancy}
-    \fancyhead[L]{Automated filter effect}
-    \fancyhead[LE,RO]{\thepage}
-    \fancyfoot{}
-    \usepackage{float}
-    \floatplacement{figure}{H}
-    \usepackage{lineno}
-    \linenumbers
-linestretch: 1.3
-fontsize: 12pt
-always_allow_html: yes
 ---
-
-1. sDiv, German Center for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Leipzig, Germany
-2. Naturalis Biodiversity Center, Leiden, The Netherlands
-3. Departamento de Genética, Ecologia e Evolução, Universidade Federal de Minas Gerais, Belo Horizonte, Brazil
-4. Departamento de Botânica e Zoologia, Universidade Federal do Rio Grande do Norte, Natal, Brazil
-5. Departamento de Botânica, Universidade de São Paulo, São Paulo, Brazil
-6. Departamento de Ecologia, Universidade de São Paulo, São Paulo, Brazil
-7. Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil
-8. Departamento de Ciências Biológicas, Universidade Estadual de Feira de Santana, Feira de Santana, Brazil
-9. Universidade Estadual de Feira de Santana, Bahia, Brazil
-10. Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden
-11. Department for Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden
-12. Royal Botanic Gardens Kew, Richmond, United Kingdom
-
-\newpage{}
 
@@
 
-# Abstract
-Species occurrence records provide the basis for many biodiversity studies.
-They derive from geo-referenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications.
-Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern.
-Automatic flagging and filtering are a scalable and reproducible means to identify potentially problematic records in datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org).
-However, it is unclear how much data may be lost by filtering, whether the same tests should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses.
-Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 16 Neotropical taxa including plants and animals, terrestrial and marine, downloaded from GBIF.
-We find that 29-67% of the records are potentially erroneous, with large variation across taxonomic groups.
-Tests for duplicated information, collection year, basis of record as well as urban areas and political centroids have the greatest effect.
-While many flagged records might not be *de facto* erroneous, they could be overly imprecise and increase uncertainty in downstream analyses.
-Automated flagging can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus.
-Our results stress the importance of thorough exploration of the meta-data associated with species records for biodiversity research.
-
-# Keywords
-Automated cleaning, automated conservation assessment, data quality, GBIF, Neotropics.
-
-\newpage{}
-
-# Introduction
+# Introduction {-}
 Publicly available species distribution data have become a crucial resource in biodiversity research, including studies in ecology, biogeography, systematics and conservation biology.
@@
 
-# Material and Methods
-## Choice of study taxa
+# Material and Methods {-}
+## Choice of study taxa {-}
 This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering graduate students, taxonomists and biogeographers working with different taxonomic groups of animals, funghi, and plants across the Neotropics.
@@
 
-## Species occurrence data
+## Species occurrence data {-}
 <!-- @GBIForg2019; @GBIForg2019a; @GBIForg2019b; @GBIForg2019c; @GBIForg2019d; @GBIForg2019e; @GBIForg2019f; @GBIForg2019g; @GBIForg2019h; @GBIForg2019i; @GBIForg2019j; @GBIForg2019k; @GBIForg2019l; @GBIForg2019m; @GBIForg2019n; @GBIForg2019o -->
@@
 
-##  Automated cleaning
+##  Automated cleaning {-}
 We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the raw data downloaded from GBIF ("raw", hereafter) using meta-data for those records for which they were available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99.
@@
 
-## Downstream analyses
+## Downstream analyses {-}
 We first generated species richness maps using 100x100 km grid cells for the raw and filtered datasets respectively, using the package `speciesgeocodeR v2.0-10` [@Topel2016].
@@
 
-## Evaluation of results
+## Evaluation of results {-}
 <!-- This session is exciting and makes the study more reliable and interesting than one just based on data mining, but how this was done is very unclear.
@@
 
-# Results
+# Results {-}
 ```{r}
@@
   select(-researcher) %>% 
-  filter(!taxon %in% c("Entomobryidae", "Opisthonema", "Eugenia", "Harpalyce"))
+  filter(!taxon %in% c("Opisthonema", "Eugenia"))
 
@@
 tab2 <- read_csv("output/summary_table_flags.csv") %>% 
-  dplyr::select(-coordinate_missing, -invalid_coords, -sea_coords, -gbif_coords,-inst_coords, -equal_coords) %>% 
+  dplyr::select(-coordinate_missing, -invalid_coords, -gbif_coords,-inst_coords, -equal_coords) %>% 
   mutate(fraction = round(fraction * 100,1))
@@
 
-# Discussion
+# Discussion {-}
 <!-- add some subheadings to discussion -->
@@
 
-<!-- 1. Which is the most relevant flagging mechanism? -->
+## The relevance of individual filters {-}
 Several reasons may explain the high proportion of records flagged as duplicates.
@@
 
-<!-- 2. Does the importance of individual errors differ among taxonomic groups? -->
+## Similarities and differences among taxa {-}
 The number of records flagged by individual tests was similar across study groups, suggesting that similar error types might be relevant for collections of plants and animals.
@@
 
+## The impact of filtering on the accuracy of automated conservation assessments {-}
 The accuracy of the automated conservation assessment was in the same range as found by previous studies [@NicLughadha2019; @Zizka2019a].
@@
 
-# Conclusions
+# Conclusions {-}
 Our results suggest that between one to two thirds of the occurrence records obtained from public databases might be unsuitable for downstream biodiversity analyses.
@@
 
-# Acknowledgements
+# Acknowledgments {-}
 This study enrolled participants of the workshop "Biodiversity data: from field to yield" led by Alexandre Antonelli and Alexander Zizka through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI.
@@
 
-# Supplementary material
-Appendix 1 - Analyses scripts
+# Supplementary material {-}
 
-Appendix 2 - Table S1, reporting the absolute number of flagged records per taxonomic group and test
+- Appendix 1 - Analyses scripts
 
-Appendix 3 - The full results of the conservation assessment
+- Appendix 2 - Table S1, reporting the absolute number of flagged records per taxonomic group and test
 
-Appendix 4 - Taxon specific richness maps and comments.
+- Appendix 3 - The full results of the conservation assessment
 
-\newpage{}
+- Appendix 4 - Taxon specific richness maps and comments.
 
-# Tables
-
+# Tables {-}
 <!-- A table presenting the study taxa -->
@@
 
-\newpage{}
+# Figures {-}
+<!-- The study species-->
+```{r species, out.width="0.9\\linewidth", fig.cap=c('Example of study taxa.
+\\textbf{A)} Pilosocereus pusillibaccatus, picture by Pamela Lavor; \\textbf{B)} Conchocarpus macrocarpus, Juliana El-Ottra; \\textbf{C)} Tillandsia recurva, EC Tomaz; \\textbf{D} \\textit{Oxyrhopus guibei} (Dipsadidae), Filipe C Serrano;\\textbf{E)} Neanuridae (\\textit{Aethiopella ricardoi}), Raiane Vital da Paz; \\textbf{F)} \\textit{Tocoyena formosa}, Fernanda GL Moreira; \\textbf{G)} \\textit{Harengula jaguana}, Thais Ferreira-Araujo; \\textbf{H)} \\textit{Gaylussacia decipiens}, Luiz Menini Neto; \\textbf{I)} \textit{Ooocephalus foliosus}, Arthur Soares; \\textbf{J)} \\textit{Tityus carvalhoi}, Renata C Santos-Costa; \\textbf{K)} \\textit{Prostheca vespa}, Tiago veira'), echo=FALSE}
+knitr::include_graphics(path = "./figures_tables/Figure_study_groups.jpg")
+```
 
-# Figures
 <!-- A map with number of species and where they have been removed -->
@@
 \newpage{}
-
-# References
+# References {-}

### 2020-02-29 manuscript_peerj.tex
peerj format

--- 2020-02-25 manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  12pt,
-]{article}
-\usepackage{lmodern}
-\usepackage{setspace}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={No one-size-fits-all solution to clean GBIF},
-  pdfauthor={Alexander Zizka1,2,; Fernanda Antunes Carvalho4,; Alice Calvente4,; Ana Paula Alves da Silva4,; Andressa Cabral5,; Arthur de Souza Soares4,; Eduardo Calisto Tomaz4,; Fernanda Gondim Lambert Moreira4,; Filipe Cabreirinha Serrano6,; Jéssica Fernanda Ramos Coelho4,; Mabel Rocio Baez-Lizarazo,7; Mariana Ramos Fantinati4,; Matheus Colli-Silva5,; Moabe Ferreira Fernandes8,; Nathália Michelly da Cunha Santos4,; Paolla Gabryelle Cavalcante de Souza4,; Renata Clicia dos Santos‐Costa4,; Thais Ferreira-Araújo4,; Tiago Andrade Borges Santos9,; Tiago Luiz Vieira9,; Valéria Fonseca Vale4,; Alexandre Antonelli10,11,12},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[margin=1in]{geometry}
-\usepackage{longtable,booktabs}
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
+%% Submissions for peer-review must enable line-numbering
+%% using the lineno option in the \documentclass command.
+%%
+%% Preprints and camera-ready submissions do not need
+%% line numbers, and should have this option removed.
+%%
+%% Please note that the line numbering option requires
+%% version 1.1 or newer of the wlpeerj.cls file, and
+%% the corresponding author info requires v1.2
+
+\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions
+
+% ZNK -- Adding headers for pandoc
+
+\setlength{\emergencystretch}{3em}
+\providecommand{\tightlist}{
 \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-\usepackage{caption}
-\usepackage{fancyhdr}
-\usepackage{booktabs}
-\pagestyle{fancy}
-\fancyhead[L]{Automated filter effect}
-\fancyhead[LE,RO]{\thepage}
-\fancyfoot{}
-\usepackage{float}
-\floatplacement{figure}{H}
-\usepackage{lineno}
-\linenumbers
+\usepackage{lipsum}
+\usepackage[unicode=true]{hyperref}
+\usepackage{longtable}
+
+% Pandoc syntax highlighting
+% See https://github.com/rstudio/rticles/issues/182
+
+% Pandoc Header
 \usepackage{booktabs}
@@
 \title{No one-size-fits-all solution to clean GBIF}
-\author{Alexander Zizka\textsuperscript{1,2}, \and Fernanda Antunes Carvalho\textsuperscript{4}, \and Alice Calvente\textsuperscript{4}, \and Ana Paula Alves da Silva\textsuperscript{4}, \and Andressa Cabral\textsuperscript{5}, \and Arthur de Souza Soares\textsuperscript{4}, \and Eduardo Calisto Tomaz\textsuperscript{4}, \and Fernanda Gondim Lambert Moreira\textsuperscript{4}, \and Filipe Cabreirinha Serrano\textsuperscript{6}, \and Jéssica Fernanda Ramos Coelho\textsuperscript{4}, \and Mabel Rocio Baez-Lizarazo,\textsuperscript{7} \and Mariana Ramos Fantinati\textsuperscript{4}, \and Matheus Colli-Silva\textsuperscript{5}, \and Moabe Ferreira Fernandes\textsuperscript{8}, \and Nathália Michelly da Cunha Santos\textsuperscript{4}, \and Paolla Gabryelle Cavalcante de Souza\textsuperscript{4}, \and Renata Clicia dos Santos‐Costa\textsuperscript{4}, \and Thais Ferreira-Araújo\textsuperscript{4}, \and Tiago Andrade Borges Santos\textsuperscript{9}, \and Tiago Luiz Vieira\textsuperscript{9}, \and Valéria Fonseca Vale\textsuperscript{4}, \and Alexandre Antonelli\textsuperscript{10,11,12}}
-\date{}
 
-\begin{document}
-\maketitle
+\author[1,2]{Alexander Zizka}
 
-\setstretch{1.3}
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  sDiv, German Center for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Leipzig, Germany
-\item
-  Naturalis Biodiversity Center, Leiden, The Netherlands
-\item
-  Departamento de Genética, Ecologia e Evolução, Universidade Federal de Minas Gerais, Belo Horizonte, Brazil
-\item
-  Departamento de Botânica e Zoologia, Universidade Federal do Rio Grande do Norte, Natal, Brazil
-\item
-  Departamento de Botânica, Universidade de São Paulo, São Paulo, Brazil
-\item
-  Departamento de Ecologia, Universidade de São Paulo, São Paulo, Brazil
-\item
-  Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil
-\item
-  Departamento de Ciências Biológicas, Universidade Estadual de Feira de Santana, Feira de Santana, Brazil
-\item
-  Universidade Estadual de Feira de Santana, Bahia, Brazil
-\item
-  Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden
-\item
-  Department for Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden
-\item
-  Royal Botanic Gardens Kew, Richmond, United Kingdom
-\end{enumerate}
+\corrauthor[1,2]{Alexander Zizka}{\href{mailto:alexander.zizka@idiv.de}{\nolinkurl{alexander.zizka@idiv.de}}}
+\author[3]{Fernanda Antunes Carvalho}
 
-\newpage{}
+\author[4]{Alice Calvente}
 
-\hypertarget{abstract}{%
-\section{Abstract}\label{abstract}}
+\author[4]{Ana Paula Alves da Silva}
 
+\author[5]{Andressa Cabral}
+
+\author[4]{Arthur de Souza Soares}
+
+\author[4]{Eduardo Calisto Tomaz}
+
+\author[4]{Fernanda Gondim Lambert Moreira}
+
+\author[6]{Filipe Cabreirinha Serrano}
+
+\author[4]{Jéssica Fernanda Ramos Coelho}
+
+\author[7]{Mabel Rocio Baez-Lizarazo}
+
+\author[4]{Mariana Ramos Fantinati}
+
+\author[5]{Matheus Colli-Silva}
+
+\author[8]{Moabe Ferreira Fernandes}
+
+\author[4]{Nathália Michelly da Cunha Santos}
+
+\author[4]{Paolla Gabryelle Cavalcante de Souza}
+
+\author[4]{Renata Clicia dos Santos‐Costa}
+
+\author[4]{Thais Ferreira-Araújo}
+
+\author[9]{Tiago Andrade Borges Santos}
+
+\author[9]{Tiago Luiz Vieira}
+
+\author[4]{Valéria Fonseca Vale}
+
+\author[10,11,12]{Alexandre Antonelli}
+
+\affil[1]{sDiv, German Center for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Leipzig, Germany}
+\affil[2]{Naturalis Biodiversity Center, Leiden, The Netherlands}
+\affil[3]{Departamento de Genética, Ecologia e Evolução, Universidade Federal de Minas Gerais, Belo Horizonte, Brazil}
+\affil[4]{Departamento de Botânica e Zoologia, Universidade Federal do Rio Grande do Norte, Natal, Brazil}
+\affil[5]{Departamento de Botânica, Universidade de São Paulo, São Paulo, Brazil}
+\affil[6]{Departamento de Ecologia, Universidade de São Paulo, São Paulo, Brazil}
+\affil[7]{Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil}
+\affil[8]{Departamento de Ciências Biológicas, Universidade Estadual de Feira de Santana, Feira de Santana, Brazil}
+\affil[9]{Universidade Estadual de Feira de Santana, Bahia, Brazil}
+\affil[10]{Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden}
+\affil[11]{Department for Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden}
+\affil[12]{Royal Botanic Gardens Kew, Richmond, United Kingdom}
+
+%
+% \author[1]{First Author}
+% \author[2]{Second Author}
+% \affil[1]{Address of first author}
+% \affil[2]{Address of second author}
+% \corrauthor[1]{First Author}{f.author@email.com}
+
+%
+
+\begin{abstract}
 Species occurrence records provide the basis for many biodiversity studies.
@@
 Our results stress the importance of thorough exploration of the meta-data associated with species records for biodiversity research.
+% Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+\end{abstract}
 
-\hypertarget{keywords}{%
-\section{Keywords}\label{keywords}}
+\begin{document}
 
-Automated cleaning, automated conservation assessment, data quality, GBIF, Neotropics.
-
-\newpage{}
+\flushbottom
+\maketitle
+\thispagestyle{empty}
 
 \hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section*{Introduction}\label{introduction}}
+\addcontentsline{toc}{section}{Introduction}
 
@@
 
-A central challenge to the use of these publicly available species occurrence data in research are erroneous geographic coordinates (Anderson et al., 2016).
+A central challenge to the use of these publicly available species occurrence data in research are erroneous geographic coordinates (Anderson et al.
+2016).
 Errors mostly arise because public databases integrate records collected with different methodologies in different places, at different times; often without centralized curation and only rudimentary meta-data.
-For instance, erroneous coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common (Maldonado et al., 2015; Yesson et al., 2007) and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in captivity (Zizka, Silvestro, et al., 2019).
+For instance, erroneous coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common (Maldonado et al.
+2015; Yesson et al.
+2007) and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in captivity (Zizka, Silvestro, et al.
+2019).
 
@@
 As an alternative, automated flagging methods to identify potentially problematic records have been proposed as a scalable option, as they are able to deal with datasets containing up to millions of records and many different taxa.
-Those methods are usually based on geographic gazetteers (e.g., Chamberlain, 2016; Jin \& Yang, 2020; Zizka, Silvestro, et al., 2019) or on additional data, such as environmental variables (Robertson, Visser, \& Hui, 2016).
-Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biodiveristy research (Zizka, Silvestro, et al., 2019).
+Those methods are usually based on geographic gazetteers (e.g., Chamberlain 2016; Zizka, Silvestro, et al.
+2019; Jin and Yang 2020) or on additional data, such as environmental variables (Robertson, Visser, and Hui 2016).
+Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biodiveristy research (Zizka, Silvestro, et al.
+2019).
 
 Erroneous records are especially problematic in conservation, where stakes are high.
-Recently proposed methods for automated conservation assessments could support the formal assessment procedures for the global Red List of the International Union for the Conservation of Nature (IUCN) (Bachman et al., 2011; Dauby et al., 2017; Pelletier et al., 2018).
+Recently proposed methods for automated conservation assessments could support the formal assessment procedures for the global Red List of the International Union for the Conservation of Nature (IUCN) (Dauby et al.
+2017; Bachman et al.
+2011; Pelletier et al.
+2018).
 These methods approximate species' range size, namely the Extent of Occurrence (EOO, which is the area of a convex hull polygon comprising all records of a species), the Area of Occupancy (AOO, which is the sum of the area actually occupied by a species, calculated based on a small-scale regular grid), and the number of locations for a preliminary conservation assessment following IUCN Criterion B (``Geographic range'').
-These methods have been used topropose preliminary global (Stévart et al., 2019; Zizka, Azevedo, et al., 2019) and regional (Cosiaux et al., 2018; Schmidt et al., 2017) Red List assessments.
+These methods have been used topropose preliminary global (Stévart et al.
+2019; Zizka, Azevedo, et al.
+2019) and regional (Schmidt et al.
+2017; Cosiaux et al.
+2018) Red List assessments.
 However, all metrics but especially EOO, are sensitive to individual records with erroneous coordinates \textbf{(REFERENCE)}.
@@
 \hypertarget{material-and-methods}{%
-\section{Material and Methods}\label{material-and-methods}}
+\section*{Material and Methods}\label{material-and-methods}}
+\addcontentsline{toc}{section}{Material and Methods}
 
 \hypertarget{choice-of-study-taxa}{%
-\subsection{Choice of study taxa}\label{choice-of-study-taxa}}
+\subsection*{Choice of study taxa}\label{choice-of-study-taxa}}
+\addcontentsline{toc}{subsection}{Choice of study taxa}
 
@@
 \hypertarget{species-occurrence-data}{%
-\subsection{Species occurrence data}\label{species-occurrence-data}}
+\subsection*{Species occurrence data}\label{species-occurrence-data}}
+\addcontentsline{toc}{subsection}{Species occurrence data}
 
-We downloaded occurrence information for all study groups from www.gbif.org using the \texttt{rgbif} v1.4.0 package (Chamberlain, 2017) in R (GBIF.org, 2019a-p).
-We downloaded only records with geographic coordinates and limited the study area to latitudes smaller than 33\(^\circ\) N and longitudes smaller than 35\(^\circ\) W and larger than 120\(^\circ\) W reflecting the Neotropics, our main area of expertise (Morrone, 2014).
+We downloaded occurrence information for all study groups from www.gbif.org using the \texttt{rgbif} v1.4.0 package (Chamberlain 2017) in R (GBIF.org, 2019a-p).
+We downloaded only records with geographic coordinates and limited the study area to latitudes smaller than 33\(^\circ\) N and longitudes smaller than 35\(^\circ\) W and larger than 120\(^\circ\) W reflecting the Neotropics, our main area of expertise (Morrone 2014).
 The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, \emph{Gaylussacia}, Iridaceae, Neanuridae, and \emph{Tillandsia} for which we only obtained the Neotropical occurrences.
@@
 \hypertarget{automated-cleaning}{%
-\subsection{Automated cleaning}\label{automated-cleaning}}
+\subsection*{Automated cleaning}\label{automated-cleaning}}
+\addcontentsline{toc}{subsection}{Automated cleaning}
 
@@
 Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e.~test for duplicated records).
-In a second step, we used the \texttt{clean\_coordinates} function of the \texttt{CoordinateCleaner\ v2.0-11} package (Zizka, Silvestro, et al., 2019) with default options to flag errors that are common to biological data sets (``filtered'', hereafter).
+In a second step, we used the \texttt{clean\_coordinates} function of the \texttt{CoordinateCleaner\ v2.0-11} package (Zizka, Silvestro, et al.
+2019) with default options to flag errors that are common to biological data sets (``filtered'', hereafter).
 These including: coordinates in the sea for terrestrial taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions.
@@
 \hypertarget{downstream-analyses}{%
-\subsection{Downstream analyses}\label{downstream-analyses}}
+\subsection*{Downstream analyses}\label{downstream-analyses}}
+\addcontentsline{toc}{subsection}{Downstream analyses}
 
-We first generated species richness maps using 100x100 km grid cells for the raw and filtered datasets respectively, using the package \texttt{speciesgeocodeR\ v2.0-10} (Töpel et al., 2016).
-We then performed an automated conservation assessment for all study groups based on both datasets using the \texttt{ConR\ v1.2.4} package (Dauby et al., 2017).
+We first generated species richness maps using 100x100 km grid cells for the raw and filtered datasets respectively, using the package \texttt{speciesgeocodeR\ v2.0-10} (Töpel et al.
+2016).
+We then performed an automated conservation assessment for all study groups based on both datasets using the \texttt{ConR\ v1.2.4} package (Dauby et al.
+2017).
 \texttt{ConR} estimates the EOO, AOO, and the number of locations, and then suggests a preliminary conservation status based on Criterion B of the global IUCN Red List.
-While these assessments are preliminary (see IUCN Standards and Petitions Subcommittee, 2017), they can be a proxy used by the IUCN to speed up full assessments.
-We then benchmarked the preliminary conservation assessments against the global IUCN Red List assessments as far as available for the same taxa, which we obtained from www.iucn.org via the \texttt{rredlist\ v.0.5.0} package (Chamberlain, 2018).
+While these assessments are preliminary (see IUCN Standards and Petitions Subcommittee 2017), they can be a proxy used by the IUCN to speed up full assessments.
+We then benchmarked the preliminary conservation assessments against the global IUCN Red List assessments as far as available for the same taxa, which we obtained from www.iucn.org via the \texttt{rredlist\ v.0.5.0} package (Chamberlain 2018).
 
 \hypertarget{evaluation-of-results}{%
-\subsection{Evaluation of results}\label{evaluation-of-results}}
+\subsection*{Evaluation of results}\label{evaluation-of-results}}
+\addcontentsline{toc}{subsection}{Evaluation of results}
 
@@
 
-We did all analyses in the R computing environment (R Core Team, 2019), using standard libraries for data handling and visualization (Garnier, 2018; Hijmans, 2019; Ooms, 2014, 2019; Wickham, 2018).
+We did all analyses in the R computing environment (R Core Team 2019), using standard libraries for data handling and visualization (Wickham 2018; Garnier 2018; Ooms 2014, 2019; Hijmans 2019).
 All scripts are available from \url{https://github.com/idiv-biodiversity/effects_of_automated_cleaning}, or from the Appendix 1 in the electronic supplement of this study.
@@
 \hypertarget{results}{%
-\section{Results}\label{results}}
+\section*{Results}\label{results}}
+\addcontentsline{toc}{section}{Results}
 
-We retrieved a total of 215,262 species occurrence records, with a median of 3,460 records per study group and 11 records per species (Table 3, Appendix 2).
+We retrieved a total of 218,899 species occurrence records, with a median of 2,844.5 records per study group and 10 records per species (Table 3, Appendix 2).
 We obtained most records for Dipsadidae (64,249) and fewest for \emph{Thozetella} (51).
@@
 
-Our automated tests flagged a total of 88,648 records (Fig.
-\ref{fig:total}), with a median of 40.3\% per group.
-Overall, the most important test was for duplicated records (on average 32.9\% per taxonomic group).
-The filtering steps based on record meta-data that flagged the largest number of records were the basis of records (2.1\%) and the collection year (3.3\%).
-The most important automated tests were for urban area (5\%), the capital coordinates (1.1\%) and the centroids of political units (0.5\%, see Table 3 and Supplementary material 1 for further details and the absolute numbers).
+Our automated tests flagged a total of 91,434 records (Fig.
+\ref{fig:total}), with a median of 42.7\% per group.
+Overall, the most important test was for duplicated records (on average 35.5\% per taxonomic group).
+The filtering steps based on record meta-data that flagged the largest number of records were the basis of records (5.9\%) and the collection year (3.4\%).
+The most important automated tests were for urban area (8.6\%), the capital coordinates (1\%) and the centroids of political units (0.5\%, see Table 3 and Supplementary material 1 for further details and the absolute numbers).
 Only few records were flagged by the zero coordinates and biodiversity institution tests (Fig.
@@
 \hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\section*{Discussion}\label{discussion}}
+\addcontentsline{toc}{section}{Discussion}
 
-Automated flagging based on meta-data and automatic tests flagged on average 40.3\% of the records per taxonomic group.
+Automated flagging based on meta-data and automatic tests flagged on average 42.7\% of the records per taxonomic group.
 The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (Question 1).
@@
 \ref{fig:speciesrichness}, Supplementary material 1).
-The filter based on political centroids had an important impact on species richness patterns, which is congruent with the results from a previous study in the coffee family (Maldonado et al., 2015).
+The filter based on political centroids had an important impact on species richness patterns, which is congruent with the results from a previous study in the coffee family (Maldonado et al.
+2015).
 Records assigned to country or province centroids are often old records, which are geo-referenced at a later point based on vague locality descriptions.
@@
 
-The accuracy of the automated conservation assessment was in the same range as found by previous studies (Nic Lughadha et al., 2019; Zizka, Azevedo, et al., 2019).
-The similar accuracy of the raw and filtered dataset for the automated conservation assessment was surprising, in particular given the EOO and AOO reduction observed in the filtered dataset (Table 4) and the impact of errors on spatial analyses observed in previous studies (Gueta \& Carmel, 2016).
+The accuracy of the automated conservation assessment was in the same range as found by previous studies (Nic Lughadha et al.
+2019; Zizka, Azevedo, et al.
+2019).
+The similar accuracy of the raw and filtered dataset for the automated conservation assessment was surprising, in particular given the EOO and AOO reduction observed in the filtered dataset (Table 4) and the impact of errors on spatial analyses observed in previous studies (Gueta and Carmel 2016).
 The robustness of the automated assessment was likely due to the fact that the EOO for most species was large, even after the considerable reduction caused by filtering.
@@
 Those inclusion criteria are likely to have, biased the datasets towards species with large ranges, since generally more data for them are available.
-The robustness of automated conservation assessments to data quality is encouraging, although these methods are only an approximation, especially for species with few collection records (Rivers et al., 2011).
+The robustness of automated conservation assessments to data quality is encouraging, although these methods are only an approximation, especially for species with few collection records (Rivers et al.
+2011).
 
 \hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
+\section*{Conclusions}\label{conclusions}}
+\addcontentsline{toc}{section}{Conclusions}
 
@@
 
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
+\hypertarget{acknowledgments}{%
+\section*{Acknowledgments}\label{acknowledgments}}
+\addcontentsline{toc}{section}{Acknowledgments}
 
@@
 \hypertarget{supplementary-material}{%
-\section{Supplementary material}\label{supplementary-material}}
+\section*{Supplementary material}\label{supplementary-material}}
+\addcontentsline{toc}{section}{Supplementary material}
 
+\begin{itemize}
+\item
   Appendix 1 - Analyses scripts
-
+\item
   Appendix 2 - Table S1, reporting the absolute number of flagged records per taxonomic group and test
-
+\item
   Appendix 3 - The full results of the conservation assessment
-
+\item
   Appendix 4 - Taxon specific richness maps and comments.
-
-\newpage{}
+\end{itemize}
 
 \hypertarget{tables}{%
-\section{Tables}\label{tables}}
+\section*{Tables}\label{tables}}
+\addcontentsline{toc}{section}{Tables}
 
@@
 From column four onwards the numbers show the percentage of records flagged by the respective test.
-Only tests that flagged at least 0.1\% of the records in any group are shown.}
+Only tests that flagged at least 0.1\% of the records in any group are shown.
+Individual records can be flagged by multiple tests, therefore the sum of percentge from all tests can supersede the total percentage.}
 \centering
@@
 
-\newpage{}
+\hypertarget{figures}{%
+\section*{Figures}\label{figures}}
+\addcontentsline{toc}{section}{Figures}
 
-\hypertarget{figures}{%
-\section{Figures}\label{figures}}
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{./figures_tables/Figure_study_groups} \caption{Example of study taxa.
+\textbf{A)} Pilosocereus pusillibaccatus, picture by Pamela Lavor; \textbf{B)} Conchocarpus macrocarpus, Juliana El-Ottra; \textbf{C)} Tillandsia recurva, EC Tomaz; \textbf{D} \textit{Oxyrhopus guibei} (Dipsadidae), Filipe C Serrano;\textbf{E)} Neanuridae (\textit{Aethiopella ricardoi}), Raiane Vital da Paz; \textbf{F)} \textit{Tocoyena formosa}, Fernanda GL Moreira; \textbf{G)} \textit{Harengula jaguana}, Thais Ferreira-Araujo; \textbf{H)} \textit{Gaylussacia decipiens}, Luiz Menini Neto; \textbf{I)}  extit{Ooocephalus foliosus}, Arthur Soares; \textbf{J)} \textit{Tityus carvalhoi}, Renata C Santos-Costa; \textbf{K)} \textit{Prostheca vespa}, Tiago veira}\label{fig:species}
+\end{figure}
 
@@
 
-\newpage{}
-
 \hypertarget{references}{%
@@
 \leavevmode\hypertarget{ref-Anderson2016}{}%
-Anderson, R.
-P., Araújo, M., Guisan, A., Lobo, J.
-M., Martínez-Meyer, E., Peterson, T., \& Soberón, J.
-(2016).
-\emph{Final Report of the Task Group on GBIF Data Fitness for Use in Distribution Modelling - Are species occurrence data in global online repositories fit for modeling species distributions?
-The case of the Global Biodiversity Information Facility (GBIF)} (p.
-27pp).
-GBIF.
+Anderson, Robert P, Miguel Araújo, Antoine Guisan, Jorge M Lobo, Enrique Martínez-Meyer, Townsend Peterson, and Jorge Soberón.
+2016.
+``Final Report of the Task Group on GBIF Data Fitness for Use in Distribution Modelling - Are species occurrence data in global online repositories fit for modeling species distributions?
+The case of the Global Biodiversity Information Facility (GBIF).'' GBIF.
 
 \leavevmode\hypertarget{ref-Bachman2011}{}%
-Bachman, S.
-P., Moat, J., Hill, A., Torre, J.
-de la, \& Scott, B.
-(2011).
-Supporting Red List threat assessments with GeoCAT: geospatial conservation assessment tool.
-\emph{ZooKeys}, \emph{150}, 117--126.
-\url{https://doi.org/10.3897/zookeys.150.2109}
+Bachman, Steven P., Justin Moat, Andrew Hill, Javier de la Torre, and Ben Scott.
+2011.
+``Supporting Red List threat assessments with GeoCAT: geospatial conservation assessment tool.'' \emph{ZooKeys} 150 (November): 117--26.
+\url{https://doi.org/10.3897/zookeys.150.2109}.
 
 \leavevmode\hypertarget{ref-Chamberlain2016}{}%
-Chamberlain, S. (2016). scrubr: Clean Biological Occurrence Records. Retrieved from \url{https://cran.r-project.org/package=scrubr}
+Chamberlain, Scott. 2016. ``scrubr: Clean Biological Occurrence Records.'' \url{https://cran.r-project.org/package=scrubr}.
 
 \leavevmode\hypertarget{ref-Chamberlain2018}{}%
-Chamberlain, S. (2018). \emph{rredlist: 'IUCN' Red List Client}. Retrieved from \url{https://cran.r-project.org/package=rredlist}
+---------. 2018. \emph{rredlist: 'IUCN' Red List Client}. \url{https://cran.r-project.org/package=rredlist}.
 
 \leavevmode\hypertarget{ref-Chamberlain2017}{}%
-Chamberlain, S. A. (2017). rgbif: Interface to the Global Biodiversity Information Facility API. R package version 0.9.9. Retrieved from \url{https://github.com/ropensci/rgbif}
+Chamberlain, Scott A. 2017. ``rgbif: Interface to the Global Biodiversity Information Facility API. R package version 0.9.9.'' \url{https://github.com/ropensci/rgbif}.
 
 \leavevmode\hypertarget{ref-Cosiaux2018}{}%
-Cosiaux, A., Gardiner, L.
-M., Stauffer, F.
-W., Bachman, S.
-P., Sonké, B., Baker, W.
-J., \& Couvreur, T.
-L.
+Cosiaux, Ariane, Lauren M.
+Gardiner, Fred W.
+Stauffer, Steven P.
+Bachman, Bonaventure Sonké, William J.
+Baker, and Thomas L.
 P.
-(2018).
-Low extinction risk for an important plant resource: Conservation assessments of continental African palms (Arecaceae/Palmae).
-\emph{Biological Conservation}, \emph{221}, 323--333.
-\url{https://doi.org/10.1016/j.biocon.2018.02.025}
+Couvreur.
+2018.
+``Low extinction risk for an important plant resource: Conservation assessments of continental African palms (Arecaceae/Palmae).'' \emph{Biological Conservation} 221 (May): 323--33.
+\url{https://doi.org/10.1016/j.biocon.2018.02.025}.
 
 \leavevmode\hypertarget{ref-Dauby2017}{}%
-Dauby, G., Stévart, T., Droissart, V., Cosiaux, A., Deblauwe, V., Simo-Droissart, M., \ldots{} Couvreur, T.
-L.
-P.
-(2017).
-ConR: An R package to assist large-scale multispecies preliminary conservation assessments using distribution data.
-\emph{Ecology and Evolution}, \emph{7}(24), 11292--11303.
-\url{https://doi.org/10.1002/ece3.3704}
+Dauby, Gilles, Tariq Stévart, Vincent Droissart, Ariane Cosiaux, Vincent Deblauwe, Murielle Simo-Droissart, Marc S.
+M.
+Sosef, et al.
+2017.
+``ConR: An R package to assist large-scale multispecies preliminary conservation assessments using distribution data.'' \emph{Ecology and Evolution} 7 (24): 11292--11303.
+\url{https://doi.org/10.1002/ece3.3704}.
 
 \leavevmode\hypertarget{ref-Garnier2018}{}%
-Garnier, S. (2018). \emph{viridis: Default color maps from 'matplotlib'}. Retrieved from \url{https://cran.r-project.org/package=viridis}
+Garnier, Simon. 2018. \emph{viridis: Default color maps from 'matplotlib'}. \url{https://cran.r-project.org/package=viridis}.
 
 \leavevmode\hypertarget{ref-GBIForg2019}{}%
-GBIF.org. (2019a). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.sojrfp}
+GBIF.org. 2019a. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.sojrfp}.
 
 \leavevmode\hypertarget{ref-GBIForg2019a}{}%
-GBIF.org. (2019b). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.bx0jjw}
+---------. 2019b. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.bx0jjw}.
 
 \leavevmode\hypertarget{ref-GBIForg2019b}{}%
-GBIF.org. (2019c). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.zv6kuq}
+---------. 2019c. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.zv6kuq}.
 
 \leavevmode\hypertarget{ref-GBIForg2019c}{}%
-GBIF.org. (2019d). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.uutyb6}
+---------. 2019d. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.uutyb6}.
 
 \leavevmode\hypertarget{ref-GBIForg2019d}{}%
-GBIF.org. (2019e). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.8hnzfo}
+---------. 2019e. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.8hnzfo}.
 
 \leavevmode\hypertarget{ref-GBIForg2019e}{}%
-GBIF.org. (2019f). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.zznjbv}
+---------. 2019f. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.zznjbv}.
 
 \leavevmode\hypertarget{ref-GBIForg2019f}{}%
-GBIF.org. (2019g). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.zjjpmh}
+---------. 2019g. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.zjjpmh}.
 
 \leavevmode\hypertarget{ref-GBIForg2019g}{}%
-GBIF.org. (2019h). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.4srw8a}
+---------. 2019h. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.4srw8a}.
 
 \leavevmode\hypertarget{ref-GBIForg2019h}{}%
-GBIF.org. (2019i). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.rpkjsh}
+---------. 2019i. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.rpkjsh}.
 
 \leavevmode\hypertarget{ref-GBIForg2019i}{}%
-GBIF.org. (2019j). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.762543}
+---------. 2019j. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.762543}.
 
 \leavevmode\hypertarget{ref-GBIForg2019j}{}%
-GBIF.org. (2019k). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.nmzgi9}
+---------. 2019k. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.nmzgi9}.
 
 \leavevmode\hypertarget{ref-GBIForg2019k}{}%
-GBIF.org. (2019l). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.scmkx5}
+---------. 2019l. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.scmkx5}.
 
 \leavevmode\hypertarget{ref-GBIForg2019l}{}%
-GBIF.org. (2019m). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.d34gos}
+---------. 2019m. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.d34gos}.
 
 \leavevmode\hypertarget{ref-GBIForg2019m}{}%
-GBIF.org. (2019n). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.6bzfz4}
+---------. 2019n. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.6bzfz4}.
 
 \leavevmode\hypertarget{ref-GBIForg2019n}{}%
-GBIF.org. (2019o). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.wkwque}
+---------. 2019o. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.wkwque}.
 
 \leavevmode\hypertarget{ref-GBIForg2019o}{}%
-GBIF.org. (2019p). (29 December 2019) GBIF Occurrence Download. \url{https://doi.org/doi.org/10.15468/dl.zj2cyj}
+---------. 2019p. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.zj2cyj}.
 
 \leavevmode\hypertarget{ref-Gueta2016}{}%
-Gueta, T., \& Carmel, Y.
-(2016).
-Quantifying the value of user-level data cleaning for big data: A case study using mammal distribution models.
-\emph{Ecological Informatics}, \emph{34}, 139--145.
-\url{https://doi.org/10.1016/j.ecoinf.2016.06.001}
+Gueta, Tomer, and Yohay Carmel.
+2016.
+``Quantifying the value of user-level data cleaning for big data: A case study using mammal distribution models.'' \emph{Ecological Informatics} 34: 139--45.
+\url{https://doi.org/10.1016/j.ecoinf.2016.06.001}.
 
 \leavevmode\hypertarget{ref-Hijmans2019}{}%
-Hijmans, R. J. (2019). raster: Geographic data analysis and modeling. Retrieved from \url{https://cran.r-project.org/package=raster}
+Hijmans, Robert J. 2019. ``raster: Geographic data analysis and modeling.'' \url{https://cran.r-project.org/package=raster}.
 
@@
 IUCN Standards and Petitions Subcommittee.
-(2017).
-\emph{Guidelines for Using the IUCN Red List - Categories and Criteria.
+2017.
+``Guidelines for Using the IUCN Red List - Categories and Criteria.
 Version 13.
 Prepared by the Standards and Petitions Subcommittee.
-Downloadable from http://www.iucnredlist.org/documents/RedListGuidelines.pdf} (pp.
-1--60).
+Downloadable from http://www.iucnredlist.org/documents/RedListGuidelines.pdf.''
 
 \leavevmode\hypertarget{ref-Jin2020}{}%
-Jin, J., \& Yang, J.
-(2020).
-BDcleaner: A workflow for cleaning taxonomic and geographic errors in occurrence data archived in biodiversity databases.
-\emph{Global Ecology and Conservation}, \emph{21}, e00852.
-\url{https://doi.org/10.1016/j.gecco.2019.e00852}
+Jin, Jing, and Jun Yang.
+2020.
+``BDcleaner: A workflow for cleaning taxonomic and geographic errors in occurrence data archived in biodiversity databases.'' \emph{Global Ecology and Conservation} 21 (March): e00852.
+\url{https://doi.org/10.1016/j.gecco.2019.e00852}.
 
 \leavevmode\hypertarget{ref-Maldonado2015}{}%
-Maldonado, C., Molina, C.
-I., Zizka, A., Persson, C., Taylor, C.
-M., Albán, J., \ldots{} Antonelli, A.
-(2015).
-Estimating species diversity and distribution in the era of Big Data: to what extent can we trust public databases?
-\emph{Global Ecology and Biogeography}, \emph{24}(8), 973--984.
-\url{https://doi.org/10.1111/geb.12326}
+Maldonado, Carla, Carlos I.
+Molina, Alexander Zizka, Claes Persson, Charlotte M.
+Taylor, Joaquina Albán, Eder Chilquillo, Nina Rønsted, and Alexandre Antonelli.
+2015.
+``Estimating species diversity and distribution in the era of Big Data: to what extent can we trust public databases?'' \emph{Global Ecology and Biogeography} 24 (8): 973--84.
+\url{https://doi.org/10.1111/geb.12326}.
 
 \leavevmode\hypertarget{ref-Morrone2014}{}%
-Morrone, J. J. (2014). Biogeographical regionalisation of the Neotropical region. \emph{Zootaxa}, \emph{3782}(1), 1. \url{https://doi.org/10.11646/zootaxa.3782.1.1}
+Morrone, Juan J. 2014. ``Biogeographical regionalisation of the Neotropical region.'' \emph{Zootaxa} 3782 (1): 1. \url{https://doi.org/10.11646/zootaxa.3782.1.1}.
 
 \leavevmode\hypertarget{ref-NicLughadha2019}{}%
-Nic Lughadha, E., Walker, B.
-E., Canteiro, C., Chadburn, H., Davis, A.
-P., Hargreaves, S., \ldots{} Rivers, M.
-C.
-(2019).
-The use and misuse of herbarium specimens in evaluating plant extinction risks.
-\emph{Philosophical Transactions of the Royal Society B: Biological Sciences}, \emph{374}(1763), 20170402.
-\url{https://doi.org/10.1098/rstb.2017.0402}
+Nic Lughadha, Eimear, Barnaby E.
+Walker, Cátia Canteiro, Helen Chadburn, Aaron P.
+Davis, Serene Hargreaves, Eve J.
+Lucas, et al.
+2019.
+``The use and misuse of herbarium specimens in evaluating plant extinction risks.'' \emph{Philosophical Transactions of the Royal Society B: Biological Sciences} 374 (1763): 20170402.
+\url{https://doi.org/10.1098/rstb.2017.0402}.
 
 \leavevmode\hypertarget{ref-Ooms2014}{}%
-Ooms, J. (2014). The jsonlite Package: A Practical and Consistent Mapping Between JSON Data and R Objects. \emph{arXiv}. Retrieved from \url{https://arxiv.org/abs/1403.2805}
+Ooms, Jeroen. 2014. ``The jsonlite Package: A Practical and Consistent Mapping Between JSON Data and R Objects.'' \emph{arXiv}. \url{https://arxiv.org/abs/1403.2805}.
 
 \leavevmode\hypertarget{ref-Ooms2019}{}%
-Ooms, J. (2019). \emph{writexl: Export Data Frames to Excel 'xlsx' Format}. Retrieved from \url{https://cran.r-project.org/package=writexl}
+---------. 2019. \emph{writexl: Export Data Frames to Excel 'xlsx' Format}. \url{https://cran.r-project.org/package=writexl}.
 
 \leavevmode\hypertarget{ref-Pelletier2018}{}%
-Pelletier, T.
-A., Carstens, B.
-C., Tank, D.
-C., Sullivan, J., \& Espíndola, A.
-(2018).
-Predicting plant conservation priorities on a global scale.
-\emph{Proceedings of the National Academy of Sciences}, \emph{115}(51), 13027--13032.
-\url{https://doi.org/10.1073/pnas.1804098115}
+Pelletier, Tara A., Bryan C.
+Carstens, David C.
+Tank, Jack Sullivan, and Anahí Espíndola.
+2018.
+``Predicting plant conservation priorities on a global scale.'' \emph{Proceedings of the National Academy of Sciences} 115 (51): 13027--32.
+\url{https://doi.org/10.1073/pnas.1804098115}.
 
 \leavevmode\hypertarget{ref-rcoreteam2019}{}%
-R Core Team. (2019). R: A language and environment for statistical computing. Austria, Vienna: R Foundation for Statistical Computing. Retrieved from \url{https://www.r-project.org/}
+R Core Team. 2019. ``R: A language and environment for statistical computing.'' Austria, Vienna: R Foundation for Statistical Computing. \url{https://www.r-project.org/}.
 
 \leavevmode\hypertarget{ref-Rivers2011}{}%
-Rivers, M.
-C., Taylor, L., Brummitt, N.
-a., Meagher, T.
-R., Roberts, D.
-L., \& Lughadha, E.
-N.
-(2011).
-How many herbarium specimens are needed to detect threatened species?
-\emph{Biological Conservation}, \emph{144}(10), 2541--2547.
-\url{https://doi.org/10.1016/j.biocon.2011.07.014}
+Rivers, Malin C., Lin Taylor, Neil a.
+Brummitt, Thomas R.
+Meagher, David L.
+Roberts, and Eimear Nic Lughadha.
+2011.
+``How many herbarium specimens are needed to detect threatened species?'' \emph{Biological Conservation} 144 (10): 2541--7.
+\url{https://doi.org/10.1016/j.biocon.2011.07.014}.
 
 \leavevmode\hypertarget{ref-Robertson2016}{}%
-Robertson, M.
-P., Visser, V., \& Hui, C.
-(2016).
-Biogeo: an R package for assessing and improving data quality of occurrence record datasets.
-\emph{Ecography}, \emph{39}, 394--401.
-\url{https://doi.org/10.1111/ecog.02118}
+Robertson, Mark P, Vernon Visser, and Cang Hui.
+2016.
+``Biogeo: an R package for assessing and improving data quality of occurrence record datasets.'' \emph{Ecography} 39: 394--401.
+\url{https://doi.org/10.1111/ecog.02118}.
 
 \leavevmode\hypertarget{ref-Schmidt2017}{}%
-Schmidt, M., Zizka, A., Traoré, S., Ataholo, M., Chatelain, C., Daget, P., \ldots{} Thiombiano, A.
-(2017).
-Diversity, distribution and preliminary conservation status of the flora of Burkina Faso.
-\emph{Phytotaxa Monographs}, \emph{304}(1), 1--215.
+Schmidt, Marco, Alexander Zizka, Salifou Traoré, Mandingo Ataholo, Cyrille Chatelain, Philippe Daget, Stefan Dressler, et al.
+2017.
+``Diversity, distribution and preliminary conservation status of the flora of Burkina Faso.'' \emph{Phytotaxa Monographs} 304 (1): 1--215.
 
 \leavevmode\hypertarget{ref-Stevart2019}{}%
-Stévart, T., Dauby, G., Lowry, P.
-P., Blach-Overgaard, A., Droissart, V., Harris, D.
-J., \ldots{} Couvreur, T.
-L.
+Stévart, T., G.
+Dauby, P.
 P.
-(2019).
-A third of the tropical African flora is potentially threatened with extinction.
-\emph{Science Advances}, \emph{5}(11), eaax9444.
-\url{https://doi.org/10.1126/sciadv.aax9444}
+Lowry, A.
+Blach-Overgaard, V.
+Droissart, D.
+J.
+Harris, B.
+A.
+Mackinder, et al.
+2019.
+``A third of the tropical African flora is potentially threatened with extinction.'' \emph{Science Advances} 5 (11): eaax9444.
+\url{https://doi.org/10.1126/sciadv.aax9444}.
 
 \leavevmode\hypertarget{ref-Topel2016}{}%
-Töpel, M., Zizka, A., Calió, M.
-F., Scharn, R., Silvestro, D., \& Antonelli, A.
-(2016).
-SpeciesGeoCoder: Fast Categorization of Species Occurrences for Analyses of Biodiversity, Biogeography, Ecology, and Evolution.
-\emph{Systematic Biology}, \emph{66}(2), syw064.
-\url{https://doi.org/10.1093/sysbio/syw064}
+Töpel, Mats, Alexander Zizka, Maria Fernanda Calió, Ruud Scharn, Daniele Silvestro, and Alexandre Antonelli.
+2016.
+``SpeciesGeoCoder: Fast Categorization of Species Occurrences for Analyses of Biodiversity, Biogeography, Ecology, and Evolution.'' \emph{Systematic Biology} 66 (2): syw064.
+\url{https://doi.org/10.1093/sysbio/syw064}.
 
 \leavevmode\hypertarget{ref-Wickham2018}{}%
-Wickham, H. (2018). tidyverse: Easily install and load the 'Tidyverse'. Retrieved from \url{https://cran.r-project.org/package=tidyverse}
+Wickham, Hadley. 2018. ``tidyverse: Easily install and load the 'Tidyverse'.'' \url{https://cran.r-project.org/package=tidyverse}.
 
 \leavevmode\hypertarget{ref-Yesson2007}{}%
-Yesson, C., Brewer, P.
-W., Sutton, T., Caithness, N., Pahwa, J.
-S., Burgess, M., \ldots{} Culham, A.
-(2007).
-How Global Is the Global Biodiversity Information Facility?
-\emph{PLoS ONE}, \emph{2}(11), e1124.
-\url{https://doi.org/10.1371/journal.pone.0001124}
+Yesson, Chris, Peter W Brewer, Tim Sutton, Neil Caithness, Jaspreet S Pahwa, Mikhaila Burgess, W Alec Gray, et al.
+2007.
+``How Global Is the Global Biodiversity Information Facility?'' Edited by James Beach.
+\emph{PLoS ONE} 2 (11): e1124.
+\url{https://doi.org/10.1371/journal.pone.0001124}.
 
 \leavevmode\hypertarget{ref-Zizka2019a}{}%
-Zizka, A., Azevedo, J., Leme, E., Neves, B., Ferreira, A., Caceres, D., \& Zizka, G.
-(2019).
-Biogeography and conservation status of the pineapple family ( Bromeliaceae ), (June), 1--13.
-\url{https://doi.org/10.1111/ddi.13004}
+Zizka, Alexander, Josue Azevedo, Elton Leme, Beatriz Neves, Andrea Ferreira, Daniel Caceres, and Georg Zizka.
+2019.
+``Biogeography and conservation status of the pineapple family ( Bromeliaceae ),'' no.
+June: 1--13.
+\url{https://doi.org/10.1111/ddi.13004}.
 
 \leavevmode\hypertarget{ref-Zizka2019}{}%
-Zizka, A., Silvestro, D., Andermann, T., Azevedo, J., Duarte Ritter, C., Edler, D., \ldots{} Antonelli, A.
-(2019).
-CoordinateCleaner: Standardized cleaning of occurrence records from biological collection databases.
-\emph{Methods in Ecology and Evolution}, \emph{10}(5), 744--751.
-\url{https://doi.org/10.1111/2041-210X.13152}
+Zizka, Alexander, Daniele Silvestro, Tobias Andermann, Josué Azevedo, Camila Duarte Ritter, Daniel Edler, Harith Farooq, et al.
+2019.
+``CoordinateCleaner: Standardized cleaning of occurrence records from biological collection databases.'' Edited by Tiago Quental.
+\emph{Methods in Ecology and Evolution} 10 (5): 744--51.
+\url{https://doi.org/10.1111/2041-210X.13152}.
 

### 2020-02-29 02_automated_cleaning.R

@@ -16,7 +16,7 @@ for(i in 1:length(li)){
     mutate_all(as.character)
   if(nrow(sub) > 0){
     sub$taxon <- str_split(li[i], pattern = "_")[[1]][1]
-    sub <- select(sub, taxon, decimalLongitude, decimalLatitude,
+    sub <- dplyr::select(sub, taxon, decimalLongitude, decimalLatitude,
                   coordinateUncertaintyInMeters, 
                   countryCode, gbifID, species, class, family, genus, taxonRank,
                   year, basisOfRecord, individualCount)
@@ -65,8 +65,17 @@ dat <- clean_coordinates(x = dat,
                            seas_ref = buffland) # most test are on by default
 
 # convert .seas flag for the seas taxa
+sea2 <- cc_sea(x = as.data.frame(dat), 
+               lon = "decimalLongitude", 
+               lat = "decimalLatitude", value = "flagged")
+
 dat <- dat %>% 
-  mutate(.sea = ifelse(taxon == "Arhynchobatidae" | taxon == "Harengula" | taxon == "Opisthonema", TRUE, TRUE))
+  mutate(.sea2 = !sea2) %>% 
+  mutate(.sea = ifelse(taxon == "Arhynchobatidae" |
+                         taxon == "Harengula" |
+                         taxon == "Opisthonema" |
+                         taxon == "Diogenidae", .sea2, .sea)) %>% 
+  dplyr::select(-.sea2)
 
 # summarize the records and taxa remaining
 out <- dat %>% 

### 2020-02-29 03_automated_conservation_Assessment.R

@@ -33,39 +33,38 @@ cl_in <- dat_cl%>%
 
 
 # AUtomated assessment
-
 ## raw
-splist <- unique(rw_in$tax)
-
-for(i in 1:length(splist)){
-  print(paste(i, length(splist), sep = "/"))
-  sub <- rw_in %>% filter(tax == splist[i])
-  
-  out <- IUCN.eval(sub)
-  
-  if(i == 1){
-    write_csv(out, "output/automated_assessment_raw.csv")
-  }else{
-    write_csv(out, "output/automated_assessment_raw.csv", append = TRUE)
-  }
-}
-
-## filtered
-splist <- unique(cl_in$tax)
-
-for(i in 1:length(splist)){
-  print(paste(i, length(splist), sep = "/"))
-  sub <- cl_in %>% filter(tax == splist[i])
-  
-  out <- IUCN.eval(sub)
-  
-  if(i == 1){
-    write_csv(out, "output/automated_assessment_clean.csv")
-  }else{
-    write_csv(out, "output/automated_assessment_clean.csv", append = TRUE)
-  }
-}
-
+# splist <- unique(rw_in$tax)
+# 
+# for(i in 1:length(splist)){
+#   print(paste(i, length(splist), sep = "/"))
+#   sub <- rw_in %>% filter(tax == splist[i])
+#   
+#   out <- IUCN.eval(sub)
+#   
+#   if(i == 1){
+#     write_csv(out, "output/automated_assessment_raw.csv")
+#   }else{
+#     write_csv(out, "output/automated_assessment_raw.csv", append = TRUE)
+#   }
+# }
+# 
+# ## filtered
+# splist <- unique(cl_in$tax)
+# 
+# for(i in 1:length(splist)){
+#   print(paste(i, length(splist), sep = "/"))
+#   sub <- cl_in %>% filter(tax == splist[i])
+#   
+#   out <- IUCN.eval(sub)
+#   
+#   if(i == 1){
+#     write_csv(out, "output/automated_assessment_clean.csv")
+#   }else{
+#     write_csv(out, "output/automated_assessment_clean.csv", append = TRUE)
+#   }
+# }
+# 
 
 # IUCN assessment for the species
 iucn.key <- "01524b67f4972521acd1ded2d8b3858e7fedc7da5fd75b8bb2c5456ea18b01ba"

### 2020-03-02 manuscript_peerj.Rmd
co-author version

@@
     affiliation: 4
+  - name: Tiago Andrade Borges Santos
+    affiliation: 5
   - name: Andressa Cabral
-    affiliation: 5
-  - name: Arthur de Souza Soares
-    affiliation: 4
+    affiliation: 6
+  - name: Filipe Cabreirinha Serrano
+    affiliation: 7
   - name: Eduardo Calisto Tomaz
-    affiliation: 4
-  - name: Fernanda Gondim Lambert Moreira
-    affiliation: 4
-  - name: Filipe Cabreirinha Serrano
-    affiliation: 6
-  - name: Jéssica Fernanda Ramos Coelho
-    affiliation: 4
-  - name: Mabel Rocio Baez-Lizarazo
-    affiliation: 7
-  - name: Mariana Ramos Fantinati
-    affiliation: 4
-  - name: Matheus Colli-Silva
-    affiliation: 5
-  - name: Moabe Ferreira Fernandes
-    affiliation: 8
-  - name: Nathália Michelly da Cunha Santos
     affiliation: 4
@@
     affiliation: 4
+  - name: Matheus Colli-Silva
+    affiliation: 6
+  - name: Nathália Michelly da Cunha Santos
+    affiliation: 4
+  - name: Arthur de Souza Soares
+    affiliation: 4
   - name: Thais Ferreira-Araújo
     affiliation: 4
-  - name: Tiago Andrade Borges Santos
-    affiliation: 9
-  - name: Tiago Luiz Vieira
+  - name: Moabe Ferreira Fernandes
     affiliation: 9
@@
     affiliation: 4
+  - name: Fernanda Gondim Lambert Moreira
+    affiliation: 4
+  - name: Jéssica Fernanda Ramos Coelho
+    affiliation: 4
+  - name: Mariana Ramos Fantinati
+    affiliation: 4
+  - name: Mabel Rocio Baez-Lizarazo
+    affiliation: 8
+  - name: Tiago Luiz Vieira
+    affiliation: 6
   - name: Alexandre Antonelli
@@
   - code: 1
-    address: sDiv, German Center for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Leipzig, Germany
+    address: sDiv, German Center for Integrative Biodiversity Research Halle-Jena-Leipzig (iDiv), Leipzig, Germany
   - code: 2
@@
   - code: 5
+    address: Universidade Estadual de Feira de Santana, Feira de Santana, Brazil
+  - code: 6
     address: Departamento de Botânica, Universidade de São Paulo, São Paulo, Brazil
-  - code: 6
+  - code: 7
     address: Departamento de Ecologia, Universidade de São Paulo, São Paulo, Brazil
-  - code: 7
+  - code: 8
     address: Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil
-  - code: 8
+  - code: 9
     address: Departamento de Ciências Biológicas, Universidade Estadual de Feira de Santana, Feira de Santana, Brazil
-  - code: 9
-    address: Universidade Estadual de Feira de Santana, Bahia, Brazil
   - code: 10
@@
 However, it is unclear how much data may be lost by filtering, whether the same tests should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses.
-Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 16 Neotropical taxa including plants and animals, terrestrial and marine, downloaded from GBIF.
-We find that 29-67% of the records are potentially erroneous, with large variation across taxonomic groups.
-Tests for duplicated information, collection year, basis of record as well as urban areas and political centroids have the greatest effect.
+Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 18 Neotropical taxa including animals, fungi, and plants, terrestrial and marine, downloaded from GBIF.
+We find that 29-90% of the records are potentially erroneous, with large variation across taxonomic groups.
+Tests for duplicated information, collection year, basis of record as well as urban areas and coordinates for terrestrial taxa in the sea or marine taxa on land have the greatest effect.
 While many flagged records might not be *de facto* erroneous, they could be overly imprecise and increase uncertainty in downstream analyses.
@@
 nocite: |
-    @GBIForg2019; @GBIForg2019a; @GBIForg2019b; @GBIForg2019c; @GBIForg2019d; @GBIForg2019e; @GBIForg2019f; @GBIForg2019g; @GBIForg2019h; @GBIForg2019i; @GBIForg2019j; @GBIForg2019k; @GBIForg2019l; @GBIForg2019m; @GBIForg2019n; @GBIForg2019o;@GBIForg2019p;@GBIForg2019q
+    @GBIForg2019; @GBIForg2019a; @GBIForg2019b; @GBIForg2019c; @GBIForg2019d; @GBIForg2019e; @GBIForg2019f; @GBIForg2019g; @GBIForg2019h; @GBIForg2019i; @GBIForg2019j; @GBIForg2019k; @GBIForg2019l; @GBIForg2019m; @GBIForg2019n; @GBIForg2019o;@GBIForg2019p;@GBIForg2020, @GBIForg2020a
 ---
@@
 Those methods are usually based on geographic gazetteers [e.g., @Chamberlain2016; @Zizka2019; @Jin2020] or on additional data, such as environmental variables [@Robertson2016].
-Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biodiveristy research [@Zizka2019].
+Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the suitability of publicly available occurrence records for biodiversity research [@Zizka2019].
 
@@
 These methods approximate species' range size, namely the Extent of Occurrence (EOO, which is the area of a convex hull polygon comprising all records of a species), the Area of Occupancy (AOO, which is the sum of the area actually occupied by a species, calculated based on a small-scale regular grid), and the number of locations for a preliminary conservation assessment following IUCN Criterion B ("Geographic range").
-These methods have been used topropose preliminary global [@Stevart2019; @Zizka2019a] and regional [@Schmidt2017; @Cosiaux2018] Red List assessments.
-However, all metrics but especially EOO, are sensitive to individual records with erroneous coordinates **(REFERENCE)**.
+These methods have been used to propose preliminary global [@Stevart2019; @Zizka2019a] and regional [@Schmidt2017; @Cosiaux2018] Red List assessments.
+However, all metrics but especially EOO, are sensitive to individual records with erroneous coordinates.
 Automated conservation assessments may therefore be biased, particularly if the number of records are few, as it is the case for many tropical species.
@@
 While newly proposed methods for automated cleaning of species records hold great promise for biodiversity research, their use across taxonomic groups and datasets remains poorly explored.
-Here, we test the effect of automated filtering of species geographic occurrence records on the number of records available in different groups of animals, funghi, and plants.
+Here, we test the effect of automated filtering of species geographic occurrence records on the number of records available in different groups of animals, fungi, and plants.
 Furthermore, we test the impact of automated filtering procedures for the accuracy of automated conservation assessments compared to full IUCN assessments.
@@
 ## Choice of study taxa {-}
-This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering graduate students, taxonomists and biogeographers working with different taxonomic groups of animals, funghi, and plants across the Neotropics.
-Each participant analysed geographic occurrence data from their taxonomic group of interest and provided an informed assessment of the results for their group.
-Hence, we include groups based on the expertise of the participants wrather than following some arbitrary choice of taxa and taxonomic ranks.
+This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering students and researchers working with different taxonomic groups of animals, fungi, and plants across the Neotropics.
+Each participant analysed geographic occurrence data from their taxonomic group of interest and commented on the results for their group.
+Hence, we include groups based on the expertise of the participants rather than following some arbitrary choice of taxa and taxonomic ranks.
 We acknowledge a varying degree in documented expertise and number of years working on each group.
-We obtained public occurrence records for 16 taxa, including one plant family, eight plant genera, one genus of fungi, two families and one genus of terrestrial arthropods, one family of snakes, one family of skates, and one genus of bony fish (Table 1).
+We obtained public occurrence records for 18 taxa, including one plant family, nine plant genera, one genus of fungi, three families and one genus of terrestrial arthropods, one family of snakes, one family of skates, and one genus of bony fish (Table 1).
 
@@
 
-We downloaded occurrence information for all study groups from www.gbif.org using the `rgbif` v1.4.0 package [@Chamberlain2017] in R (GBIF.org, 2019a-p).
+We downloaded occurrence information for all study groups from www.gbif.org using the `rgbif` v1.4.0 package [@Chamberlain2017] in R (GBIF.org, 2019a-p,2020a,b).
 We downloaded only records with geographic coordinates and limited the study area to latitudes smaller than 33$^\circ$ N and longitudes smaller than 35$^\circ$ W and larger than 120$^\circ$ W reflecting the Neotropics, our main area of expertise [@Morrone2014].
-The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, *Gaylussacia*, Iridaceae, Neanuridae, and *Tillandsia* for which we only obtained the Neotropical occurrences.
+The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, Entomobryidae, *Gaylussacia*, Iridaceae, Neanuridae, and *Tillandsia* for which we only obtained the Neotropical occurrences.
 
@@
 We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the raw data downloaded from GBIF ("raw", hereafter) using meta-data for those records for which they were available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99.
-Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e.
-test for duplicated records).
+Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e., test for duplicated records).
 In a second step, we used the `clean_coordinates` function of the `CoordinateCleaner v2.0-11` package [@Zizka2019] with default options to flag errors that are common to biological data sets (“filtered”, hereafter).
-These including: coordinates in the sea for terrestrial taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions.
+These include: coordinates in the sea for terrestrial taxa and on land for marine taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions.
 See Table 2 for a summary of all flagging tests we used.
@@
 -->
-Each specialist provided an informed assessment on the performance of the raw and cleaned datasets, respectively concerning: 1) the number of flagged records; 2) the importance of specific tests; 3) the accuracy of the overall species richness maps; 4) the accuracy of the preliminary conservation assessments.
+Each author provided an informed comment on the performance of the raw and cleaned datasets, respectively concerning the number of flagged records and the accuracy of the overall species richness maps and we compared the agreement between automated conservation assessments based on raw and filtered occurrences with the global IUCN Red List (www.iucn.org).
 
 We did all analyses in the R computing environment [@rcoreteam2019], using standard libraries for data handling and visualization [@Wickham2018; @Garnier2018; @Ooms2014; @Ooms2019; @Hijmans2019].
-All scripts are available from https://github.com/idiv-biodiversity/effects_of_automated_cleaning, or from the Appendix 1 in the electronic supplement of this study.
+All scripts are available from a zenodo repository (**DOI**).
 
@@
 tab2 <- read_csv("output/summary_table_flags.csv") %>% 
-  dplyr::select(-coordinate_missing, -invalid_coords, -gbif_coords,-inst_coords, -equal_coords) %>% 
+  dplyr::select(-coordinate_missing, -invalid_coords, -gbif_coords, -equal_coords) %>% 
   mutate(fraction = round(fraction * 100,1))
 
-tab2 <- tab2[,c(1:3,14,5,7,4,8,6,10,13,11,12,9)]
+tab2 <- tab2[,c(1:3,16,5,7,4,8,6,14,10,15,11:13,9)]
 
@@
   select(taxon) %>%  
+  bind_rows(data.frame(taxon = "Total")) %>% 
   left_join(tab2, by = "taxon") %>% 
@@
 
-We retrieved a total of `r nrow(dat)` species occurrence records, with a median of `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per study group and `r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per species (Table 3, Appendix 2).
+We retrieved a total of `r nrow(dat)` species occurrence records, with a median of `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per study group and `r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per species (Table 3, Appendix 1).
 We obtained most records for `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(taxon) %>%  unlist()` (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs) %>%  unlist()`) and fewest for *`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(taxon) %>%  unlist()`* (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(nrecs) %>%  unlist()`).
@@
 The filtering steps based on record meta-data that flagged the largest number of records were the basis of records (`r dat %>% group_by(taxon) %>% summarize(num = mean(!coordinate_base)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the collection year (`r dat %>% group_by(taxon) %>% summarize(num = mean(!record_age)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%).
-The most important automated tests were for urban area (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.urb)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%), the capital coordinates (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.cap)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the centroids of political units (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.cen)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%, see Table 3 and Supplementary material 1 for further details and the absolute numbers).
-Only few records were flagged by the zero coordinates and biodiversity institution tests (Fig.
-\ref{fig:split}), and no records were flagged by the sea test.
+The most important automated tests were for urban area (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.urb)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the occurrence from records of terrestrial taxa in the sea and marine taxa on land (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.sea)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%, see Table 3 and Appendix 1 in the electronic supplement for further details and the absolute numbers).
+Only few records were flagged by the coordinate precision, zero coordinates and biodiversity institution tests (Fig.
+\ref{fig:split}).
 
-*`r tab2 %>% arrange(fraction) %>% top_n(1) %>% select(taxon) %>% unlist()`*, *`r tab2 %>% arrange(fraction) %>% top_n(3) %>% select(taxon) %>% slice(2) %>% unlist()`*, and *`r tab2 %>% arrange(fraction) %>% top_n(3) %>% select(taxon) %>% slice(3) %>% unlist()`* had the highest fraction of flagged records (Table 3).
+*`r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(1) %>% unlist()`*, *`r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(2) %>% unlist()`*, and *`r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(3) %>% unlist()`* had the highest fraction of flagged records (Table 3).
 In general, the different filters we tested were of similar importance for different study groups.
-There were few outstanding exceptions, including the particularly high proportions of records flagged on the "basis of record test" for *Tityus* (7.0%), Dipsadidae (5.6%), *Prosthechea* (5.0%) and *Tillandsia* (4.9%), by the collection year for Dipsadidae (11.3%), by the taxonomic identification level for  *Tityus* (1.6%), by the capital coordinates for *Oocephalus* (6.1%) and *Gaylussacia* (3.2%), and by the urban areas test for *Oocephalus* (13.3%) and Iridaceae (12.3%).
+There were few outstanding exceptions, including the particularly high proportions of records flagged on the "basis of record test" for *Tityus* (7.0%), Dipsadidae (5.6%), *Prosthechea* (5.0%) and *Tillandsia* (4.9%), by the collection year for Dipsadidae (11.3%), by the taxonomic identification level for  *Tityus* (1.6%), by the capital coordinates for *Oocephalus* (6.1%) and *Gaylussacia* (3.2%), by the seas/land test for Diogenidae and *Thozetella*, and by the urban areas test for *Oocephalus* (13.3%) and Iridaceae (12.3%).
+Furthermore, Entomobryidae differed considerably from all other study taxa with exceptionally high numbers of records flagged by the "basis of record, "level of identification' and urban areas" tests.
 
@@
 \ref{fig:split}).
-See Supplementary material 2 for species richness maps using the raw and cleaned data for all study groups.
+See Appendix 2 for species richness maps using the raw and cleaned data for all study groups.
 
-We found IUCN assessments for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>% nrow()` species that were also included in our distribution data from 10 of our study groups (Table 4, Appendix 3).
-The fraction of species evaluated varied among the study group, with a maximum of `r tab3 %>% select(iucn_frac_eval) %>% unlist() %>%  max()`% for *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>% slice(1L) %>% select(taxon) %>%  unlist()`* and *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>%  select(taxon) %>% slice(2) %>% unlist()`* and a minimum of `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(iucn_frac_eval) %>%  unlist()`% for `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(taxon) %>%  unlist()` (note that the number of species varied considerably among groups).
+We found IUCN assessments for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>% nrow()` species that were also included in our distribution data from 11 of our study groups (Table 4, Appendix 3).
+The fraction of species evaluated varied among the study group, with a maximum of `r tab3 %>% select(iucn_frac_eval) %>% unlist() %>%  max()`% for *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>% slice(1L) %>% select(taxon) %>%  unlist()`* and *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>%  select(taxon) %>% slice(2) %>% unlist()`* and a minimum of `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(iucn_frac_eval) %>%  unlist()`% for `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(taxon) %>%  unlist()` (note that the number of total species varied considerably among groups).
 The median percentage of species per study group with an IUCN assessment was `r tab3 %>% select(iucn_frac_eval) %>% unlist() %>%  median()`%.
@@
 We obtained automated conservation assessments for `r aa %>% select(flag_Category_CriteriaB) %>% filter(!is.na(flag_Category_CriteriaB)) %>% nrow()` species in the filtered dataset.
-Based on the filtered data, the automated conservation assessment evaluated `r sum(aa$flag_status == "possibly_threatened", na.rm = TRUE)` species as possibly threatened (`r round(sum(aa$flag_status == "possibly_threatened", na.rm=TRUE) / aa %>% filter(!is.na(flag_status)) %>% nrow() * 100, 1)`%, CR = `r aa %>% filter(flag_Category_CriteriaB == "CR") %>% nrow()`, EN = `r aa %>% filter(flag_Category_CriteriaB == "EN") %>% nrow()`, VU = `r aa %>% filter(flag_Category_CriteriaB == "VU") %>% nrow()`).
+Based on the filtered data, the automated conservation assessment evaluated `r sum(aa$flag_status == "possibly_threatened", na.rm = TRUE)` species as possibly threatened (`r round(sum(aa$flag_status == "possibly_threatened", na.rm=TRUE) / aa %>% filter(!is.na(flag_status)) %>% nrow() * 100, 1)`%, CR = `r aa %>% filter(flag_Category_CriteriaB == "CR") %>% nrow()`, EN = `r aa %>% filter(flag_Category_CriteriaB == "EN") %>% nrow()`, VU = `r aa %>% filter(flag_Category_CriteriaB == "VU") %>% nrow()`, see Appendix 3 for assessments of all species).
 The automated assessment based on the filtered dataset agreed with the IUCN assessment for identifying species as possibly threatened (CR, EN, VU) for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status)) %>% mutate(agreement = iucn_status == flag_status) %>% select(agreement) %>% unlist() %>% sum() ` species (`r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status))  %>% mutate(agreement = iucn_status == flag_status) %>% select(agreement) %>% unlist() %>% mean() %>%  round(3) * 100`%; Table 4).
@@
 The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (Question 1).
-The importance of different tests was similar across taxonomic groups, with few exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (Question 2).
+The importance of different tests was similar across taxonomic groups, with exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (Question 2).
 The results for species richness were similar between the raw and filtered data with some improvements by the filters.
@@
 Third, low coordinate precision, for instance based on automated geo-referencing from locality descriptions, may lump records from nearby localities.
-Duplicated records do not represent errors per se, but an unnecessary computational burden, which is often a bottleneck for analyses of datasets with millions of records.
+Duplicated records do not represent errors *per se*, but an unnecessary computational burden, which is often a bottleneck for analyses of datasets with millions of records.
 
 The aim of automated filtering is to identify possibly problematic records that are unsuitable for particular downstream analyses.
-Hence, our tests may identify true errors  (e.g., wrong coordinates such as those with exclusively zeros, equal latitude and longitude and records of terestrial species in the sea) as well as records that are not necessarily erroneous, but have a low precision (e.g., collection year, coordinate precision, centroids of political units) or are redundant for certain analyses (duplicates).
+Hence, our tests may identify true errors (e.g., wrong coordinates such as those with exclusively zeros, equal latitude and longitude and records of terrestrial species in the sea or marine species on land) as well as records that are not necessarily erroneous, but have a low precision (e.g., collection year, coordinate precision, centroids of political units) or are redundant for certain analyses (duplicates).
 The importance of each test and the linked thresholds must be judged based on the specific downstream analyses.
@@
 Some notable exceptions stress the need to adapt each test to the taxonomic study group to balance data quality and data availability.
-The high fraction of records flagged by the "basis of record test" for *Tityus*, Dipsadidae, *Prosthechea* and *Tillandsia*, were mostly caused by a high number of records in these groups based on unknown collection methods (See Appendix 4 for the individual groups), which might be caused by the contribution of specific datasets lacking this information for these groups.
+The high fraction of records flagged by the "basis of record test" for *Tityus*, Dipsadidae, *Prosthechea* and *Tillandsia*, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups.
 The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection effort already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700.
@@
 The general agreement between the species richness maps based on raw and filtered data was encouraging, for the use of this data for large-scale biogeographic research (Fig.
-\ref{fig:speciesrichness}, Supplementary material 1).
+\ref{fig:speciesrichness}, Appendix 2).
 The filter based on political centroids had an important impact on species richness patterns, which is congruent with the results from a previous study in the coffee family [@Maldonado2015].
@@
 These records are at the same time more likely to represent dubious species names, since they might be old synonyms or type specimens of species that have only been collected and described once, which are erroneously increasing species numbers.
+
+We overall consider the effect of the automated filters as positive since they identified the above-mentioned issues and increased the data precision and reduced computational burden (Table 3, Appendix 2).
+However, in some cases filters failed to remove major issues, often due to incomplete metadata.
+For instance, for Diogenidae we found at least two records of an extinct species (*Paguristes mexicanus*) which slipped the "basis of record" test because they were marked as "preserved specimen" rather than "fossil specimen".
+Furthermore, for Entomobryidae we found that for `r dat %>% filter(taxon == "Entomobryidae") %>%  filter(!is.na(species)) %>%  filter(taxonRank == "UNRANKED") %>%  nrow()` records the meta-data on taxonomic rank was "UNRANKED" despite all of them being identified to species level, leading to a high fraction of records removed by the "Identification level" filter.
+Additionally automated filters might be overly strict  or unsuitable for certain taxa.
+For instance, in Entomobryidae, `r dat %>% filter(taxon == "Entomobryidae") %>% filter( basisOfRecord == "MATERIAL_SAMPLE") %>% nrow()` samples were marked as material samples, and therefore removed by our global filter retaining only specimen and observation data, which in this case was overly strict.
+It should be noted, that, while in this study we focused on meta-data and geographic filtering, taxonomic cleaning---the resolution of synonymies and identification of accepted names---is another important part of data curation, but depends on taxon-specific taxonomic backbones and synonymy lists which are not readily available for many groups.
 
@@
 Our results suggest that between one to two thirds of the occurrence records obtained from public databases might be unsuitable for downstream biodiversity analyses.
-While the majority of these records might not be erroneous per se, they might be overly imprecise and thereby increase uncertainty of downstream results or add computational burden on big data analyses.
+While the majority of these records might not be erroneous *per se*, they might be overly imprecise and thereby increase uncertainty of downstream results or add computational burden on big data analyses.
 
@@
 
-# Acknowledgments {-}
+# Acknowledgements {-}
 This study enrolled participants of the workshop "Biodiversity data: from field to yield" led by Alexandre Antonelli and Alexander Zizka through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI.
@@
 
-- Appendix 1 - Analyses scripts
+- Appendix 1 - Table S1, reporting the absolute number of flagged records per taxonomic group and test
 
-- Appendix 2 - Table S1, reporting the absolute number of flagged records per taxonomic group and test
+- Appendix 2 - Taxon specific richness maps and comments.
 
 - Appendix 3 - The full results of the conservation assessment
-
-- Appendix 4 - Taxon specific richness maps and comments.
 
@@
 ```{r tabletaxa, warning=FALSE}
-kable(tab1, booktabs = T, escape = T, format="latex",
+
+gens <- tab1 %>% filter(taxon_rank == "Genus") %>% select(taxon) %>% unlist()
+
+tab1 <- tab1 %>% 
+  mutate(taxon =  cell_spec(taxon, "latex", italic = ifelse(.$taxon %in% gens, TRUE, FALSE)))
+
+kable(tab1, booktabs = T, escape = F, format="latex",
       col.names = linebreak(c("Taxon",
@@
         caption = "The study groups and their taxonomy.
-This study includes three marine and 13 terrestrial taxa, six of them animals, one group of fungi and nine plants, belonging to 16 different orders.")%>%
+This study includes three marine and 15 terrestrial taxa, seven of them animals, one group of fungi and ten plants, belonging to 16 different orders.")%>%
   landscape()%>%
   kable_styling(font_size = 11) %>%
-   row_spec(6, hline_after = TRUE) %>%
    row_spec(7, hline_after = TRUE) %>%
+   row_spec(8, hline_after = TRUE) %>%
    column_spec(1, width = "2.5cm", bold = FALSE, italic = FALSE)%>%
@@
 tab2 <- tab2 %>% 
+  mutate(taxon =  cell_spec(taxon, "latex", italic = ifelse(.$taxon %in% gens, TRUE, FALSE)))%>%
   mutate_all(linebreak)
 
-kable(tab2, booktabs = T, escape = T, format="latex",
+kable(tab2, booktabs = T, escape = F, format="latex",
       col.names = linebreak(c("Taxon",
@@
                               "Records flagged",
-                              "Fraction flagged [%]",
-                              "Basis of record [%]",
-                              "Collection year [%]",
-                              "Coordinate precision [%]",
-                              "Id-level [%]",
-                              "Individual count [%]",
-                              "Capitals [%]",
-                              "Duplicates [%]",
-                              "Political centroids [%]",
-                              "Urban areas [%]",
-                              "Zeros [%]")),
-        caption = "The impact of automated flagging on occurrence records for 16 Neotropical taxa downloaded from www.gbif.org.
+                              "Fraction flagged [\\%]",
+                              "Basis of record [\\%]",
+                              "Collection year [\\%]",
+                              "Coordinate precision [\\%]",
+                              "Id-level [\\%]",
+                              "Individual count [\\%]",
+                              "Biodiversity Institutions [\\%]",
+                              "Capitals [\\%]",
+                              "Duplicates [\\%]",
+                              "Political centroids [\\%]",
+                              "Sea/land area [\\%]",
+                              "Urban areas [\\%]",
+                              "Zeros [\\%]")),
+        caption = "The impact of automated filtering on occurrence records for 18 Neotropical taxa downloaded from www.gbif.org.
 From column four onwards the numbers show the percentage of records flagged by the respective test.
 Only tests that flagged at least 0.1\\% of the records in any group are shown.
-Individual records can be flagged by multiple tests, therefore the sum of percentge from all tests can supersede the total percentage.", format.args = list(big.mark = ","))%>%
+Individual records can be flagged by multiple tests, therefore the sum of percentage from all tests can supersede the total percentage.", format.args = list(big.mark = ","))%>%
   landscape()%>%
   kable_styling(font_size = 9) %>%
-   row_spec(6, hline_after = TRUE) %>%
    row_spec(7, hline_after = TRUE) %>%
+   row_spec(8, hline_after = TRUE) %>%
+  row_spec(18, hline_after = TRUE) %>%
    column_spec(1, width = "2cm", bold = FALSE, italic = FALSE)%>%
-   column_spec(2:6, width = "1.1cm", bold = FALSE, italic = FALSE)%>%
-   column_spec(7:12, width = "1.3cm", bold = FALSE, italic = FALSE)%>%
-  column_spec(13:14, width = "1cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(2:9, width = "1cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(10, width = "1.2cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(11:12, width = "1cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(13:15, width = "1cm", bold = FALSE, italic = FALSE)%>%
    add_header_above(c(" " = 1,
@@
                      "Meta data filters" = 5,
-                     "Automated filters" = 5))
+                     "Gazetteer-based filters" = 7))
 ```
@@
 tab3 <- tab3 %>%
+  mutate(taxon =  cell_spec(taxon, "latex", italic = ifelse(.$taxon %in% gens, TRUE, FALSE)))%>%
   mutate_all(linebreak)
 
-kable(tab3, booktabs = T, escape = T, format="latex",
+kable(tab3, booktabs = T, escape = F, format="latex",
       col.names = linebreak(c("Taxon",
                               "n taxa",
-                              "Evaluated [%]",
-                              "Threatened [%]",
+                              "Evaluated [\\%]",
+                              "Threatened [\\%]",
                               "n taxa",
-                              "Threatened [%]",
-                              "Match with IUCN [%]",
+                              "Threatened [\\%]",
+                              "Match with IUCN [\\%]",
                               "n taxa",
-                              "Threatened [%]",
-                              "Match with IUCN [%]",
-                              "EOO change compared to raw [%]",
-                              "AOO change compared to raw [%]")),
-        caption = "Conservation assessment for 10 Neotropical taxa of plants and animals based on three datasets.
+                              "Threatened [\\%]",
+                              "Match with IUCN [\\%]",
+                              "EOO change compared to raw [\\%]",
+                              "AOO change compared to raw [\\%]")),
+        caption = "Conservation assessment for 11 Neotropical taxa of plants and animals based on three datasets.
 IUCN: global red list assessment obtained from www.iucn.org; GBIF Raw: Preliminary conservation assessment based on IUCN Criterion B using ConR and the raw dataset from GBIF; GBIF filtered: Preliminary conservation assessment based on IUCN Criterion B using ConR and the filtered dataset.
@@
   kable_styling(font_size = 9) %>%
+   row_spec(3, hline_after = TRUE) %>%
    column_spec(2:10, width = "1.2cm", bold = FALSE, italic = FALSE)%>%
@@
 <!-- The study species-->
-```{r species, out.width="0.9\\linewidth", fig.cap=c('Example of study taxa.
-\\textbf{A)} Pilosocereus pusillibaccatus, picture by Pamela Lavor; \\textbf{B)} Conchocarpus macrocarpus, Juliana El-Ottra; \\textbf{C)} Tillandsia recurva, EC Tomaz; \\textbf{D} \\textit{Oxyrhopus guibei} (Dipsadidae), Filipe C Serrano;\\textbf{E)} Neanuridae (\\textit{Aethiopella ricardoi}), Raiane Vital da Paz; \\textbf{F)} \\textit{Tocoyena formosa}, Fernanda GL Moreira; \\textbf{G)} \\textit{Harengula jaguana}, Thais Ferreira-Araujo; \\textbf{H)} \\textit{Gaylussacia decipiens}, Luiz Menini Neto; \\textbf{I)} \textit{Ooocephalus foliosus}, Arthur Soares; \\textbf{J)} \\textit{Tityus carvalhoi}, Renata C Santos-Costa; \\textbf{K)} \\textit{Prostheca vespa}, Tiago veira'), echo=FALSE}
+```{r species, out.width="0.9\\linewidth", fig.cap=c('Examples of study taxa.
+\\textbf{A)} Pilosocereus pusillibaccatus, picture by Pamela Lavor; \\textbf{B)} Conchocarpus macrocarpus, Juliana El-Ottra; \\textbf{C)} Tillandsia recurva, EC Tomaz; \\textbf{D)} \\textit{Oxyrhopus guibei} (Dipsadidae), Filipe C Serrano; \\textbf{E)} Neanuridae (\\textit{Aethiopella ricardoi}), Raiane Vital da Paz; \\textbf{F)} \\textit{Tocoyena formosa}, Fernanda GL Moreira; \\textbf{G)} \\textit{Harengula jaguana}, Thais Ferreira-Araujo; \\textbf{H)} \\textit{Gaylussacia decipiens}, Luiz Menini Neto; \\textbf{I)} \\textit{Ooocephalus foliosus}, Arthur Soares; \\textbf{J)} \\textit{Tityus carvalhoi}, Renata C Santos-Costa; \\textbf{K)} \\textit{Prostheca vespa}, Tiago Veira'), echo=FALSE}
 knitr::include_graphics(path = "./figures_tables/Figure_study_groups.jpg")
@@
 <!-- A map with number of species and where they have been removed -->
-```{r total, out.width="\\linewidth", fig.cap=c('The difference in records available between the raw and filtered dataset of 16 Neotropical taxa including plants and animals, plotted in a 100x100 km grid across the Neotropics.'), echo=FALSE}
+```{r total, out.width="\\linewidth", fig.cap=c('The difference in records available between the raw and filtered dataset of 18 Neotropical taxa including animals, funghi, and plants, plotted in a 100x100 km grid across the Neotropics.'), echo=FALSE}
 knitr::include_graphics(path = "./output/figure_number_of_records.jpg")
@@
 The concentration of records flagged on the basis of record test and individual count test in Central America and southern North America.
-Only filters that flagged at least one record are shown.'), echo=FALSE}
+Only filters that flagged at least 0.1\\% of records in any taxon are shown.'), echo=FALSE}
 knitr::include_graphics(path = "./output/figure_number_of_records_split.jpg")
@@
 
-```{r speciesrichness, out.width="\\linewidth", fig.cap=c('The difference in species richness derived from the raw and filtered datasets (raw - filtered), for 16 taxa of animals, funghi, and plants.'), echo=FALSE}
-knitr::include_graphics(path = "./output/figure_number_species_richness_difference.jpg")
+```{r speciesrichness, out.width="\\linewidth", fig.cap=c('Illustrative examples for the difference in species richness between the raw and filtered dataset (raw - filtered) from four of the study taxa.
+Photo credits for C) by Tiago Andrade Borges Santos, otherwise as in Figure 1.'), echo=FALSE}
+knitr::include_graphics(path = "./figures_tables/figure_number_species_richness_difference.jpg")
 ```

### 2020-03-02 manuscript_peerj.tex
co-author version

@@
 
-\author[5]{Andressa Cabral}
+\author[5]{Tiago Andrade Borges Santos}
 
-\author[4]{Arthur de Souza Soares}
+\author[6]{Andressa Cabral}
+
+\author[7]{Filipe Cabreirinha Serrano}
 
 \author[4]{Eduardo Calisto Tomaz}
-
-\author[4]{Fernanda Gondim Lambert Moreira}
-
-\author[6]{Filipe Cabreirinha Serrano}
-
-\author[4]{Jéssica Fernanda Ramos Coelho}
-
-\author[7]{Mabel Rocio Baez-Lizarazo}
-
-\author[4]{Mariana Ramos Fantinati}
-
-\author[5]{Matheus Colli-Silva}
-
-\author[8]{Moabe Ferreira Fernandes}
-
-\author[4]{Nathália Michelly da Cunha Santos}
 
@@
 
+\author[6]{Matheus Colli-Silva}
+
+\author[4]{Nathália Michelly da Cunha Santos}
+
+\author[4]{Arthur de Souza Soares}
+
 \author[4]{Thais Ferreira-Araújo}
 
-\author[9]{Tiago Andrade Borges Santos}
-
-\author[9]{Tiago Luiz Vieira}
+\author[9]{Moabe Ferreira Fernandes}
 
@@
 
+\author[4]{Fernanda Gondim Lambert Moreira}
+
+\author[4]{Jéssica Fernanda Ramos Coelho}
+
+\author[4]{Mariana Ramos Fantinati}
+
+\author[8]{Mabel Rocio Baez-Lizarazo}
+
+\author[6]{Tiago Luiz Vieira}
+
 \author[10,11,12]{Alexandre Antonelli}
 
-\affil[1]{sDiv, German Center for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Leipzig, Germany}
+\affil[1]{sDiv, German Center for Integrative Biodiversity Research Halle-Jena-Leipzig (iDiv), Leipzig, Germany}
 \affil[2]{Naturalis Biodiversity Center, Leiden, The Netherlands}
@@
 \affil[4]{Departamento de Botânica e Zoologia, Universidade Federal do Rio Grande do Norte, Natal, Brazil}
-\affil[5]{Departamento de Botânica, Universidade de São Paulo, São Paulo, Brazil}
-\affil[6]{Departamento de Ecologia, Universidade de São Paulo, São Paulo, Brazil}
-\affil[7]{Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil}
-\affil[8]{Departamento de Ciências Biológicas, Universidade Estadual de Feira de Santana, Feira de Santana, Brazil}
-\affil[9]{Universidade Estadual de Feira de Santana, Bahia, Brazil}
+\affil[5]{Universidade Estadual de Feira de Santana, Feira de Santana, Brazil}
+\affil[6]{Departamento de Botânica, Universidade de São Paulo, São Paulo, Brazil}
+\affil[7]{Departamento de Ecologia, Universidade de São Paulo, São Paulo, Brazil}
+\affil[8]{Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil}
+\affil[9]{Departamento de Ciências Biológicas, Universidade Estadual de Feira de Santana, Feira de Santana, Brazil}
 \affil[10]{Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden}
@@
 However, it is unclear how much data may be lost by filtering, whether the same tests should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses.
-Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 16 Neotropical taxa including plants and animals, terrestrial and marine, downloaded from GBIF.
-We find that 29-67\% of the records are potentially erroneous, with large variation across taxonomic groups.
-Tests for duplicated information, collection year, basis of record as well as urban areas and political centroids have the greatest effect.
+Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 18 Neotropical taxa including animals, fungi, and plants, terrestrial and marine, downloaded from GBIF.
+We find that 29-90\% of the records are potentially erroneous, with large variation across taxonomic groups.
+Tests for duplicated information, collection year, basis of record as well as urban areas and coordinates for terrestrial taxa in the sea or marine taxa on land have the greatest effect.
 While many flagged records might not be \emph{de facto} erroneous, they could be overly imprecise and increase uncertainty in downstream analyses.
@@
 2019; Jin and Yang 2020) or on additional data, such as environmental variables (Robertson, Visser, and Hui 2016).
-Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biodiveristy research (Zizka, Silvestro, et al.
+Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the suitability of publicly available occurrence records for biodiversity research (Zizka, Silvestro, et al.
 2019).
@@
 These methods approximate species' range size, namely the Extent of Occurrence (EOO, which is the area of a convex hull polygon comprising all records of a species), the Area of Occupancy (AOO, which is the sum of the area actually occupied by a species, calculated based on a small-scale regular grid), and the number of locations for a preliminary conservation assessment following IUCN Criterion B (``Geographic range'').
-These methods have been used topropose preliminary global (Stévart et al.
+These methods have been used to propose preliminary global (Stévart et al.
 2019; Zizka, Azevedo, et al.
@@
 2018) Red List assessments.
-However, all metrics but especially EOO, are sensitive to individual records with erroneous coordinates \textbf{(REFERENCE)}.
+However, all metrics but especially EOO, are sensitive to individual records with erroneous coordinates.
 Automated conservation assessments may therefore be biased, particularly if the number of records are few, as it is the case for many tropical species.
@@
 
-This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering graduate students, taxonomists and biogeographers working with different taxonomic groups of animals, funghi, and plants across the Neotropics.
-Each participant analysed geographic occurrence data from their taxonomic group of interest and provided an informed assessment of the results for their group.
-Hence, we include groups based on the expertise of the participants wrather than following some arbitrary choice of taxa and taxonomic ranks.
+This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering students and researchers working with different taxonomic groups of animals, fungi, and plants across the Neotropics.
+Each participant analysed geographic occurrence data from their taxonomic group of interest and commented on the results for their group.
+Hence, we include groups based on the expertise of the participants rather than following some arbitrary choice of taxa and taxonomic ranks.
 We acknowledge a varying degree in documented expertise and number of years working on each group.
-We obtained public occurrence records for 16 taxa, including one plant family, eight plant genera, one genus of fungi, two families and one genus of terrestrial arthropods, one family of snakes, one family of skates, and one genus of bony fish (Table 1).
+We obtained public occurrence records for 18 taxa, including one plant family, nine plant genera, one genus of fungi, three families and one genus of terrestrial arthropods, one family of snakes, one family of skates, and one genus of bony fish (Table 1).
 
@@
 
-We downloaded occurrence information for all study groups from www.gbif.org using the \texttt{rgbif} v1.4.0 package (Chamberlain 2017) in R (GBIF.org, 2019a-p).
+We downloaded occurrence information for all study groups from www.gbif.org using the \texttt{rgbif} v1.4.0 package (Chamberlain 2017) in R (GBIF.org, 2019a-p,2020a,b).
 We downloaded only records with geographic coordinates and limited the study area to latitudes smaller than 33\(^\circ\) N and longitudes smaller than 35\(^\circ\) W and larger than 120\(^\circ\) W reflecting the Neotropics, our main area of expertise (Morrone 2014).
-The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, \emph{Gaylussacia}, Iridaceae, Neanuridae, and \emph{Tillandsia} for which we only obtained the Neotropical occurrences.
+The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, Entomobryidae, \emph{Gaylussacia}, Iridaceae, Neanuridae, and \emph{Tillandsia} for which we only obtained the Neotropical occurrences.
 
@@
 (2019) and first filtered the raw data downloaded from GBIF (``raw'', hereafter) using meta-data for those records for which they were available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99.
-Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e.~test for duplicated records).
+Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e., test for duplicated records).
 In a second step, we used the \texttt{clean\_coordinates} function of the \texttt{CoordinateCleaner\ v2.0-11} package (Zizka, Silvestro, et al.
 2019) with default options to flag errors that are common to biological data sets (``filtered'', hereafter).
-These including: coordinates in the sea for terrestrial taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions.
+These include: coordinates in the sea for terrestrial taxa and on land for marine taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions.
 See Table 2 for a summary of all flagging tests we used.
@@
 
-Each specialist provided an informed assessment on the performance of the raw and cleaned datasets, respectively concerning: 1) the number of flagged records; 2) the importance of specific tests; 3) the accuracy of the overall species richness maps; 4) the accuracy of the preliminary conservation assessments.
+Each author provided an informed comment on the performance of the raw and cleaned datasets, respectively concerning the number of flagged records and the accuracy of the overall species richness maps and we compared the agreement between automated conservation assessments based on raw and filtered occurrences with the global IUCN Red List (www.iucn.org).
 
 We did all analyses in the R computing environment (R Core Team 2019), using standard libraries for data handling and visualization (Wickham 2018; Garnier 2018; Ooms 2014, 2019; Hijmans 2019).
-All scripts are available from \url{https://github.com/idiv-biodiversity/effects_of_automated_cleaning}, or from the Appendix 1 in the electronic supplement of this study.
+All scripts are available from a zenodo repository (\textbf{DOI}).
 
@@
 
-We retrieved a total of 218,899 species occurrence records, with a median of 2,844.5 records per study group and 10 records per species (Table 3, Appendix 2).
+We retrieved a total of 218,899 species occurrence records, with a median of 2,844.5 records per study group and 10 records per species (Table 3, Appendix 1).
 We obtained most records for Dipsadidae (64,249) and fewest for \emph{Thozetella} (51).
@@
 
-Our automated tests flagged a total of 91,434 records (Fig.
-\ref{fig:total}), with a median of 42.7\% per group.
+Our automated tests flagged a total of 97,018 records (Fig.
+\ref{fig:total}), with a median of 45\% per group.
 Overall, the most important test was for duplicated records (on average 35.5\% per taxonomic group).
 The filtering steps based on record meta-data that flagged the largest number of records were the basis of records (5.9\%) and the collection year (3.4\%).
-The most important automated tests were for urban area (8.6\%), the capital coordinates (1\%) and the centroids of political units (0.5\%, see Table 3 and Supplementary material 1 for further details and the absolute numbers).
-Only few records were flagged by the zero coordinates and biodiversity institution tests (Fig.
-\ref{fig:split}), and no records were flagged by the sea test.
+The most important automated tests were for urban area (8.6\%) and the occurrence from records of terrestrial taxa in the sea and marine taxa on land (4.3\%, see Table 3 and Appendix 1 in the electronic supplement for further details and the absolute numbers).
+Only few records were flagged by the coordinate precision, zero coordinates and biodiversity institution tests (Fig.
+\ref{fig:split}).
 
-\emph{Prosthechea}, \emph{Harengula}, and \emph{Thozetella} had the highest fraction of flagged records (Table 3).
+\emph{Entomobryidae}, \emph{Diogenidae}, and \emph{Neanuridae} had the highest fraction of flagged records (Table 3).
 In general, the different filters we tested were of similar importance for different study groups.
-There were few outstanding exceptions, including the particularly high proportions of records flagged on the ``basis of record test'' for \emph{Tityus} (7.0\%), Dipsadidae (5.6\%), \emph{Prosthechea} (5.0\%) and \emph{Tillandsia} (4.9\%), by the collection year for Dipsadidae (11.3\%), by the taxonomic identification level for \emph{Tityus} (1.6\%), by the capital coordinates for \emph{Oocephalus} (6.1\%) and \emph{Gaylussacia} (3.2\%), and by the urban areas test for \emph{Oocephalus} (13.3\%) and Iridaceae (12.3\%).
+There were few outstanding exceptions, including the particularly high proportions of records flagged on the ``basis of record test'' for \emph{Tityus} (7.0\%), Dipsadidae (5.6\%), \emph{Prosthechea} (5.0\%) and \emph{Tillandsia} (4.9\%), by the collection year for Dipsadidae (11.3\%), by the taxonomic identification level for \emph{Tityus} (1.6\%), by the capital coordinates for \emph{Oocephalus} (6.1\%) and \emph{Gaylussacia} (3.2\%), by the seas/land test for Diogenidae and \emph{Thozetella}, and by the urban areas test for \emph{Oocephalus} (13.3\%) and Iridaceae (12.3\%).
+Furthermore, Entomobryidae differed considerably from all other study taxa with exceptionally high numbers of records flagged by the ``basis of record,''level of identification' and urban areas" tests.
 
@@
 \ref{fig:split}).
-See Supplementary material 2 for species richness maps using the raw and cleaned data for all study groups.
+See Appendix 2 for species richness maps using the raw and cleaned data for all study groups.
 
-We found IUCN assessments for 575 species that were also included in our distribution data from 10 of our study groups (Table 4, Appendix 3).
-The fraction of species evaluated varied among the study group, with a maximum of 100\% for \emph{Harengula} and \emph{Lepismium} and a minimum of 2.3\% for Iridaceae (note that the number of species varied considerably among groups).
-The median percentage of species per study group with an IUCN assessment was 32.45\%.
-A total of 101 species were listed as Threatened by the IUCN global Red List (CR = 19, EN = 40, VU = 42) and 474 as Not Threatened.
+We found IUCN assessments for 579 species that were also included in our distribution data from 11 of our study groups (Table 4, Appendix 3).
+The fraction of species evaluated varied among the study group, with a maximum of 100\% for \emph{Harengula} and \emph{Lepismium} and a minimum of 2.3\% for Iridaceae (note that the number of total species varied considerably among groups).
+The median percentage of species per study group with an IUCN assessment was 15\%.
+A total of 102 species were listed as Threatened by the IUCN global Red List (CR = 19, EN = 40, VU = 43) and 477 as Not Threatened.
 
-We obtained automated conservation assessments for 2,092 species in the filtered dataset.
-Based on the filtered data, the automated conservation assessment evaluated 1,299 species as possibly threatened (62.1\%, CR = 464, EN = 543, VU = 292).
-The automated assessment based on the filtered dataset agreed with the IUCN assessment for identifying species as possibly threatened (CR, EN, VU) for 356 species (64\%; Table 4).
-Filtering reduced the EOO by -16.1\% and the AOO by -9.3\% on median per group.
-For the raw dataset the agreement with IUCN was higher at 378 species (65.6\%).
+We obtained automated conservation assessments for 2,181 species in the filtered dataset.
+Based on the filtered data, the automated conservation assessment evaluated 1,382 species as possibly threatened (63.4\%, CR = 495, EN = 577, VU = 310, see Appendix 3 for assessments of all species).
+The automated assessment based on the filtered dataset agreed with the IUCN assessment for identifying species as possibly threatened (CR, EN, VU) for 358 species (64\%; Table 4).
+Filtering reduced the EOO by -18.4\% and the AOO by -9.9\% on median per group.
+For the raw dataset the agreement with IUCN was higher at 381 species (65.7\%).
 
@@
 
-Automated flagging based on meta-data and automatic tests flagged on average 42.7\% of the records per taxonomic group.
+Automated flagging based on meta-data and automatic tests flagged on average 45\% of the records per taxonomic group.
 The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (Question 1).
-The importance of different tests was similar across taxonomic groups, with few exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (Question 2).
+The importance of different tests was similar across taxonomic groups, with exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (Question 2).
 The results for species richness were similar between the raw and filtered data with some improvements by the filters.
 We found little impact of filtering on the accuracy of the automated conservation assessments (Question 3).
+
+\hypertarget{the-relevance-of-individual-filters}{%
+\subsection*{The relevance of individual filters}\label{the-relevance-of-individual-filters}}
+\addcontentsline{toc}{subsection}{The relevance of individual filters}
 
@@
 Third, low coordinate precision, for instance based on automated geo-referencing from locality descriptions, may lump records from nearby localities.
-Duplicated records do not represent errors per se, but an unnecessary computational burden, which is often a bottleneck for analyses of datasets with millions of records.
+Duplicated records do not represent errors \emph{per se}, but an unnecessary computational burden, which is often a bottleneck for analyses of datasets with millions of records.
 
 The aim of automated filtering is to identify possibly problematic records that are unsuitable for particular downstream analyses.
-Hence, our tests may identify true errors (e.g., wrong coordinates such as those with exclusively zeros, equal latitude and longitude and records of terestrial species in the sea) as well as records that are not necessarily erroneous, but have a low precision (e.g., collection year, coordinate precision, centroids of political units) or are redundant for certain analyses (duplicates).
+Hence, our tests may identify true errors (e.g., wrong coordinates such as those with exclusively zeros, equal latitude and longitude and records of terrestrial species in the sea or marine species on land) as well as records that are not necessarily erroneous, but have a low precision (e.g., collection year, coordinate precision, centroids of political units) or are redundant for certain analyses (duplicates).
 The importance of each test and the linked thresholds must be judged based on the specific downstream analyses.
@@
 
+\hypertarget{similarities-and-differences-among-taxa}{%
+\subsection*{Similarities and differences among taxa}\label{similarities-and-differences-among-taxa}}
+\addcontentsline{toc}{subsection}{Similarities and differences among taxa}
+
 The number of records flagged by individual tests was similar across study groups, suggesting that similar error types might be relevant for collections of plants and animals.
@@
 Some notable exceptions stress the need to adapt each test to the taxonomic study group to balance data quality and data availability.
-The high fraction of records flagged by the ``basis of record test'' for \emph{Tityus}, Dipsadidae, \emph{Prosthechea} and \emph{Tillandsia}, were mostly caused by a high number of records in these groups based on unknown collection methods (See Appendix 4 for the individual groups), which might be caused by the contribution of specific datasets lacking this information for these groups.
+The high fraction of records flagged by the ``basis of record test'' for \emph{Tityus}, Dipsadidae, \emph{Prosthechea} and \emph{Tillandsia}, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups.
 The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection effort already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700.
@@
 The general agreement between the species richness maps based on raw and filtered data was encouraging, for the use of this data for large-scale biogeographic research (Fig.
-\ref{fig:speciesrichness}, Supplementary material 1).
+\ref{fig:speciesrichness}, Appendix 2).
 The filter based on political centroids had an important impact on species richness patterns, which is congruent with the results from a previous study in the coffee family (Maldonado et al.
@@
 These records are at the same time more likely to represent dubious species names, since they might be old synonyms or type specimens of species that have only been collected and described once, which are erroneously increasing species numbers.
+
+We overall consider the effect of the automated filters as positive since they identified the above-mentioned issues and increased the data precision and reduced computational burden (Table 3, Appendix 2).
+However, in some cases filters failed to remove major issues, often due to incomplete metadata.
+For instance, for Diogenidae we found at least two records of an extinct species (\emph{Paguristes mexicanus}) which slipped the ``basis of record'' test because they were marked as ``preserved specimen'' rather than ``fossil specimen''.
+Furthermore, for Entomobryidae we found that for 1,996 records the meta-data on taxonomic rank was ``UNRANKED'' despite all of them being identified to species level, leading to a high fraction of records removed by the ``Identification level'' filter.
+Additionally automated filters might be overly strict or unsuitable for certain taxa.
+For instance, in Entomobryidae, 2,004 samples were marked as material samples, and therefore removed by our global filter retaining only specimen and observation data, which in this case was overly strict.
+It should be noted, that, while in this study we focused on meta-data and geographic filtering, taxonomic cleaning---the resolution of synonymies and identification of accepted names---is another important part of data curation, but depends on taxon-specific taxonomic backbones and synonymy lists which are not readily available for many groups.
+
+\hypertarget{the-impact-of-filtering-on-the-accuracy-of-automated-conservation-assessments}{%
+\subsection*{The impact of filtering on the accuracy of automated conservation assessments}\label{the-impact-of-filtering-on-the-accuracy-of-automated-conservation-assessments}}
+\addcontentsline{toc}{subsection}{The impact of filtering on the accuracy of automated conservation assessments}
 
@@
 Our results suggest that between one to two thirds of the occurrence records obtained from public databases might be unsuitable for downstream biodiversity analyses.
-While the majority of these records might not be erroneous per se, they might be overly imprecise and thereby increase uncertainty of downstream results or add computational burden on big data analyses.
+While the majority of these records might not be erroneous \emph{per se}, they might be overly imprecise and thereby increase uncertainty of downstream results or add computational burden on big data analyses.
 
@@
 
-\hypertarget{acknowledgments}{%
-\section*{Acknowledgments}\label{acknowledgments}}
-\addcontentsline{toc}{section}{Acknowledgments}
+\hypertarget{acknowledgements}{%
+\section*{Acknowledgements}\label{acknowledgements}}
+\addcontentsline{toc}{section}{Acknowledgements}
 
@@
 \item
-  Appendix 1 - Analyses scripts
+  Appendix 1 - Table S1, reporting the absolute number of flagged records per taxonomic group and test
 \item
-  Appendix 2 - Table S1, reporting the absolute number of flagged records per taxonomic group and test
+  Appendix 2 - Taxon specific richness maps and comments.
 \item
   Appendix 3 - The full results of the conservation assessment
-\item
-  Appendix 4 - Taxon specific richness maps and comments.
 \end{itemize}
@@
 \caption{\label{tab:tabletaxa}The study groups and their taxonomy.
-This study includes three marine and 13 terrestrial taxa, six of them animals, one group of fungi and nine plants, belonging to 16 different orders.}
+This study includes three marine and 15 terrestrial taxa, seven of them animals, one group of fungi and ten plants, belonging to 16 different orders.}
 \centering
@@
 Diogenidae & Family & Marine & Hermit crabs & Arthropoda & Decapoda & Diogenidae\\
+Entomobryidae & Family & Terrestrial & Springtails & Arthropoda & Poduromorpha & Entomobryidae\\
 Neanuridae & Family & Terrestrial & Springtails & Arthropoda & Poduromorpha & Neanuridae\\
-Tityus & Genus & Terrestrial & Scorpions & Arthropoda & Scorpiones & Buthidae\\
+\em{Tityus} & Genus & Terrestrial & Scorpions & Arthropoda & Scorpiones & Buthidae\\
 Arhynchobatidae & Family & Marine & Skates & Chordata & Rajiformes & Arhynchobatidae\\
+\addlinespace
 Dipsadidae & Family & Terrestrial & Snakes & Chordata & Squamata & Dipsadidae\\
+\em{Harengula} & Genus & Marine & Herrings & Chordata & Clupeiformes & Clupeidae\\
+\hline
+\em{Thozetella} & Genus & Terrestrial & Sac fungi & Ascomycota & Chaetosphaeriales & Chaetosphaeriaceae\\
+\hline
+\em{Conchocarpus} & Genus & Terrestrial & NA & Angiosperms & Sapindales & Rutaceae\\
+\em{Gaylussacia} & Genus & Terrestrial & Huckleberries & Angiosperms & Ericales & Ericaceae\\
 \addlinespace
-Harengula & Genus & Marine & Herrings & Chordata & Clupeiformes & Clupeidae\\
-\hline
-Thozetella & Genus & Terrestrial & Sac fungi & Ascomycota & Chaetosphaeriales & Chaetosphaeriaceae\\
-\hline
-Conchocarpus & Genus & Terrestrial & NA & Angiosperms & Sapindales & Rutaceae\\
-Gaylussacia & Genus & Terrestrial & Huckleberries & Angiosperms & Ericales & Ericaceae\\
+\em{Harpalyce} & Genus & Terrestrial & NA & Angiosperms & Fabales & Fabaceae\\
 Iridaceae & Family & Terrestrial & NA & Angiosperms & Asparagales & Iridaceae\\
+\em{Lepismium} & Genus & Terrestrial & Cacti & Angiosperms & Caryophyllales & Cactaceae\\
+\em{Oocephalus} & Genus & Terrestrial & NA & Angiosperms & Lamiales & Lamiaceae\\
+\em{Pilosocereus} & Genus & Terrestrial & NA & Angiosperms & Caryophyllales & Cactaceae\\
 \addlinespace
-Lepismium & Genus & Terrestrial & Cacti & Angiosperms & Caryophyllales & Cactaceae\\
-Oocephalus & Genus & Terrestrial & NA & Angiosperms & Lamiales & Lamiaceae\\
-Pilosocereus & Genus & Terrestrial & NA & Angiosperms & Caryophyllales & Cactaceae\\
-Prosthechea & Genus & Terrestrial & Orchids & Angiosperms & Asparagales & Orchidaceae\\
-Tillandsia & Genus & Terrestrial & Bromeliads & Angiosperms & Poales & Bromeliaceae\\
-\addlinespace
-Tocoyena & Genus & Terrestrial & NA & Angiosperms & Gentianales & Rubiaceae\\
+\em{Prosthechea} & Genus & Terrestrial & Orchids & Angiosperms & Asparagales & Orchidaceae\\
+\em{Tillandsia} & Genus & Terrestrial & Bromeliads & Angiosperms & Poales & Bromeliaceae\\
+\em{Tocoyena} & Genus & Terrestrial & NA & Angiosperms & Gentianales & Rubiaceae\\
 \bottomrule
@@
 
-\caption{\label{tab:tablecoords}The impact of automated flagging on occurrence records for 16 Neotropical taxa downloaded from www.gbif.org.
+\caption{\label{tab:tablecoords}The impact of automated filtering on occurrence records for 18 Neotropical taxa downloaded from www.gbif.org.
 From column four onwards the numbers show the percentage of records flagged by the respective test.
 Only tests that flagged at least 0.1\% of the records in any group are shown.
-Individual records can be flagged by multiple tests, therefore the sum of percentge from all tests can supersede the total percentage.}
+Individual records can be flagged by multiple tests, therefore the sum of percentage from all tests can supersede the total percentage.}
 \centering
 \fontsize{9}{11}\selectfont
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{1.1cm}>{\raggedleft\arraybackslash}p{1.1cm}>{\raggedleft\arraybackslash}p{1.1cm}>{\raggedleft\arraybackslash}p{1.1cm}>{\raggedleft\arraybackslash}p{1.1cm}>{\raggedleft\arraybackslash}p{1.3cm}>{\raggedleft\arraybackslash}p{1.3cm}>{\raggedleft\arraybackslash}p{1.3cm}>{\raggedleft\arraybackslash}p{1.3cm}>{\raggedleft\arraybackslash}p{1.3cm}>{\raggedleft\arraybackslash}p{1.3cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}}
+\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}r}
 \toprule
-\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Summary} & \multicolumn{5}{c}{Meta data filters} & \multicolumn{5}{c}{Automated filters} \\
-\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-9} \cmidrule(l{3pt}r{3pt}){10-14}
-Taxon & Total records & Records flagged & Fraction flagged [\%] & Basis of record [\%] & Collection year [\%] & Coordinate precision [\%] & Id-level [\%] & Individual count [\%] & Capitals [\%] & Duplicates [\%] & Political centroids [\%] & Urban areas [\%] & Zeros [\%]\\
+\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Summary} & \multicolumn{5}{c}{Meta data filters} & \multicolumn{7}{c}{Gazetteer-based filters} \\
+\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-9} \cmidrule(l{3pt}r{3pt}){10-16}
+Taxon & Total records & Records flagged & Fraction flagged [\%] & Basis of record [\%] & Collection year [\%] & Coordinate precision [\%] & Id-level [\%] & Individual count [\%] & Biodiversity Institutions [\%] & Capitals [\%] & Duplicates [\%] & Political centroids [\%] & Sea/land area [\%] & Urban areas [\%] & Zeros [\%]\\
 \midrule
-Diogenidae & 13,840 & 5,294 & 38.3 & 1.7 & 2.5 & 0.0 & 0.0 & 0.0 & 0.7 & 33.8 & 0.2 & 1.3 & 0.0\\
-Neanuridae & 689 & 461 & 66.9 & 2.9 & 1.3 & 0.0 & 0.0 & 0.0 & 0.0 & 62.4 & 0.0 & 2.0 & 0.0\\
-Tityus & 1,018 & 562 & 55.2 & 7.0 & 0.4 & 1.8 & 1.6 & 0.0 & 1.2 & 43.5 & 0.1 & 6.9 & 0.0\\
-Arhynchobatidae & 14,633 & 5,469 & 37.4 & 1.7 & 1.3 & 0.0 & 0.9 & 0.0 & 0.0 & 35.4 & 0.0 & 1.9 & 0.0\\
-Dipsadidae & 64,249 & 37,054 & 57.7 & 5.6 & 11.3 & 0.8 & 0.0 & 0.1 & 1.8 & 46.3 & 0.4 & 8.5 & 0.0\\
+Diogenidae & 13,840 & 9,508 & 68.7 & 1.7 & 2.5 & 0.0 & 0.0 & 0.0 & 0.0 & 0.7 & 33.8 & 0.2 & 44.3 & 1.3 & 0.0\\
+Entomobryidae & 2,767 & 2,498 & 90.3 & 72.9 & 2.0 & 0.0 & 72.1 & 0.0 & 0.1 & 0.1 & 85.5 & 0.0 & 0.0 & 70.1 & 0.0\\
+Neanuridae & 689 & 461 & 66.9 & 2.9 & 1.3 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 62.4 & 0.0 & 0.0 & 2.0 & 0.0\\
+\em{Tityus} & 1,018 & 562 & 55.2 & 7.0 & 0.4 & 1.8 & 1.6 & 0.0 & 0.5 & 1.2 & 43.5 & 0.1 & 0.0 & 6.9 & 0.0\\
+Arhynchobatidae & 14,633 & 5,635 & 38.5 & 1.7 & 1.3 & 0.0 & 0.9 & 0.0 & 0.0 & 0.0 & 35.4 & 0.0 & 3.8 & 1.9 & 0.0\\
 \addlinespace
-Harengula & 36,697 & 10,230 & 27.9 & 1.0 & 0.4 & 0.0 & 0.3 & 0.0 & 0.2 & 27.0 & 0.1 & 0.2 & 0.0\\
+Dipsadidae & 64,249 & 37,058 & 57.7 & 5.6 & 11.3 & 0.8 & 0.0 & 0.1 & 0.3 & 1.8 & 46.3 & 0.4 & 0.0 & 8.5 & 0.0\\
+\em{Harengula} & 36,697 & 11,384 & 31.0 & 1.0 & 0.4 & 0.0 & 0.3 & 0.0 & 0.0 & 0.2 & 27.0 & 0.1 & 5.5 & 0.2 & 0.0\\
 \hline
-Thozetella & 51 & 15 & 29.4 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 27.5 & 0.0 & 2.0 & 0.0\\
+\em{Thozetella} & 51 & 18 & 35.3 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 27.5 & 0.0 & 23.5 & 2.0 & 0.0\\
 \hline
-Conchocarpus & 1,551 & 665 & 42.9 & 0.5 & 1.9 & 0.1 & 0.0 & 0.0 & 0.0 & 39.6 & 0.9 & 2.3 & 0.0\\
-Gaylussacia & 3,998 & 1,884 & 47.1 & 0.7 & 4.4 & 0.6 & 0.0 & 0.0 & 3.2 & 41.8 & 1.1 & 5.2 & 0.0\\
-Iridaceae & 23,127 & 7,760 & 33.6 & 0.9 & 4.7 & 0.1 & 0.0 & 1.3 & 1.0 & 17.1 & 0.4 & 12.3 & 0.0\\
+\em{Conchocarpus} & 1,551 & 670 & 43.2 & 0.5 & 1.9 & 0.1 & 0.0 & 0.0 & 0.1 & 0.0 & 39.6 & 0.9 & 0.4 & 2.3 & 0.0\\
+\em{Gaylussacia} & 3,998 & 1,887 & 47.2 & 0.7 & 4.4 & 0.6 & 0.0 & 0.0 & 0.1 & 3.2 & 41.8 & 1.1 & 0.1 & 5.2 & 0.0\\
 \addlinespace
-Lepismium & 825 & 245 & 29.7 & 0.0 & 2.1 & 0.0 & 0.0 & 0.0 & 0.1 & 21.9 & 0.1 & 7.8 & 0.0\\
-Oocephalus & 883 & 435 & 49.3 & 0.0 & 0.7 & 0.3 & 0.0 & 0.1 & 6.1 & 41.9 & 0.8 & 13.3 & 0.0\\
-Pilosocereus & 1,940 & 501 & 25.8 & 1.8 & 7.0 & 0.0 & 0.0 & 0.9 & 0.5 & 16.8 & 0.5 & 2.1 & 0.0\\
-Prosthechea & 6,617 & 2,088 & 31.6 & 5.0 & 8.3 & 0.1 & 0.0 & 0.2 & 0.4 & 19.6 & 1.7 & 0.9 & 0.1\\
-Tillandsia & 42,222 & 14,893 & 35.3 & 4.9 & 5.1 & 0.1 & 0.0 & 1.0 & 0.7 & 19.8 & 0.7 & 9.2 & 0.0\\
+\em{Harpalyce} & 870 & 288 & 33.1 & 0.5 & 5.5 & 0.7 & 0.0 & 0.9 & 0.0 & 1.0 & 26.0 & 1.3 & 0.0 & 3.8 & 0.0\\
+Iridaceae & 23,127 & 7,774 & 33.6 & 0.9 & 4.7 & 0.1 & 0.0 & 1.3 & 0.4 & 1.0 & 17.1 & 0.4 & 0.1 & 12.3 & 0.0\\
+\em{Lepismium} & 825 & 245 & 29.7 & 0.0 & 2.1 & 0.0 & 0.0 & 0.0 & 0.0 & 0.1 & 21.9 & 0.1 & 0.0 & 7.8 & 0.0\\
+\em{Oocephalus} & 883 & 435 & 49.3 & 0.0 & 0.7 & 0.3 & 0.0 & 0.1 & 0.0 & 6.1 & 41.9 & 0.8 & 0.0 & 13.3 & 0.0\\
+\em{Pilosocereus} & 1,940 & 501 & 25.8 & 1.8 & 7.0 & 0.0 & 0.0 & 0.9 & 0.2 & 0.5 & 16.8 & 0.5 & 0.0 & 2.1 & 0.0\\
 \addlinespace
-Tocoyena & 2,922 & 1,092 & 37.4 & 0.1 & 1.9 & 0.2 & 0.0 & 0.5 & 0.8 & 32.3 & 0.8 & 5.0 & 0.0\\
+\em{Prosthechea} & 6,617 & 2,088 & 31.6 & 5.0 & 8.3 & 0.1 & 0.0 & 0.2 & 0.0 & 0.4 & 19.6 & 1.7 & 0.0 & 0.9 & 0.1\\
+\em{Tillandsia} & 42,222 & 14,908 & 35.3 & 4.9 & 5.1 & 0.1 & 0.0 & 1.0 & 0.3 & 0.7 & 19.8 & 0.7 & 0.0 & 9.2 & 0.0\\
+\em{Tocoyena} & 2,922 & 1,098 & 37.6 & 0.1 & 1.9 & 0.2 & 0.0 & 0.5 & 0.0 & 0.8 & 32.3 & 0.8 & 0.2 & 5.0 & 0.0\\
+\hline
+Total & 218,899 & 97,018 & 44.3 & 4.2 & 5.6 & 0.3 & 1.0 & 0.4 & 0.2 & 1.0 & 32.3 & 0.4 & 4.0 & 7.1 & 0.0\\
 \bottomrule
@@
 
-\caption{\label{tab:unnamed-chunk-3}Conservation assessment for 10 Neotropical taxa of plants and animals based on three datasets.
+\caption{\label{tab:unnamed-chunk-3}Conservation assessment for 11 Neotropical taxa of plants and animals based on three datasets.
 IUCN: global red list assessment obtained from www.iucn.org; GBIF Raw: Preliminary conservation assessment based on IUCN Criterion B using ConR and the raw dataset from GBIF; GBIF filtered: Preliminary conservation assessment based on IUCN Criterion B using ConR and the filtered dataset.
@@
 \midrule
-Arhynchobatidae & 37 & 51.3 & 17.9 & 39 & 35.9 & 45.0 & 39 & 41.0 & 40.0 & -21.0 & -13.6\\
-Dipsadidae & 519 & 68.0 & 8.8 & 637 & 58.4 & 63.0 & 597 & 60.0 & 61.2 & -2.1 & -15.6\\
-Harengula & 4 & 100.0 & 0.0 & 4 & 0.0 & 100.0 & 4 & 0.0 & 100.0 & -20.3 & -13.2\\
-Conchocarpus & 4 & 8.7 & 0.0 & 46 & 63.0 & 100.0 & 45 & 62.2 & 100.0 & -12.0 & -6.6\\
-Gaylussacia & 2 & 3.3 & 0.0 & 61 & 59.0 & 50.0 & 58 & 60.3 & 50.0 & -20.7 & -8.2\\
+Arhynchobatidae & 37 & 51.3 & 17.9 & 39 & 35.9 & 45.0 & 39 & 41.0 & 40.0 & -32.7 & -18.5\\
+Dipsadidae & 520 & 68.0 & 8.8 & 638 & 58.3 & 63.0 & 598 & 59.9 & 61.2 & -2.3 & -15.6\\
+\em{Harengula} & 4 & 100.0 & 0.0 & 4 & 0.0 & 100.0 & 4 & 0.0 & 100.0 & -38.0 & -36.9\\
+\hline
+\em{Conchocarpus} & 4 & 8.7 & 0.0 & 46 & 63.0 & 100.0 & 45 & 62.2 & 100.0 & -15.3 & -7.1\\
+\em{Gaylussacia} & 2 & 3.3 & 0.0 & 61 & 59.0 & 50.0 & 58 & 60.3 & 50.0 & -22.5 & -8.6\\
 \addlinespace
-Iridaceae & 13 & 2.3 & 0.2 & 531 & 64.4 & 50.0 & 466 & 62.9 & 62.5 & -17.5 & -12.3\\
-Lepismium & 6 & 100.0 & 0.0 & 6 & 16.7 & 83.3 & 6 & 16.7 & 83.3 & -33.9 & -7.9\\
-Pilosocereus & 41 & 80.9 & 19.1 & 47 & 55.3 & 73.7 & 46 & 56.5 & 71.1 & -8.5 & -5.8\\
-Tillandsia & 54 & 11.6 & 6.0 & 464 & 61.4 & 85.2 & 453 & 62.7 & 83.3 & -12.8 & -9.8\\
-Tocoyena & 3 & 13.6 & 4.5 & 22 & 31.8 & 66.7 & 21 & 38.1 & 66.7 & -14.7 & -8.8\\
+\em{Harpalyce} & 3 & 15.0 & 5.0 & 20 & 65.0 & 66.7 & 17 & 58.8 & 50.0 & -18.4 & -16.5\\
+Iridaceae & 13 & 2.3 & 0.2 & 531 & 64.4 & 50.0 & 466 & 62.9 & 62.5 & -18.2 & -12.3\\
+\em{Lepismium} & 6 & 100.0 & 0.0 & 6 & 16.7 & 83.3 & 6 & 16.7 & 83.3 & -33.9 & -7.9\\
+\em{Pilosocereus} & 41 & 80.9 & 19.1 & 47 & 55.3 & 73.7 & 46 & 56.5 & 71.1 & -8.5 & -5.8\\
+\em{Tillandsia} & 54 & 11.6 & 6.0 & 464 & 61.4 & 85.2 & 453 & 62.7 & 83.3 & -13.7 & -9.9\\
+\addlinespace
+\em{Tocoyena} & 3 & 13.6 & 4.5 & 22 & 31.8 & 66.7 & 21 & 38.1 & 66.7 & -23.0 & -9.5\\
 \bottomrule
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{./figures_tables/Figure_study_groups} \caption{Example of study taxa.
-\textbf{A)} Pilosocereus pusillibaccatus, picture by Pamela Lavor; \textbf{B)} Conchocarpus macrocarpus, Juliana El-Ottra; \textbf{C)} Tillandsia recurva, EC Tomaz; \textbf{D} \textit{Oxyrhopus guibei} (Dipsadidae), Filipe C Serrano;\textbf{E)} Neanuridae (\textit{Aethiopella ricardoi}), Raiane Vital da Paz; \textbf{F)} \textit{Tocoyena formosa}, Fernanda GL Moreira; \textbf{G)} \textit{Harengula jaguana}, Thais Ferreira-Araujo; \textbf{H)} \textit{Gaylussacia decipiens}, Luiz Menini Neto; \textbf{I)}  extit{Ooocephalus foliosus}, Arthur Soares; \textbf{J)} \textit{Tityus carvalhoi}, Renata C Santos-Costa; \textbf{K)} \textit{Prostheca vespa}, Tiago veira}\label{fig:species}
+\includegraphics[width=0.9\linewidth]{./figures_tables/Figure_study_groups} \caption{Examples of study taxa.
+\textbf{A)} Pilosocereus pusillibaccatus, picture by Pamela Lavor; \textbf{B)} Conchocarpus macrocarpus, Juliana El-Ottra; \textbf{C)} Tillandsia recurva, EC Tomaz; \textbf{D)} \textit{Oxyrhopus guibei} (Dipsadidae), Filipe C Serrano; \textbf{E)} Neanuridae (\textit{Aethiopella ricardoi}), Raiane Vital da Paz; \textbf{F)} \textit{Tocoyena formosa}, Fernanda GL Moreira; \textbf{G)} \textit{Harengula jaguana}, Thais Ferreira-Araujo; \textbf{H)} \textit{Gaylussacia decipiens}, Luiz Menini Neto; \textbf{I)} \textit{Ooocephalus foliosus}, Arthur Soares; \textbf{J)} \textit{Tityus carvalhoi}, Renata C Santos-Costa; \textbf{K)} \textit{Prostheca vespa}, Tiago Veira}\label{fig:species}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\linewidth]{./output/figure_number_of_records} \caption{The difference in records available between the raw and filtered dataset of 16 Neotropical taxa including plants and animals, plotted in a 100x100 km grid across the Neotropics.}\label{fig:total}
+\includegraphics[width=\linewidth]{./output/figure_number_of_records} \caption{The difference in records available between the raw and filtered dataset of 18 Neotropical taxa including animals, funghi, and plants, plotted in a 100x100 km grid across the Neotropics.}\label{fig:total}
 \end{figure}
@@
 The concentration of records flagged on the basis of record test and individual count test in Central America and southern North America.
-Only filters that flagged at least one record are shown.}\label{fig:split}
+Only filters that flagged at least 0.1\% of records in any taxon are shown.}\label{fig:split}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\linewidth]{./output/figure_number_species_richness_difference} \caption{The difference in species richness derived from the raw and filtered datasets (raw - filtered), for 16 taxa of animals, funghi, and plants.}\label{fig:speciesrichness}
+\includegraphics[width=\linewidth]{./figures_tables/figure_number_species_richness_difference} \caption{Illustrative examples for the difference in species richness between the raw and filtered dataset (raw - filtered) from four of the study taxa.
+Photo credits for C) by Tiago Andrade Borges Santos, otherwise as in Figure 1.}\label{fig:speciesrichness}
 \end{figure}
+
+\newpage{}
 
@@
 ``Final Report of the Task Group on GBIF Data Fitness for Use in Distribution Modelling - Are species occurrence data in global online repositories fit for modeling species distributions?
-The case of the Global Biodiversity Information Facility (GBIF).'' GBIF.
+The case of the Global Biodiversity Information Facility (GBIF).'' Copenhagen, Denmark: GBIF.
 
@@
 ---------. 2019p. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.zj2cyj}.
+
+\leavevmode\hypertarget{ref-GBIForg2020}{}%
+---------. 2020a. ``(25 February 2020) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.qazjh4}.
+
+\leavevmode\hypertarget{ref-GBIForg2020a}{}%
+---------. 2020b. ``(25 February 2020) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.ixq7wh}.
 

### 2020-03-02 02_automated_cleaning.R

@@ -96,6 +96,28 @@ out <- dat %>%
                             .inst & 
                             .dpl, TRUE, FALSE))
 
+tot <- out %>% 
+  summarize(
+    total_records = length(summary),
+    total_flags = sum(!summary),
+    coordinate_missing = sum(!coordinate_missing),
+    coordinate_precision = sum(!coordinate_precision),
+    coordinate_base = sum(!coordinate_base),
+    individual_count = sum(!individual_count),
+    record_age = sum(!record_age),
+    record_id = sum(!record_id),
+    invalid_coords = sum(!.val),
+    equal_coords = sum(!.equ),
+    zero_coords = sum(!.zer),
+    capital_coords = sum(!.cap),
+    centroid_coords = sum(!.cen),
+    sea_coords = sum(!.sea),
+    urban_coords = sum(!.urb),
+    gbif_coords = sum(!.gbf),
+    inst_coords = sum(!.inst),
+    dupl_coords = sum(!.dpl)) %>% 
+  mutate(taxon = "Total")
+
 summary_table <- out %>% 
   group_by(taxon) %>% 
   summarize(
@@ -117,7 +139,8 @@ summary_table <- out %>%
     gbif_coords = sum(!.gbf),
     inst_coords = sum(!.inst),
     dupl_coords = sum(!.dpl)) %>% 
-mutate(fraction = total_flags / total_records)
+  bind_rows(tot) %>% 
+  mutate(fraction = total_flags / total_records)
 
 # write to disk
 write_csv(summary_table, path = "output/summary_table_flags.csv")

### 2020-03-02 03_automated_conservation_Assessment.R

@@ -67,44 +67,44 @@ cl_in <- dat_cl%>%
 # 
 
 # IUCN assessment for the species
-iucn.key <- "01524b67f4972521acd1ded2d8b3858e7fedc7da5fd75b8bb2c5456ea18b01ba"
-
-sp.list <- rw_in$tax %>% 
-  as.character() %>% 
-  unique()
-
-#get conservation status from IUCN
-for(i in 1:length(sp.list)){
-  print(i)
-  pick <- jsonlite::fromJSON(rl_search_(sp.list[i], key = iucn.key))$result
-  
-  if(length(pick) >0){
-    if(i == 1){
-      write_csv(pick, "output/iucn_assessment.csv")
-    }else{
-      write_csv(pick, "output/iucn_assessment.csv", append = TRUE)
-    }
-  }
-
-  Sys.sleep(1)
-}
+# iucn.key <- "01524b67f4972521acd1ded2d8b3858e7fedc7da5fd75b8bb2c5456ea18b01ba"
+# 
+# sp.list <- rw_in$tax %>% 
+#   as.character() %>% 
+#   unique()
+# 
+# #get conservation status from IUCN
+# for(i in 1:length(sp.list)){
+#   print(i)
+#   pick <- jsonlite::fromJSON(rl_search_(sp.list[i], key = iucn.key))$result
+#   
+#   if(length(pick) >0){
+#     if(i == 1){
+#       write_csv(pick, "output/iucn_assessment.csv")
+#     }else{
+#       write_csv(pick, "output/iucn_assessment.csv", append = TRUE)
+#     }
+#   }
+# 
+#   Sys.sleep(1)
+# }
 
 # Load assessments
 rw <- read_csv("output/automated_assessment_raw.csv") %>% 
-  select(taxa, EOO, AOO, Category_CriteriaB, Category_code)
+  dplyr::select(taxa, EOO, AOO, Category_CriteriaB, Category_code)
 names(rw) <- paste("raw", names(rw), sep = "_")
 
 cl <- read_csv("output/automated_assessment_clean.csv")
 names(cl) <- gsub("filtered_", "", names(cl))
-cl <- select(cl, taxa, EOO, AOO, Category_CriteriaB, Category_code)
+cl <- dplyr::select(cl, taxa, EOO, AOO, Category_CriteriaB, Category_code)
 names(cl) <- paste("flag", names(cl), sep = "_")
 
 iucn <- read_csv("output/iucn_assessment.csv") %>% 
[…]

### 2020-03-02 04_figure_number_of_occurrences.R

@@ -102,8 +102,11 @@ plo <- dat %>%
                       .inst = "Biodiversity institutions",
                       .cen = "Political centroids",
                       individual_count = "Individual count",
+                      .sea = "Sea/land area",
                       .zer = "Zeros",
-                      .equ = "Equal lat/lon"))
+                      .equ = "Equal lat/lon")) %>% 
+  filter(test != "Equal lat/lon")
+  
 
 
 plo <- split(plo, f = plo$test)
@@ -134,6 +137,7 @@ plo <- plo %>%
                                         "Duplicates", 
                                         "Political centroids",
                                         "Equal lat/lon",
+                                        "Sea/land area",
                                         "Urban areas", 
                                         "Zeros")))
 

### 2020-03-02 06_species_richness_maps.R

@@ -78,19 +78,26 @@ plo <- full_join(plo_rw, plo_cl, by = c("taxon", "x", "y")) %>%
 write_csv(plo, "output/specis_richness.csv")
 
 # plots of difference for the main manuscript
+
+## select illustrative taxa
+
+plo <- plo %>% 
+  filter(taxon %in% c("Thozetella", "Tillandsia", "Dipsadidae", "Harengula"))
+
+# plot
 ggplot()+
   geom_polygon(data = world.behr,
                aes(x = long, y = lat, group = group), fill = "transparent", color = "black")+
   geom_tile(data = plo, aes(x = x, y = y, fill = log(difference)), alpha = 0.8)+
-  scale_fill_viridis(name = "Number of\nspecies", direction = 1, na.value = "transparent",
+  scale_fill_viridis(name = "Number of\nremoved\nspecies", direction = 1, na.value = "transparent",
                      breaks = c(log(1), log(5), log(10), log(20), log(30), log(40)),
                      labels = c(1, 5, 10, 20, 30, 40))+
   xlim(-12000000, -3000000)+
   ylim(-6500000, 4500000)+
   coord_fixed()+
   theme_bw()+
-  theme(legend.position = "bottom",
-        legend.key.width = unit(1.5, "cm"),
+  theme(legend.position = "right",
+        legend.key.height = unit(2.5, "cm"),
         axis.title = element_blank(),
         axis.ticks = element_blank(),
         axis.text = element_blank(),
@@ -98,36 +105,36 @@ ggplot()+
         panel.grid.minor = element_blank())+
   facet_wrap(.~ taxon)
 
-ggsave("output/figure_number_species_richness_difference.jpg", height = 10, width=8)
-
+ggsave("output/figure_number_species_richness_difference.pdf", height = 10, width=8)
+# edit the facet titles seperately to the genera in italics and add the figures
 
 # per species plots for the supplement
-li <- unique(plo$taxon)
-
-for(i in 1:length(li)){
-  sub <- filter(plo, taxon == li[i]) %>% 
-    pivot_longer(contains("layer"), values_to = "species", names_to = "dataset") %>% 
-    mutate(dataset = recode(dataset, layer_rw = "Raw", layer_cl = "Filtered")) %>% 
-    mutate(dataset = factor(dataset, levels = c("Raw", "Filtered")))
-  
-  ggplot()+
-    geom_polygon(data = world.behr,
[…]

### 2020-03-02 07_supplement_per_taxon_occurrence_maps.R

@@ -1,0 +1,104 @@
+# a script to calculate the difference in occurrence numbers per taxon for the supplement
+
+# a script to visualize the raw and cleaned records
+library(tidyverse)
+library(raster)
+library(viridis)
+library(ggthemes)
+library(speciesgeocodeR)
+
+# load data
+dat <- read_csv("output/all_records.csv") %>% 
+  filter(!is.na(decimalLongitude)) %>% 
+  filter(!is.na(decimalLatitude))
+
+be <- raster("input/ABROCOMIDAE_ABROCOMIDAE.tif")
+
+# define projections
+behr <- '+proj=cea +lon_0=0 +lat_ts=30 +x_0=0 +y_0=0 +datum=WGS84 +ellps=WGS84 +units=m +no_defs'
+wgs1984 <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
+
+# background map
+world.inp  <- suppressWarnings(rnaturalearth::ne_download(scale = 50, 
+                                                          type = 'land', 
+                                                          category = 'physical',
+                                                          load = TRUE))
+world.behr <- spTransform(world.inp, CRS(behr)) %>% fortify()
+
+#cleaned data for main mansucript
+plo <- dat %>%
+  filter(summary) %>% 
+  select(species, taxon, decimalLongitude, decimalLatitude)
+
+plo <- split(plo, f = plo$taxon)
+
+plo <-  lapply(plo, 
+               function(k){
+                 pts <- k[, c("decimalLongitude", "decimalLatitude")]%>%
+                   SpatialPoints(proj4string = CRS(wgs1984))%>%
+                   spTransform(behr) %>% 
+                   coordinates()
+                 pts <-  data.frame(species = k$species,
+                                    pts)
+                 out <- pts %>% 
+                   RichnessGrid(ras = be, type = "spnum") %>% 
+                   rasterToPoints() %>% 
+                   data.frame()})
+
+plo_cl <- bind_rows(plo, .id = "taxon")
+names(plo_cl)[4] <- "layer_cl"
+
+# unfiltered for the supplement
+plo <- dat %>%
+  select(species, taxon, decimalLongitude, decimalLatitude)
+
+plo <- split(plo, f = plo$taxon)
+
+plo <-  lapply(plo, 
+               function(k){
+                 pts <- k[, c("decimalLongitude", "decimalLatitude")]%>%
[…]

### 2020-06-04 manuscript_peerj.Rmd
send to coauthors

@@
     affiliation: 4
-  - name: Ana Paula Alves da Silva
-    affiliation: 4
-  - name: Tiago Andrade Borges Santos
+  - name: Mabel Rocio Baez-Lizarazo
     affiliation: 5
@@
     affiliation: 6    
+  - name: Jéssica Fernanda Ramos Coelho
+    affiliation: 4    
+  - name: Matheus Colli-Silva
+    affiliation: 6    
+  - name: Mariana Ramos Fantinati
+    affiliation: 4
+  - name: Moabe Ferreira Fernandes
+    affiliation: 7   
+  - name: Thais Ferreira-Araújo
+    affiliation: 4        
+  - name: Fernanda Gondim Lambert Moreira
+    affiliation: 4            
+  - name: Nathália Michelly da Cunha Santos
+    affiliation: 4
+  - name: Tiago Andrade Borges Santos
+    affiliation: 7
+  - name: Renata Clicia dos Santos‐Costa
+    affiliation: 4                                    
   - name: Filipe Cabreirinha Serrano
-    affiliation: 7
-  - name: Eduardo Calisto Tomaz
+    affiliation: 8
+  - name: Ana Paula Alves da Silva
+    affiliation: 4    
+  - name: Arthur de Souza Soares
     affiliation: 4                                            
@@
     affiliation: 4                            
-  - name: Renata Clicia dos Santos‐Costa
+  - name: Eduardo Calisto Tomaz
     affiliation: 4
-  - name: Matheus Colli-Silva
-    affiliation: 6
-  - name: Nathália Michelly da Cunha Santos
-    affiliation: 4
-  - name: Arthur de Souza Soares
-    affiliation: 4
-  - name: Thais Ferreira-Araújo
-    affiliation: 4
-  - name: Moabe Ferreira Fernandes
-    affiliation: 9
   - name: Valéria Fonseca Vale
     affiliation: 4
-  - name: Fernanda Gondim Lambert Moreira
-    affiliation: 4
-  - name: Jéssica Fernanda Ramos Coelho
-    affiliation: 4
-  - name: Mariana Ramos Fantinati
-    affiliation: 4
-  - name: Mabel Rocio Baez-Lizarazo
-    affiliation: 8
   - name: Tiago Luiz Vieira
-    affiliation: 6
+    affiliation: 7
   - name: Alexandre Antonelli
-    affiliation: 10,11,12
+    affiliation: 9,10,11
 affiliation:
@@
   - code: 5
-    address: Universidade Estadual de Feira de Santana, Feira de Santana, Brazil
+    address: Departamento de Botânica, Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil
   - code: 6
@@
   - code: 7
+    address: Departamento de Ciências Biológicas, Universidade Estadual de Feira de Santana, Feira de Santana, Brazil
+  - code: 8
     address: Departamento de Ecologia, Universidade de São Paulo, São Paulo, Brazil
-  - code: 8
-    address: Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil
   - code: 9
-    address: Departamento de Ciências Biológicas, Universidade Estadual de Feira de Santana, Feira de Santana, Brazil
+    address: Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden
   - code: 10
-    address: Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden
+    address: Department for Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden
   - code: 11
-    address: Department for Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden
-  - code: 12
     address: Royal Botanic Gardens Kew, Richmond, United Kingdom
@@
 Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern.
-Automatic flagging and filtering are a scalable and reproducible means to identify potentially problematic records in datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org).
-However, it is unclear how much data may be lost by filtering, whether the same tests should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses.
+Automatic filtering is a scalable and reproducible means to identify potentially problematic records and tailor datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org), for biodiversity analyses.
+However, it is unclear how much data may be lost by filtering, whether the same filters should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses.
 Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 18 Neotropical taxa including animals, fungi, and plants, terrestrial and marine, downloaded from GBIF.
-We find that 29-90% of the records are potentially erroneous, with large variation across taxonomic groups.
-Tests for duplicated information, collection year, basis of record as well as urban areas and coordinates for terrestrial taxa in the sea or marine taxa on land have the greatest effect.
-While many flagged records might not be *de facto* erroneous, they could be overly imprecise and increase uncertainty in downstream analyses.
-Automated flagging can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus.
+We find that a total of 44.3% of the records are potentially problematic, with large variation across taxonomic groups (25 - 90%).
+A small fraction of records was identified as erroneous in the strict sense (4.2%), the majority of filtered records that were rather unfit for downstream analyses (41.7%).
+Filters of duplicated information, collection year, basis of record as well as urban areas and coordinates for terrestrial taxa in the sea or marine taxa on land have the greatest effect.
+Automated filtering can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus.
 Our results stress the importance of thorough exploration of the meta-data associated with species records for biodiversity research.
@@
   rticles::peerj_article: default
+header-includes:
+  - \usepackage[export]{adjustbox}
 nocite: |
@@
 ---
+
+<!-- # Keywords -->
+<!-- Automated cleaning, automated conservation assessment, data quality, GBIF, Neotropics. -->
 
@@
 In particular, the availability of digitized collections from museums and herbaria, and citizen science observations has increased drastically over the last few years.
-As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.3 billion geo-referenced occurrence records for species from across the globe and the tree of life.
+As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.4 billion geo-referenced occurrence records for species from across the globe and the tree of life.
 
-A central challenge to the use of these publicly available species occurrence data in research are erroneous geographic coordinates [@Anderson2016].
-Errors mostly arise because public databases integrate records collected with different methodologies in different places, at different times; often without centralized curation and only rudimentary meta-data.
-For instance, erroneous coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common [@Maldonado2015; @Yesson2007] and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in captivity [@Zizka2019].
+A central challenge to the use of these publicly available species occurrence data in research are problematic geographic coordinates, which are either erroneous or unfit for downstream analyses [for instance because they are overly imprecise, @Anderson2016].
+Problems mostly arise because public databases integrate records collected with different methodologies in different places, at different times; often without centralized curation and only rudimentary meta-data.
+For instance, problematic coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common [@Maldonado2015; @Yesson2007] and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in captivity [@Zizka2019].
 
 Manual data cleaning based on expert knowledge can detect these issues, but it is only feasible on small taxonomic or geographic scales, and it is time-consuming and difficult to reproduce.
-As an alternative, automated flagging methods to identify potentially problematic records have been proposed as a scalable option, as they are able to deal with datasets containing up to millions of records and many different taxa.
+As an alternative, automated filtering methods to identify potentially problematic records have been proposed as a scalable option, as they are able to deal with datasets containing up to millions of records and many different taxa.
 Those methods are usually based on geographic gazetteers [e.g., @Chamberlain2016; @Zizka2019; @Jin2020] or on additional data, such as environmental variables [@Robertson2016].
@@
 
-Erroneous records are especially problematic in conservation, where stakes are high.
+Problematic records are especially critical in conservation, where stakes are high.
 Recently proposed methods for automated conservation assessments could support the formal assessment procedures for the global Red List of the International Union for the Conservation of Nature (IUCN) [@Dauby2017; @Bachman2011; @Pelletier2018].
 These methods approximate species' range size, namely the Extent of Occurrence (EOO, which is the area of a convex hull polygon comprising all records of a species), the Area of Occupancy (AOO, which is the sum of the area actually occupied by a species, calculated based on a small-scale regular grid), and the number of locations for a preliminary conservation assessment following IUCN Criterion B ("Geographic range").
-These methods have been used to propose preliminary global [@Stevart2019; @Zizka2019a] and regional [@Schmidt2017; @Cosiaux2018] Red List assessments.
-However, all metrics but especially EOO, are sensitive to individual records with erroneous coordinates.
+These methods have been used to propose preliminary global [@Stevart2019; @Zizka2020b] and regional [@Schmidt2017; @Cosiaux2018] Red List assessments.
+However, all metrics and especially EOO, are sensitive to individual records with problematic coordinates.
 Automated conservation assessments may therefore be biased, particularly if the number of records are few, as it is the case for many tropical species.
 
-While newly proposed methods for automated cleaning of species records hold great promise for biodiversity research, their use across taxonomic groups and datasets remains poorly explored.
+While automated filters hold great promise for biodiversity research, their use across taxonomic groups and datasets remains poorly explored.
 Here, we test the effect of automated filtering of species geographic occurrence records on the number of records available in different groups of animals, fungi, and plants.
-Furthermore, we test the impact of automated filtering procedures for the accuracy of automated conservation assessments compared to full IUCN assessments.
-Specifically, we evaluate a pipeline of 13 automated filters to flag possibly erroneous records by using 1) record meta-data and 2) geographic gazetteers.
+Furthermore, we test the impact of automated filtering procedures for the accuracy of preliminary automated conservation assessments compared to full IUCN assessments.
+Specifically, we evaluate a pipeline of 13 automated filters to flag possibly problematic records by using record meta-data and geographic gazetteers in two categories: 1) erroneous (coordinates, that are likely wrong, irrespective of the downstream analyses, for instance due to data entry errors) and 2) unfit for purpose (coordinates that are not wrong *per se*, but likely unfit for the planned downstream analyses, for instance because they are overly imprecise).
 We address three questions:
@@
 We downloaded occurrence information for all study groups from www.gbif.org using the `rgbif` v1.4.0 package [@Chamberlain2017] in R (GBIF.org, 2019a-p,2020a,b).
-We downloaded only records with geographic coordinates and limited the study area to latitudes smaller than 33$^\circ$ N and longitudes smaller than 35$^\circ$ W and larger than 120$^\circ$ W reflecting the Neotropics, our main area of expertise [@Morrone2014].
+We downloaded GBIF-interpreted data including only records with geographic coordinates and limited the study area to a rectangle between 90$^\circ$S - 33$^\circ$ N and 35$^\circ$ W - 120$^\circ$ W reflecting the Neotropics [@Morrone2014], our main area of expertise.
 The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, Entomobryidae, *Gaylussacia*, Iridaceae, Neanuridae, and *Tillandsia* for which we only obtained the Neotropical occurrences.
+We consider GBIF data generally of high quality and use them as study case because GBIF is the largest, most widely-used and taxonomically most comprehensive data source for species occurrence records, but many more exist [e.g., https://bien.nceas.ucsb.edu/bien/, www.fishbase.de or @Guedes2018].
+The variety of existing databases is also the reason why we download data without accounting for GBIF-internal quality testing ("Geospatial issues"), since quality control mechanisms might differ among databases and sometimes no quality control might be done (for instance in *de novo* databases compiled by groups of researchers).
 
 ##  Automated cleaning {-}
-We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the raw data downloaded from GBIF ("raw", hereafter) using meta-data for those records for which they were available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99.
+We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the raw data downloaded from GBIF ("raw", hereafter) using meta-data for those records for which they were available [although this is often not the case, @Peterson2018], removing: 1) records with a coordinate precision below 100 km (as this represents the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99.
 Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e., test for duplicated records).
@@
 These include: coordinates in the sea for terrestrial taxa and on land for marine taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions.
-See Table 2 for a summary of all flagging tests we used.
+See Table 2 for a summary of all filters we used and their classification into "erroneous" and "unfit".
 
 ## Downstream analyses {-}
-We first generated species richness maps using 100x100 km grid cells for the raw and filtered datasets respectively, using the package `speciesgeocodeR v2.0-10` [@Topel2016].
+We first generated species richness maps using 100x100 km grid cells for the raw and filtered datasets respectively, using the package `speciesgeocodeR v2.0-10` [@Topel2017].
 We then performed an automated conservation assessment for all study groups based on both datasets using the `ConR v1.2.4` package [@Dauby2017].
@@
 ## Evaluation of results {-}
-<!-- This session is exciting and makes the study more reliable and interesting than one just based on data mining, but how this was done is very unclear.
-Did people provide a specific score for those, or how were their assessments recorded?
-I think we need to clarify this, as it’s the weakest part of the paper at the moment.
-We don’t report the results for this evaluation nor what they mean.
--->
-Each author provided an informed comment on the performance of the raw and cleaned datasets, respectively concerning the number of flagged records and the accuracy of the overall species richness maps and we compared the agreement between automated conservation assessments based on raw and filtered occurrences with the global IUCN Red List (www.iucn.org).
+Each author provided an informed comment on the performance of the raw and cleaned datasets, concerning the number of removed records and the accuracy of the overall species richness maps.
+We then compared the agreement between automated conservation assessments based on raw and filtered occurrences with the global IUCN Red List for those taxa where IUCN assessments were available (www.iucn.org).
 
 We did all analyses in the R computing environment [@rcoreteam2019], using standard libraries for data handling and visualization [@Wickham2018; @Garnier2018; @Ooms2014; @Ooms2019; @Hijmans2019].
-All scripts are available from a zenodo repository (**DOI**).
+All scripts are available from a zenodo repository (doi:10.5281/zenodo.3695102).
 
@@
   mutate_all(linebreak) %>%
-  select(-researcher) %>% 
+  dplyr::select(-researcher) %>% 
   filter(!taxon %in% c("Opisthonema", "Eugenia"))
@@
 
-tab2 <- tab2[,c(1:3,16,5,7,4,8,6,14,10,15,11:13,9)]
+tab2 <- tab2[,c(1:3,16:17,14,12,9,10,15,11,13,5,7,4,8,6,18)]
 
-tab2[,-c(1:4)] <- round((tab2[,-c(1:4)] / unlist(tab2[,2]) * 100), 1)
+err <- tab2[19,4] <-  base::colSums(tab2[, 4], na.rm = TRUE)
+unfit <- tab2[19,5] <-  base::colSums(tab2[, 5], na.rm = TRUE)
+
+tab2[,-c(1:2,18)] <- round((tab2[,-c(1:2,18)] / unlist(tab2[,2]) * 100), 1)
 
 tab2 <- tab1 %>% 
-  select(taxon) %>%  
+  dplyr::select(taxon) %>%  
   bind_rows(data.frame(taxon = "Total")) %>% 
@@
 tab3 <- tab1 %>% 
-  select(taxon) %>%  
+  dplyr::select(taxon) %>%  
   left_join(tab3, by = "taxon") %>% 
@@
 
-We retrieved a total of `r nrow(dat)` species occurrence records, with a median of `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per study group and `r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per species (Table 3, Appendix 1).
+We retrieved a total of `r nrow(dat)` species occurrence records, with a median of `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist() %>%  round(0)` records per study group and `r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per species (Table 3, Appendix 1).
 We obtained most records for `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(taxon) %>%  unlist()` (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs) %>%  unlist()`) and fewest for *`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(taxon) %>%  unlist()`* (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(nrecs) %>%  unlist()`).
@@
 
-Our automated tests flagged a total of `r sum(!dat$summary)` records (Fig.
-\ref{fig:total}), with a median of `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per group.
+Our automated tests filtered a total of `r sum(!dat$summary)` records (Fig.
+\ref{fig:total}, erroneous: `r err`, unfit: `r unfit`), with a median of `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per group (erroneous: `r median(tab2$total_error)`%, unfit: `r median(tab2$total_unfit)`%).
 Overall, the most important test was for duplicated records (on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!.dpl)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per taxonomic group).
-The filtering steps based on record meta-data that flagged the largest number of records were the basis of records (`r dat %>% group_by(taxon) %>% summarize(num = mean(!coordinate_base)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the collection year (`r dat %>% group_by(taxon) %>% summarize(num = mean(!record_age)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%).
+The filtering steps based on record meta-data that filtered the largest number of records were the basis of records (`r dat %>% group_by(taxon) %>% summarize(num = mean(!coordinate_base)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the collection year (`r dat %>% group_by(taxon) %>% summarize(num = mean(!record_age)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%).
 The most important automated tests were for urban area (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.urb)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the occurrence from records of terrestrial taxa in the sea and marine taxa on land (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.sea)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%, see Table 3 and Appendix 1 in the electronic supplement for further details and the absolute numbers).
-Only few records were flagged by the coordinate precision, zero coordinates and biodiversity institution tests (Fig.
+Only a few records were filtered by the coordinate precision, zero coordinates and biodiversity institution tests (Fig.
 \ref{fig:split}).
 
-*`r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(1) %>% unlist()`*, *`r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(2) %>% unlist()`*, and *`r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(3) %>% unlist()`* had the highest fraction of flagged records (Table 3).
+`r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(1) %>% unlist()`, `r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(2) %>% unlist()`, and `r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(3) %>% unlist()` had the highest fraction of filtered records (Table 3).
 In general, the different filters we tested were of similar importance for different study groups.
-There were few outstanding exceptions, including the particularly high proportions of records flagged on the "basis of record test" for *Tityus* (7.0%), Dipsadidae (5.6%), *Prosthechea* (5.0%) and *Tillandsia* (4.9%), by the collection year for Dipsadidae (11.3%), by the taxonomic identification level for  *Tityus* (1.6%), by the capital coordinates for *Oocephalus* (6.1%) and *Gaylussacia* (3.2%), by the seas/land test for Diogenidae and *Thozetella*, and by the urban areas test for *Oocephalus* (13.3%) and Iridaceae (12.3%).
-Furthermore, Entomobryidae differed considerably from all other study taxa with exceptionally high numbers of records flagged by the "basis of record, "level of identification' and urban areas" tests.
+There were few outstanding exceptions, including the particularly high proportions of records filtered on the "basis of record test" for *Tityus* (7.0%), Dipsadidae (5.6%), *Prosthechea* (5.0%) and *Tillandsia* (4.9%), by the collection year for Dipsadidae (11.3%), by the taxonomic identification level for  *Tityus* (1.6%), by the capital coordinates for *Oocephalus* (6.1%) and *Gaylussacia* (3.2%), by the seas/land test for Diogenidae and *Thozetella*, and by the urban areas test for *Oocephalus* (13.3%) and Iridaceae (12.3%).
+Furthermore, Entomobryidae differed considerably from all other study taxa with exceptionally high numbers of records filtered by the "basis of record, "level of identification' and urban areas" tests.
 
-Geographically, the records flagged by the "basis of record" and "individual count" tests were concentrated in Central America and southern North America, and a relatively high number of records flagged due to their proximity to the centroids of political entities were located on Caribbean islands (Fig.
+Geographically, the records filtered by the "basis of record" and "individual count" tests were concentrated in Central America and southern North America, and a relatively high number of records filtered due to their proximity to the centroids of political entities were located on Caribbean islands (Fig.
 \ref{fig:split}).
@@
 <!-- add some subheadings to discussion -->
-Automated flagging based on meta-data and automatic tests flagged on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% of the records per taxonomic group.
-The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (Question 1).
-The importance of different tests was similar across taxonomic groups, with exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (Question 2).
+Automated flagging based on meta-data and automatic tests filtered on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% of the records per taxonomic group; `r min(tab2$total_unfit)`% - `r max(tab2$total_unfit)`% as "unfit" and `r min(tab2$total_error)`% - `r max(tab2$total_error)`% as "erroneous".
+The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (**Question 1**).
+The importance of different tests was similar across taxonomic groups, with exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (**Question 2**).
 The results for species richness were similar between the raw and filtered data with some improvements by the filters.
-We found little impact of filtering on the accuracy of the automated conservation assessments (Question 3).
+We found little impact of filtering on the accuracy of the automated conservation assessments (**Question 3**).
 
 ## The relevance of individual filters {-}
+The aim of automated filtering is to identify possibly problematic records that are unsuitable for particular downstream analyses.
+While those records filtered as "erroneous" will likely cause problems for most biodiversity research, those filtered as "unfit" might have varying impact, depending on the type and spatial resolution of the downstream analyses, ranging from an unnecessary computational burden which can be a bottle neck for large-scale analyses [i.e.
+duplicates, @Antonelli2018] over increasing uncertainty (due to low precision) to compromising results.
+For instance, records assigned to country centroids might be acceptable for inter-continental comparisons, but likely are erroneous for species distribution modelling on a local scale.
+The importance of each test and the linked thresholds must be judged based on the specific downstream analyses.
+As our results show, adapting automated tests to the geographic study area or the taxonomic study group can be warranted.
+For instance, the high number of records flagged for centroids on the Lesser Antilles (Fig.
+\ref{fig:split}) might be overly strict (https://data-blog.gbif.org/post/country-centroids/), although we chose a conservative distance for the Political centroid test (1 km).
+
 Several reasons may explain the high proportion of records flagged as duplicates.
@@
 Third, low coordinate precision, for instance based on automated geo-referencing from locality descriptions, may lump records from nearby localities.
-Duplicated records do not represent errors *per se*, but an unnecessary computational burden, which is often a bottleneck for analyses of datasets with millions of records.
-
-The aim of automated filtering is to identify possibly problematic records that are unsuitable for particular downstream analyses.
-Hence, our tests may identify true errors (e.g., wrong coordinates such as those with exclusively zeros, equal latitude and longitude and records of terrestrial species in the sea or marine species on land) as well as records that are not necessarily erroneous, but have a low precision (e.g., collection year, coordinate precision, centroids of political units) or are redundant for certain analyses (duplicates).
-The importance of each test and the linked thresholds must be judged based on the specific downstream analyses.
-For instance, records assigned to country centroids might be acceptable for inter-continental comparisons, but can be considered erroneous for species distribution modelling on a local scale.
-As our results show, adapting automated tests to the geographic study area or the taxonomic study group can be warranted.
-For instance, the high number of records flagged for centroids on the Lesser Antilles (Fig.
-\ref{fig:split}) might be overly strict, although we chose a conservative distance for the Political centroid test (1 km).
+Fourth, different data contributors might add the same record to GBIF, if their sources overlap, as can for instance be the case for the Barcode of Life and Plazi databases.
 
 ## Similarities and differences among taxa {-}
-The number of records flagged by individual tests was similar across study groups, suggesting that similar error types might be relevant for collections of plants and animals.
-Therefore, the same tests can be used across taxonomic groups.
-Some notable exceptions stress the need to adapt each test to the taxonomic study group to balance data quality and data availability.
-The high fraction of records flagged by the "basis of record test" for *Tityus*, Dipsadidae, *Prosthechea* and *Tillandsia*, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups.
+The number of records flagged by individual tests was similar across study groups, suggesting that similar problems might be relevant for collections of plants and animals.
+Therefore, the same filters can be used across taxonomic groups.
+Some notable exceptions stress the need to adapt each filter to the taxonomic study group to balance data quality and data availability.
+The high fraction of records filtered by the "basis of record filter" for *Tityus*, Dipsadidae, *Prosthechea* and *Tillandsia*, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups.
 The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection effort already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700.
 The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700.
-The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae, and *Oocephalus* might be related to horticulture since at least some species in those groups are commonly cultivated as ornamentals.
+The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae and *Oocephalus* might be related to horticulture since at least some species in those groups are commonly cultivated as ornamentals.
 
@@
 For instance, in Entomobryidae, `r dat %>% filter(taxon == "Entomobryidae") %>% filter( basisOfRecord == "MATERIAL_SAMPLE") %>% nrow()` samples were marked as material samples, and therefore removed by our global filter retaining only specimen and observation data, which in this case was overly strict.
-It should be noted, that, while in this study we focused on meta-data and geographic filtering, taxonomic cleaning---the resolution of synonymies and identification of accepted names---is another important part of data curation, but depends on taxon-specific taxonomic backbones and synonymy lists which are not readily available for many groups.
+
+The filters we included in this study address a set of important but relatively easy to identify problems.
+In fact, the internal quality control of GBIF does flag some of the problems we tested for (i.e., zero coordinates, equal lat/lon) while others might be implemented in the near future (country centroids, https://data-blog.gbif.org/post/country-centroids/, https://data-blog.gbif.org/post/country-centroids/).
+While this internal quality is very helpful, we see a huge potential to overcome issues with data quality in a user-feedback system allowing users to provide expert assessments, i.e.
+a meta-annotation of records being challenged (and why).
+Such a system would not need to change the original data and could include multiple levels to account for differing opinions.
+
+As a next step for automated filtering, it is possible to detect additional problematic records with the help of external species distribution information, which is available for some taxa for example from www.iucn.org (vertebrates) or https://wcsp.science.kew.org/ (selected seed plant families).
+Furthermore, while in this study we focused on meta-data and geographic filtering, taxonomic cleaning---the resolution of synonymies and identification of accepted names---is another important part of data curation, but depends on taxon-specific taxonomic backbones and synonymy lists which are not readily available for many groups and often are contradictory within individual taxa.
 
 ## The impact of filtering on the accuracy of automated conservation assessments {-}
-The accuracy of the automated conservation assessment was in the same range as found by previous studies [@NicLughadha2019; @Zizka2019a].
+The accuracy of the automated conservation assessment was in the same range as found by previous studies [@NicLughadha2019; @Zizka2020b].
 The similar accuracy of the raw and filtered dataset for the automated conservation assessment was surprising, in particular given the EOO and AOO reduction observed in the filtered dataset (Table 4) and the impact of errors on spatial analyses observed in previous studies [@Gueta2016].
 The robustness of the automated assessment was likely due to the fact that the EOO for most species was large, even after the considerable reduction caused by filtering.
-This might be caused by the structure of our comparison, which only included species that were evaluated by the IUCN Red List (and not considered as Data Deficient) and at the same time had occurrences recorded in GBIF.
-Those inclusion criteria are likely to have, biased the datasets towards species with large ranges, since generally more data for them are available.
-The robustness of automated conservation assessments to data quality is encouraging, although these methods are only an approximation, especially for species with few collection records [@Rivers2011].
+This might be caused by the structure of our comparison, which only included species that were evaluated by the IUCN Red List (and not considered as *Data Deficient*) and at the same time had occurrences recorded in GBIF.
+Those inclusion criteria are likely to have biased the datasets towards species with large ranges, since generally more data for them are available.
+The robustness of automated conservation assessments to data quality is encouraging, although these methods are only an approximation (and not replacements) of full IUCN Red List assessments, especially for species with few collection records [@Rivers2011].
 
@@
 
-Our results stress the importance of (meta-)data exploration prior to most biodiversity analyses.
-Automated flagging can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the specific taxonomic groups and geographic area of interest.
-The putative errors we encountered point to the importance to train researchers and students to curate species occurrence datasets and to allow users to provide feedback for particular records, thus contributing to overall accuracy and precision of public biodiversity databases.
+While our results suggest that large-scale species richness patterns and automated conservation assessments are largely robust to problematic occurrence records, they also stress the importance of (meta-)data exploration prior to most biodiversity analyses.
+Automated filtering can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the specific taxonomic groups and geographic area of interest.
+The putative problems we encountered point to the importance to train researchers and students to curate species occurrence datasets and to allow users to provide data aggregators such as GBIF with feedback for particular records, thus contributing to overall accuracy and precision of public biodiversity databases.
 
 # Acknowledgements {-}
-This study enrolled participants of the workshop "Biodiversity data: from field to yield" led by Alexandre Antonelli and Alexander Zizka through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI.
+This study enrolled participants of the workshop "Biodiversity data: from field to yield" led by Alice Calvente, Fernanda Carvalho, Alexander Zizka, and Alexandre Antonelli through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI.
 We thank the Pró-reitoria de Pesquisa and the Pró-reitoria de Pós-graduação of UFRN for financial support (edital 02/2016 - internacionalização).
@@
 AA is supported by the Swedish Research Council, the Knut and Alice Wallenberg Foundation, the Swedish Foundation for Strategic Research and the Royal Botanic Gardens, Kew.
+FS was financed by the Coordenação de Aperfeiçoamento de Pessoal de Nível Superior - Brasil (CAPES) - Finance Code 001 and Fundação de Amparo  à Pesquisa do estado de São Paulo (FAPESP) (FAPESP, process 2015/20215-7).
 
@@
 
-- Appendix 1 - Table S1, reporting the absolute number of flagged records per taxonomic group and test
+- Appendix 1 - Absolute number of flagged records per taxonomic group and test
 
-- Appendix 2 - Taxon specific richness maps and comments.
+- Appendix 2 - Taxon specific richness maps and comments
 
-- Appendix 3 - The full results of the conservation assessment
+- Appendix 3 - Full results of the conservation assessment
 
@@
 
-gens <- tab1 %>% filter(taxon_rank == "Genus") %>% select(taxon) %>% unlist()
+gens <- tab1 %>% filter(taxon_rank == "Genus") %>% dplyr::select(taxon) %>% unlist()
 
 tab1 <- tab1 %>% 
-  mutate(taxon =  cell_spec(taxon, "latex", italic = ifelse(.$taxon %in% gens, TRUE, FALSE)))
+  mutate(taxon =  cell_spec(taxon, "latex", italic = ifelse(.$taxon %in% gens, TRUE, FALSE))) %>% 
+  dplyr::select(-contains("pp")) %>% 
+  mutate(phylopic = sprintf('\\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/%s.png}', phylopic)) %>% 
+  dplyr::select(phylopic, everything()) %>% 
+  select(-order)
+
+  #mutate(phylopic = paste("input/phylopics/", phylopic, ".jpg", sep = "")%>% pander::pandoc.image.return())
 
 kable(tab1, booktabs = T, escape = F, format="latex",
-      col.names = linebreak(c("Taxon",
+      col.names = linebreak(c("",
+                              "Taxon",
                               "Taxon rank",
@@
                               "'Phylum'",
-                              "Order",
+#                              "Order",
                               "Family")),
@@
 This study includes three marine and 15 terrestrial taxa, seven of them animals, one group of fungi and ten plants, belonging to 16 different orders.")%>%
-  landscape()%>%
+  #landscape()%>%
   kable_styling(font_size = 11) %>%
@@
    row_spec(8, hline_after = TRUE) %>%
-   column_spec(1, width = "2.5cm", bold = FALSE, italic = FALSE)%>%
-   column_spec(2, width = "2cm", bold = FALSE, italic = FALSE)%>%
-   column_spec(3, width = "2cm", bold = FALSE, italic = FALSE)%>%
-   column_spec(4:5, width = "2.5cm", bold = FALSE, italic = FALSE)
+   column_spec(1, width = "1cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(2, width = "2.5cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(3, width = "1.5cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(4, width = "2cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(5, width = "2cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(6, width = "2cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(7, width = "3cm", bold = FALSE, italic = FALSE)
 ```
@@
    column_spec(1, width = "2cm", bold = FALSE, italic = FALSE)%>%
-   column_spec(2, width = "2.5cm", bold = FALSE, italic = FALSE)%>%
-   column_spec(3, width = "10cm", bold = FALSE, italic = FALSE)
+   column_spec(2, width = "1cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(3, width = "2.5cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(4, width = "9cm", bold = FALSE, italic = FALSE)
 ```
@@
 tab2 <- tab2 %>% 
+  select(-fraction) %>% 
   mutate(taxon =  cell_spec(taxon, "latex", italic = ifelse(.$taxon %in% gens, TRUE, FALSE)))%>%
@@
                               "Total records",
-                              "Records flagged",
                               "Fraction flagged [\\%]",
+                              "Fraction error [\\%]",
+                              "Fraction unfit [\\%]",
+                              "Biodiversity Institutions [\\%]",
+                              "Sea/land area [\\%]",
+                              "Zeros [\\%]",
+                              "Capitals [\\%]",
+                              "Duplicates [\\%]",
+                              "Political centroids [\\%]",
+                              "Urban areas [\\%]",
                               "Basis of record [\\%]",
@@
                               "Id-level [\\%]",
-                              "Individual count [\\%]",
-                              "Biodiversity Institutions [\\%]",
-                              "Capitals [\\%]",
-                              "Duplicates [\\%]",
-                              "Political centroids [\\%]",
-                              "Sea/land area [\\%]",
-                              "Urban areas [\\%]",
-                              "Zeros [\\%]")),
+                              "Individual count [\\%]"
+                              )),
         caption = "The impact of automated filtering on occurrence records for 18 Neotropical taxa downloaded from www.gbif.org.
@@
   landscape()%>%
-  kable_styling(font_size = 9) %>%
+  kable_styling(font_size = 8) %>%
    row_spec(7, hline_after = TRUE) %>%
@@
   row_spec(18, hline_after = TRUE) %>%
-   column_spec(1, width = "2cm", bold = FALSE, italic = FALSE)%>%
-   column_spec(2:9, width = "1cm", bold = FALSE, italic = FALSE)%>%
-   column_spec(10, width = "1.2cm", bold = FALSE, italic = FALSE)%>%
-   column_spec(11:12, width = "1cm", bold = FALSE, italic = FALSE)%>%
-   column_spec(13:15, width = "1cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(1, width = "1.9cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(2:5, width = "0.9cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(6, width = "1.1cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(7:9, width = "0.9cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(10, width = "1.1cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(11:12, width = "0.9cm", bold = FALSE, italic = FALSE)%>%
+   column_spec(13:17, width = "0.9cm", bold = FALSE, italic = FALSE)%>%
    add_header_above(c(" " = 1,
-                      "Summary" = 3,
-                     "Meta data filters" = 5,
-                     "Gazetteer-based filters" = 7))
+                      "Summary" = 4,
+                      "Errors" = 3,
+                      "Unfit" = 9))
 ```
@@
 <!-- The study species-->
-```{r species, out.width="0.9\\linewidth", fig.cap=c('Examples of study taxa.
-\\textbf{A)} Pilosocereus pusillibaccatus, picture by Pamela Lavor; \\textbf{B)} Conchocarpus macrocarpus, Juliana El-Ottra; \\textbf{C)} Tillandsia recurva, EC Tomaz; \\textbf{D)} \\textit{Oxyrhopus guibei} (Dipsadidae), Filipe C Serrano; \\textbf{E)} Neanuridae (\\textit{Aethiopella ricardoi}), Raiane Vital da Paz; \\textbf{F)} \\textit{Tocoyena formosa}, Fernanda GL Moreira; \\textbf{G)} \\textit{Harengula jaguana}, Thais Ferreira-Araujo; \\textbf{H)} \\textit{Gaylussacia decipiens}, Luiz Menini Neto; \\textbf{I)} \\textit{Ooocephalus foliosus}, Arthur Soares; \\textbf{J)} \\textit{Tityus carvalhoi}, Renata C Santos-Costa; \\textbf{K)} \\textit{Prostheca vespa}, Tiago Veira'), echo=FALSE}
-knitr::include_graphics(path = "./figures_tables/Figure_study_groups.jpg")
+```{r species, out.width="0.9\\linewidth", fig.cap=c('Examples for taxa included in this study.
+\\textbf{A)} \\textit{Pilosocereus pusillibaccatus} (\\textit{Pilosocereus}), \\textbf{B)} \\textit{Conchocarpus macrocarpus} (\\textit{Conchocarpus}); \\textbf{C)} \\textit{Tillandsia recurva} (\\textit{Tillandsia}); \\textbf{D)} \\textit{Oxyrhopus guibei} (Dipsadidae); \\textbf{E)} \\textit{Aethiopella ricardoi} (Neanuridae); \\textbf{F)} \\textit{Tocoyena formosa} (\\textit{Tocoyena}); \\textbf{G)} \\textit{Harengula jaguana} (\\textit{Harengula}); \\textbf{H)} \\textit{Gaylussacia decipiens} (\\textit{Gaylussacia}); \\textbf{I)} \\textit{Oocephalus foliosus} (\\textit{Oocephalus}); \\textbf{J)} \\textit{Tityus carvalhoi} (\\textit{Tityus}); \\textbf{K)} \\textit{Prosthechea vespa} (\\textit{Prosthechea}), Image credits: A) Pamela Lavor, B) Juliana El-Ottra, C) Eduardo Tomaz, D) Filipe C Serrano, E) Raiane Vital da Paz, F) Fernanda GL Moreira, G) Thais Ferreira-Araujo, H) Luiz Menini Neto, I) Arthur Soares, J) Renata C Santos-Costa, K) Tiago Vieira'), echo=FALSE}
+knitr::include_graphics(path = "./figures_tables/Figure_study_groups_reduced.jpg")
 ```
@@
 <!-- A map with number of species and where they have been removed -->
-```{r total, out.width="\\linewidth", fig.cap=c('The difference in records available between the raw and filtered dataset of 18 Neotropical taxa including animals, funghi, and plants, plotted in a 100x100 km grid across the Neotropics.'), echo=FALSE}
+```{r total, out.width="\\linewidth", fig.cap=c('The difference in records available between the raw and filtered dataset of 18 Neotropical taxa including animals, fungi, and plants, plotted in a 100x100 km grid across the Neotropics.'), echo=FALSE}
 knitr::include_graphics(path = "./output/figure_number_of_records.jpg")
@@
 ```{r split, out.width="\\linewidth", fig.cap=c('Geographic location of the occurrence records flagged by the automated tests applied in this study.
-The concentration of records flagged on the basis of record test and individual count test in Central America and southern North America.
 Only filters that flagged at least 0.1\\% of records in any taxon are shown.'), echo=FALSE}

### 2020-06-04 manuscript_peerj.tex
send to coauthors

@@
 % Pandoc Header
+\usepackage[export]{adjustbox}
 \usepackage{booktabs}
@@
 
-\author[4]{Ana Paula Alves da Silva}
-
-\author[5]{Tiago Andrade Borges Santos}
+\author[5]{Mabel Rocio Baez-Lizarazo}
 
@@
 
-\author[7]{Filipe Cabreirinha Serrano}
+\author[4]{Jéssica Fernanda Ramos Coelho}
+
+\author[6]{Matheus Colli-Silva}
+
+\author[4]{Mariana Ramos Fantinati}
+
+\author[7]{Moabe Ferreira Fernandes}
+
+\author[4]{Thais Ferreira-Araújo}
+
+\author[4]{Fernanda Gondim Lambert Moreira}
+
+\author[4]{Nathália Michelly da Cunha Santos}
+
+\author[7]{Tiago Andrade Borges Santos}
+
+\author[4]{Renata Clicia dos Santos‐Costa}
+
+\author[8]{Filipe Cabreirinha Serrano}
+
+\author[4]{Ana Paula Alves da Silva}
+
+\author[4]{Arthur de Souza Soares}
+
+\author[4]{Paolla Gabryelle Cavalcante de Souza}
 
@@
 
-\author[4]{Paolla Gabryelle Cavalcante de Souza}
-
-\author[4]{Renata Clicia dos Santos‐Costa}
-
-\author[6]{Matheus Colli-Silva}
-
-\author[4]{Nathália Michelly da Cunha Santos}
-
-\author[4]{Arthur de Souza Soares}
-
-\author[4]{Thais Ferreira-Araújo}
-
-\author[9]{Moabe Ferreira Fernandes}
-
 \author[4]{Valéria Fonseca Vale}
 
-\author[4]{Fernanda Gondim Lambert Moreira}
+\author[7]{Tiago Luiz Vieira}
 
-\author[4]{Jéssica Fernanda Ramos Coelho}
-
-\author[4]{Mariana Ramos Fantinati}
-
-\author[8]{Mabel Rocio Baez-Lizarazo}
-
-\author[6]{Tiago Luiz Vieira}
-
-\author[10,11,12]{Alexandre Antonelli}
+\author[9,10,11]{Alexandre Antonelli}
 
@@
 \affil[4]{Departamento de Botânica e Zoologia, Universidade Federal do Rio Grande do Norte, Natal, Brazil}
-\affil[5]{Universidade Estadual de Feira de Santana, Feira de Santana, Brazil}
+\affil[5]{Departamento de Botânica, Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil}
 \affil[6]{Departamento de Botânica, Universidade de São Paulo, São Paulo, Brazil}
-\affil[7]{Departamento de Ecologia, Universidade de São Paulo, São Paulo, Brazil}
-\affil[8]{Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil}
-\affil[9]{Departamento de Ciências Biológicas, Universidade Estadual de Feira de Santana, Feira de Santana, Brazil}
-\affil[10]{Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden}
-\affil[11]{Department for Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden}
-\affil[12]{Royal Botanic Gardens Kew, Richmond, United Kingdom}
+\affil[7]{Departamento de Ciências Biológicas, Universidade Estadual de Feira de Santana, Feira de Santana, Brazil}
+\affil[8]{Departamento de Ecologia, Universidade de São Paulo, São Paulo, Brazil}
+\affil[9]{Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden}
+\affil[10]{Department for Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden}
+\affil[11]{Royal Botanic Gardens Kew, Richmond, United Kingdom}
 
@@
 Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern.
-Automatic flagging and filtering are a scalable and reproducible means to identify potentially problematic records in datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org).
-However, it is unclear how much data may be lost by filtering, whether the same tests should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses.
+Automatic filtering is a scalable and reproducible means to identify potentially problematic records and tailor datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org), for biodiversity analyses.
+However, it is unclear how much data may be lost by filtering, whether the same filters should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses.
 Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 18 Neotropical taxa including animals, fungi, and plants, terrestrial and marine, downloaded from GBIF.
-We find that 29-90\% of the records are potentially erroneous, with large variation across taxonomic groups.
-Tests for duplicated information, collection year, basis of record as well as urban areas and coordinates for terrestrial taxa in the sea or marine taxa on land have the greatest effect.
-While many flagged records might not be \emph{de facto} erroneous, they could be overly imprecise and increase uncertainty in downstream analyses.
-Automated flagging can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus.
+We find that a total of 44.3\% of the records are potentially problematic, with large variation across taxonomic groups (25 - 90\%).
+A small fraction of records was identified as erroneous in the strict sense (4.2\%), the majority of filtered records that were rather unfit for downstream analyses (41.7\%).
+Filters of duplicated information, collection year, basis of record as well as urban areas and coordinates for terrestrial taxa in the sea or marine taxa on land have the greatest effect.
+Automated filtering can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus.
 Our results stress the importance of thorough exploration of the meta-data associated with species records for biodiversity research.
@@
 In particular, the availability of digitized collections from museums and herbaria, and citizen science observations has increased drastically over the last few years.
-As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.3 billion geo-referenced occurrence records for species from across the globe and the tree of life.
+As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.4 billion geo-referenced occurrence records for species from across the globe and the tree of life.
 
-A central challenge to the use of these publicly available species occurrence data in research are erroneous geographic coordinates (Anderson et al.
+A central challenge to the use of these publicly available species occurrence data in research are problematic geographic coordinates, which are either erroneous or unfit for downstream analyses (for instance because they are overly imprecise, Anderson et al.
 2016).
-Errors mostly arise because public databases integrate records collected with different methodologies in different places, at different times; often without centralized curation and only rudimentary meta-data.
-For instance, erroneous coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common (Maldonado et al.
+Problems mostly arise because public databases integrate records collected with different methodologies in different places, at different times; often without centralized curation and only rudimentary meta-data.
+For instance, problematic coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common (Maldonado et al.
 2015; Yesson et al.
-2007) and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in captivity (Zizka, Silvestro, et al.
+2007) and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in captivity (Zizka et al.
 2019).
@@
 Manual data cleaning based on expert knowledge can detect these issues, but it is only feasible on small taxonomic or geographic scales, and it is time-consuming and difficult to reproduce.
-As an alternative, automated flagging methods to identify potentially problematic records have been proposed as a scalable option, as they are able to deal with datasets containing up to millions of records and many different taxa.
-Those methods are usually based on geographic gazetteers (e.g., Chamberlain 2016; Zizka, Silvestro, et al.
+As an alternative, automated filtering methods to identify potentially problematic records have been proposed as a scalable option, as they are able to deal with datasets containing up to millions of records and many different taxa.
+Those methods are usually based on geographic gazetteers (e.g., Chamberlain 2016; Zizka et al.
 2019; Jin and Yang 2020) or on additional data, such as environmental variables (Robertson, Visser, and Hui 2016).
-Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the suitability of publicly available occurrence records for biodiversity research (Zizka, Silvestro, et al.
+Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the suitability of publicly available occurrence records for biodiversity research (Zizka et al.
 2019).
 
-Erroneous records are especially problematic in conservation, where stakes are high.
+Problematic records are especially critical in conservation, where stakes are high.
 Recently proposed methods for automated conservation assessments could support the formal assessment procedures for the global Red List of the International Union for the Conservation of Nature (IUCN) (Dauby et al.
@@
 These methods have been used to propose preliminary global (Stévart et al.
-2019; Zizka, Azevedo, et al.
-2019) and regional (Schmidt et al.
+2019; Zizka et al.
+2020) and regional (Schmidt et al.
 2017; Cosiaux et al.
 2018) Red List assessments.
-However, all metrics but especially EOO, are sensitive to individual records with erroneous coordinates.
+However, all metrics and especially EOO, are sensitive to individual records with problematic coordinates.
 Automated conservation assessments may therefore be biased, particularly if the number of records are few, as it is the case for many tropical species.
 
-While newly proposed methods for automated cleaning of species records hold great promise for biodiversity research, their use across taxonomic groups and datasets remains poorly explored.
+While automated filters hold great promise for biodiversity research, their use across taxonomic groups and datasets remains poorly explored.
 Here, we test the effect of automated filtering of species geographic occurrence records on the number of records available in different groups of animals, fungi, and plants.
-Furthermore, we test the impact of automated filtering procedures for the accuracy of automated conservation assessments compared to full IUCN assessments.
-Specifically, we evaluate a pipeline of 13 automated filters to flag possibly erroneous records by using 1) record meta-data and 2) geographic gazetteers.
+Furthermore, we test the impact of automated filtering procedures for the accuracy of preliminary automated conservation assessments compared to full IUCN assessments.
+Specifically, we evaluate a pipeline of 13 automated filters to flag possibly problematic records by using record meta-data and geographic gazetteers in two categories: 1) erroneous (coordinates, that are likely wrong, irrespective of the downstream analyses, for instance due to data entry errors) and 2) unfit for purpose (coordinates that are not wrong \emph{per se}, but likely unfit for the planned downstream analyses, for instance because they are overly imprecise).
 We address three questions:
@@
 We downloaded occurrence information for all study groups from www.gbif.org using the \texttt{rgbif} v1.4.0 package (Chamberlain 2017) in R (GBIF.org, 2019a-p,2020a,b).
-We downloaded only records with geographic coordinates and limited the study area to latitudes smaller than 33\(^\circ\) N and longitudes smaller than 35\(^\circ\) W and larger than 120\(^\circ\) W reflecting the Neotropics, our main area of expertise (Morrone 2014).
+We downloaded GBIF-interpreted data including only records with geographic coordinates and limited the study area to a rectangle between 90\(^\circ\)S - 33\(^\circ\) N and 35\(^\circ\) W - 120\(^\circ\) W reflecting the Neotropics (Morrone 2014), our main area of expertise.
 The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, Entomobryidae, \emph{Gaylussacia}, Iridaceae, Neanuridae, and \emph{Tillandsia} for which we only obtained the Neotropical occurrences.
+We consider GBIF data generally of high quality and use them as study case because GBIF is the largest, most widely-used and taxonomically most comprehensive data source for species occurrence records, but many more exist (e.g., \url{https://bien.nceas.ucsb.edu/bien/}, www.fishbase.de or Guedes et al.
+2018).
+The variety of existing databases is also the reason why we download data without accounting for GBIF-internal quality testing (``Geospatial issues''), since quality control mechanisms might differ among databases and sometimes no quality control might be done (for instance in \emph{de novo} databases compiled by groups of researchers).
 
@@
 
-We followed the cleaning pipeline outlined by Zizka, Silvestro, et al.
-(2019) and first filtered the raw data downloaded from GBIF (``raw'', hereafter) using meta-data for those records for which they were available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99.
+We followed the cleaning pipeline outlined by Zizka et al.
+(2019) and first filtered the raw data downloaded from GBIF (``raw'', hereafter) using meta-data for those records for which they were available (although this is often not the case, Peterson et al.
+2018), removing: 1) records with a coordinate precision below 100 km (as this represents the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99.
 Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e., test for duplicated records).
-In a second step, we used the \texttt{clean\_coordinates} function of the \texttt{CoordinateCleaner\ v2.0-11} package (Zizka, Silvestro, et al.
+In a second step, we used the \texttt{clean\_coordinates} function of the \texttt{CoordinateCleaner\ v2.0-11} package (Zizka et al.
 2019) with default options to flag errors that are common to biological data sets (``filtered'', hereafter).
 These include: coordinates in the sea for terrestrial taxa and on land for marine taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions.
-See Table 2 for a summary of all flagging tests we used.
+See Table 2 for a summary of all filters we used and their classification into ``erroneous'' and ``unfit''.
 
@@
 We first generated species richness maps using 100x100 km grid cells for the raw and filtered datasets respectively, using the package \texttt{speciesgeocodeR\ v2.0-10} (Töpel et al.
-2016).
+2017).
 We then performed an automated conservation assessment for all study groups based on both datasets using the \texttt{ConR\ v1.2.4} package (Dauby et al.
@@
 
-Each author provided an informed comment on the performance of the raw and cleaned datasets, respectively concerning the number of flagged records and the accuracy of the overall species richness maps and we compared the agreement between automated conservation assessments based on raw and filtered occurrences with the global IUCN Red List (www.iucn.org).
+Each author provided an informed comment on the performance of the raw and cleaned datasets, concerning the number of removed records and the accuracy of the overall species richness maps.
+We then compared the agreement between automated conservation assessments based on raw and filtered occurrences with the global IUCN Red List for those taxa where IUCN assessments were available (www.iucn.org).
 
 We did all analyses in the R computing environment (R Core Team 2019), using standard libraries for data handling and visualization (Wickham 2018; Garnier 2018; Ooms 2014, 2019; Hijmans 2019).
-All scripts are available from a zenodo repository (\textbf{DOI}).
+All scripts are available from a zenodo repository (\url{doi:10.5281/zenodo.3695102}).
 
@@
 
-We retrieved a total of 218,899 species occurrence records, with a median of 2,844.5 records per study group and 10 records per species (Table 3, Appendix 1).
+We retrieved a total of 218,899 species occurrence records, with a median of 2,844 records per study group and 10 records per species (Table 3, Appendix 1).
 We obtained most records for Dipsadidae (64,249) and fewest for \emph{Thozetella} (51).
@@
 
-Our automated tests flagged a total of 97,018 records (Fig.
-\ref{fig:total}), with a median of 45\% per group.
+Our automated tests filtered a total of 97,004 records (Fig.
+\ref{fig:total}, erroneous: 9,254, unfit: 91,298), with a median of 45\% per group (erroneous: 0.3\%, unfit: 37.4\%).
 Overall, the most important test was for duplicated records (on average 35.5\% per taxonomic group).
-The filtering steps based on record meta-data that flagged the largest number of records were the basis of records (5.9\%) and the collection year (3.4\%).
+The filtering steps based on record meta-data that filtered the largest number of records were the basis of records (5.9\%) and the collection year (3.4\%).
 The most important automated tests were for urban area (8.6\%) and the occurrence from records of terrestrial taxa in the sea and marine taxa on land (4.3\%, see Table 3 and Appendix 1 in the electronic supplement for further details and the absolute numbers).
-Only few records were flagged by the coordinate precision, zero coordinates and biodiversity institution tests (Fig.
+Only a few records were filtered by the coordinate precision, zero coordinates and biodiversity institution tests (Fig.
 \ref{fig:split}).
 
-\emph{Entomobryidae}, \emph{Diogenidae}, and \emph{Neanuridae} had the highest fraction of flagged records (Table 3).
+Entomobryidae, Diogenidae, and Neanuridae had the highest fraction of filtered records (Table 3).
 In general, the different filters we tested were of similar importance for different study groups.
-There were few outstanding exceptions, including the particularly high proportions of records flagged on the ``basis of record test'' for \emph{Tityus} (7.0\%), Dipsadidae (5.6\%), \emph{Prosthechea} (5.0\%) and \emph{Tillandsia} (4.9\%), by the collection year for Dipsadidae (11.3\%), by the taxonomic identification level for \emph{Tityus} (1.6\%), by the capital coordinates for \emph{Oocephalus} (6.1\%) and \emph{Gaylussacia} (3.2\%), by the seas/land test for Diogenidae and \emph{Thozetella}, and by the urban areas test for \emph{Oocephalus} (13.3\%) and Iridaceae (12.3\%).
-Furthermore, Entomobryidae differed considerably from all other study taxa with exceptionally high numbers of records flagged by the ``basis of record,''level of identification' and urban areas" tests.
+There were few outstanding exceptions, including the particularly high proportions of records filtered on the ``basis of record test'' for \emph{Tityus} (7.0\%), Dipsadidae (5.6\%), \emph{Prosthechea} (5.0\%) and \emph{Tillandsia} (4.9\%), by the collection year for Dipsadidae (11.3\%), by the taxonomic identification level for \emph{Tityus} (1.6\%), by the capital coordinates for \emph{Oocephalus} (6.1\%) and \emph{Gaylussacia} (3.2\%), by the seas/land test for Diogenidae and \emph{Thozetella}, and by the urban areas test for \emph{Oocephalus} (13.3\%) and Iridaceae (12.3\%).
+Furthermore, Entomobryidae differed considerably from all other study taxa with exceptionally high numbers of records filtered by the ``basis of record,''level of identification' and urban areas" tests.
 
-Geographically, the records flagged by the ``basis of record'' and ``individual count'' tests were concentrated in Central America and southern North America, and a relatively high number of records flagged due to their proximity to the centroids of political entities were located on Caribbean islands (Fig.
+Geographically, the records filtered by the ``basis of record'' and ``individual count'' tests were concentrated in Central America and southern North America, and a relatively high number of records filtered due to their proximity to the centroids of political entities were located on Caribbean islands (Fig.
 \ref{fig:split}).
@@
 
-Automated flagging based on meta-data and automatic tests flagged on average 45\% of the records per taxonomic group.
-The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (Question 1).
-The importance of different tests was similar across taxonomic groups, with exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (Question 2).
+Automated flagging based on meta-data and automatic tests filtered on average 45\% of the records per taxonomic group; 25.9\% - 90.3\% as ``unfit'' and 0\% - 44.3\% as ``erroneous''.
+The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (\textbf{Question 1}).
+The importance of different tests was similar across taxonomic groups, with exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (\textbf{Question 2}).
 The results for species richness were similar between the raw and filtered data with some improvements by the filters.
-We found little impact of filtering on the accuracy of the automated conservation assessments (Question 3).
+We found little impact of filtering on the accuracy of the automated conservation assessments (\textbf{Question 3}).
 
@@
 
+The aim of automated filtering is to identify possibly problematic records that are unsuitable for particular downstream analyses.
+While those records filtered as ``erroneous'' will likely cause problems for most biodiversity research, those filtered as ``unfit'' might have varying impact, depending on the type and spatial resolution of the downstream analyses, ranging from an unnecessary computational burden which can be a bottle neck for large-scale analyses (i.e.~duplicates, Antonelli et al.
+2018) over increasing uncertainty (due to low precision) to compromising results.
+For instance, records assigned to country centroids might be acceptable for inter-continental comparisons, but likely are erroneous for species distribution modelling on a local scale.
+The importance of each test and the linked thresholds must be judged based on the specific downstream analyses.
+As our results show, adapting automated tests to the geographic study area or the taxonomic study group can be warranted.
+For instance, the high number of records flagged for centroids on the Lesser Antilles (Fig.
+\ref{fig:split}) might be overly strict (\url{https://data-blog.gbif.org/post/country-centroids/}), although we chose a conservative distance for the Political centroid test (1 km).
+
 Several reasons may explain the high proportion of records flagged as duplicates.
@@
 Third, low coordinate precision, for instance based on automated geo-referencing from locality descriptions, may lump records from nearby localities.
-Duplicated records do not represent errors \emph{per se}, but an unnecessary computational burden, which is often a bottleneck for analyses of datasets with millions of records.
-
-The aim of automated filtering is to identify possibly problematic records that are unsuitable for particular downstream analyses.
-Hence, our tests may identify true errors (e.g., wrong coordinates such as those with exclusively zeros, equal latitude and longitude and records of terrestrial species in the sea or marine species on land) as well as records that are not necessarily erroneous, but have a low precision (e.g., collection year, coordinate precision, centroids of political units) or are redundant for certain analyses (duplicates).
-The importance of each test and the linked thresholds must be judged based on the specific downstream analyses.
-For instance, records assigned to country centroids might be acceptable for inter-continental comparisons, but can be considered erroneous for species distribution modelling on a local scale.
-As our results show, adapting automated tests to the geographic study area or the taxonomic study group can be warranted.
-For instance, the high number of records flagged for centroids on the Lesser Antilles (Fig.
-\ref{fig:split}) might be overly strict, although we chose a conservative distance for the Political centroid test (1 km).
+Fourth, different data contributors might add the same record to GBIF, if their sources overlap, as can for instance be the case for the Barcode of Life and Plazi databases.
 
@@
 
-The number of records flagged by individual tests was similar across study groups, suggesting that similar error types might be relevant for collections of plants and animals.
-Therefore, the same tests can be used across taxonomic groups.
-Some notable exceptions stress the need to adapt each test to the taxonomic study group to balance data quality and data availability.
-The high fraction of records flagged by the ``basis of record test'' for \emph{Tityus}, Dipsadidae, \emph{Prosthechea} and \emph{Tillandsia}, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups.
+The number of records flagged by individual tests was similar across study groups, suggesting that similar problems might be relevant for collections of plants and animals.
+Therefore, the same filters can be used across taxonomic groups.
+Some notable exceptions stress the need to adapt each filter to the taxonomic study group to balance data quality and data availability.
+The high fraction of records filtered by the ``basis of record filter'' for \emph{Tityus}, Dipsadidae, \emph{Prosthechea} and \emph{Tillandsia}, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups.
 The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection effort already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700.
 The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700.
-The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae, and \emph{Oocephalus} might be related to horticulture since at least some species in those groups are commonly cultivated as ornamentals.
+The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae and \emph{Oocephalus} might be related to horticulture since at least some species in those groups are commonly cultivated as ornamentals.
 
@@
 For instance, in Entomobryidae, 2,004 samples were marked as material samples, and therefore removed by our global filter retaining only specimen and observation data, which in this case was overly strict.
-It should be noted, that, while in this study we focused on meta-data and geographic filtering, taxonomic cleaning---the resolution of synonymies and identification of accepted names---is another important part of data curation, but depends on taxon-specific taxonomic backbones and synonymy lists which are not readily available for many groups.
+
+The filters we included in this study address a set of important but relatively easy to identify problems.
+In fact, the internal quality control of GBIF does flag some of the problems we tested for (i.e., zero coordinates, equal lat/lon) while others might be implemented in the near future (country centroids, \url{https://data-blog.gbif.org/post/country-centroids/}, \url{https://data-blog.gbif.org/post/country-centroids/}).
+While this internal quality is very helpful, we see a huge potential to overcome issues with data quality in a user-feedback system allowing users to provide expert assessments, i.e.~a meta-annotation of records being challenged (and why).
+Such a system would not need to change the original data and could include multiple levels to account for differing opinions.
+
+As a next step for automated filtering, it is possible to detect additional problematic records with the help of external species distribution information, which is available for some taxa for example from www.iucn.org (vertebrates) or \url{https://wcsp.science.kew.org/} (selected seed plant families).
+Furthermore, while in this study we focused on meta-data and geographic filtering, taxonomic cleaning---the resolution of synonymies and identification of accepted names---is another important part of data curation, but depends on taxon-specific taxonomic backbones and synonymy lists which are not readily available for many groups and often are contradictory within individual taxa.
 
@@
 The accuracy of the automated conservation assessment was in the same range as found by previous studies (Nic Lughadha et al.
-2019; Zizka, Azevedo, et al.
-2019).
+2019; Zizka et al.
+2020).
 The similar accuracy of the raw and filtered dataset for the automated conservation assessment was surprising, in particular given the EOO and AOO reduction observed in the filtered dataset (Table 4) and the impact of errors on spatial analyses observed in previous studies (Gueta and Carmel 2016).
 The robustness of the automated assessment was likely due to the fact that the EOO for most species was large, even after the considerable reduction caused by filtering.
-This might be caused by the structure of our comparison, which only included species that were evaluated by the IUCN Red List (and not considered as Data Deficient) and at the same time had occurrences recorded in GBIF.
-Those inclusion criteria are likely to have, biased the datasets towards species with large ranges, since generally more data for them are available.
-The robustness of automated conservation assessments to data quality is encouraging, although these methods are only an approximation, especially for species with few collection records (Rivers et al.
+This might be caused by the structure of our comparison, which only included species that were evaluated by the IUCN Red List (and not considered as \emph{Data Deficient}) and at the same time had occurrences recorded in GBIF.
+Those inclusion criteria are likely to have biased the datasets towards species with large ranges, since generally more data for them are available.
+The robustness of automated conservation assessments to data quality is encouraging, although these methods are only an approximation (and not replacements) of full IUCN Red List assessments, especially for species with few collection records (Rivers et al.
 2011).
@@
 
-Our results stress the importance of (meta-)data exploration prior to most biodiversity analyses.
-Automated flagging can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the specific taxonomic groups and geographic area of interest.
-The putative errors we encountered point to the importance to train researchers and students to curate species occurrence datasets and to allow users to provide feedback for particular records, thus contributing to overall accuracy and precision of public biodiversity databases.
+While our results suggest that large-scale species richness patterns and automated conservation assessments are largely robust to problematic occurrence records, they also stress the importance of (meta-)data exploration prior to most biodiversity analyses.
+Automated filtering can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the specific taxonomic groups and geographic area of interest.
+The putative problems we encountered point to the importance to train researchers and students to curate species occurrence datasets and to allow users to provide data aggregators such as GBIF with feedback for particular records, thus contributing to overall accuracy and precision of public biodiversity databases.
 
@@
 
-This study enrolled participants of the workshop ``Biodiversity data: from field to yield'' led by Alexandre Antonelli and Alexander Zizka through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI.
+This study enrolled participants of the workshop ``Biodiversity data: from field to yield'' led by Alice Calvente, Fernanda Carvalho, Alexander Zizka, and Alexandre Antonelli through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI.
 We thank the Pró-reitoria de Pesquisa and the Pró-reitoria de Pós-graduação of UFRN for financial support (edital 02/2016 - internacionalização).
@@
 AA is supported by the Swedish Research Council, the Knut and Alice Wallenberg Foundation, the Swedish Foundation for Strategic Research and the Royal Botanic Gardens, Kew.
+FS was financed by the Coordenação de Aperfeiçoamento de Pessoal de Nível Superior - Brasil (CAPES) - Finance Code 001 and Fundação de Amparo à Pesquisa do estado de São Paulo (FAPESP) (FAPESP, process 2015/20215-7).
 
@@
 \item
-  Appendix 1 - Table S1, reporting the absolute number of flagged records per taxonomic group and test
+  Appendix 1 - Absolute number of flagged records per taxonomic group and test
 \item
-  Appendix 2 - Taxon specific richness maps and comments.
+  Appendix 2 - Taxon specific richness maps and comments
 \item
-  Appendix 3 - The full results of the conservation assessment
+  Appendix 3 - Full results of the conservation assessment
 \end{itemize}
@@
 
-\begin{landscape}\begin{table}
+\begin{table}
 
@@
 \fontsize{11}{13}\selectfont
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}ll}
+\begin{tabular}[t]{>{\raggedright\arraybackslash}p{1cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{1.5cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{3cm}}
 \toprule
-Taxon & Taxon rank & Realm & Common name & 'Phylum' & Order & Family\\
+ & Taxon & Taxon rank & Realm & Common name & 'Phylum' & Family\\
 \midrule
-Diogenidae & Family & Marine & Hermit crabs & Arthropoda & Decapoda & Diogenidae\\
-Entomobryidae & Family & Terrestrial & Springtails & Arthropoda & Poduromorpha & Entomobryidae\\
-Neanuridae & Family & Terrestrial & Springtails & Arthropoda & Poduromorpha & Neanuridae\\
-\em{Tityus} & Genus & Terrestrial & Scorpions & Arthropoda & Scorpiones & Buthidae\\
-Arhynchobatidae & Family & Marine & Skates & Chordata & Rajiformes & Arhynchobatidae\\
+\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/diogenidae.png} & Diogenidae & Family & Marine & Hermit crabs & Arthropoda & Diogenidae\\
+\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/entomobryidae.png} & Entomobryidae & Family & Terrestrial & Springtails & Arthropoda & Entomobryidae\\
+\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/neanuridae.png} & Neanuridae & Family & Terrestrial & Springtails & Arthropoda & Neanuridae\\
+\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/tityus.png} & \em{Tityus} & Genus & Terrestrial & Scorpions & Arthropoda & Buthidae\\
+\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/arhynchobatidae.png} & Arhynchobatidae & Family & Marine & Skates & Chordata & Arhynchobatidae\\
 \addlinespace
-Dipsadidae & Family & Terrestrial & Snakes & Chordata & Squamata & Dipsadidae\\
-\em{Harengula} & Genus & Marine & Herrings & Chordata & Clupeiformes & Clupeidae\\
+\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/dipsadidae.png} & Dipsadidae & Family & Terrestrial & Snakes & Chordata & Dipsadidae\\
+\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/harengula.png} & \em{Harengula} & Genus & Marine & Herrings & Chordata & Clupeidae\\
 \hline
-\em{Thozetella} & Genus & Terrestrial & Sac fungi & Ascomycota & Chaetosphaeriales & Chaetosphaeriaceae\\
+\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/thozetella.png} & \em{Thozetella} & Genus & Terrestrial & Sac fungi & Ascomycota & Chaetosphaeriaceae\\
 \hline
-\em{Conchocarpus} & Genus & Terrestrial & NA & Angiosperms & Sapindales & Rutaceae\\
-\em{Gaylussacia} & Genus & Terrestrial & Huckleberries & Angiosperms & Ericales & Ericaceae\\
+\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/conchocarpus.png} & \em{Conchocarpus} & Genus & Terrestrial & NA & Angiosperms & Rutaceae\\
+\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/gaylussacia.png} & \em{Gaylussacia} & Genus & Terrestrial & Huckleberries & Angiosperms & Ericaceae\\
 \addlinespace
-\em{Harpalyce} & Genus & Terrestrial & NA & Angiosperms & Fabales & Fabaceae\\
-Iridaceae & Family & Terrestrial & NA & Angiosperms & Asparagales & Iridaceae\\
-\em{Lepismium} & Genus & Terrestrial & Cacti & Angiosperms & Caryophyllales & Cactaceae\\
-\em{Oocephalus} & Genus & Terrestrial & NA & Angiosperms & Lamiales & Lamiaceae\\
-\em{Pilosocereus} & Genus & Terrestrial & NA & Angiosperms & Caryophyllales & Cactaceae\\
+\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/harpalyce.png} & \em{Harpalyce} & Genus & Terrestrial & NA & Angiosperms & Fabaceae\\
+\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/iridaceae.png} & Iridaceae & Family & Terrestrial & NA & Angiosperms & Iridaceae\\
+\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/lepismium.png} & \em{Lepismium} & Genus & Terrestrial & Cacti & Angiosperms & Cactaceae\\
+\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/oocephalus.png} & \em{Oocephalus} & Genus & Terrestrial & NA & Angiosperms & Lamiaceae\\
+\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/pilosocereus.png} & \em{Pilosocereus} & Genus & Terrestrial & Cacti & Angiosperms & Cactaceae\\
 \addlinespace
-\em{Prosthechea} & Genus & Terrestrial & Orchids & Angiosperms & Asparagales & Orchidaceae\\
-\em{Tillandsia} & Genus & Terrestrial & Bromeliads & Angiosperms & Poales & Bromeliaceae\\
-\em{Tocoyena} & Genus & Terrestrial & NA & Angiosperms & Gentianales & Rubiaceae\\
+\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/prosthechea.png} & \em{Prosthechea} & Genus & Terrestrial & Orchids & Angiosperms & Orchidaceae\\
+\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/tillandsia.png} & \em{Tillandsia} & Genus & Terrestrial & Bromeliads & Angiosperms & Bromeliaceae\\
+\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/tocoyena.png} & \em{Tocoyena} & Genus & Terrestrial & NA & Angiosperms & Rubiaceae\\
 \bottomrule
@@
 \end{table}
-\end{landscape}
 
@@
 \fontsize{9}{11}\selectfont
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{10cm}}
+\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{1cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{9cm}}
 \toprule
-Test & Type & Rationale\\
+Test & Type & Basis & Rationale\\
 \midrule
-\rowcolor{gray!6}  Basis of record & Meta-data & Records might be unsuitable or unreliable for certain analyses dependent on their source, e.g. "fossil" or "unknown"\\
-Collection year & Meta-data & Coordinates from old records are more likely to be imprecise or erroneous coordinates since they are derived from  geo-referencing based on the locality description.
+\rowcolor{gray!6}  Biodiversity institutions & Error & Gazetteer-based & Records may have coordinates at the location of biodiversity institutions, e.g.
+because they were erroneously entered with the physical location of the specimen or because they represent individuals from captivity or horticulture\\
+Equal lat/lon & Error & Gazetteer-based & Coordinates with equal latitude and longitude are usually indicative of data entry errors\\
+\rowcolor{gray!6}  Sea & Error & Gazetteer-based & Coordinates from terrestrial organisms in the sea are usually indicative of data entry errors, e.g. swapped latitude and longitude\\
+Zeros & Error & Gazetteer-based & Coordinates with plain zeros are often indicative of data entry errors\\
+\rowcolor{gray!6}  Capitals & Unfit & Gazetteer-based & Records may be assigned to the coordinates of country capitals based on a vague locality description\\
+\addlinespace
+Duplicates & Unfit & Gazetteer-based & Duplicated records may add unnecessary computational burden, in particular for large scale biodiversity analyses and distribution modelling for many species\\
+\rowcolor{gray!6}  Political centroids & Unfit & Gazetteer-based & Records may be assigned to the coordinates of the centroids of political entities based on a vague locality description\\
+Urban areas & Unfit & Gazetteer-based & Records from urban areas are not necessarily errors, but often represent imprecise records automatically geo-referenced from vague locality descriptions or old records from different land-use types\\
+\rowcolor{gray!6}  Basis of record & Unfit & Meta-data & Records might be unsuitable or unreliable for certain analyses dependent on their source, e.g. "fossil" or "unknown"\\
+Collection year & Unfit & Meta-data & Coordinates from old records are more likely to be imprecise or erroneous coordinates since they are derived from  geo-referencing based on the locality description.
 This is more problematic for older records, since names or borders of places may change\\
-\rowcolor{gray!6}  Coordinate precision & Meta-data & Records may be unsuitable for a study if their precision is lower than the study analysis scale\\
-Identification level & Meta-data & Records may be unsuitable if they are not identified to species level.\\
-\rowcolor{gray!6}  Individual count & Meta-data & Records may be unsuitable if the number of recorded individuals is 0 (record of absence) or if the count is to high, as this is often related to records from barcoding or indicative of data entry problems.\\
 \addlinespace
-Biodiversity institutions & Gazetteer-based & Records may have coordinates at the location of biodiversity institutions, e.g.
-because they were erroneously entered with the physical location of the specimen or because they represent individuals from captivity or horticulture\\
-\rowcolor{gray!6}  Duplicates & Gazetteer-based & Duplicated records may add unnecessary computational burden to analyses\\
-Equal lat/lon & Gazetteer-based & Coordinates with equal latitude and longitude are usually indicative of data entry errors\\
-\rowcolor{gray!6}  Capitals & Gazetteer-based & Records may be assigned to the coordinates of country capitals based on a vague locality description\\
-Political centroids & Gazetteer-based & Records may be assigned to the coordinates of the centroids of political entities based on a vague locality description\\
-\addlinespace
-\rowcolor{gray!6}  Sea & Gazetteer-based & Coordinates from terrestrial organisms in the sea are usually indicative of data entry errors, e.g. swapped latitude and longitude\\
-Urban areas & Gazetteer-based & Records from urban areas are not necessarily errors, but often represent imprecise records automatically geo-referenced from vague locality descriptions or old records from different land-use types\\
-\rowcolor{gray!6}  Zeros & Gazetteer-based & Coordinates with plain zeros are often indicative of data entry errors\\
+\rowcolor{gray!6}  Coordinate precision & Unfit & Meta-data & Records may be unsuitable for a study if their precision is lower than the study analysis scale\\
+Identification level & Unfit & Meta-data & Records may be unsuitable if they are not identified to species level.\\
+\rowcolor{gray!6}  Individual count & Unfit & Meta-data & Records may be unsuitable if the number of recorded individuals is 0 (record of absence) or if the count is too high, as this is often related to records from barcoding or indicative of data entry problems.\\
 \bottomrule
@@
 \centering
-\fontsize{9}{11}\selectfont
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}r}
+\fontsize{8}{10}\selectfont
+\begin{tabular}[t]{>{\raggedright\arraybackslash}p{1.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{1.1cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{1.1cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}}
 \toprule
-\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Summary} & \multicolumn{5}{c}{Meta data filters} & \multicolumn{7}{c}{Gazetteer-based filters} \\
-\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-9} \cmidrule(l{3pt}r{3pt}){10-16}
-Taxon & Total records & Records flagged & Fraction flagged [\%] & Basis of record [\%] & Collection year [\%] & Coordinate precision [\%] & Id-level [\%] & Individual count [\%] & Biodiversity Institutions [\%] & Capitals [\%] & Duplicates [\%] & Political centroids [\%] & Sea/land area [\%] & Urban areas [\%] & Zeros [\%]\\
+\multicolumn{1}{c}{ } & \multicolumn{4}{c}{Summary} & \multicolumn{3}{c}{Errors} & \multicolumn{9}{c}{Unfit} \\
+\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-8} \cmidrule(l{3pt}r{3pt}){9-17}
+Taxon & Total records & Fraction flagged [\%] & Fraction error [\%] & Fraction unfit [\%] & Biodiversity Institutions [\%] & Sea/land area [\%] & Zeros [\%] & Capitals [\%] & Duplicates [\%] & Political centroids [\%] & Urban areas [\%] & Basis of record [\%] & Collection year [\%] & Coordinate precision [\%] & Id-level [\%] & Individual count [\%]\\
 \midrule
-Diogenidae & 13,840 & 9,508 & 68.7 & 1.7 & 2.5 & 0.0 & 0.0 & 0.0 & 0.0 & 0.7 & 33.8 & 0.2 & 44.3 & 1.3 & 0.0\\
-Entomobryidae & 2,767 & 2,498 & 90.3 & 72.9 & 2.0 & 0.0 & 72.1 & 0.0 & 0.1 & 0.1 & 85.5 & 0.0 & 0.0 & 70.1 & 0.0\\
-Neanuridae & 689 & 461 & 66.9 & 2.9 & 1.3 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 62.4 & 0.0 & 0.0 & 2.0 & 0.0\\
-\em{Tityus} & 1,018 & 562 & 55.2 & 7.0 & 0.4 & 1.8 & 1.6 & 0.0 & 0.5 & 1.2 & 43.5 & 0.1 & 0.0 & 6.9 & 0.0\\
-Arhynchobatidae & 14,633 & 5,635 & 38.5 & 1.7 & 1.3 & 0.0 & 0.9 & 0.0 & 0.0 & 0.0 & 35.4 & 0.0 & 3.8 & 1.9 & 0.0\\
+Diogenidae & 13,840 & 68.7 & 44.3 & 38.2 & 0.0 & 44.3 & 0.0 & 0.7 & 33.8 & 0.2 & 1.3 & 1.7 & 2.5 & 0.0 & 0.0 & 0.0\\
+Entomobryidae & 2,767 & 90.3 & 0.1 & 90.3 & 0.1 & 0.0 & 0.0 & 0.1 & 85.5 & 0.0 & 70.1 & 72.9 & 2.0 & 0.0 & 72.1 & 0.0\\
+Neanuridae & 689 & 66.9 & 0.0 & 66.9 & 0.0 & 0.0 & 0.0 & 0.0 & 62.4 & 0.0 & 2.0 & 2.9 & 1.3 & 0.0 & 0.0 & 0.0\\
+\em{Tityus} & 1,018 & 55.2 & 0.5 & 54.9 & 0.5 & 0.0 & 0.0 & 1.2 & 43.5 & 0.1 & 6.9 & 7.0 & 0.4 & 1.8 & 1.6 & 0.0\\
+Arhynchobatidae & 14,633 & 38.5 & 3.8 & 37.4 & 0.0 & 3.8 & 0.0 & 0.0 & 35.4 & 0.0 & 1.9 & 1.7 & 1.3 & 0.0 & 0.9 & 0.0\\
 \addlinespace
-Dipsadidae & 64,249 & 37,058 & 57.7 & 5.6 & 11.3 & 0.8 & 0.0 & 0.1 & 0.3 & 1.8 & 46.3 & 0.4 & 0.0 & 8.5 & 0.0\\
-\em{Harengula} & 36,697 & 11,384 & 31.0 & 1.0 & 0.4 & 0.0 & 0.3 & 0.0 & 0.0 & 0.2 & 27.0 & 0.1 & 5.5 & 0.2 & 0.0\\
+Dipsadidae & 64,249 & 57.7 & 0.3 & 57.6 & 0.3 & 0.0 & 0.0 & 1.8 & 46.3 & 0.4 & 8.5 & 5.6 & 11.3 & 0.8 & 0.0 & 0.1\\
+\em{Harengula} & 36,697 & 31.0 & 5.5 & 27.8 & 0.0 & 5.5 & 0.0 & 0.2 & 27.0 & 0.1 & 0.2 & 1.0 & 0.4 & 0.0 & 0.3 & 0.0\\
 \hline
-\em{Thozetella} & 51 & 18 & 35.3 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 27.5 & 0.0 & 23.5 & 2.0 & 0.0\\
+\em{Thozetella} & 51 & 35.3 & 23.5 & 29.4 & 0.0 & 23.5 & 0.0 & 0.0 & 27.5 & 0.0 & 2.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0\\
 \hline
-\em{Conchocarpus} & 1,551 & 670 & 43.2 & 0.5 & 1.9 & 0.1 & 0.0 & 0.0 & 0.1 & 0.0 & 39.6 & 0.9 & 0.4 & 2.3 & 0.0\\
-\em{Gaylussacia} & 3,998 & 1,887 & 47.2 & 0.7 & 4.4 & 0.6 & 0.0 & 0.0 & 0.1 & 3.2 & 41.8 & 1.1 & 0.1 & 5.2 & 0.0\\
+\em{Conchocarpus} & 1,551 & 43.2 & 0.5 & 42.9 & 0.1 & 0.4 & 0.0 & 0.0 & 39.6 & 0.9 & 2.3 & 0.5 & 1.9 & 0.1 & 0.0 & 0.0\\
+\em{Gaylussacia} & 3,998 & 47.2 & 0.1 & 47.1 & 0.1 & 0.1 & 0.0 & 3.2 & 41.8 & 1.1 & 5.2 & 0.7 & 4.4 & 0.6 & 0.0 & 0.0\\
 \addlinespace
-\em{Harpalyce} & 870 & 288 & 33.1 & 0.5 & 5.5 & 0.7 & 0.0 & 0.9 & 0.0 & 1.0 & 26.0 & 1.3 & 0.0 & 3.8 & 0.0\\
-Iridaceae & 23,127 & 7,774 & 33.6 & 0.9 & 4.7 & 0.1 & 0.0 & 1.3 & 0.4 & 1.0 & 17.1 & 0.4 & 0.1 & 12.3 & 0.0\\
-\em{Lepismium} & 825 & 245 & 29.7 & 0.0 & 2.1 & 0.0 & 0.0 & 0.0 & 0.0 & 0.1 & 21.9 & 0.1 & 0.0 & 7.8 & 0.0\\
-\em{Oocephalus} & 883 & 435 & 49.3 & 0.0 & 0.7 & 0.3 & 0.0 & 0.1 & 0.0 & 6.1 & 41.9 & 0.8 & 0.0 & 13.3 & 0.0\\
-\em{Pilosocereus} & 1,940 & 501 & 25.8 & 1.8 & 7.0 & 0.0 & 0.0 & 0.9 & 0.2 & 0.5 & 16.8 & 0.5 & 0.0 & 2.1 & 0.0\\
+\em{Harpalyce} & 870 & 33.1 & 0.0 & 33.1 & 0.0 & 0.0 & 0.0 & 1.0 & 26.0 & 1.3 & 3.8 & 0.5 & 5.5 & 0.7 & 0.0 & 0.9\\
+Iridaceae & 23,127 & 33.6 & 0.5 & 33.5 & 0.4 & 0.1 & 0.0 & 1.0 & 17.1 & 0.4 & 12.3 & 0.9 & 4.7 & 0.1 & 0.0 & 1.3\\
+\em{Lepismium} & 825 & 29.7 & 0.0 & 29.7 & 0.0 & 0.0 & 0.0 & 0.1 & 21.9 & 0.1 & 7.8 & 0.0 & 2.1 & 0.0 & 0.0 & 0.0\\
+\em{Oocephalus} & 883 & 49.3 & 0.0 & 49.3 & 0.0 & 0.0 & 0.0 & 6.1 & 41.9 & 0.8 & 13.3 & 0.0 & 0.7 & 0.3 & 0.0 & 0.1\\
+\em{Pilosocereus} & 1,940 & 25.9 & 0.2 & 25.9 & 0.2 & 0.0 & 0.0 & 0.5 & 16.8 & 0.5 & 2.1 & 1.8 & 7.0 & 0.0 & 0.0 & 0.9\\
 \addlinespace
-\em{Prosthechea} & 6,617 & 2,088 & 31.6 & 5.0 & 8.3 & 0.1 & 0.0 & 0.2 & 0.0 & 0.4 & 19.6 & 1.7 & 0.0 & 0.9 & 0.1\\
-\em{Tillandsia} & 42,222 & 14,908 & 35.3 & 4.9 & 5.1 & 0.1 & 0.0 & 1.0 & 0.3 & 0.7 & 19.8 & 0.7 & 0.0 & 9.2 & 0.0\\
-\em{Tocoyena} & 2,922 & 1,098 & 37.6 & 0.1 & 1.9 & 0.2 & 0.0 & 0.5 & 0.0 & 0.8 & 32.3 & 0.8 & 0.2 & 5.0 & 0.0\\
+\em{Prosthechea} & 6,617 & 31.5 & 0.1 & 31.5 & 0.0 & 0.0 & 0.1 & 0.4 & 19.6 & 1.7 & 0.9 & 5.0 & 8.3 & 0.1 & 0.0 & 0.2\\
+\em{Tillandsia} & 42,222 & 35.3 & 0.4 & 35.2 & 0.3 & 0.0 & 0.0 & 0.7 & 19.8 & 0.7 & 9.2 & 4.9 & 5.1 & 0.1 & 0.0 & 1.0\\
+\em{Tocoyena} & 2,922 & 37.6 & 0.3 & 37.4 & 0.0 & 0.2 & 0.0 & 0.8 & 32.3 & 0.8 & 5.0 & 0.1 & 1.9 & 0.2 & 0.0 & 0.5\\
 \hline
-Total & 218,899 & 97,018 & 44.3 & 4.2 & 5.6 & 0.3 & 1.0 & 0.4 & 0.2 & 1.0 & 32.3 & 0.4 & 4.0 & 7.1 & 0.0\\
+Total & 218,899 & 44.3 & 4.2 & 41.7 & 0.2 & 4.0 & 0.0 & 1.0 & 32.3 & 0.4 & 7.1 & 4.2 & 5.6 & 0.3 & 1.0 & 0.4\\
 \bottomrule
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{./figures_tables/Figure_study_groups} \caption{Examples of study taxa.
-\textbf{A)} Pilosocereus pusillibaccatus, picture by Pamela Lavor; \textbf{B)} Conchocarpus macrocarpus, Juliana El-Ottra; \textbf{C)} Tillandsia recurva, EC Tomaz; \textbf{D)} \textit{Oxyrhopus guibei} (Dipsadidae), Filipe C Serrano; \textbf{E)} Neanuridae (\textit{Aethiopella ricardoi}), Raiane Vital da Paz; \textbf{F)} \textit{Tocoyena formosa}, Fernanda GL Moreira; \textbf{G)} \textit{Harengula jaguana}, Thais Ferreira-Araujo; \textbf{H)} \textit{Gaylussacia decipiens}, Luiz Menini Neto; \textbf{I)} \textit{Ooocephalus foliosus}, Arthur Soares; \textbf{J)} \textit{Tityus carvalhoi}, Renata C Santos-Costa; \textbf{K)} \textit{Prostheca vespa}, Tiago Veira}\label{fig:species}
+\includegraphics[width=0.9\linewidth]{./figures_tables/Figure_study_groups_reduced} \caption{Examples for taxa included in this study.
+\textbf{A)} \textit{Pilosocereus pusillibaccatus} (\textit{Pilosocereus}), \textbf{B)} \textit{Conchocarpus macrocarpus} (\textit{Conchocarpus}); \textbf{C)} \textit{Tillandsia recurva} (\textit{Tillandsia}); \textbf{D)} \textit{Oxyrhopus guibei} (Dipsadidae); \textbf{E)} \textit{Aethiopella ricardoi} (Neanuridae); \textbf{F)} \textit{Tocoyena formosa} (\textit{Tocoyena}); \textbf{G)} \textit{Harengula jaguana} (\textit{Harengula}); \textbf{H)} \textit{Gaylussacia decipiens} (\textit{Gaylussacia}); \textbf{I)} \textit{Oocephalus foliosus} (\textit{Oocephalus}); \textbf{J)} \textit{Tityus carvalhoi} (\textit{Tityus}); \textbf{K)} \textit{Prosthechea vespa} (\textit{Prosthechea}), Image credits: A) Pamela Lavor, B) Juliana El-Ottra, C) Eduardo Tomaz, D) Filipe C Serrano, E) Raiane Vital da Paz, F) Fernanda GL Moreira, G) Thais Ferreira-Araujo, H) Luiz Menini Neto, I) Arthur Soares, J) Renata C Santos-Costa, K) Tiago Vieira}\label{fig:species}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\linewidth]{./output/figure_number_of_records} \caption{The difference in records available between the raw and filtered dataset of 18 Neotropical taxa including animals, funghi, and plants, plotted in a 100x100 km grid across the Neotropics.}\label{fig:total}
+\includegraphics[width=\linewidth]{./output/figure_number_of_records} \caption{The difference in records available between the raw and filtered dataset of 18 Neotropical taxa including animals, fungi, and plants, plotted in a 100x100 km grid across the Neotropics.}\label{fig:total}
 \end{figure}
@@
 \includegraphics[width=\linewidth]{./output/figure_number_of_records_split} \caption{Geographic location of the occurrence records flagged by the automated tests applied in this study.
-The concentration of records flagged on the basis of record test and individual count test in Central America and southern North America.
 Only filters that flagged at least 0.1\% of records in any taxon are shown.}\label{fig:split}
@@
 The case of the Global Biodiversity Information Facility (GBIF).'' Copenhagen, Denmark: GBIF.
+
+\leavevmode\hypertarget{ref-Antonelli2018}{}%
+Antonelli, Alexandre, Alexander Zizka, Fernanda Antunes Carvalho, Ruud Scharn, Christine D.
+Bacon, Daniele Silvestro, and Fabien L Condamine.
+2018.
+``Amazonia is the primary source of Neotropical biodiversity.'' \emph{Proceedings of the National Academy of Sciences} 115 (23): 6034--9.
+\url{https://doi.org/10.1073/pnas.1713819115}.
 
@@
 
+\leavevmode\hypertarget{ref-GBIForg2019c}{}%
+GBIF.org. 2019a. ``Arhynchobatidae (29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/10.15468/dl.uutyb6}.
+
+\leavevmode\hypertarget{ref-GBIForg2019f}{}%
+---------. 2019b. ``Conchocarpus (29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/10.15468/dl.zjjpmh}.
+
 \leavevmode\hypertarget{ref-GBIForg2019}{}%
-GBIF.org. 2019a. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.sojrfp}.
+---------. 2019c. ``Diogenidae (29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/10.15468/dl.sojrfp}.
+
+\leavevmode\hypertarget{ref-GBIForg2019d}{}%
+---------. 2019d. ``Dipsadidae (29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/10.15468/dl.8hnzfo}.
+
+\leavevmode\hypertarget{ref-GBIForg2019g}{}%
+---------. 2019e. ``Gaylussacia (29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/10.15468/dl.4srw8a}.
+
+\leavevmode\hypertarget{ref-GBIForg2019e}{}%
+---------. 2019f. ``Harengula (29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/10.15468/dl.zznjbv}.
+
+\leavevmode\hypertarget{ref-GBIForg2019j}{}%
+---------. 2019g. ``Iridaceae (29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.nmzgi9}.
+
+\leavevmode\hypertarget{ref-GBIForg2019i}{}%
+---------. 2019h. ``Lepismium (29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/10.15468/dl.762543}.
 
 \leavevmode\hypertarget{ref-GBIForg2019a}{}%
----------. 2019b. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.bx0jjw}.
+---------. 2019i. ``Neanuridae (29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/10.15468/dl.bx0jjw}.
+
+\leavevmode\hypertarget{ref-GBIForg2019n}{}%
+---------. 2019j. ``Oocephalus (29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/10.15468/dl.wkwque}.
+
+\leavevmode\hypertarget{ref-GBIForg2019k}{}%
+---------. 2019k. ``Pilosocereus (29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/10.15468/dl.scmkx5}.
+
+\leavevmode\hypertarget{ref-GBIForg2019m}{}%
+---------. 2019l. ``Prosthechea (29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/10.15468/dl.6bzfz4}.
+
+\leavevmode\hypertarget{ref-GBIForg2019h}{}%
+---------. 2019m. ``Thozetella (29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/10.15468/dl.rpkjsh}.
+
+\leavevmode\hypertarget{ref-GBIForg2019o}{}%
+---------. 2019n. ``Tillandsia (29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/10.15468/dl.zj2cyj}.
 
 \leavevmode\hypertarget{ref-GBIForg2019b}{}%
----------. 2019c. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.zv6kuq}.
-
-\leavevmode\hypertarget{ref-GBIForg2019c}{}%
----------. 2019d. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.uutyb6}.
-
-\leavevmode\hypertarget{ref-GBIForg2019d}{}%
----------. 2019e. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.8hnzfo}.
-
-\leavevmode\hypertarget{ref-GBIForg2019e}{}%
----------. 2019f. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.zznjbv}.
-
-\leavevmode\hypertarget{ref-GBIForg2019f}{}%
----------. 2019g. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.zjjpmh}.
-
-\leavevmode\hypertarget{ref-GBIForg2019g}{}%
----------. 2019h. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.4srw8a}.
-
-\leavevmode\hypertarget{ref-GBIForg2019h}{}%
----------. 2019i. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.rpkjsh}.
-
-\leavevmode\hypertarget{ref-GBIForg2019i}{}%
----------. 2019j. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.762543}.
-
-\leavevmode\hypertarget{ref-GBIForg2019j}{}%
----------. 2019k. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.nmzgi9}.
-
-\leavevmode\hypertarget{ref-GBIForg2019k}{}%
----------. 2019l. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.scmkx5}.
+---------. 2019o. ``Tityus (29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/10.15468/dl.zv6kuq}.
 
 \leavevmode\hypertarget{ref-GBIForg2019l}{}%
----------. 2019m. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.d34gos}.
-
-\leavevmode\hypertarget{ref-GBIForg2019m}{}%
----------. 2019n. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.6bzfz4}.
-
-\leavevmode\hypertarget{ref-GBIForg2019n}{}%
----------. 2019o. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.wkwque}.
-
-\leavevmode\hypertarget{ref-GBIForg2019o}{}%
----------. 2019p. ``(29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.zj2cyj}.
+---------. 2019p. ``Tocoyena (29 December 2019) GBIF Occurrence Download.'' \url{https://doi.org/10.15468/dl.d34gos}.
 
 \leavevmode\hypertarget{ref-GBIForg2020}{}%
----------. 2020a. ``(25 February 2020) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.qazjh4}.
+---------. 2020a. ``Diogenidae (25 February 2020) GBIF Occurrence Download.'' \url{https://doi.org/10.15468/dl.qazjh4}.
 
 \leavevmode\hypertarget{ref-GBIForg2020a}{}%
----------. 2020b. ``(25 February 2020) GBIF Occurrence Download.'' \url{https://doi.org/doi.org/10.15468/dl.ixq7wh}.
+---------. 2020b. ``Entomobryidae (25 February 2020) GBIF Occurrence Download.'' \url{https://doi.org/10.15468/dl.ixq7wh}.
+
+\leavevmode\hypertarget{ref-Guedes2018}{}%
+Guedes, Thaís B., Ricardo J.
+Sawaya, Alexander Zizka, Shawn Laffan, Søren Faurby, R.
+Alexander Pyron, Renato S.
+Bérnils, et al.
+2018.
+``Patterns, biases and prospects in the distribution and diversity of Neotropical snakes.'' \emph{Global Ecology and Biogeography} 27 (1): 14--21.
+\url{https://doi.org/10.1111/geb.12679}.
 
@@
 
+\leavevmode\hypertarget{ref-Peterson2018}{}%
+Peterson, A.
+Townsend, Alex Asase, Dora Canhos, Sidnei de Souza, and John Wieczorek.
+2018.
+``Data Leakage and Loss in Biodiversity Informatics.'' \emph{Biodiversity Data Journal} 6 (November): e26826.
+\url{https://doi.org/10.3897/BDJ.6.e26826}.
+
 \leavevmode\hypertarget{ref-rcoreteam2019}{}%
@@
 \leavevmode\hypertarget{ref-Rivers2011}{}%
-Rivers, Malin C., Lin Taylor, Neil a.
+Rivers, Malin C., Lin Taylor, Neil A.
 Brummitt, Thomas R.
@@
 
-\leavevmode\hypertarget{ref-Topel2016}{}%
-Töpel, Mats, Alexander Zizka, Maria Fernanda Calió, Ruud Scharn, Daniele Silvestro, and Alexandre Antonelli.
-2016.
-``SpeciesGeoCoder: Fast Categorization of Species Occurrences for Analyses of Biodiversity, Biogeography, Ecology, and Evolution.'' \emph{Systematic Biology} 66 (2): syw064.
+\leavevmode\hypertarget{ref-Topel2017}{}%
+Töpel, Mats, Alexander Zizka, M.
+F.
+Maria Fernanda Calió, Ruud Scharn, Daniele Silvestro, and Alexandre Antonelli.
+2017.
+``SpeciesGeoCoder: Fast categorization of species occurrences for analyses of biodiversity, biogeography, ecology, and evolution.'' \emph{Systematic Biology} 66 (2): 145--51.
 \url{https://doi.org/10.1093/sysbio/syw064}.
@@
 
-\leavevmode\hypertarget{ref-Zizka2019a}{}%
+\leavevmode\hypertarget{ref-Zizka2020b}{}%
 Zizka, Alexander, Josue Azevedo, Elton Leme, Beatriz Neves, Andrea Ferreira, Daniel Caceres, and Georg Zizka.
-2019.
-``Biogeography and conservation status of the pineapple family ( Bromeliaceae ),'' no.
-June: 1--13.
+2020.
+``Biogeography and conservation status of the pineapple family (Bromeliaceae).'' \emph{Diveristy and Distributions} 26 (2): 183--95.
 \url{https://doi.org/10.1111/ddi.13004}.

### 2020-06-04 02_automated_cleaning.R

@@ -120,6 +120,16 @@ tot <- out %>%
 
 summary_table <- out %>% 
   group_by(taxon) %>% 
+  mutate(total_error = !(.sea & .inst & .zer & .equ)) %>% 
+  mutate(total_unfit = !(coordinate_precision&
+                            coordinate_base&
+                            individual_count&
+                            record_age&
+                            record_id&
+                            .cap&
+                            .cen&
+                            .urb&
+                            .dpl)) %>% 
   summarize(
     total_records = length(summary),
     total_flags = sum(!summary),
@@ -138,7 +148,10 @@ summary_table <- out %>%
     urban_coords = sum(!.urb),
     gbif_coords = sum(!.gbf),
     inst_coords = sum(!.inst),
-    dupl_coords = sum(!.dpl)) %>% 
+    dupl_coords = sum(!.dpl),
+    total_error = sum(total_error),
+    total_unfit = sum(total_unfit),
+    ) %>% 
   bind_rows(tot) %>% 
   mutate(fraction = total_flags / total_records)
 

### 2020-06-04 03_automated_conservation_Assessment.R

@@ -13,6 +13,25 @@ dat <- read_csv("output/all_records.csv") %>%
 dat_cl <-  dat %>% 
   filter(summary)
 
+## table of species per group
+dat_cl %>% 
+  select(taxon, species) %>% 
+  distinct() %>% 
+  group_by(taxon) %>% 
+  count()
+
+sp <- dat_cl %>% 
+  select(taxon, species) %>% 
+  distinct() %>% 
+  filter(!taxon %in% c("Dipsadidae", "Iridaceae", "Tillandsia"))
+
+sp <- split(sp, f = sp$taxon)
+
+for(i in 1:length(sp)){
+  write_csv(sp[[i]], path = paste("output/species_lists/", names(sp)[i], ".csv", sep = ""))
+}
+
+
 # Automated assessment
 ## raw
 rw_in <- dat%>%

### 2020-06-04 04_figure_number_of_occurrences.R

@@ -4,6 +4,7 @@ library(raster)
 library(viridis)
 library(ggthemes)
 library(speciesgeocodeR)
+library(rnaturalearth)
 
 # load data
 dat <- read_csv("output/all_records.csv") %>% 
@@ -40,7 +41,8 @@ abu_cl <- RichnessGrid(x = pts_cl, ras = be, type = "abu")
 
 plo <- abu_rw - abu_cl
 
-plo <-data.frame(rasterToPoints(plo))
+plo <-data.frame(rasterToPoints(plo)) %>% 
+  filter(layer != 0)
 # 
 # plo_rw <-data.frame(rasterToPoints(abu_rw))%>%
 #   # filter(layer > 0 ) %>% 
@@ -53,17 +55,22 @@ plo <-data.frame(rasterToPoints(plo))
 # plo <- bind_rows(plo_rw, plo_cl)
 
 # background map
-world.inp  <- suppressWarnings(rnaturalearth::ne_download(scale = 50, 
+world.inp  <- suppressWarnings(rnaturalearth::ne_download(scale = 110, 
                                                           type = 'land', 
                                                           category = 'physical',
                                                           load = TRUE))
 world.behr <- spTransform(world.inp, CRS(behr)) %>% fortify()
 
+world.countries <- rnaturalearth::ne_countries(type = 'countries', 
+                                               scale = 110)
+countries.behr <- spTransform(world.countries, CRS(behr)) %>% fortify()
 
 # plot
 ggplot()+
-  geom_polygon(data = world.behr,
-               aes(x = long, y = lat, group = group), fill = "transparent", color = "black")+
+  geom_polygon(data = countries.behr,
+               aes(x = long, y = lat, group = group), 
+               fill = "transparent", 
+               color = "grey20")+
   geom_tile(data = plo, aes(x = x, y = y, fill = log(layer)), alpha = 0.8)+
   scale_fill_viridis(name = "Number of\nrecords", direction = -1, na.value = "transparent",
                      breaks = c(log(1), log(10), log(100), log(1000), log(5000)),
@@ -81,7 +88,7 @@ ggsave("output/figure_number_of_records.jpg", height = 8, width=8)
 
 # Figure 2
 plo <- dat %>% 
-  select(-coordinateUncertaintyInMeters, -taxon, -countryCode, -gbifID,
+  dplyr::select(-coordinateUncertaintyInMeters, -taxon, -countryCode, -gbifID,
          -class,
          -genus, -taxonRank,
          -basisOfRecord, 
[…]

### 2020-06-04 06_species_richness_maps.R

@@ -4,6 +4,7 @@ library(raster)
 library(viridis)
 library(ggthemes)
 library(speciesgeocodeR)
+library(rnaturalearth)
 
 # load data
 dat <- read_csv("output/all_records.csv") %>% 
@@ -20,16 +21,19 @@ behr <- '+proj=cea +lon_0=0 +lat_ts=30 +x_0=0 +y_0=0 +datum=WGS84 +ellps=WGS84 +
 wgs1984 <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
 
 # background map
-world.inp  <- suppressWarnings(rnaturalearth::ne_download(scale = 50, 
+world.inp  <- suppressWarnings(rnaturalearth::ne_download(scale = 110, 
                                                           type = 'land', 
                                                           category = 'physical',
                                                           load = TRUE))
 world.behr <- spTransform(world.inp, CRS(behr)) %>% fortify()
 
+world.countries <- rnaturalearth::ne_countries(type = 'countries', 
+                                scale = 110)
+countries.behr <- spTransform(world.countries, CRS(behr)) %>% fortify()
 #cleaned data for main mansucript
 plo <- dat %>%
   filter(summary) %>% 
-  select(species, taxon, decimalLongitude, decimalLatitude)
+  dplyr::select(species, taxon, decimalLongitude, decimalLatitude)
 
 plo <- split(plo, f = plo$taxon)
 
@@ -51,7 +55,7 @@ names(plo_cl)[4] <- "layer_cl"
 
 # unfiltered for the supplement
 plo <- dat %>%
-  select(species, taxon, decimalLongitude, decimalLatitude)
+  dplyr::select(species, taxon, decimalLongitude, decimalLatitude)
 
 plo <- split(plo, f = plo$taxon)
 
@@ -82,12 +86,21 @@ write_csv(plo, "output/specis_richness.csv")
 ## select illustrative taxa
 
 plo <- plo %>% 
-  filter(taxon %in% c("Thozetella", "Tillandsia", "Dipsadidae", "Harengula"))
+  filter(taxon %in% c("Thozetella", "Tillandsia", "Dipsadidae", "Harengula")) %>% 
+  filter(difference != 0)
 
 # plot
 ggplot()+
-  geom_polygon(data = world.behr,
-               aes(x = long, y = lat, group = group), fill = "transparent", color = "black")+
+  # geom_polygon(data = world.behr,
+  #              aes(x = long, y = lat, group = group), 
+  #              fill = "transparent", 
[…]

### 2020-06-10 latexdiff/new_manuscript_peerj.tex
latexdiff

--- 2020-06-04 manuscript_peerj.tex
@@
 Species occurrence records provide the basis for many biodiversity studies.
-They derive from geo-referenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications.
+They derive from georeferenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications.
 Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern.
 Automatic filtering is a scalable and reproducible means to identify potentially problematic records and tailor datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org), for biodiversity analyses.
-However, it is unclear how much data may be lost by filtering, whether the same filters should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses.
-Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 18 Neotropical taxa including animals, fungi, and plants, terrestrial and marine, downloaded from GBIF.
+However, it is unclear how much data may be lost by filtering, whether the same filters should be applied across all taxonomic groups, and what the effect of filtering is on common downstream analyses.
+Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 18 Neotropical taxa, including terrestrial and marine animals, fungi, and plants downloaded from GBIF.
 We find that a total of 44.3\% of the records are potentially problematic, with large variation across taxonomic groups (25 - 90\%).
-A small fraction of records was identified as erroneous in the strict sense (4.2\%), the majority of filtered records that were rather unfit for downstream analyses (41.7\%).
-Filters of duplicated information, collection year, basis of record as well as urban areas and coordinates for terrestrial taxa in the sea or marine taxa on land have the greatest effect.
+A small fraction of records was identified as erroneous in the strict sense (4.2\%), and a much larger proportion as unfit for most downstream analyses (41.7\%).
+Filters of duplicated information, collection year, and basis of record, as well as coordinates in urban areas, or for terrestrial taxa in the sea or marine taxa on land, have the greatest effect.
 Automated filtering can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus.
-Our results stress the importance of thorough exploration of the meta-data associated with species records for biodiversity research.
+Our results stress the importance of thorough recording and exploration of the meta-data associated with species records for biodiversity research.
 % Dummy abstract text.
@@
 Publicly available species distribution data have become a crucial resource in biodiversity research, including studies in ecology, biogeography, systematics and conservation biology.
-In particular, the availability of digitized collections from museums and herbaria, and citizen science observations has increased drastically over the last few years.
+In particular, the availability of digitized collections from museums and herbaria and citizen science observations has increased drastically over the last few years.
 As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.4 billion geo-referenced occurrence records for species from across the globe and the tree of life.
 
-A central challenge to the use of these publicly available species occurrence data in research are problematic geographic coordinates, which are either erroneous or unfit for downstream analyses (for instance because they are overly imprecise, Anderson et al.
+A central challenge to the use of these publicly available species occurrence data in research is problematic geographic coordinates, which are either erroneous or unfit for downstream analyses (for instance because they are overly imprecise, Anderson et al.
 2016).
-Problems mostly arise because public databases integrate records collected with different methodologies in different places, at different times; often without centralized curation and only rudimentary meta-data.
+Problems mostly arise because data aggregators such as GBIF integrate records collected with different methodologies in different places at different times--- often without centralized curation and only rudimentary meta-data.
 For instance, problematic coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common (Maldonado et al.
 2015; Yesson et al.
-2007) and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in captivity (Zizka et al.
+2007) and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in cultivation or captivity (Zizka et al.
 2019).
@@
 2019; Jin and Yang 2020) or on additional data, such as environmental variables (Robertson, Visser, and Hui 2016).
-Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the suitability of publicly available occurrence records for biodiversity research (Zizka et al.
+Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions, have been proposed to improve the suitability of publicly available occurrence records for biodiversity research (Zizka et al.
 2019).
@@
 2018) Red List assessments.
-However, all metrics and especially EOO, are sensitive to individual records with problematic coordinates.
-Automated conservation assessments may therefore be biased, particularly if the number of records are few, as it is the case for many tropical species.
+However, all metrics, and especially EOO, are sensitive to individual records with problematic coordinates.
+Automated conservation assessments may therefore be biased, particularly if the number of records is low, as it is the case for many tropical species.
 
@@
 
-This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering students and researchers working with different taxonomic groups of animals, fungi, and plants across the Neotropics.
+This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 which gathered students and researchers working with different taxonomic groups of animals, fungi, and plants across the Neotropics.
 Each participant analysed geographic occurrence data from their taxonomic group of interest and commented on the results for their group.
-Hence, we include groups based on the expertise of the participants rather than following some arbitrary choice of taxa and taxonomic ranks.
-We acknowledge a varying degree in documented expertise and number of years working on each group.
+Hence, we include groups based on the expertise of the participants rather than following an arbitrary choice of taxa and taxonomic ranks.
+We acknowledge a varying degree of documented expertise and number of years working on each group.
 We obtained public occurrence records for 18 taxa, including one plant family, nine plant genera, one genus of fungi, three families and one genus of terrestrial arthropods, one family of snakes, one family of skates, and one genus of bony fish (Table 1).
@@
 We downloaded GBIF-interpreted data including only records with geographic coordinates and limited the study area to a rectangle between 90\(^\circ\)S - 33\(^\circ\) N and 35\(^\circ\) W - 120\(^\circ\) W reflecting the Neotropics (Morrone 2014), our main area of expertise.
-The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, Entomobryidae, \emph{Gaylussacia}, Iridaceae, Neanuridae, and \emph{Tillandsia} for which we only obtained the Neotropical occurrences.
-We consider GBIF data generally of high quality and use them as study case because GBIF is the largest, most widely-used and taxonomically most comprehensive data source for species occurrence records, but many more exist (e.g., \url{https://bien.nceas.ucsb.edu/bien/}, www.fishbase.de or Guedes et al.
+The natural distributions of all included taxa are confined to the Neotropics except for Arhynchobatidae, Diogenidae, Dipsadidae, Entomobryidae, \emph{Gaylussacia}, Iridaceae, Neanuridae, and \emph{Tillandsia}, for which we only obtained the Neotropical occurrences.
+We consider GBIF data generally of high quality and use them as a case study because GBIF is the largest, most widely used and taxonomically most comprehensive data source for species occurrence records; however many more exist (e.g., \url{https://bien.nceas.ucsb.edu/bien/}, www.fishbase.de or Guedes et al.
 2018).
-The variety of existing databases is also the reason why we download data without accounting for GBIF-internal quality testing (``Geospatial issues''), since quality control mechanisms might differ among databases and sometimes no quality control might be done (for instance in \emph{de novo} databases compiled by groups of researchers).
+The variety of existing databases is also the reason why we downloaded data without accounting for GBIF-internal quality testing (``Geospatial issues''), since quality control mechanisms might differ among databases and sometimes quality control might not be carried out (for instance in \emph{de novo} databases compiled by groups of researchers).
 
@@
 We followed the cleaning pipeline outlined by Zizka et al.
-(2019) and first filtered the raw data downloaded from GBIF (``raw'', hereafter) using meta-data for those records for which they were available (although this is often not the case, Peterson et al.
+(2019) and first filtered the raw data downloaded from GBIF (``raw'', hereafter) using meta-data for those records for which they were available (although meta-data were often missing, Peterson et al.
 2018), removing: 1) records with a coordinate precision below 100 km (as this represents the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99.
-Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e., test for duplicated records).
+Furthermore, we rounded the geographic coordinates to four decimal places and retained only one record per species per location (i.e., test for duplicated records).
 In a second step, we used the \texttt{clean\_coordinates} function of the \texttt{CoordinateCleaner\ v2.0-11} package (Zizka et al.
@@
 While these assessments are preliminary (see IUCN Standards and Petitions Subcommittee 2017), they can be a proxy used by the IUCN to speed up full assessments.
-We then benchmarked the preliminary conservation assessments against the global IUCN Red List assessments as far as available for the same taxa, which we obtained from www.iucn.org via the \texttt{rredlist\ v.0.5.0} package (Chamberlain 2018).
+We then benchmarked the preliminary conservation assessments against the global IUCN Red List assessments for the same taxa (where available), which we obtained from www.iucn.org via the \texttt{rredlist\ v.0.5.0} package (Chamberlain 2018).
 
@@
 
-We did all analyses in the R computing environment (R Core Team 2019), using standard libraries for data handling and visualization (Wickham 2018; Garnier 2018; Ooms 2014, 2019; Hijmans 2019).
+We carried out all analyses in the R computing environment (R Core Team 2019), using standard libraries for data handling and visualization (Wickham 2018; Garnier 2018; Ooms 2014, 2019; Hijmans 2019).
 All scripts are available from a zenodo repository (\url{doi:10.5281/zenodo.3695102}).
@@
 In general, the different filters we tested were of similar importance for different study groups.
-There were few outstanding exceptions, including the particularly high proportions of records filtered on the ``basis of record test'' for \emph{Tityus} (7.0\%), Dipsadidae (5.6\%), \emph{Prosthechea} (5.0\%) and \emph{Tillandsia} (4.9\%), by the collection year for Dipsadidae (11.3\%), by the taxonomic identification level for \emph{Tityus} (1.6\%), by the capital coordinates for \emph{Oocephalus} (6.1\%) and \emph{Gaylussacia} (3.2\%), by the seas/land test for Diogenidae and \emph{Thozetella}, and by the urban areas test for \emph{Oocephalus} (13.3\%) and Iridaceae (12.3\%).
-Furthermore, Entomobryidae differed considerably from all other study taxa with exceptionally high numbers of records filtered by the ``basis of record,''level of identification' and urban areas" tests.
+There were few outstanding exceptions, including the particularly high proportions of records filtered by the ``basis of record test'' for \emph{Tityus} (7.0\%), Dipsadidae (5.6\%), \emph{Prosthechea} (5.0\%) and \emph{Tillandsia} (4.9\%), by the collection year for Dipsadidae (11.3\%), by the taxonomic identification level for \emph{Tityus} (1.6\%), by the capital coordinates for \emph{Oocephalus} (6.1\%) and \emph{Gaylussacia} (3.2\%), by the seas/land test for Diogenidae and \emph{Thozetella}, and by the urban areas test for \emph{Oocephalus} (13.3\%) and Iridaceae (12.3\%).
+Furthermore, Entomobryidae differed considerably from all other study taxa with exceptionally high numbers of records filtered by the ``basis of record'', ``level of identification'' and ``urban areas'' tests.
 
-Geographically, the records filtered by the ``basis of record'' and ``individual count'' tests were concentrated in Central America and southern North America, and a relatively high number of records filtered due to their proximity to the centroids of political entities were located on Caribbean islands (Fig.
+Geographically, the records filtered by the ``basis of record'' and ``individual count'' tests were concentrated in Central America and southern North America, and a relatively high number of records were filtered due to their proximity to the centroids of political entities were located on Caribbean islands (Fig.
 \ref{fig:split}).
@@
 The median percentage of species per study group with an IUCN assessment was 15\%.
-A total of 102 species were listed as Threatened by the IUCN global Red List (CR = 19, EN = 40, VU = 43) and 477 as Not Threatened.
+A total of 102 species were listed as \emph{Threatened} by the IUCN global Red List (CR = 19, EN = 40, VU = 43) and 477 as \emph{Not Threatened}.
 
@@
 The automated assessment based on the filtered dataset agreed with the IUCN assessment for identifying species as possibly threatened (CR, EN, VU) for 358 species (64\%; Table 4).
-Filtering reduced the EOO by -18.4\% and the AOO by -9.9\% on median per group.
+Filtering reduced the EOO by 18.4\% and the AOO by 9.9\% on median per group.
 For the raw dataset the agreement with IUCN was higher at 381 species (65.7\%).
@@
 
-Automated flagging based on meta-data and automatic tests filtered on average 45\% of the records per taxonomic group; 25.9\% - 90.3\% as ``unfit'' and 0\% - 44.3\% as ``erroneous''.
+Automated flagging based on meta-data and automatic tests filtered on average 45\% of the records per taxonomic group; 25.9\%-90.3\% as ``unfit'' and 0\%-44.3\% as ``erroneous''.
 The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (\textbf{Question 1}).
-The importance of different tests was similar across taxonomic groups, with exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (\textbf{Question 2}).
-The results for species richness were similar between the raw and filtered data with some improvements by the filters.
+The importance of different tests was similar across taxonomic groups, with particular exceptions for the tests on basis of record, collection year, capital coordinates, and urban areas (\textbf{Question 2}).
+The results for species richness were similar between the raw and filtered data with some improvements using the filters.
 We found little impact of filtering on the accuracy of the automated conservation assessments (\textbf{Question 3}).
@@
 The aim of automated filtering is to identify possibly problematic records that are unsuitable for particular downstream analyses.
-While those records filtered as ``erroneous'' will likely cause problems for most biodiversity research, those filtered as ``unfit'' might have varying impact, depending on the type and spatial resolution of the downstream analyses, ranging from an unnecessary computational burden which can be a bottle neck for large-scale analyses (i.e.~duplicates, Antonelli et al.
-2018) over increasing uncertainty (due to low precision) to compromising results.
-For instance, records assigned to country centroids might be acceptable for inter-continental comparisons, but likely are erroneous for species distribution modelling on a local scale.
+While those records filtered as ``erroneous'' will likely cause problems for most biodiversity research, those filtered as ``unfit'' might have varying impact, depending on the type and spatial resolution of the downstream analyses.
+Unwanted effects include an unnecessary computational burden, which can be a bottleneck for large-scale analyses (i.e.~duplicates, Antonelli et al.
+2018), and increased uncertainty (due to low precision), or completely compromising results.
+For instance, records assigned to country centroids might be acceptable for inter-continental comparisons, but are likely to be erroneous for species distribution modelling on a local scale.
 The importance of each test and the linked thresholds must be judged based on the specific downstream analyses.
-As our results show, adapting automated tests to the geographic study area or the taxonomic study group can be warranted.
+As our results show, it may be advisable to adapt automated tests to the geographic study area or the taxonomic study group.
 For instance, the high number of records flagged for centroids on the Lesser Antilles (Fig.
@@
 
-Several reasons may explain the high proportion of records flagged as duplicates.
-First, the deposition of duplicates from the same specimen hosted at different collections is common practice, especially for plants, where a specimen duplication is easily feasible.
+Several factors may explain the high proportion of records flagged as duplicates.
+First, the deposition of duplicates from the same specimen at different institutions is common practice, especially for plants, where a specimen duplication is entirely feasible.
 Second, independent collections at similar localities may occur, in particular for local endemics.
@@
 Some notable exceptions stress the need to adapt each filter to the taxonomic study group to balance data quality and data availability.
-The high fraction of records filtered by the ``basis of record filter'' for \emph{Tityus}, Dipsadidae, \emph{Prosthechea} and \emph{Tillandsia}, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups.
-The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection effort already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700.
-The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700.
-The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae and \emph{Oocephalus} might be related to horticulture since at least some species in those groups are commonly cultivated as ornamentals.
+The high fraction of records filtered by the ``basis of record'' filter for \emph{Tityus}, Dipsadidae, \emph{Prosthechea} and \emph{Tillandsia}, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups.
+The high fraction of records flagged by the ``collection year'' filter for Dispadidae was caused by a high collection effort in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700.
+The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institution's code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700.
+The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae and \emph{Oocephalus} might be related to horticulture, since at least some species in those groups are commonly cultivated as ornamentals.
 
-The general agreement between the species richness maps based on raw and filtered data was encouraging, for the use of this data for large-scale biogeographic research (Fig.
+The general agreement between the species richness maps based on raw and filtered data was encouraging, in terms of the use of this data for large-scale biogeographic research (Fig.
 \ref{fig:speciesrichness}, Appendix 2).
@@
 
-We overall consider the effect of the automated filters as positive since they identified the above-mentioned issues and increased the data precision and reduced computational burden (Table 3, Appendix 2).
+Overall, we consider the effect of the automated filters as positive since they identified the above-mentioned issues and increased the data precision and reduced computational burden (Table 3, Appendix 2).
 However, in some cases filters failed to remove major issues, often due to incomplete metadata.
@@
 Additionally automated filters might be overly strict or unsuitable for certain taxa.
-For instance, in Entomobryidae, 2,004 samples were marked as material samples, and therefore removed by our global filter retaining only specimen and observation data, which in this case was overly strict.
+For instance, in Entomobryidae, 2,004 samples were marked as material samples and therefore removed by our global filter retaining only specimen and observation data, which in this case was overly strict.
 
@@
 In fact, the internal quality control of GBIF does flag some of the problems we tested for (i.e., zero coordinates, equal lat/lon) while others might be implemented in the near future (country centroids, \url{https://data-blog.gbif.org/post/country-centroids/}, \url{https://data-blog.gbif.org/post/country-centroids/}).
-While this internal quality is very helpful, we see a huge potential to overcome issues with data quality in a user-feedback system allowing users to provide expert assessments, i.e.~a meta-annotation of records being challenged (and why).
+While this internal quality is very helpful, we see a huge potential to overcome issues with data quality in a user-feedback system that allows users to provide expert assessments, i.e.~a meta-annotation of records being challenged (and why).
 Such a system would not need to change the original data and could include multiple levels to account for differing opinions.
 
-As a next step for automated filtering, it is possible to detect additional problematic records with the help of external species distribution information, which is available for some taxa for example from www.iucn.org (vertebrates) or \url{https://wcsp.science.kew.org/} (selected seed plant families).
+As a next step for automated filtering, it is possible to detect additional problematic records with the help of external species distribution information, which is available for some taxa, for example from www.iucn.org (vertebrates) or \url{https://wcsp.science.kew.org/} (selected seed plant families).
 Furthermore, while in this study we focused on meta-data and geographic filtering, taxonomic cleaning---the resolution of synonymies and identification of accepted names---is another important part of data curation, but depends on taxon-specific taxonomic backbones and synonymy lists which are not readily available for many groups and often are contradictory within individual taxa.
@@
 This might be caused by the structure of our comparison, which only included species that were evaluated by the IUCN Red List (and not considered as \emph{Data Deficient}) and at the same time had occurrences recorded in GBIF.
-Those inclusion criteria are likely to have biased the datasets towards species with large ranges, since generally more data for them are available.
+Those inclusion criteria are likely to have biased the datasets towards species with large ranges, since generally more data are available for them.
 The robustness of automated conservation assessments to data quality is encouraging, although these methods are only an approximation (and not replacements) of full IUCN Red List assessments, especially for species with few collection records (Rivers et al.
@@
 
-Our results suggest that between one to two thirds of the occurrence records obtained from public databases might be unsuitable for downstream biodiversity analyses.
-While the majority of these records might not be erroneous \emph{per se}, they might be overly imprecise and thereby increase uncertainty of downstream results or add computational burden on big data analyses.
+Our results suggest that between one to half of the occurrence records obtained from GBIF might be unsuitable for downstream biodiversity analyses.
+While the majority of these records might not be erroneous \emph{per se}, they are overly imprecise and thereby increase uncertainty of downstream results or add computational burden on big data analyses.
 
-While our results suggest that large-scale species richness patterns and automated conservation assessments are largely robust to problematic occurrence records, they also stress the importance of (meta-)data exploration prior to most biodiversity analyses.
+While our results suggest that large-scale species richness patterns and automated conservation assessments are largely resilient to the effects of problematic occurrence records, they also stress the importance of (meta-)data exploration prior to most biodiversity analyses.
 Automated filtering can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the specific taxonomic groups and geographic area of interest.
@@
 
+We thank GBIF and all data collectors and contributors for their excellent work.
+We tahnk Town Peterson, Roderic Page and one anonymous reviewer for the helpful comments on an earlier version of this mansucript.
 This study enrolled participants of the workshop ``Biodiversity data: from field to yield'' led by Alice Calvente, Fernanda Carvalho, Alexander Zizka, and Alexandre Antonelli through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI.
@@
 \item
-  Appendix 2 - Taxon specific richness maps and comments
+  Appendix 2 - Taxon-specific richness maps and comments
 \item
@@
 \end{itemize}
-
-\hypertarget{tables}{%
-\section*{Tables}\label{tables}}
-\addcontentsline{toc}{section}{Tables}
-
-\begin{table}
-
-\caption{\label{tab:tabletaxa}The study groups and their taxonomy.
-This study includes three marine and 15 terrestrial taxa, seven of them animals, one group of fungi and ten plants, belonging to 16 different orders.}
-\centering
-\fontsize{11}{13}\selectfont
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{1cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{1.5cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{3cm}}
-\toprule
- & Taxon & Taxon rank & Realm & Common name & 'Phylum' & Family\\
-\midrule
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/diogenidae.png} & Diogenidae & Family & Marine & Hermit crabs & Arthropoda & Diogenidae\\
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/entomobryidae.png} & Entomobryidae & Family & Terrestrial & Springtails & Arthropoda & Entomobryidae\\
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/neanuridae.png} & Neanuridae & Family & Terrestrial & Springtails & Arthropoda & Neanuridae\\
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/tityus.png} & \em{Tityus} & Genus & Terrestrial & Scorpions & Arthropoda & Buthidae\\
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/arhynchobatidae.png} & Arhynchobatidae & Family & Marine & Skates & Chordata & Arhynchobatidae\\
-\addlinespace
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/dipsadidae.png} & Dipsadidae & Family & Terrestrial & Snakes & Chordata & Dipsadidae\\
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/harengula.png} & \em{Harengula} & Genus & Marine & Herrings & Chordata & Clupeidae\\
-\hline
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/thozetella.png} & \em{Thozetella} & Genus & Terrestrial & Sac fungi & Ascomycota & Chaetosphaeriaceae\\
-\hline
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/conchocarpus.png} & \em{Conchocarpus} & Genus & Terrestrial & NA & Angiosperms & Rutaceae\\
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/gaylussacia.png} & \em{Gaylussacia} & Genus & Terrestrial & Huckleberries & Angiosperms & Ericaceae\\
-\addlinespace
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/harpalyce.png} & \em{Harpalyce} & Genus & Terrestrial & NA & Angiosperms & Fabaceae\\
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/iridaceae.png} & Iridaceae & Family & Terrestrial & NA & Angiosperms & Iridaceae\\
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/lepismium.png} & \em{Lepismium} & Genus & Terrestrial & Cacti & Angiosperms & Cactaceae\\
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/oocephalus.png} & \em{Oocephalus} & Genus & Terrestrial & NA & Angiosperms & Lamiaceae\\
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/pilosocereus.png} & \em{Pilosocereus} & Genus & Terrestrial & Cacti & Angiosperms & Cactaceae\\
-\addlinespace
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/prosthechea.png} & \em{Prosthechea} & Genus & Terrestrial & Orchids & Angiosperms & Orchidaceae\\
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/tillandsia.png} & \em{Tillandsia} & Genus & Terrestrial & Bromeliads & Angiosperms & Bromeliaceae\\
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/tocoyena.png} & \em{Tocoyena} & Genus & Terrestrial & NA & Angiosperms & Rubiaceae\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{table}[!h]
-
-\caption{\label{tab:tabletests}The automated filters used in this study.}
-\centering
-\fontsize{9}{11}\selectfont
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{1cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{9cm}}
-\toprule
-Test & Type & Basis & Rationale\\
-\midrule
-\rowcolor{gray!6}  Biodiversity institutions & Error & Gazetteer-based & Records may have coordinates at the location of biodiversity institutions, e.g.
-because they were erroneously entered with the physical location of the specimen or because they represent individuals from captivity or horticulture\\
-Equal lat/lon & Error & Gazetteer-based & Coordinates with equal latitude and longitude are usually indicative of data entry errors\\
-\rowcolor{gray!6}  Sea & Error & Gazetteer-based & Coordinates from terrestrial organisms in the sea are usually indicative of data entry errors, e.g. swapped latitude and longitude\\
-Zeros & Error & Gazetteer-based & Coordinates with plain zeros are often indicative of data entry errors\\
-\rowcolor{gray!6}  Capitals & Unfit & Gazetteer-based & Records may be assigned to the coordinates of country capitals based on a vague locality description\\
-\addlinespace
-Duplicates & Unfit & Gazetteer-based & Duplicated records may add unnecessary computational burden, in particular for large scale biodiversity analyses and distribution modelling for many species\\
-\rowcolor{gray!6}  Political centroids & Unfit & Gazetteer-based & Records may be assigned to the coordinates of the centroids of political entities based on a vague locality description\\
-Urban areas & Unfit & Gazetteer-based & Records from urban areas are not necessarily errors, but often represent imprecise records automatically geo-referenced from vague locality descriptions or old records from different land-use types\\
-\rowcolor{gray!6}  Basis of record & Unfit & Meta-data & Records might be unsuitable or unreliable for certain analyses dependent on their source, e.g. "fossil" or "unknown"\\
-Collection year & Unfit & Meta-data & Coordinates from old records are more likely to be imprecise or erroneous coordinates since they are derived from  geo-referencing based on the locality description.
-This is more problematic for older records, since names or borders of places may change\\
-\addlinespace
-\rowcolor{gray!6}  Coordinate precision & Unfit & Meta-data & Records may be unsuitable for a study if their precision is lower than the study analysis scale\\
-Identification level & Unfit & Meta-data & Records may be unsuitable if they are not identified to species level.\\
-\rowcolor{gray!6}  Individual count & Unfit & Meta-data & Records may be unsuitable if the number of recorded individuals is 0 (record of absence) or if the count is too high, as this is often related to records from barcoding or indicative of data entry problems.\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{landscape}\begin{table}
-
-\caption{\label{tab:tablecoords}The impact of automated filtering on occurrence records for 18 Neotropical taxa downloaded from www.gbif.org.
-From column four onwards the numbers show the percentage of records flagged by the respective test.
-Only tests that flagged at least 0.1\% of the records in any group are shown.
-Individual records can be flagged by multiple tests, therefore the sum of percentage from all tests can supersede the total percentage.}
-\centering
-\fontsize{8}{10}\selectfont
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{1.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{1.1cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{1.1cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}>{\raggedleft\arraybackslash}p{0.9cm}}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{4}{c}{Summary} & \multicolumn{3}{c}{Errors} & \multicolumn{9}{c}{Unfit} \\
-\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-8} \cmidrule(l{3pt}r{3pt}){9-17}
-Taxon & Total records & Fraction flagged [\%] & Fraction error [\%] & Fraction unfit [\%] & Biodiversity Institutions [\%] & Sea/land area [\%] & Zeros [\%] & Capitals [\%] & Duplicates [\%] & Political centroids [\%] & Urban areas [\%] & Basis of record [\%] & Collection year [\%] & Coordinate precision [\%] & Id-level [\%] & Individual count [\%]\\
-\midrule
-Diogenidae & 13,840 & 68.7 & 44.3 & 38.2 & 0.0 & 44.3 & 0.0 & 0.7 & 33.8 & 0.2 & 1.3 & 1.7 & 2.5 & 0.0 & 0.0 & 0.0\\
-Entomobryidae & 2,767 & 90.3 & 0.1 & 90.3 & 0.1 & 0.0 & 0.0 & 0.1 & 85.5 & 0.0 & 70.1 & 72.9 & 2.0 & 0.0 & 72.1 & 0.0\\
-Neanuridae & 689 & 66.9 & 0.0 & 66.9 & 0.0 & 0.0 & 0.0 & 0.0 & 62.4 & 0.0 & 2.0 & 2.9 & 1.3 & 0.0 & 0.0 & 0.0\\
-\em{Tityus} & 1,018 & 55.2 & 0.5 & 54.9 & 0.5 & 0.0 & 0.0 & 1.2 & 43.5 & 0.1 & 6.9 & 7.0 & 0.4 & 1.8 & 1.6 & 0.0\\
-Arhynchobatidae & 14,633 & 38.5 & 3.8 & 37.4 & 0.0 & 3.8 & 0.0 & 0.0 & 35.4 & 0.0 & 1.9 & 1.7 & 1.3 & 0.0 & 0.9 & 0.0\\
-\addlinespace
-Dipsadidae & 64,249 & 57.7 & 0.3 & 57.6 & 0.3 & 0.0 & 0.0 & 1.8 & 46.3 & 0.4 & 8.5 & 5.6 & 11.3 & 0.8 & 0.0 & 0.1\\
-\em{Harengula} & 36,697 & 31.0 & 5.5 & 27.8 & 0.0 & 5.5 & 0.0 & 0.2 & 27.0 & 0.1 & 0.2 & 1.0 & 0.4 & 0.0 & 0.3 & 0.0\\
-\hline
-\em{Thozetella} & 51 & 35.3 & 23.5 & 29.4 & 0.0 & 23.5 & 0.0 & 0.0 & 27.5 & 0.0 & 2.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0\\
-\hline
-\em{Conchocarpus} & 1,551 & 43.2 & 0.5 & 42.9 & 0.1 & 0.4 & 0.0 & 0.0 & 39.6 & 0.9 & 2.3 & 0.5 & 1.9 & 0.1 & 0.0 & 0.0\\
-\em{Gaylussacia} & 3,998 & 47.2 & 0.1 & 47.1 & 0.1 & 0.1 & 0.0 & 3.2 & 41.8 & 1.1 & 5.2 & 0.7 & 4.4 & 0.6 & 0.0 & 0.0\\
-\addlinespace
-\em{Harpalyce} & 870 & 33.1 & 0.0 & 33.1 & 0.0 & 0.0 & 0.0 & 1.0 & 26.0 & 1.3 & 3.8 & 0.5 & 5.5 & 0.7 & 0.0 & 0.9\\
-Iridaceae & 23,127 & 33.6 & 0.5 & 33.5 & 0.4 & 0.1 & 0.0 & 1.0 & 17.1 & 0.4 & 12.3 & 0.9 & 4.7 & 0.1 & 0.0 & 1.3\\
-\em{Lepismium} & 825 & 29.7 & 0.0 & 29.7 & 0.0 & 0.0 & 0.0 & 0.1 & 21.9 & 0.1 & 7.8 & 0.0 & 2.1 & 0.0 & 0.0 & 0.0\\
-\em{Oocephalus} & 883 & 49.3 & 0.0 & 49.3 & 0.0 & 0.0 & 0.0 & 6.1 & 41.9 & 0.8 & 13.3 & 0.0 & 0.7 & 0.3 & 0.0 & 0.1\\
-\em{Pilosocereus} & 1,940 & 25.9 & 0.2 & 25.9 & 0.2 & 0.0 & 0.0 & 0.5 & 16.8 & 0.5 & 2.1 & 1.8 & 7.0 & 0.0 & 0.0 & 0.9\\
-\addlinespace
-\em{Prosthechea} & 6,617 & 31.5 & 0.1 & 31.5 & 0.0 & 0.0 & 0.1 & 0.4 & 19.6 & 1.7 & 0.9 & 5.0 & 8.3 & 0.1 & 0.0 & 0.2\\
-\em{Tillandsia} & 42,222 & 35.3 & 0.4 & 35.2 & 0.3 & 0.0 & 0.0 & 0.7 & 19.8 & 0.7 & 9.2 & 4.9 & 5.1 & 0.1 & 0.0 & 1.0\\
-\em{Tocoyena} & 2,922 & 37.6 & 0.3 & 37.4 & 0.0 & 0.2 & 0.0 & 0.8 & 32.3 & 0.8 & 5.0 & 0.1 & 1.9 & 0.2 & 0.0 & 0.5\\
-\hline
-Total & 218,899 & 44.3 & 4.2 & 41.7 & 0.2 & 4.0 & 0.0 & 1.0 & 32.3 & 0.4 & 7.1 & 4.2 & 5.6 & 0.3 & 1.0 & 0.4\\
-\bottomrule
-\end{tabular}
-\end{table}
-\end{landscape}
-
-\begin{landscape}\begin{table}
-
-\caption{\label{tab:unnamed-chunk-3}Conservation assessment for 11 Neotropical taxa of plants and animals based on three datasets.
-IUCN: global red list assessment obtained from www.iucn.org; GBIF Raw: Preliminary conservation assessment based on IUCN Criterion B using ConR and the raw dataset from GBIF; GBIF filtered: Preliminary conservation assessment based on IUCN Criterion B using ConR and the filtered dataset.
-Only taxa with at least on species evaluated by IUCN shown.}
-\centering
-\fontsize{9}{11}\selectfont
-\begin{tabular}[t]{l>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{3}{c}{IUCN} & \multicolumn{3}{c}{GBIF Raw} & \multicolumn{5}{c}{GBIF Filtered} \\
-\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7} \cmidrule(l{3pt}r{3pt}){8-12}
-Taxon & n taxa & Evaluated [\%] & Threatened [\%] & n taxa & Threatened [\%] & Match with IUCN [\%] & n taxa & Threatened [\%] & Match with IUCN [\%] & EOO change compared to raw [\%] & AOO change compared to raw [\%]\\
-\midrule
-Arhynchobatidae & 37 & 51.3 & 17.9 & 39 & 35.9 & 45.0 & 39 & 41.0 & 40.0 & -32.7 & -18.5\\
-Dipsadidae & 520 & 68.0 & 8.8 & 638 & 58.3 & 63.0 & 598 & 59.9 & 61.2 & -2.3 & -15.6\\
-\em{Harengula} & 4 & 100.0 & 0.0 & 4 & 0.0 & 100.0 & 4 & 0.0 & 100.0 & -38.0 & -36.9\\
-\hline
-\em{Conchocarpus} & 4 & 8.7 & 0.0 & 46 & 63.0 & 100.0 & 45 & 62.2 & 100.0 & -15.3 & -7.1\\
-\em{Gaylussacia} & 2 & 3.3 & 0.0 & 61 & 59.0 & 50.0 & 58 & 60.3 & 50.0 & -22.5 & -8.6\\
-\addlinespace
-\em{Harpalyce} & 3 & 15.0 & 5.0 & 20 & 65.0 & 66.7 & 17 & 58.8 & 50.0 & -18.4 & -16.5\\
-Iridaceae & 13 & 2.3 & 0.2 & 531 & 64.4 & 50.0 & 466 & 62.9 & 62.5 & -18.2 & -12.3\\
-\em{Lepismium} & 6 & 100.0 & 0.0 & 6 & 16.7 & 83.3 & 6 & 16.7 & 83.3 & -33.9 & -7.9\\
-\em{Pilosocereus} & 41 & 80.9 & 19.1 & 47 & 55.3 & 73.7 & 46 & 56.5 & 71.1 & -8.5 & -5.8\\
-\em{Tillandsia} & 54 & 11.6 & 6.0 & 464 & 61.4 & 85.2 & 453 & 62.7 & 83.3 & -13.7 & -9.9\\
-\addlinespace
-\em{Tocoyena} & 3 & 13.6 & 4.5 & 22 & 31.8 & 66.7 & 21 & 38.1 & 66.7 & -23.0 & -9.5\\
-\bottomrule
-\end{tabular}
-\end{table}
-\end{landscape}
 
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{./figures_tables/Figure_study_groups_reduced} \caption{Examples for taxa included in this study.
-\textbf{A)} \textit{Pilosocereus pusillibaccatus} (\textit{Pilosocereus}), \textbf{B)} \textit{Conchocarpus macrocarpus} (\textit{Conchocarpus}); \textbf{C)} \textit{Tillandsia recurva} (\textit{Tillandsia}); \textbf{D)} \textit{Oxyrhopus guibei} (Dipsadidae); \textbf{E)} \textit{Aethiopella ricardoi} (Neanuridae); \textbf{F)} \textit{Tocoyena formosa} (\textit{Tocoyena}); \textbf{G)} \textit{Harengula jaguana} (\textit{Harengula}); \textbf{H)} \textit{Gaylussacia decipiens} (\textit{Gaylussacia}); \textbf{I)} \textit{Oocephalus foliosus} (\textit{Oocephalus}); \textbf{J)} \textit{Tityus carvalhoi} (\textit{Tityus}); \textbf{K)} \textit{Prosthechea vespa} (\textit{Prosthechea}), Image credits: A) Pamela Lavor, B) Juliana El-Ottra, C) Eduardo Tomaz, D) Filipe C Serrano, E) Raiane Vital da Paz, F) Fernanda GL Moreira, G) Thais Ferreira-Araujo, H) Luiz Menini Neto, I) Arthur Soares, J) Renata C Santos-Costa, K) Tiago Vieira}\label{fig:species}
+\includegraphics[width=0.9\linewidth]{./figures_tables/Figure_study_groups_reduced} \caption{Examples of taxa included in this study.
+\textbf{A)} \textit{Pilosocereus pusillibaccatus} (\textit{Pilosocereus}), \textbf{B)} \textit{Conchocarpus macrocarpus} (\textit{Conchocarpus}); \textbf{C)} \textit{Tillandsia recurva} (\textit{Tillandsia}); \textbf{D)} \textit{Oxyrhopus guibei} (Dipsadidae); \textbf{E)} \textit{Aethiopella ricardoi} (Neanuridae); \textbf{F)} \textit{Tocoyena formosa} (\textit{Tocoyena}); \textbf{G)} \textit{Harengula jaguana} (\textit{Harengula}); \textbf{H)} \textit{Gaylussacia decipiens} (\textit{Gaylussacia}); \textbf{I)} \textit{Oocephalus foliosus} (\textit{Oocephalus}); \textbf{J)} \textit{Tityus carvalhoi} (\textit{Tityus}); \textbf{K)} \textit{Prosthechea vespa} (\textit{Prosthechea}), Image credits: A) Pamela Lavor, B) Juliana El-Ottra, C) Eduardo Tomaz, D) Filipe C Serrano, E) Raiane Vital da Paz, F) Fernanda GL Moreira, G) Thais Ferreira-Araujo, H) Luiz Menini Neto, I) Arthur Soares, J) Renata C Santos-Costa, K) Tiago Vieira.}\label{fig:species}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\linewidth]{./output/figure_number_of_records} \caption{The difference in records available between the raw and filtered dataset of 18 Neotropical taxa including animals, fungi, and plants, plotted in a 100x100 km grid across the Neotropics.}\label{fig:total}
+\includegraphics[width=\linewidth]{./output/figure_number_of_records} \caption{The difference in records available between the raw and filtered dataset of 18 Neotropical taxa including animals, fungi, and plants, plotted in a 100 x 100 km grid across the Neotropics.}\label{fig:total}
 \end{figure}

### 2020-06-10 latexdiff/old_manuscript.tex
latexdiff

--- 2020-03-02 manuscript_peerj.tex
@@
 Here, we test the effect of automated filtering of species geographic occurrence records on the number of records available in different groups of animals, fungi, and plants.
-Furthermore, we test the impact of automated filtering procedures for the accuracy of automated conservation assessments compared to full IUCN assessments.
+Furthermore, we test the impact of automated filtering procedures for the accuracy of preliminary automated conservation assessments compared to full IUCN assessments.
 Specifically, we evaluate a pipeline of 13 automated filters to flag possibly erroneous records by using 1) record meta-data and 2) geographic gazetteers.
@@
 We downloaded occurrence information for all study groups from www.gbif.org using the \texttt{rgbif} v1.4.0 package (Chamberlain 2017) in R (GBIF.org, 2019a-p,2020a,b).
-We downloaded only records with geographic coordinates and limited the study area to latitudes smaller than 33\(^\circ\) N and longitudes smaller than 35\(^\circ\) W and larger than 120\(^\circ\) W reflecting the Neotropics, our main area of expertise (Morrone 2014).
+We downloaded only records with geographic coordinates and limited the study area to latitudes smaller than 33\(^\circ\) N and longitudes smaller than 35\(^\circ\) W and larger than 120\(^\circ\) W reflecting the Neotropics (Morrone 2014), our main area of expertise.
 The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, Entomobryidae, \emph{Gaylussacia}, Iridaceae, Neanuridae, and \emph{Tillandsia} for which we only obtained the Neotropical occurrences.
@@
 
-Each author provided an informed comment on the performance of the raw and cleaned datasets, respectively concerning the number of flagged records and the accuracy of the overall species richness maps and we compared the agreement between automated conservation assessments based on raw and filtered occurrences with the global IUCN Red List (www.iucn.org).
+Each author provided an informed comment on the performance of the raw and cleaned datasets, concerning the number of flagged records and the accuracy of the overall species richness maps.
+We then compared the agreement between automated conservation assessments based on raw and filtered occurrences with the global IUCN Red List for those taxa where IUCN assessments were available (www.iucn.org).
 
 We did all analyses in the R computing environment (R Core Team 2019), using standard libraries for data handling and visualization (Wickham 2018; Garnier 2018; Ooms 2014, 2019; Hijmans 2019).
-All scripts are available from a zenodo repository (\textbf{DOI}).
+All scripts are available from a zenodo repository (\url{doi:10.5281/zenodo.3695102}).
 
@@
 The most important automated tests were for urban area (8.6\%) and the occurrence from records of terrestrial taxa in the sea and marine taxa on land (4.3\%, see Table 3 and Appendix 1 in the electronic supplement for further details and the absolute numbers).
-Only few records were flagged by the coordinate precision, zero coordinates and biodiversity institution tests (Fig.
+Only a few records were flagged by the coordinate precision, zero coordinates and biodiversity institution tests (Fig.
 \ref{fig:split}).
 
-\emph{Entomobryidae}, \emph{Diogenidae}, and \emph{Neanuridae} had the highest fraction of flagged records (Table 3).
+Entomobryidae, Diogenidae, and Neanuridae had the highest fraction of flagged records (Table 3).
 In general, the different filters we tested were of similar importance for different study groups.
@@
 
-Several reasons may explain the high proportion of records flagged as duplicates.
-First, the deposition of duplicates from the same specimen hosted at different collections is common practice, especially for plants, where a specimen duplication is easily feasible.
-Second, independent collections at similar localities may occur, in particular for local endemics.
-Third, low coordinate precision, for instance based on automated geo-referencing from locality descriptions, may lump records from nearby localities.
-Duplicated records do not represent errors \emph{per se}, but an unnecessary computational burden, which is often a bottleneck for analyses of datasets with millions of records.
-
 The aim of automated filtering is to identify possibly problematic records that are unsuitable for particular downstream analyses.
@@
 \ref{fig:split}) might be overly strict, although we chose a conservative distance for the Political centroid test (1 km).
+
+Several reasons may explain the high proportion of records flagged as duplicates.
+First, the deposition of duplicates from the same specimen hosted at different collections is common practice, especially for plants, where a specimen duplication is easily feasible.
+Second, independent collections at similar localities may occur, in particular for local endemics.
+Third, low coordinate precision, for instance based on automated geo-referencing from locality descriptions, may lump records from nearby localities.
+Duplicated records do not represent errors \emph{per se}, but may constitute an unnecessary computational burden for large-scale biodiversity analyses (Antonelli et al.
+2018), which is often a bottleneck for analyses of datasets with millions of records.
 
@@
 The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700.
-The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae, and \emph{Oocephalus} might be related to horticulture since at least some species in those groups are commonly cultivated as ornamentals.
+The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae and \emph{Oocephalus} might be related to horticulture since at least some species in those groups are commonly cultivated as ornamentals.
 
@@
 For instance, in Entomobryidae, 2,004 samples were marked as material samples, and therefore removed by our global filter retaining only specimen and observation data, which in this case was overly strict.
-It should be noted, that, while in this study we focused on meta-data and geographic filtering, taxonomic cleaning---the resolution of synonymies and identification of accepted names---is another important part of data curation, but depends on taxon-specific taxonomic backbones and synonymy lists which are not readily available for many groups.
+
+It should be noted, that while in this study we focused on meta-data and geographic filtering, taxonomic cleaning---the resolution of synonymies and identification of accepted names---is another important part of data curation, but depends on taxon-specific taxonomic backbones and synonymy lists which are not readily available for many groups.
 
@@
 This might be caused by the structure of our comparison, which only included species that were evaluated by the IUCN Red List (and not considered as Data Deficient) and at the same time had occurrences recorded in GBIF.
-Those inclusion criteria are likely to have, biased the datasets towards species with large ranges, since generally more data for them are available.
-The robustness of automated conservation assessments to data quality is encouraging, although these methods are only an approximation, especially for species with few collection records (Rivers et al.
+Those inclusion criteria are likely to have biased the datasets towards species with large ranges, since generally more data for them are available.
+The robustness of automated conservation assessments to data quality is encouraging, although these methods are only an approximation (and not replacements) of full IUCN Red List assessments, especially for species with few collection records (Rivers et al.
 2011).
@@
 Automated flagging can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the specific taxonomic groups and geographic area of interest.
-The putative errors we encountered point to the importance to train researchers and students to curate species occurrence datasets and to allow users to provide feedback for particular records, thus contributing to overall accuracy and precision of public biodiversity databases.
+The putative errors we encountered point to the importance to train researchers and students to curate species occurrence datasets and to allow users to provide data aggregators such as GBIF with feedback for particular records, thus contributing to overall accuracy and precision of public biodiversity databases.
 
@@
 \item
-  Appendix 1 - Table S1, reporting the absolute number of flagged records per taxonomic group and test
+  Appendix 1 - Absolute number of flagged records per taxonomic group and test
 \item
-  Appendix 2 - Taxon specific richness maps and comments.
+  Appendix 2 - Taxon specific richness maps and comments
 \item
-  Appendix 3 - The full results of the conservation assessment
+  Appendix 3 - Full results of the conservation assessment
 \end{itemize}
-
-\hypertarget{tables}{%
-\section*{Tables}\label{tables}}
-\addcontentsline{toc}{section}{Tables}
-
-\begin{landscape}\begin{table}
-
-\caption{\label{tab:tabletaxa}The study groups and their taxonomy.
-This study includes three marine and 15 terrestrial taxa, seven of them animals, one group of fungi and ten plants, belonging to 16 different orders.}
-\centering
-\fontsize{11}{13}\selectfont
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}ll}
-\toprule
-Taxon & Taxon rank & Realm & Common name & 'Phylum' & Order & Family\\
-\midrule
-Diogenidae & Family & Marine & Hermit crabs & Arthropoda & Decapoda & Diogenidae\\
-Entomobryidae & Family & Terrestrial & Springtails & Arthropoda & Poduromorpha & Entomobryidae\\
-Neanuridae & Family & Terrestrial & Springtails & Arthropoda & Poduromorpha & Neanuridae\\
-\em{Tityus} & Genus & Terrestrial & Scorpions & Arthropoda & Scorpiones & Buthidae\\
-Arhynchobatidae & Family & Marine & Skates & Chordata & Rajiformes & Arhynchobatidae\\
-\addlinespace
-Dipsadidae & Family & Terrestrial & Snakes & Chordata & Squamata & Dipsadidae\\
-\em{Harengula} & Genus & Marine & Herrings & Chordata & Clupeiformes & Clupeidae\\
-\hline
-\em{Thozetella} & Genus & Terrestrial & Sac fungi & Ascomycota & Chaetosphaeriales & Chaetosphaeriaceae\\
-\hline
-\em{Conchocarpus} & Genus & Terrestrial & NA & Angiosperms & Sapindales & Rutaceae\\
-\em{Gaylussacia} & Genus & Terrestrial & Huckleberries & Angiosperms & Ericales & Ericaceae\\
-\addlinespace
-\em{Harpalyce} & Genus & Terrestrial & NA & Angiosperms & Fabales & Fabaceae\\
-Iridaceae & Family & Terrestrial & NA & Angiosperms & Asparagales & Iridaceae\\
-\em{Lepismium} & Genus & Terrestrial & Cacti & Angiosperms & Caryophyllales & Cactaceae\\
-\em{Oocephalus} & Genus & Terrestrial & NA & Angiosperms & Lamiales & Lamiaceae\\
-\em{Pilosocereus} & Genus & Terrestrial & NA & Angiosperms & Caryophyllales & Cactaceae\\
-\addlinespace
-\em{Prosthechea} & Genus & Terrestrial & Orchids & Angiosperms & Asparagales & Orchidaceae\\
-\em{Tillandsia} & Genus & Terrestrial & Bromeliads & Angiosperms & Poales & Bromeliaceae\\
-\em{Tocoyena} & Genus & Terrestrial & NA & Angiosperms & Gentianales & Rubiaceae\\
-\bottomrule
-\end{tabular}
-\end{table}
-\end{landscape}
-
-\begin{table}[!h]
-
-\caption{\label{tab:tabletests}The automated filters used in this study.}
-\centering
-\fontsize{9}{11}\selectfont
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{10cm}}
-\toprule
-Test & Type & Rationale\\
-\midrule
-\rowcolor{gray!6}  Basis of record & Meta-data & Records might be unsuitable or unreliable for certain analyses dependent on their source, e.g. "fossil" or "unknown"\\
-Collection year & Meta-data & Coordinates from old records are more likely to be imprecise or erroneous coordinates since they are derived from  geo-referencing based on the locality description.
-This is more problematic for older records, since names or borders of places may change\\
-\rowcolor{gray!6}  Coordinate precision & Meta-data & Records may be unsuitable for a study if their precision is lower than the study analysis scale\\
-Identification level & Meta-data & Records may be unsuitable if they are not identified to species level.\\
-\rowcolor{gray!6}  Individual count & Meta-data & Records may be unsuitable if the number of recorded individuals is 0 (record of absence) or if the count is to high, as this is often related to records from barcoding or indicative of data entry problems.\\
-\addlinespace
-Biodiversity institutions & Gazetteer-based & Records may have coordinates at the location of biodiversity institutions, e.g.
-because they were erroneously entered with the physical location of the specimen or because they represent individuals from captivity or horticulture\\
-\rowcolor{gray!6}  Duplicates & Gazetteer-based & Duplicated records may add unnecessary computational burden to analyses\\
-Equal lat/lon & Gazetteer-based & Coordinates with equal latitude and longitude are usually indicative of data entry errors\\
-\rowcolor{gray!6}  Capitals & Gazetteer-based & Records may be assigned to the coordinates of country capitals based on a vague locality description\\
-Political centroids & Gazetteer-based & Records may be assigned to the coordinates of the centroids of political entities based on a vague locality description\\
-\addlinespace
-\rowcolor{gray!6}  Sea & Gazetteer-based & Coordinates from terrestrial organisms in the sea are usually indicative of data entry errors, e.g. swapped latitude and longitude\\
-Urban areas & Gazetteer-based & Records from urban areas are not necessarily errors, but often represent imprecise records automatically geo-referenced from vague locality descriptions or old records from different land-use types\\
-\rowcolor{gray!6}  Zeros & Gazetteer-based & Coordinates with plain zeros are often indicative of data entry errors\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{landscape}\begin{table}
-
-\caption{\label{tab:tablecoords}The impact of automated filtering on occurrence records for 18 Neotropical taxa downloaded from www.gbif.org.
-From column four onwards the numbers show the percentage of records flagged by the respective test.
-Only tests that flagged at least 0.1\% of the records in any group are shown.
-Individual records can be flagged by multiple tests, therefore the sum of percentage from all tests can supersede the total percentage.}
-\centering
-\fontsize{9}{11}\selectfont
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}>{\raggedleft\arraybackslash}p{1cm}r}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Summary} & \multicolumn{5}{c}{Meta data filters} & \multicolumn{7}{c}{Gazetteer-based filters} \\
-\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-9} \cmidrule(l{3pt}r{3pt}){10-16}
-Taxon & Total records & Records flagged & Fraction flagged [\%] & Basis of record [\%] & Collection year [\%] & Coordinate precision [\%] & Id-level [\%] & Individual count [\%] & Biodiversity Institutions [\%] & Capitals [\%] & Duplicates [\%] & Political centroids [\%] & Sea/land area [\%] & Urban areas [\%] & Zeros [\%]\\
-\midrule
-Diogenidae & 13,840 & 9,508 & 68.7 & 1.7 & 2.5 & 0.0 & 0.0 & 0.0 & 0.0 & 0.7 & 33.8 & 0.2 & 44.3 & 1.3 & 0.0\\
-Entomobryidae & 2,767 & 2,498 & 90.3 & 72.9 & 2.0 & 0.0 & 72.1 & 0.0 & 0.1 & 0.1 & 85.5 & 0.0 & 0.0 & 70.1 & 0.0\\
-Neanuridae & 689 & 461 & 66.9 & 2.9 & 1.3 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 62.4 & 0.0 & 0.0 & 2.0 & 0.0\\
-\em{Tityus} & 1,018 & 562 & 55.2 & 7.0 & 0.4 & 1.8 & 1.6 & 0.0 & 0.5 & 1.2 & 43.5 & 0.1 & 0.0 & 6.9 & 0.0\\
-Arhynchobatidae & 14,633 & 5,635 & 38.5 & 1.7 & 1.3 & 0.0 & 0.9 & 0.0 & 0.0 & 0.0 & 35.4 & 0.0 & 3.8 & 1.9 & 0.0\\
-\addlinespace
-Dipsadidae & 64,249 & 37,058 & 57.7 & 5.6 & 11.3 & 0.8 & 0.0 & 0.1 & 0.3 & 1.8 & 46.3 & 0.4 & 0.0 & 8.5 & 0.0\\
-\em{Harengula} & 36,697 & 11,384 & 31.0 & 1.0 & 0.4 & 0.0 & 0.3 & 0.0 & 0.0 & 0.2 & 27.0 & 0.1 & 5.5 & 0.2 & 0.0\\
-\hline
-\em{Thozetella} & 51 & 18 & 35.3 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 27.5 & 0.0 & 23.5 & 2.0 & 0.0\\
-\hline
-\em{Conchocarpus} & 1,551 & 670 & 43.2 & 0.5 & 1.9 & 0.1 & 0.0 & 0.0 & 0.1 & 0.0 & 39.6 & 0.9 & 0.4 & 2.3 & 0.0\\
-\em{Gaylussacia} & 3,998 & 1,887 & 47.2 & 0.7 & 4.4 & 0.6 & 0.0 & 0.0 & 0.1 & 3.2 & 41.8 & 1.1 & 0.1 & 5.2 & 0.0\\
-\addlinespace
-\em{Harpalyce} & 870 & 288 & 33.1 & 0.5 & 5.5 & 0.7 & 0.0 & 0.9 & 0.0 & 1.0 & 26.0 & 1.3 & 0.0 & 3.8 & 0.0\\
-Iridaceae & 23,127 & 7,774 & 33.6 & 0.9 & 4.7 & 0.1 & 0.0 & 1.3 & 0.4 & 1.0 & 17.1 & 0.4 & 0.1 & 12.3 & 0.0\\
-\em{Lepismium} & 825 & 245 & 29.7 & 0.0 & 2.1 & 0.0 & 0.0 & 0.0 & 0.0 & 0.1 & 21.9 & 0.1 & 0.0 & 7.8 & 0.0\\
-\em{Oocephalus} & 883 & 435 & 49.3 & 0.0 & 0.7 & 0.3 & 0.0 & 0.1 & 0.0 & 6.1 & 41.9 & 0.8 & 0.0 & 13.3 & 0.0\\
-\em{Pilosocereus} & 1,940 & 501 & 25.8 & 1.8 & 7.0 & 0.0 & 0.0 & 0.9 & 0.2 & 0.5 & 16.8 & 0.5 & 0.0 & 2.1 & 0.0\\
-\addlinespace
-\em{Prosthechea} & 6,617 & 2,088 & 31.6 & 5.0 & 8.3 & 0.1 & 0.0 & 0.2 & 0.0 & 0.4 & 19.6 & 1.7 & 0.0 & 0.9 & 0.1\\
-\em{Tillandsia} & 42,222 & 14,908 & 35.3 & 4.9 & 5.1 & 0.1 & 0.0 & 1.0 & 0.3 & 0.7 & 19.8 & 0.7 & 0.0 & 9.2 & 0.0\\
-\em{Tocoyena} & 2,922 & 1,098 & 37.6 & 0.1 & 1.9 & 0.2 & 0.0 & 0.5 & 0.0 & 0.8 & 32.3 & 0.8 & 0.2 & 5.0 & 0.0\\
-\hline
-Total & 218,899 & 97,018 & 44.3 & 4.2 & 5.6 & 0.3 & 1.0 & 0.4 & 0.2 & 1.0 & 32.3 & 0.4 & 4.0 & 7.1 & 0.0\\
-\bottomrule
-\end{tabular}
-\end{table}
-\end{landscape}
-
-\begin{landscape}\begin{table}
-
-\caption{\label{tab:unnamed-chunk-3}Conservation assessment for 11 Neotropical taxa of plants and animals based on three datasets.
-IUCN: global red list assessment obtained from www.iucn.org; GBIF Raw: Preliminary conservation assessment based on IUCN Criterion B using ConR and the raw dataset from GBIF; GBIF filtered: Preliminary conservation assessment based on IUCN Criterion B using ConR and the filtered dataset.
-Only taxa with at least on species evaluated by IUCN shown.}
-\centering
-\fontsize{9}{11}\selectfont
-\begin{tabular}[t]{l>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{3}{c}{IUCN} & \multicolumn{3}{c}{GBIF Raw} & \multicolumn{5}{c}{GBIF Filtered} \\
-\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7} \cmidrule(l{3pt}r{3pt}){8-12}
-Taxon & n taxa & Evaluated [\%] & Threatened [\%] & n taxa & Threatened [\%] & Match with IUCN [\%] & n taxa & Threatened [\%] & Match with IUCN [\%] & EOO change compared to raw [\%] & AOO change compared to raw [\%]\\
-\midrule
-Arhynchobatidae & 37 & 51.3 & 17.9 & 39 & 35.9 & 45.0 & 39 & 41.0 & 40.0 & -32.7 & -18.5\\
-Dipsadidae & 520 & 68.0 & 8.8 & 638 & 58.3 & 63.0 & 598 & 59.9 & 61.2 & -2.3 & -15.6\\
-\em{Harengula} & 4 & 100.0 & 0.0 & 4 & 0.0 & 100.0 & 4 & 0.0 & 100.0 & -38.0 & -36.9\\
-\hline
-\em{Conchocarpus} & 4 & 8.7 & 0.0 & 46 & 63.0 & 100.0 & 45 & 62.2 & 100.0 & -15.3 & -7.1\\
-\em{Gaylussacia} & 2 & 3.3 & 0.0 & 61 & 59.0 & 50.0 & 58 & 60.3 & 50.0 & -22.5 & -8.6\\
-\addlinespace
-\em{Harpalyce} & 3 & 15.0 & 5.0 & 20 & 65.0 & 66.7 & 17 & 58.8 & 50.0 & -18.4 & -16.5\\
-Iridaceae & 13 & 2.3 & 0.2 & 531 & 64.4 & 50.0 & 466 & 62.9 & 62.5 & -18.2 & -12.3\\
-\em{Lepismium} & 6 & 100.0 & 0.0 & 6 & 16.7 & 83.3 & 6 & 16.7 & 83.3 & -33.9 & -7.9\\
-\em{Pilosocereus} & 41 & 80.9 & 19.1 & 47 & 55.3 & 73.7 & 46 & 56.5 & 71.1 & -8.5 & -5.8\\
-\em{Tillandsia} & 54 & 11.6 & 6.0 & 464 & 61.4 & 85.2 & 453 & 62.7 & 83.3 & -13.7 & -9.9\\
-\addlinespace
-\em{Tocoyena} & 3 & 13.6 & 4.5 & 22 & 31.8 & 66.7 & 21 & 38.1 & 66.7 & -23.0 & -9.5\\
-\bottomrule
-\end{tabular}
-\end{table}
-\end{landscape}
 
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{./figures_tables/Figure_study_groups} \caption{Examples of study taxa.
-\textbf{A)} Pilosocereus pusillibaccatus, picture by Pamela Lavor; \textbf{B)} Conchocarpus macrocarpus, Juliana El-Ottra; \textbf{C)} Tillandsia recurva, EC Tomaz; \textbf{D)} \textit{Oxyrhopus guibei} (Dipsadidae), Filipe C Serrano; \textbf{E)} Neanuridae (\textit{Aethiopella ricardoi}), Raiane Vital da Paz; \textbf{F)} \textit{Tocoyena formosa}, Fernanda GL Moreira; \textbf{G)} \textit{Harengula jaguana}, Thais Ferreira-Araujo; \textbf{H)} \textit{Gaylussacia decipiens}, Luiz Menini Neto; \textbf{I)} \textit{Ooocephalus foliosus}, Arthur Soares; \textbf{J)} \textit{Tityus carvalhoi}, Renata C Santos-Costa; \textbf{K)} \textit{Prostheca vespa}, Tiago Veira}\label{fig:species}
+\includegraphics[width=0.9\linewidth]{./figures_tables/Figure_study_groups} \caption{Examples for taxa included in this study.
+\textbf{A)} \textit{Pilosocereus pusillibaccatus} (\textit{Pilosocereus}), \textbf{B)} \textit{Conchocarpus macrocarpus} (\textit{Conchocarpus}); \textbf{C)} \textit{Tillandsia recurva} (\textit{Tillandsia}); \textbf{D)} \textit{Oxyrhopus guibei} (Dipsadidae); \textbf{E)} \textit{Aethiopella ricardoi} (Neanuridae); \textbf{F)} \textit{Tocoyena formosa} (\textit{Tocoyena}); \textbf{G)} \textit{Harengula jaguana} (\textit{Harengula}); \textbf{H)} \textit{Gaylussacia decipiens} (\textit{Gaylussacia}); \textbf{I)} \textit{Oocephalus foliosus} (\textit{Oocephalus}); \textbf{J)} \textit{Tityus carvalhoi} (\textit{Tityus}); \textbf{K)} \textit{Prosthechea vespa} (\textit{Prosthechea}), Image credits: A) Pamela Lavor, B) Juliana El-Ottra, C) Eduardo Tomaz, D) Filipe C Serrano, E) Raiane Vital da Paz, F) Fernanda GL Moreira, G) Thais Ferreira-Araujo, H) Luiz Menini Neto, I) Arthur Soares, J) Renata C Santos-Costa, K) Tiago Vieira}\label{fig:species}
 \end{figure}

### 2020-06-10 manuscript_peerj.Rmd
latexdiff

@@
   Species occurrence records provide the basis for many biodiversity studies.
-They derive from geo-referenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications.
+They derive from georeferenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications.
 Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern.
 Automatic filtering is a scalable and reproducible means to identify potentially problematic records and tailor datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org), for biodiversity analyses.
-However, it is unclear how much data may be lost by filtering, whether the same filters should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses.
-Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 18 Neotropical taxa including animals, fungi, and plants, terrestrial and marine, downloaded from GBIF.
+However, it is unclear how much data may be lost by filtering, whether the same filters should be applied across all taxonomic groups, and what the effect of filtering is on common downstream analyses.
+Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 18 Neotropical taxa, including terrestrial and marine animals, fungi, and plants downloaded from GBIF.
 We find that a total of 44.3% of the records are potentially problematic, with large variation across taxonomic groups (25 - 90%).
-A small fraction of records was identified as erroneous in the strict sense (4.2%), the majority of filtered records that were rather unfit for downstream analyses (41.7%).
-Filters of duplicated information, collection year, basis of record as well as urban areas and coordinates for terrestrial taxa in the sea or marine taxa on land have the greatest effect.
+A small fraction of records was identified as erroneous in the strict sense (4.2%), and a much larger proportion as unfit for most downstream analyses (41.7%).
+Filters of duplicated information, collection year, and basis of record, as well as coordinates in urban areas, or for terrestrial taxa in the sea or marine taxa on land, have the greatest effect.
 Automated filtering can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus.
-Our results stress the importance of thorough exploration of the meta-data associated with species records for biodiversity research.
+Our results stress the importance of thorough recording and exploration of the meta-data associated with species records for biodiversity research.
 bibliography: course_natal.bib
@@
     base_format: rticles::peerj_article # for using bookdown features like \@ref()
+    keep_tex: true
   rticles::peerj_article: default
@@
 Publicly available species distribution data have become a crucial resource in biodiversity research, including studies in ecology, biogeography, systematics and conservation biology.
-In particular, the availability of digitized collections from museums and herbaria, and citizen science observations has increased drastically over the last few years.
+In particular, the availability of digitized collections from museums and herbaria and citizen science observations has increased drastically over the last few years.
 As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.4 billion geo-referenced occurrence records for species from across the globe and the tree of life.
 
-A central challenge to the use of these publicly available species occurrence data in research are problematic geographic coordinates, which are either erroneous or unfit for downstream analyses [for instance because they are overly imprecise, @Anderson2016].
-Problems mostly arise because public databases integrate records collected with different methodologies in different places, at different times; often without centralized curation and only rudimentary meta-data.
-For instance, problematic coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common [@Maldonado2015; @Yesson2007] and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in captivity [@Zizka2019].
+A central challenge to the use of these publicly available species occurrence data in research is problematic geographic coordinates, which are either erroneous or unfit for downstream analyses [for instance because they are overly imprecise, @Anderson2016].
+Problems mostly arise because data aggregators such as GBIF integrate records collected with different methodologies in different places at different times--- often without centralized curation and only rudimentary meta-data.
+For instance, problematic coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common [@Maldonado2015; @Yesson2007] and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in cultivation or captivity [@Zizka2019].
 
@@
 Those methods are usually based on geographic gazetteers [e.g., @Chamberlain2016; @Zizka2019; @Jin2020] or on additional data, such as environmental variables [@Robertson2016].
-Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the suitability of publicly available occurrence records for biodiversity research [@Zizka2019].
+Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions, have been proposed to improve the suitability of publicly available occurrence records for biodiversity research [@Zizka2019].
 
@@
 These methods have been used to propose preliminary global [@Stevart2019; @Zizka2020b] and regional [@Schmidt2017; @Cosiaux2018] Red List assessments.
-However, all metrics and especially EOO, are sensitive to individual records with problematic coordinates.
-Automated conservation assessments may therefore be biased, particularly if the number of records are few, as it is the case for many tropical species.
+However, all metrics, and especially EOO, are sensitive to individual records with problematic coordinates.
+Automated conservation assessments may therefore be biased, particularly if the number of records is low, as it is the case for many tropical species.
 
@@
 ## Choice of study taxa {-}
-This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering students and researchers working with different taxonomic groups of animals, fungi, and plants across the Neotropics.
+This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 which gathered students and researchers working with different taxonomic groups of animals, fungi, and plants across the Neotropics.
 Each participant analysed geographic occurrence data from their taxonomic group of interest and commented on the results for their group.
-Hence, we include groups based on the expertise of the participants rather than following some arbitrary choice of taxa and taxonomic ranks.
-We acknowledge a varying degree in documented expertise and number of years working on each group.
+Hence, we include groups based on the expertise of the participants rather than following an arbitrary choice of taxa and taxonomic ranks.
+We acknowledge a varying degree of documented expertise and number of years working on each group.
 We obtained public occurrence records for 18 taxa, including one plant family, nine plant genera, one genus of fungi, three families and one genus of terrestrial arthropods, one family of snakes, one family of skates, and one genus of bony fish (Table 1).
@@
 We downloaded GBIF-interpreted data including only records with geographic coordinates and limited the study area to a rectangle between 90$^\circ$S - 33$^\circ$ N and 35$^\circ$ W - 120$^\circ$ W reflecting the Neotropics [@Morrone2014], our main area of expertise.
-The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, Entomobryidae, *Gaylussacia*, Iridaceae, Neanuridae, and *Tillandsia* for which we only obtained the Neotropical occurrences.
-We consider GBIF data generally of high quality and use them as study case because GBIF is the largest, most widely-used and taxonomically most comprehensive data source for species occurrence records, but many more exist [e.g., https://bien.nceas.ucsb.edu/bien/, www.fishbase.de or @Guedes2018].
-The variety of existing databases is also the reason why we download data without accounting for GBIF-internal quality testing ("Geospatial issues"), since quality control mechanisms might differ among databases and sometimes no quality control might be done (for instance in *de novo* databases compiled by groups of researchers).
+The natural distributions of all included taxa are confined to the Neotropics except for Arhynchobatidae, Diogenidae, Dipsadidae, Entomobryidae, *Gaylussacia*, Iridaceae, Neanuridae, and *Tillandsia*, for which we only obtained the Neotropical occurrences.
+We consider GBIF data generally of high quality and use them as a case study because GBIF is the largest, most widely used and taxonomically most comprehensive data source for species occurrence records; however many more exist [e.g., https://bien.nceas.ucsb.edu/bien/, www.fishbase.de or @Guedes2018].
+The variety of existing databases is also the reason why we downloaded data without accounting for GBIF-internal quality testing ("Geospatial issues"), since quality control mechanisms might differ among databases and sometimes quality control might not be carried out (for instance in *de novo* databases compiled by groups of researchers).
 
 ##  Automated cleaning {-}
-We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the raw data downloaded from GBIF ("raw", hereafter) using meta-data for those records for which they were available [although this is often not the case, @Peterson2018], removing: 1) records with a coordinate precision below 100 km (as this represents the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99.
-Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e., test for duplicated records).
+We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the raw data downloaded from GBIF ("raw", hereafter) using meta-data for those records for which they were available [although meta-data were often missing, @Peterson2018], removing: 1) records with a coordinate precision below 100 km (as this represents the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99.
+Furthermore, we rounded the geographic coordinates to four decimal places and retained only one record per species per location (i.e., test for duplicated records).
 In a second step, we used the `clean_coordinates` function of the `CoordinateCleaner v2.0-11` package [@Zizka2019] with default options to flag errors that are common to biological data sets (“filtered”, hereafter).
@@
 While these assessments are preliminary [see @IUCN2017], they can be a proxy used by the IUCN to speed up full assessments.
-We then benchmarked the preliminary conservation assessments against the global IUCN Red List assessments as far as available for the same taxa, which we obtained from www.iucn.org via the `rredlist v.0.5.0` package [@Chamberlain2018].
+We then benchmarked the preliminary conservation assessments against the global IUCN Red List assessments for the same taxa (where available), which we obtained from www.iucn.org via the `rredlist v.0.5.0` package [@Chamberlain2018].
 
@@
 
-We did all analyses in the R computing environment [@rcoreteam2019], using standard libraries for data handling and visualization [@Wickham2018; @Garnier2018; @Ooms2014; @Ooms2019; @Hijmans2019].
+We carried out all analyses in the R computing environment [@rcoreteam2019], using standard libraries for data handling and visualization [@Wickham2018; @Garnier2018; @Ooms2014; @Ooms2019; @Hijmans2019].
 All scripts are available from a zenodo repository (doi:10.5281/zenodo.3695102).
@@
 In general, the different filters we tested were of similar importance for different study groups.
-There were few outstanding exceptions, including the particularly high proportions of records filtered on the "basis of record test" for *Tityus* (7.0%), Dipsadidae (5.6%), *Prosthechea* (5.0%) and *Tillandsia* (4.9%), by the collection year for Dipsadidae (11.3%), by the taxonomic identification level for  *Tityus* (1.6%), by the capital coordinates for *Oocephalus* (6.1%) and *Gaylussacia* (3.2%), by the seas/land test for Diogenidae and *Thozetella*, and by the urban areas test for *Oocephalus* (13.3%) and Iridaceae (12.3%).
-Furthermore, Entomobryidae differed considerably from all other study taxa with exceptionally high numbers of records filtered by the "basis of record, "level of identification' and urban areas" tests.
+There were few outstanding exceptions, including the particularly high proportions of records filtered by the "basis of record test" for *Tityus* (7.0%), Dipsadidae (5.6%), *Prosthechea* (5.0%) and *Tillandsia* (4.9%), by the collection year for Dipsadidae (11.3%), by the taxonomic identification level for  *Tityus* (1.6%), by the capital coordinates for *Oocephalus* (6.1%) and *Gaylussacia* (3.2%), by the seas/land test for Diogenidae and *Thozetella*, and by the urban areas test for *Oocephalus* (13.3%) and Iridaceae (12.3%).
+Furthermore, Entomobryidae differed considerably from all other study taxa with exceptionally high numbers of records filtered by the "basis of record", "level of identification" and "urban areas" tests.
 
-Geographically, the records filtered by the "basis of record" and "individual count" tests were concentrated in Central America and southern North America, and a relatively high number of records filtered due to their proximity to the centroids of political entities were located on Caribbean islands (Fig.
+Geographically, the records filtered by the "basis of record" and "individual count" tests were concentrated in Central America and southern North America, and a relatively high number of records were filtered due to their proximity to the centroids of political entities were located on Caribbean islands (Fig.
 \ref{fig:split}).
@@
 The median percentage of species per study group with an IUCN assessment was `r tab3 %>% select(iucn_frac_eval) %>% unlist() %>%  median()`%.
-A total of `r aa %>% filter(!is.na(iucn_category)) %>%  filter(iucn_category != "DD") %>%filter(iucn_category != "EX") %>%  mutate(test = iucn_status == "possibly_threatened") %>% select(test) %>% unlist() %>%  sum()` species were listed as Threatened by the IUCN global Red List (CR = `r aa %>% filter(iucn_category == "CR") %>% nrow()`, EN = `r aa %>% filter(iucn_category == "EN") %>% nrow()`, VU = `r aa %>% filter(iucn_category == "VU") %>% nrow()`) and `r  aa %>% filter(!is.na(iucn_category)) %>%  filter(iucn_category != "DD") %>%filter(iucn_category != "EX") %>%  mutate(test = iucn_status == "not_threatened") %>% select(test) %>% unlist() %>%  sum()` as Not Threatened.
+A total of `r aa %>% filter(!is.na(iucn_category)) %>%  filter(iucn_category != "DD") %>%filter(iucn_category != "EX") %>%  mutate(test = iucn_status == "possibly_threatened") %>% select(test) %>% unlist() %>%  sum()` species were listed as *Threatened* by the IUCN global Red List (CR = `r aa %>% filter(iucn_category == "CR") %>% nrow()`, EN = `r aa %>% filter(iucn_category == "EN") %>% nrow()`, VU = `r aa %>% filter(iucn_category == "VU") %>% nrow()`) and `r  aa %>% filter(!is.na(iucn_category)) %>%  filter(iucn_category != "DD") %>%filter(iucn_category != "EX") %>%  mutate(test = iucn_status == "not_threatened") %>% select(test) %>% unlist() %>%  sum()` as *Not Threatened*.
 
@@
 The automated assessment based on the filtered dataset agreed with the IUCN assessment for identifying species as possibly threatened (CR, EN, VU) for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status)) %>% mutate(agreement = iucn_status == flag_status) %>% select(agreement) %>% unlist() %>% sum() ` species (`r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status))  %>% mutate(agreement = iucn_status == flag_status) %>% select(agreement) %>% unlist() %>% mean() %>%  round(3) * 100`%; Table 4).
-Filtering reduced the EOO by `r tab3 %>%  select(flag_median_EOO_change) %>%  unlist() %>%  median()`% and the AOO by `r tab3 %>%  select(flag_median_AOO_change) %>%  unlist() %>%  median()`% on median per group.
+Filtering reduced the EOO by `r tab3 %>%  select(flag_median_EOO_change) %>%  unlist() %>%  median() %>%  abs()`% and the AOO by `r tab3 %>%  select(flag_median_AOO_change) %>%  unlist() %>%  median() %>% abs()`% on median per group.
 For the raw dataset the agreement with IUCN was higher at `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% mutate(agreement = iucn_status == raw_status) %>% select(agreement) %>% unlist() %>% sum() ` species (`r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% mutate(agreement = iucn_status == raw_status) %>% select(agreement) %>% unlist() %>% mean() %>%  round(3) * 100`%).
@@
 <!-- add some subheadings to discussion -->
-Automated flagging based on meta-data and automatic tests filtered on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% of the records per taxonomic group; `r min(tab2$total_unfit)`% - `r max(tab2$total_unfit)`% as "unfit" and `r min(tab2$total_error)`% - `r max(tab2$total_error)`% as "erroneous".
+Automated flagging based on meta-data and automatic tests filtered on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% of the records per taxonomic group; `r min(tab2$total_unfit)`%-`r max(tab2$total_unfit)`% as "unfit" and `r min(tab2$total_error)`%-`r max(tab2$total_error)`% as "erroneous".
 The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (**Question 1**).
-The importance of different tests was similar across taxonomic groups, with exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (**Question 2**).
-The results for species richness were similar between the raw and filtered data with some improvements by the filters.
+The importance of different tests was similar across taxonomic groups, with particular exceptions for the tests on basis of record, collection year, capital coordinates, and urban areas (**Question 2**).
+The results for species richness were similar between the raw and filtered data with some improvements using the filters.
 We found little impact of filtering on the accuracy of the automated conservation assessments (**Question 3**).
@@
 The aim of automated filtering is to identify possibly problematic records that are unsuitable for particular downstream analyses.
-While those records filtered as "erroneous" will likely cause problems for most biodiversity research, those filtered as "unfit" might have varying impact, depending on the type and spatial resolution of the downstream analyses, ranging from an unnecessary computational burden which can be a bottle neck for large-scale analyses [i.e.
-duplicates, @Antonelli2018] over increasing uncertainty (due to low precision) to compromising results.
-For instance, records assigned to country centroids might be acceptable for inter-continental comparisons, but likely are erroneous for species distribution modelling on a local scale.
+While those records filtered as "erroneous" will likely cause problems for most biodiversity research, those filtered as "unfit" might have varying impact, depending on the type and spatial resolution of the downstream analyses.
+Unwanted effects include an unnecessary computational burden, which can be a bottleneck for large-scale analyses [i.e.
+duplicates, @Antonelli2018], and increased uncertainty (due to low precision), or completely compromising results.
+For instance, records assigned to country centroids might be acceptable for inter-continental comparisons, but are likely to be erroneous for species distribution modelling on a local scale.
 The importance of each test and the linked thresholds must be judged based on the specific downstream analyses.
-As our results show, adapting automated tests to the geographic study area or the taxonomic study group can be warranted.
+As our results show, it may be advisable to adapt automated tests to the geographic study area or the taxonomic study group.
 For instance, the high number of records flagged for centroids on the Lesser Antilles (Fig.
@@
 Some notable exceptions stress the need to adapt each filter to the taxonomic study group to balance data quality and data availability.
-The high fraction of records filtered by the "basis of record filter" for *Tityus*, Dipsadidae, *Prosthechea* and *Tillandsia*, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups.
-The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection effort already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700.
-The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700.
-The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae and *Oocephalus* might be related to horticulture since at least some species in those groups are commonly cultivated as ornamentals.
+The high fraction of records filtered by the "basis of record" filter for *Tityus*, Dipsadidae, *Prosthechea* and *Tillandsia*, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups.
+The high fraction of records flagged by the "collection year" filter for Dispadidae was caused by a high collection effort in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700.
+The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institution's code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700.
+The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae and *Oocephalus* might be related to horticulture, since at least some species in those groups are commonly cultivated as ornamentals.
 
-The general agreement between the species richness maps based on raw and filtered data was encouraging, for the use of this data for large-scale biogeographic research (Fig.
+The general agreement between the species richness maps based on raw and filtered data was encouraging, in terms of the use of this data for large-scale biogeographic research (Fig.
 \ref{fig:speciesrichness}, Appendix 2).
@@
 Additionally automated filters might be overly strict  or unsuitable for certain taxa.
-For instance, in Entomobryidae, `r dat %>% filter(taxon == "Entomobryidae") %>% filter( basisOfRecord == "MATERIAL_SAMPLE") %>% nrow()` samples were marked as material samples, and therefore removed by our global filter retaining only specimen and observation data, which in this case was overly strict.
+For instance, in Entomobryidae, `r dat %>% filter(taxon == "Entomobryidae") %>% filter( basisOfRecord == "MATERIAL_SAMPLE") %>% nrow()` samples were marked as material samples and therefore removed by our global filter retaining only specimen and observation data, which in this case was overly strict.
 
@@
 In fact, the internal quality control of GBIF does flag some of the problems we tested for (i.e., zero coordinates, equal lat/lon) while others might be implemented in the near future (country centroids, https://data-blog.gbif.org/post/country-centroids/, https://data-blog.gbif.org/post/country-centroids/).
-While this internal quality is very helpful, we see a huge potential to overcome issues with data quality in a user-feedback system allowing users to provide expert assessments, i.e.
+While this internal quality is very helpful, we see a huge potential to overcome issues with data quality in a user-feedback system that allows users to provide expert assessments, i.e.
 a meta-annotation of records being challenged (and why).
@@
 
-As a next step for automated filtering, it is possible to detect additional problematic records with the help of external species distribution information, which is available for some taxa for example from www.iucn.org (vertebrates) or https://wcsp.science.kew.org/ (selected seed plant families).
+As a next step for automated filtering, it is possible to detect additional problematic records with the help of external species distribution information, which is available for some taxa, for example from www.iucn.org (vertebrates) or https://wcsp.science.kew.org/ (selected seed plant families).
 Furthermore, while in this study we focused on meta-data and geographic filtering, taxonomic cleaning---the resolution of synonymies and identification of accepted names---is another important part of data curation, but depends on taxon-specific taxonomic backbones and synonymy lists which are not readily available for many groups and often are contradictory within individual taxa.
@@
 This might be caused by the structure of our comparison, which only included species that were evaluated by the IUCN Red List (and not considered as *Data Deficient*) and at the same time had occurrences recorded in GBIF.
-Those inclusion criteria are likely to have biased the datasets towards species with large ranges, since generally more data for them are available.
+Those inclusion criteria are likely to have biased the datasets towards species with large ranges, since generally more data are available for them.
 The robustness of automated conservation assessments to data quality is encouraging, although these methods are only an approximation (and not replacements) of full IUCN Red List assessments, especially for species with few collection records [@Rivers2011].
@@
 # Conclusions {-}
-Our results suggest that between one to two thirds of the occurrence records obtained from public databases might be unsuitable for downstream biodiversity analyses.
-While the majority of these records might not be erroneous *per se*, they might be overly imprecise and thereby increase uncertainty of downstream results or add computational burden on big data analyses.
+Our results suggest that between one to half of the occurrence records obtained from GBIF might be unsuitable for downstream biodiversity analyses.
+While the majority of these records might not be erroneous *per se*, they are overly imprecise and thereby increase uncertainty of downstream results or add computational burden on big data analyses.
 
-While our results suggest that large-scale species richness patterns and automated conservation assessments are largely robust to problematic occurrence records, they also stress the importance of (meta-)data exploration prior to most biodiversity analyses.
+While our results suggest that large-scale species richness patterns and automated conservation assessments are largely resilient to the effects of problematic occurrence records, they also stress the importance of (meta-)data exploration prior to most biodiversity analyses.
 Automated filtering can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the specific taxonomic groups and geographic area of interest.
@@
 # Acknowledgements {-}
+We thank GBIF and all data collectors and contributors for their excellent work.
+We tahnk Town Peterson, Roderic Page and one anonymous reviewer for the helpful comments on an earlier version of this mansucript.
 This study enrolled participants of the workshop "Biodiversity data: from field to yield" led by Alice Calvente, Fernanda Carvalho, Alexander Zizka, and Alexandre Antonelli through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI.
@@
 
-- Appendix 2 - Taxon specific richness maps and comments
+- Appendix 2 - Taxon-specific richness maps and comments
 
@@
   mutate(phylopic = sprintf('\\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/%s.png}', phylopic)) %>% 
+  mutate(phylopic = ifelse(grepl("harengula", phylopic), gsub("scale=0.25", "scale=0.18", phylopic), phylopic)) %>%  
+  mutate(phylopic = ifelse(grepl("neanuridae", phylopic), gsub("scale=0.25", "scale=0.2", phylopic), phylopic)) %>% 
+  mutate(phylopic = ifelse(grepl("tocoyena", phylopic), gsub("scale=0.25", "scale=0.2", phylopic), phylopic)) %>% 
   dplyr::select(phylopic, everything()) %>% 
@@
         caption = "The study groups and their taxonomy.
-This study includes three marine and 15 terrestrial taxa, seven of them animals, one group of fungi and ten plants, belonging to 16 different orders.")%>%
+This study includes three marine and 15 terrestrial taxa, seven of them animals, one group of fungi and ten plants, belonging to 16 different orders.
+The outlines illustrate the broad taxonomic group.
+Pictures from www.phylopic.org.
+Phylopic credits: Diogenidae (Ekaterina Kopeykina (vectorized by T.
+Michael Keesey), Creative Commons Attribution-NonCommercial-ShareAlike 3.0; Entomobryidae and Neanuridae: Birgit Lang, Creative Commons Attribution 3.0; Tityus: Gareth Monger, Creative Commons Attribution 3.0; \\textit{Gaylussacia}, \\textit{Lepismium}, \\textit{Pilosocereus}: T.
+Michael Keesey (vectorization) and Nadiatalent (photography), Creative Commons Attribution-ShareAlike 3.0; \\textit{Oocephalus}: Alexander Schmidt-Lebhuhn, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; \\textit{Tillandsia}: Maija Karala, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; all others in the public domain.")%>%
   #landscape()%>%
@@
 Only tests that flagged at least 0.1\\% of the records in any group are shown.
-Individual records can be flagged by multiple tests, therefore the sum of percentage from all tests can supersede the total percentage.", format.args = list(big.mark = ","))%>%
+Individual records can be flagged by multiple tests, therefore the sum of percentages from all tests can supersede the total percentage.", format.args = list(big.mark = ","))%>%
   landscape()%>%
@@
         caption = "Conservation assessment for 11 Neotropical taxa of plants and animals based on three datasets.
-IUCN: global red list assessment obtained from www.iucn.org; GBIF Raw: Preliminary conservation assessment based on IUCN Criterion B using ConR and the raw dataset from GBIF; GBIF filtered: Preliminary conservation assessment based on IUCN Criterion B using ConR and the filtered dataset.
-Only taxa with at least on species evaluated by IUCN shown.",
+IUCN: global red list assessment obtained from www.iucn.org; GBIF Raw: preliminary conservation assessment based on IUCN Criterion B using ConR and the raw dataset from GBIF; GBIF filtered: preliminary conservation assessment based on IUCN Criterion B using ConR and the filtered dataset.
+Only taxa with at least one species evaluated by IUCN shown.",
       format.args = list(big.mark = ","))%>%
@@
 <!-- The study species-->
-```{r species, out.width="0.9\\linewidth", fig.cap=c('Examples for taxa included in this study.
-\\textbf{A)} \\textit{Pilosocereus pusillibaccatus} (\\textit{Pilosocereus}), \\textbf{B)} \\textit{Conchocarpus macrocarpus} (\\textit{Conchocarpus}); \\textbf{C)} \\textit{Tillandsia recurva} (\\textit{Tillandsia}); \\textbf{D)} \\textit{Oxyrhopus guibei} (Dipsadidae); \\textbf{E)} \\textit{Aethiopella ricardoi} (Neanuridae); \\textbf{F)} \\textit{Tocoyena formosa} (\\textit{Tocoyena}); \\textbf{G)} \\textit{Harengula jaguana} (\\textit{Harengula}); \\textbf{H)} \\textit{Gaylussacia decipiens} (\\textit{Gaylussacia}); \\textbf{I)} \\textit{Oocephalus foliosus} (\\textit{Oocephalus}); \\textbf{J)} \\textit{Tityus carvalhoi} (\\textit{Tityus}); \\textbf{K)} \\textit{Prosthechea vespa} (\\textit{Prosthechea}), Image credits: A) Pamela Lavor, B) Juliana El-Ottra, C) Eduardo Tomaz, D) Filipe C Serrano, E) Raiane Vital da Paz, F) Fernanda GL Moreira, G) Thais Ferreira-Araujo, H) Luiz Menini Neto, I) Arthur Soares, J) Renata C Santos-Costa, K) Tiago Vieira'), echo=FALSE}
+```{r species, out.width="0.9\\linewidth", fig.cap=c('Examples of taxa included in this study.
+\\textbf{A)} \\textit{Pilosocereus pusillibaccatus} (\\textit{Pilosocereus}), \\textbf{B)} \\textit{Conchocarpus macrocarpus} (\\textit{Conchocarpus}); \\textbf{C)} \\textit{Tillandsia recurva} (\\textit{Tillandsia}); \\textbf{D)} \\textit{Oxyrhopus guibei} (Dipsadidae); \\textbf{E)} \\textit{Aethiopella ricardoi} (Neanuridae); \\textbf{F)} \\textit{Tocoyena formosa} (\\textit{Tocoyena}); \\textbf{G)} \\textit{Harengula jaguana} (\\textit{Harengula}); \\textbf{H)} \\textit{Gaylussacia decipiens} (\\textit{Gaylussacia}); \\textbf{I)} \\textit{Oocephalus foliosus} (\\textit{Oocephalus}); \\textbf{J)} \\textit{Tityus carvalhoi} (\\textit{Tityus}); \\textbf{K)} \\textit{Prosthechea vespa} (\\textit{Prosthechea}), Image credits: A) Pamela Lavor, B) Juliana El-Ottra, C) Eduardo Tomaz, D) Filipe C Serrano, E) Raiane Vital da Paz, F) Fernanda GL Moreira, G) Thais Ferreira-Araujo, H) Luiz Menini Neto, I) Arthur Soares, J) Renata C Santos-Costa, K) Tiago Vieira.'), echo=FALSE}
 knitr::include_graphics(path = "./figures_tables/Figure_study_groups_reduced.jpg")
@@
 <!-- A map with number of species and where they have been removed -->
-```{r total, out.width="\\linewidth", fig.cap=c('The difference in records available between the raw and filtered dataset of 18 Neotropical taxa including animals, fungi, and plants, plotted in a 100x100 km grid across the Neotropics.'), echo=FALSE}
+```{r total, out.width="\\linewidth", fig.cap=c('The difference in records available between the raw and filtered dataset of 18 Neotropical taxa including animals, fungi, and plants, plotted in a 100 x 100 km grid across the Neotropics.'), echo=FALSE}
 knitr::include_graphics(path = "./output/figure_number_of_records.jpg")

### 2020-06-14 manuscript_peerj.Rmd
manuscript resubmission

@@
 nocite: |
-    @GBIForg2019; @GBIForg2019a; @GBIForg2019b; @GBIForg2019c; @GBIForg2019d; @GBIForg2019e; @GBIForg2019f; @GBIForg2019g; @GBIForg2019h; @GBIForg2019i; @GBIForg2019j; @GBIForg2019k; @GBIForg2019l; @GBIForg2019m; @GBIForg2019n; @GBIForg2019o;@GBIForg2019p;@GBIForg2020, @GBIForg2020a
+    @GBIForg2019; @GBIForg2019a; @GBIForg2019b; @GBIForg2019c; @GBIForg2019d; @GBIForg2019e; @GBIForg2019f; @GBIForg2019g; @GBIForg2019h; @GBIForg2019i; @GBIForg2019j; @GBIForg2019k; @GBIForg2019l; @GBIForg2019m; @GBIForg2019n; @GBIForg2019o;@GBIForg2019p;@GBIForg2020, @GBIForg2020a, Goldblatt2008
 ---
@@
 The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae and *Oocephalus* might be related to horticulture, since at least some species in those groups are commonly cultivated as ornamentals.
+This was supported by the detailed examination of the data for Iridaceae, which showed that after filtering 1605 records from 69 exotic species remained in the dataset, stressing the importance to address these species in certain taxonomic groups.
 
@@
 We thank GBIF and all data collectors and contributors for their excellent work.
-We tahnk Town Peterson, Roderic Page and one anonymous reviewer for the helpful comments on an earlier version of this mansucript.
+We thank Town Peterson, Roderic Page and one anonymous reviewer for the helpful comments on an earlier version of this manuscript.
 This study enrolled participants of the workshop "Biodiversity data: from field to yield" led by Alice Calvente, Fernanda Carvalho, Alexander Zizka, and Alexandre Antonelli through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI.
@@
   mutate(phylopic = ifelse(grepl("tocoyena", phylopic), gsub("scale=0.25", "scale=0.2", phylopic), phylopic)) %>%
+  mutate(phylopic = ifelse(grepl("iridaceae", phylopic), gsub("scale=0.25", "scale=0.075", phylopic), phylopic)) %>% 
   dplyr::select(phylopic, everything()) %>% 
@@
 Michael Keesey), Creative Commons Attribution-NonCommercial-ShareAlike 3.0; Entomobryidae and Neanuridae: Birgit Lang, Creative Commons Attribution 3.0; Tityus: Gareth Monger, Creative Commons Attribution 3.0; \\textit{Gaylussacia}, \\textit{Lepismium}, \\textit{Pilosocereus}: T.
-Michael Keesey (vectorization) and Nadiatalent (photography), Creative Commons Attribution-ShareAlike 3.0; \\textit{Oocephalus}: Alexander Schmidt-Lebhuhn, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; \\textit{Tillandsia}: Maija Karala, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; all others in the public domain.")%>%
+Michael Keesey (vectorization) and Nadiatalent (photography), Creative Commons Attribution-ShareAlike 3.0; \\textit{Iridaceae} adapted from Goldblatt and Manning (2008); \\textit{Oocephalus}: Alexander Schmidt-Lebhuhn, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; \\textit{Tillandsia}: Maija Karala, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; all others in the public domain.")%>%
   #landscape()%>%

### 2020-06-14 manuscript_peerj.tex
manuscript resubmission

@@
 Some notable exceptions stress the need to adapt each filter to the taxonomic study group to balance data quality and data availability.
-The high fraction of records filtered by the ``basis of record filter'' for \emph{Tityus}, Dipsadidae, \emph{Prosthechea} and \emph{Tillandsia}, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups.
-The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection effort already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700.
-The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700.
-The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae and \emph{Oocephalus} might be related to horticulture since at least some species in those groups are commonly cultivated as ornamentals.
+The high fraction of records filtered by the ``basis of record'' filter for \emph{Tityus}, Dipsadidae, \emph{Prosthechea} and \emph{Tillandsia}, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups.
+The high fraction of records flagged by the ``collection year'' filter for Dispadidae was caused by a high collection effort in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700.
+The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institution's code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700.
+The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae and \emph{Oocephalus} might be related to horticulture, since at least some species in those groups are commonly cultivated as ornamentals.
+This was supported by the detailed examination of the data for Iridaceae, which showed that after filtering 1605 records from 69 exotic species remained in the dataset, stressing the importance to address these species in certain taxonomic groups.
 
-The general agreement between the species richness maps based on raw and filtered data was encouraging, for the use of this data for large-scale biogeographic research (Fig.
+The general agreement between the species richness maps based on raw and filtered data was encouraging, in terms of the use of this data for large-scale biogeographic research (Fig.
 \ref{fig:speciesrichness}, Appendix 2).
@@
 
+We thank GBIF and all data collectors and contributors for their excellent work.
+We thank Town Peterson, Roderic Page and one anonymous reviewer for the helpful comments on an earlier version of this manuscript.
 This study enrolled participants of the workshop ``Biodiversity data: from field to yield'' led by Alice Calvente, Fernanda Carvalho, Alexander Zizka, and Alexandre Antonelli through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI.
@@
 \caption{\label{tab:tabletaxa}The study groups and their taxonomy.
-This study includes three marine and 15 terrestrial taxa, seven of them animals, one group of fungi and ten plants, belonging to 16 different orders.}
+This study includes three marine and 15 terrestrial taxa, seven of them animals, one group of fungi and ten plants, belonging to 16 different orders.
+The outlines illustrate the broad taxonomic group.
+Pictures from www.phylopic.org.
+Phylopic credits: Diogenidae (Ekaterina Kopeykina (vectorized by T.
+Michael Keesey), Creative Commons Attribution-NonCommercial-ShareAlike 3.0; Entomobryidae and Neanuridae: Birgit Lang, Creative Commons Attribution 3.0; Tityus: Gareth Monger, Creative Commons Attribution 3.0; \textit{Gaylussacia}, \textit{Lepismium}, \textit{Pilosocereus}: T.
+Michael Keesey (vectorization) and Nadiatalent (photography), Creative Commons Attribution-ShareAlike 3.0; \textit{Iridaceae} adapted from Goldblatt and Manning (2008); \textit{Oocephalus}: Alexander Schmidt-Lebhuhn, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; \textit{Tillandsia}: Maija Karala, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; all others in the public domain.}
 \centering
@@
 \includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/entomobryidae.png} & Entomobryidae & Family & Terrestrial & Springtails & Arthropoda & Entomobryidae\\
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/neanuridae.png} & Neanuridae & Family & Terrestrial & Springtails & Arthropoda & Neanuridae\\
+\includegraphics[valign=T, scale=0.2, raise= 2mm]{input/phylopics/converted/neanuridae.png} & Neanuridae & Family & Terrestrial & Springtails & Arthropoda & Neanuridae\\
 \includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/tityus.png} & \em{Tityus} & Genus & Terrestrial & Scorpions & Arthropoda & Buthidae\\
@@
 \includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/dipsadidae.png} & Dipsadidae & Family & Terrestrial & Snakes & Chordata & Dipsadidae\\
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/harengula.png} & \em{Harengula} & Genus & Marine & Herrings & Chordata & Clupeidae\\
+\includegraphics[valign=T, scale=0.18, raise= 2mm]{input/phylopics/converted/harengula.png} & \em{Harengula} & Genus & Marine & Herrings & Chordata & Clupeidae\\
 \hline
@@
 \includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/harpalyce.png} & \em{Harpalyce} & Genus & Terrestrial & NA & Angiosperms & Fabaceae\\
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/iridaceae.png} & Iridaceae & Family & Terrestrial & NA & Angiosperms & Iridaceae\\
+\includegraphics[valign=T, scale=0.075, raise= 2mm]{input/phylopics/converted/iridaceae.png} & Iridaceae & Family & Terrestrial & NA & Angiosperms & Iridaceae\\
 \includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/lepismium.png} & \em{Lepismium} & Genus & Terrestrial & Cacti & Angiosperms & Cactaceae\\
@@
 \includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/tillandsia.png} & \em{Tillandsia} & Genus & Terrestrial & Bromeliads & Angiosperms & Bromeliaceae\\
-\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/tocoyena.png} & \em{Tocoyena} & Genus & Terrestrial & NA & Angiosperms & Rubiaceae\\
+\includegraphics[valign=T, scale=0.2, raise= 2mm]{input/phylopics/converted/tocoyena.png} & \em{Tocoyena} & Genus & Terrestrial & NA & Angiosperms & Rubiaceae\\
 \bottomrule
@@
 Urban areas & Unfit & Gazetteer-based & Records from urban areas are not necessarily errors, but often represent imprecise records automatically geo-referenced from vague locality descriptions or old records from different land-use types\\
-\rowcolor{gray!6}  Basis of record & Unfit & Meta-data & Records might be unsuitable or unreliable for certain analyses dependent on their source, e.g. "fossil" or "unknown"\\
+\rowcolor{gray!6}  Basis of record & Unfit & Meta-data & Records might be unsuitable or unreliable for certain analyses dependent on their source, e.g. 'fossil' or 'unknown'\\
 Collection year & Unfit & Meta-data & Coordinates from old records are more likely to be imprecise or erroneous coordinates since they are derived from  geo-referencing based on the locality description.
@@
 Only tests that flagged at least 0.1\% of the records in any group are shown.
-Individual records can be flagged by multiple tests, therefore the sum of percentage from all tests can supersede the total percentage.}
+Individual records can be flagged by multiple tests, therefore the sum of percentages from all tests can supersede the total percentage.}
 \centering
@@
 \caption{\label{tab:unnamed-chunk-3}Conservation assessment for 11 Neotropical taxa of plants and animals based on three datasets.
-IUCN: global red list assessment obtained from www.iucn.org; GBIF Raw: Preliminary conservation assessment based on IUCN Criterion B using ConR and the raw dataset from GBIF; GBIF filtered: Preliminary conservation assessment based on IUCN Criterion B using ConR and the filtered dataset.
-Only taxa with at least on species evaluated by IUCN shown.}
+IUCN: global red list assessment obtained from www.iucn.org; GBIF Raw: preliminary conservation assessment based on IUCN Criterion B using ConR and the raw dataset from GBIF; GBIF filtered: preliminary conservation assessment based on IUCN Criterion B using ConR and the filtered dataset.
+Only taxa with at least one species evaluated by IUCN shown.}
 \centering

### 2020-08-04 manuscript_peerj.Rmd
2nd resubmission

@@
 We find that a total of 44.3% of the records are potentially problematic, with large variation across taxonomic groups (25 - 90%).
-A small fraction of records was identified as erroneous in the strict sense (4.2%), and a much larger proportion as unfit for most downstream analyses (41.7%).
+A small fraction of records are identified as erroneous in the strict sense (4.2%), and a much larger proportion as unfit for most downstream analyses (41.7%).
 Filters of duplicated information, collection year, and basis of record, as well as coordinates in urban areas, or for terrestrial taxa in the sea or marine taxa on land, have the greatest effect.
@@
   - \usepackage[export]{adjustbox}
+  - \usepackage{float}
+  - \floatplacement{figure}{H} 
 nocite: |
@@
 ---
-
-<!-- # Keywords -->
-<!-- Automated cleaning, automated conservation assessment, data quality, GBIF, Neotropics. -->
 
@@
 In particular, the availability of digitized collections from museums and herbaria and citizen science observations has increased drastically over the last few years.
-As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.4 billion geo-referenced occurrence records for species from across the globe and the tree of life.
+As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.5 billion geo-referenced occurrence records for species from across the globe and the tree of life.
 
 A central challenge to the use of these publicly available species occurrence data in research is problematic geographic coordinates, which are either erroneous or unfit for downstream analyses [for instance because they are overly imprecise, @Anderson2016].
-Problems mostly arise because data aggregators such as GBIF integrate records collected with different methodologies in different places at different times--- often without centralized curation and only rudimentary meta-data.
+Problems mostly arise because data aggregators such as GBIF integrate records collected with different methodologies in different places at different times---often without centralized curation and only rudimentary meta-data.
 For instance, problematic coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common [@Maldonado2015; @Yesson2007] and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in cultivation or captivity [@Zizka2019].
@@
 ## Choice of study taxa {-}
-This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 which gathered students and researchers working with different taxonomic groups of animals, fungi, and plants across the Neotropics.
+This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 which gathered students and researchers working with different taxonomic groups of animals, fungi, and plants across the Neotropics (Fig.
+\ref{fig:species}).
 Each participant analysed geographic occurrence data from their taxonomic group of interest and commented on the results for their group.
@@
 We consider GBIF data generally of high quality and use them as a case study because GBIF is the largest, most widely used and taxonomically most comprehensive data source for species occurrence records; however many more exist [e.g., https://bien.nceas.ucsb.edu/bien/, www.fishbase.de or @Guedes2018].
-The variety of existing databases is also the reason why we downloaded data without accounting for GBIF-internal quality testing ("Geospatial issues"), since quality control mechanisms might differ among databases and sometimes quality control might not be carried out (for instance in *de novo* databases compiled by groups of researchers).
+GBIF provides information on the internal consistency of records, among others including information on decimal rounding of coordinates, geographic projection and date validity and geospatial issues.
+Since we specifically aimed to test the effect of user-level filtering we included records flagged with issues by GBIF (this was also the default option).
+Geospatial issues flagged by GBIF only concerned 0.4% of the records used in this study and including them had the added benefit to make our results directly comparable to other databases, which may use different internal consistency checks or none at all.
 
 ##  Automated cleaning {-}
-We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the raw data downloaded from GBIF ("raw", hereafter) using meta-data for those records for which they were available [although meta-data were often missing, @Peterson2018], removing: 1) records with a coordinate precision below 100 km (as this represents the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99.
+We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the data as downloaded from GBIF ("raw", hereafter) using meta-data for those records for which they were available [although meta-data were often missing, @Peterson2018], removing: 1) records with a coordinate precision below 100 km (as this represents the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99.
 Furthermore, we rounded the geographic coordinates to four decimal places and retained only one record per species per location (i.e., test for duplicated records).
@@
 We retrieved a total of `r nrow(dat)` species occurrence records, with a median of `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist() %>%  round(0)` records per study group and `r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per species (Table 3, Appendix 1).
-We obtained most records for `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(taxon) %>%  unlist()` (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs) %>%  unlist()`) and fewest for *`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(taxon) %>%  unlist()`* (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(nrecs) %>%  unlist()`).
-The species with most records was *`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(species)`* (`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs)`).
+We obtained most records for `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% dplyr::select(taxon) %>%  unlist()` (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% dplyr::select(nrecs) %>%  unlist()`) and fewest for *`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% dplyr::select(taxon) %>%  unlist()`* (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% dplyr::select(nrecs) %>%  unlist()`).
+The species with most records was *`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% dplyr::select(species)`* (`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% dplyr::select(nrecs)`).
 
 Our automated tests filtered a total of `r sum(!dat$summary)` records (Fig.
-\ref{fig:total}, erroneous: `r err`, unfit: `r unfit`), with a median of `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per group (erroneous: `r median(tab2$total_error)`%, unfit: `r median(tab2$total_unfit)`%).
-Overall, the most important test was for duplicated records (on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!.dpl)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per taxonomic group).
-The filtering steps based on record meta-data that filtered the largest number of records were the basis of records (`r dat %>% group_by(taxon) %>% summarize(num = mean(!coordinate_base)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the collection year (`r dat %>% group_by(taxon) %>% summarize(num = mean(!record_age)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%).
-The most important automated tests were for urban area (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.urb)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the occurrence from records of terrestrial taxa in the sea and marine taxa on land (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.sea)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%, see Table 3 and Appendix 1 in the electronic supplement for further details and the absolute numbers).
+\ref{fig:total}, erroneous: `r err`, unfit: `r unfit`), with a median of `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  dplyr::select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per group (erroneous: `r median(tab2$total_error)`%, unfit: `r median(tab2$total_unfit)`%).
+Overall, the most important test was for duplicated records (on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!.dpl)) %>%  dplyr::select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per taxonomic group).
+The filtering steps based on record meta-data that filtered the largest number of records were the basis of records (`r dat %>% group_by(taxon) %>% summarize(num = mean(!coordinate_base)) %>%  dplyr::select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the collection year (`r dat %>% group_by(taxon) %>% summarize(num = mean(!record_age)) %>%  dplyr::select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%).
+The most important automated tests were for urban area (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.urb)) %>%  dplyr::select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the occurrence from records of terrestrial taxa in the sea and marine taxa on land (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.sea)) %>%  dplyr::select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%, see Table 3 and Appendix 1 in the electronic supplement for further details and the absolute numbers).
 Only a few records were filtered by the coordinate precision, zero coordinates and biodiversity institution tests (Fig.
@@
 
-`r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(1) %>% unlist()`, `r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(2) %>% unlist()`, and `r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(3) %>% unlist()` had the highest fraction of filtered records (Table 3).
+`r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% dplyr::select(taxon) %>% slice(1) %>% unlist()`, `r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% dplyr::select(taxon) %>% slice(2) %>% unlist()`, and `r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% dplyr::select(taxon) %>% slice(3) %>% unlist()` had the highest fraction of filtered records (Table 3).
 In general, the different filters we tested were of similar importance for different study groups.
@@
 We found IUCN assessments for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>% nrow()` species that were also included in our distribution data from 11 of our study groups (Table 4, Appendix 3).
-The fraction of species evaluated varied among the study group, with a maximum of `r tab3 %>% select(iucn_frac_eval) %>% unlist() %>%  max()`% for *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>% slice(1L) %>% select(taxon) %>%  unlist()`* and *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>%  select(taxon) %>% slice(2) %>% unlist()`* and a minimum of `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(iucn_frac_eval) %>%  unlist()`% for `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(taxon) %>%  unlist()` (note that the number of total species varied considerably among groups).
-The median percentage of species per study group with an IUCN assessment was `r tab3 %>% select(iucn_frac_eval) %>% unlist() %>%  median()`%.
-A total of `r aa %>% filter(!is.na(iucn_category)) %>%  filter(iucn_category != "DD") %>%filter(iucn_category != "EX") %>%  mutate(test = iucn_status == "possibly_threatened") %>% select(test) %>% unlist() %>%  sum()` species were listed as *Threatened* by the IUCN global Red List (CR = `r aa %>% filter(iucn_category == "CR") %>% nrow()`, EN = `r aa %>% filter(iucn_category == "EN") %>% nrow()`, VU = `r aa %>% filter(iucn_category == "VU") %>% nrow()`) and `r  aa %>% filter(!is.na(iucn_category)) %>%  filter(iucn_category != "DD") %>%filter(iucn_category != "EX") %>%  mutate(test = iucn_status == "not_threatened") %>% select(test) %>% unlist() %>%  sum()` as *Not Threatened*.
+The fraction of species evaluated varied among the study group, with a maximum of `r tab3 %>% dplyr::select(iucn_frac_eval) %>% unlist() %>%  max()`% for *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>% slice(1L) %>% dplyr::select(taxon) %>%  unlist()`* and *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>%  dplyr::select(taxon) %>% slice(2) %>% unlist()`* and a minimum of `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  dplyr::select(iucn_frac_eval) %>%  unlist()`% for `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  dplyr::select(taxon) %>%  unlist()` (note that the number of total species varied considerably among groups).
+The median percentage of species per study group with an IUCN assessment was `r tab3 %>% dplyr::select(iucn_frac_eval) %>% unlist() %>%  median()`%.
+A total of `r aa %>% filter(!is.na(iucn_category)) %>%  filter(iucn_category != "DD") %>%filter(iucn_category != "EX") %>%  mutate(test = iucn_status == "possibly_threatened") %>% dplyr::select(test) %>% unlist() %>%  sum()` species were listed as *Threatened* by the IUCN global Red List (CR = `r aa %>% filter(iucn_category == "CR") %>% nrow()`, EN = `r aa %>% filter(iucn_category == "EN") %>% nrow()`, VU = `r aa %>% filter(iucn_category == "VU") %>% nrow()`) and `r  aa %>% filter(!is.na(iucn_category)) %>%  filter(iucn_category != "DD") %>%filter(iucn_category != "EX") %>%  mutate(test = iucn_status == "not_threatened") %>% dplyr::select(test) %>% unlist() %>%  sum()` as *Not Threatened*.
 
-We obtained automated conservation assessments for `r aa %>% select(flag_Category_CriteriaB) %>% filter(!is.na(flag_Category_CriteriaB)) %>% nrow()` species in the filtered dataset.
+We obtained automated conservation assessments for `r aa %>% dplyr::select(flag_Category_CriteriaB) %>% filter(!is.na(flag_Category_CriteriaB)) %>% nrow()` species in the filtered dataset.
 Based on the filtered data, the automated conservation assessment evaluated `r sum(aa$flag_status == "possibly_threatened", na.rm = TRUE)` species as possibly threatened (`r round(sum(aa$flag_status == "possibly_threatened", na.rm=TRUE) / aa %>% filter(!is.na(flag_status)) %>% nrow() * 100, 1)`%, CR = `r aa %>% filter(flag_Category_CriteriaB == "CR") %>% nrow()`, EN = `r aa %>% filter(flag_Category_CriteriaB == "EN") %>% nrow()`, VU = `r aa %>% filter(flag_Category_CriteriaB == "VU") %>% nrow()`, see Appendix 3 for assessments of all species).
-The automated assessment based on the filtered dataset agreed with the IUCN assessment for identifying species as possibly threatened (CR, EN, VU) for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status)) %>% mutate(agreement = iucn_status == flag_status) %>% select(agreement) %>% unlist() %>% sum() ` species (`r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status))  %>% mutate(agreement = iucn_status == flag_status) %>% select(agreement) %>% unlist() %>% mean() %>%  round(3) * 100`%; Table 4).
-Filtering reduced the EOO by `r tab3 %>%  select(flag_median_EOO_change) %>%  unlist() %>%  median() %>%  abs()`% and the AOO by `r tab3 %>%  select(flag_median_AOO_change) %>%  unlist() %>%  median() %>% abs()`% on median per group.
-For the raw dataset the agreement with IUCN was higher at `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% mutate(agreement = iucn_status == raw_status) %>% select(agreement) %>% unlist() %>% sum() ` species (`r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% mutate(agreement = iucn_status == raw_status) %>% select(agreement) %>% unlist() %>% mean() %>%  round(3) * 100`%).
+The automated assessment based on the filtered dataset agreed with the IUCN assessment for identifying species as possibly threatened (CR, EN, VU) for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status)) %>% mutate(agreement = iucn_status == flag_status) %>% dplyr::select(agreement) %>% unlist() %>% sum() ` species (`r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status))  %>% mutate(agreement = iucn_status == flag_status) %>% dplyr::select(agreement) %>% unlist() %>% mean() %>%  round(3) * 100`%; Table 4).
+Filtering reduced the EOO by `r tab3 %>%  dplyr::select(flag_median_EOO_change) %>%  unlist() %>%  median() %>%  abs()`% and the AOO by `r tab3 %>%  dplyr::select(flag_median_AOO_change) %>%  unlist() %>%  median() %>% abs()`% on median per group.
+For the raw dataset the agreement with IUCN was higher at `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% mutate(agreement = iucn_status == raw_status) %>% dplyr::select(agreement) %>% unlist() %>% sum() ` species (`r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% mutate(agreement = iucn_status == raw_status) %>% dplyr::select(agreement) %>% unlist() %>% mean() %>%  round(3) * 100`%).
 
@@
 <!-- add some subheadings to discussion -->
-Automated flagging based on meta-data and automatic tests filtered on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% of the records per taxonomic group; `r min(tab2$total_unfit)`%-`r max(tab2$total_unfit)`% as "unfit" and `r min(tab2$total_error)`%-`r max(tab2$total_error)`% as "erroneous".
+Automated flagging based on meta-data and automatic tests filtered on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  dplyr::select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% of the records per taxonomic group; `r min(tab2$total_unfit)`%-`r max(tab2$total_unfit)`% as "unfit" and `r min(tab2$total_error)`%-`r max(tab2$total_error)`% as "erroneous".
 The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (**Question 1**).
@@
 Overall, we consider the effect of the automated filters as positive since they identified the above-mentioned issues and increased the data precision and reduced computational burden (Table 3, Appendix 2).
-However, in some cases filters failed to remove major issues, often due to incomplete metadata.
-For instance, for Diogenidae we found at least two records of an extinct species (*Paguristes mexicanus*) which slipped the "basis of record" test because they were marked as "preserved specimen" rather than "fossil specimen".
+However, in some cases filters failed to remove major issues, often due to incomplete meta-data.
+For instance, for Diogenidae we found at least two records of an species known only from Eocene fossils (*Paguristes mexicanus*) which slipped the "basis of record" test because they were marked as "preserved specimen" rather than "fossil specimen".
 Furthermore, for Entomobryidae we found that for `r dat %>% filter(taxon == "Entomobryidae") %>%  filter(!is.na(species)) %>%  filter(taxonRank == "UNRANKED") %>%  nrow()` records the meta-data on taxonomic rank was "UNRANKED" despite all of them being identified to species level, leading to a high fraction of records removed by the "Identification level" filter.
@@
 The filters we included in this study address a set of important but relatively easy to identify problems.
-In fact, the internal quality control of GBIF does flag some of the problems we tested for (i.e., zero coordinates, equal lat/lon) while others might be implemented in the near future (country centroids, https://data-blog.gbif.org/post/country-centroids/, https://data-blog.gbif.org/post/country-centroids/).
+In fact, the internal quality control of GBIF does flag some of the problems we tested for (i.e., zero coordinates, equal lat/lon) while others might be implemented in the near future (country centroids, https://data-blog.gbif.org/post/country-centroids/).
 While this internal quality is very helpful, we see a huge potential to overcome issues with data quality in a user-feedback system that allows users to provide expert assessments, i.e.
@@
 
-As a next step for automated filtering, it is possible to detect additional problematic records with the help of external species distribution information, which is available for some taxa, for example from www.iucn.org (vertebrates) or https://wcsp.science.kew.org/ (selected seed plant families).
+As next steps for automated filtering, tests for intrinsic consistency and support by external data (if available) can help to detect additional problematic records.
+For instance, testing if records' coordinates fall within the state or province of collection noted for a record (intrinsic) or testing if agree with external species distribution information, for example from www.iucn.org (vertebrates) or https://wcsp.science.kew.org/ (selected seed plant families; extrinsic) can further corroborate the accuracy of a record's geographic referencing.
+If such tests are included it is essential to account for the sampling year, in particular for older records, since the names of political entities may change and the ranges of species may shift.
 Furthermore, while in this study we focused on meta-data and geographic filtering, taxonomic cleaning---the resolution of synonymies and identification of accepted names---is another important part of data curation, but depends on taxon-specific taxonomic backbones and synonymy lists which are not readily available for many groups and often are contradictory within individual taxa.
@@
 # Conclusions {-}
-Our results suggest that between one to half of the occurrence records obtained from GBIF might be unsuitable for downstream biodiversity analyses.
+Our results suggest that between one quarter to half of the occurrence records obtained from GBIF might be unsuitable for downstream biodiversity analyses.
 While the majority of these records might not be erroneous *per se*, they are overly imprecise and thereby increase uncertainty of downstream results or add computational burden on big data analyses.
@@
 Automated filtering can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the specific taxonomic groups and geographic area of interest.
-The putative problems we encountered point to the importance to train researchers and students to curate species occurrence datasets and to allow users to provide data aggregators such as GBIF with feedback for particular records, thus contributing to overall accuracy and precision of public biodiversity databases.
+The putative problems we encountered point to the importance to train researchers and students to curate species occurrence datasets and to visibly associate user-feedback with individual records on aggregator platforms such as GBIF so that it can contribute to the overall accuracy and precision of public biodiversity databases.
 
@@
 We thank GBIF and all data collectors and contributors for their excellent work.
-We thank Town Peterson, Roderic Page and one anonymous reviewer for the helpful comments on an earlier version of this manuscript.
+We thank Town Peterson, Roderic Page and two anonymous reviewers for the helpful comments on an earlier version of this manuscript.
 This study enrolled participants of the workshop "Biodiversity data: from field to yield" led by Alice Calvente, Fernanda Carvalho, Alexander Zizka, and Alexandre Antonelli through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI.
@@
 
+\newpage{}
+
 # Tables {-}
+\newpage{}
 <!-- A table presenting the study taxa -->
@@
 The outlines illustrate the broad taxonomic group.
-Pictures from www.phylopic.org.
-Phylopic credits: Diogenidae (Ekaterina Kopeykina (vectorized by T.
-Michael Keesey), Creative Commons Attribution-NonCommercial-ShareAlike 3.0; Entomobryidae and Neanuridae: Birgit Lang, Creative Commons Attribution 3.0; Tityus: Gareth Monger, Creative Commons Attribution 3.0; \\textit{Gaylussacia}, \\textit{Lepismium}, \\textit{Pilosocereus}: T.
-Michael Keesey (vectorization) and Nadiatalent (photography), Creative Commons Attribution-ShareAlike 3.0; \\textit{Iridaceae} adapted from Goldblatt and Manning (2008); \\textit{Oocephalus}: Alexander Schmidt-Lebhuhn, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; \\textit{Tillandsia}: Maija Karala, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; all others in the public domain.")%>%
+Pictures from www.phylopic.org if available under a Public Domain Dedication license, otherwise generated from photographs available in the public domain.")%>%
   #landscape()%>%
@@
 ```{r species, out.width="0.9\\linewidth", fig.cap=c('Examples of taxa included in this study.
-\\textbf{A)} \\textit{Pilosocereus pusillibaccatus} (\\textit{Pilosocereus}), \\textbf{B)} \\textit{Conchocarpus macrocarpus} (\\textit{Conchocarpus}); \\textbf{C)} \\textit{Tillandsia recurva} (\\textit{Tillandsia}); \\textbf{D)} \\textit{Oxyrhopus guibei} (Dipsadidae); \\textbf{E)} \\textit{Aethiopella ricardoi} (Neanuridae); \\textbf{F)} \\textit{Tocoyena formosa} (\\textit{Tocoyena}); \\textbf{G)} \\textit{Harengula jaguana} (\\textit{Harengula}); \\textbf{H)} \\textit{Gaylussacia decipiens} (\\textit{Gaylussacia}); \\textbf{I)} \\textit{Oocephalus foliosus} (\\textit{Oocephalus}); \\textbf{J)} \\textit{Tityus carvalhoi} (\\textit{Tityus}); \\textbf{K)} \\textit{Prosthechea vespa} (\\textit{Prosthechea}), Image credits: A) Pamela Lavor, B) Juliana El-Ottra, C) Eduardo Tomaz, D) Filipe C Serrano, E) Raiane Vital da Paz, F) Fernanda GL Moreira, G) Thais Ferreira-Araujo, H) Luiz Menini Neto, I) Arthur Soares, J) Renata C Santos-Costa, K) Tiago Vieira.'), echo=FALSE}
-knitr::include_graphics(path = "./figures_tables/Figure_study_groups_reduced.jpg")
+\\textbf{A)} \\textit{Pilosocereus pusillibaccatus} (\\textit{Pilosocereus}), \\textbf{B)} \\textit{Conchocarpus macrocarpus} (\\textit{Conchocarpus}); \\textbf{C)} \\textit{Tillandsia recurva} (\\textit{Tillandsia}); \\textbf{D)} \\textit{Oxyrhopus guibei} (Dipsadidae); \\textbf{E)} \\textit{Aethiopella ricardoi} (Neanuridae); \\textbf{F)} \\textit{Tocoyena formosa} (\\textit{Tocoyena}); \\textbf{G)} \\textit{Harengula jaguana} (\\textit{Harengula}); \\textbf{H)} \\textit{Gaylussacia decipiens} (\\textit{Gaylussacia}); \\textbf{I)} \\textit{Oocephalus foliosus} (\\textit{Oocephalus}); \\textbf{J)} \\textit{Tityus carvalhoi} (\\textit{Tityus}); \\textbf{K)} \\textit{Prosthechea vespa} (\\textit{Prosthechea}), Image credits: A) Pamela Lavor, B) Juliana El-Ottra, C) Eduardo Calisto Tomaz, D) Filipe C Serrano, E) Raiane Vital da Paz, F) Fernanda GL Moreira, G) Thais Ferreira-Araujo, H) Luiz Menini Neto, I) Arthur de Souza Soares, J) Renata C Santos-Costa, K) Tiago Vieira.'), echo=FALSE}
+knitr::include_graphics(path = "./figures_tables/Fig1_study_groups.png")
 ```
@@
 <!-- A map with number of species and where they have been removed -->
-```{r total, out.width="\\linewidth", fig.cap=c('The difference in records available between the raw and filtered dataset of 18 Neotropical taxa including animals, fungi, and plants, plotted in a 100 x 100 km grid across the Neotropics.'), echo=FALSE}
-knitr::include_graphics(path = "./output/figure_number_of_records.jpg")
+```{r total, out.width="\\linewidth", fig.cap=c('The absolute number of records flagged as erroneous or unfit by automated geographic filters in a dataset of 18 Neotropical taxa including animals, fungi, and plants, plotted in a 100 x 100 km grid across the Neotropics (Behrmann projection).'), echo=FALSE}
+knitr::include_graphics(path = "./figures_tables/Fig2_number_of_records.png")
 ```
@@
 Only filters that flagged at least 0.1\\% of records in any taxon are shown.'), echo=FALSE}
-knitr::include_graphics(path = "./output/figure_number_of_records_split.jpg")
+knitr::include_graphics(path = "./figures_tables/Fig3_number_of_records_split.png")
 ```
 
-```{r speciesrichness, out.width="\\linewidth", fig.cap=c('Illustrative examples for the difference in species richness between the raw and filtered dataset (raw - filtered) from four of the study taxa.
+```{r speciesrichness, out.width="\\linewidth", fig.cap=c('Illustrative examples of the difference in species richness between the raw and filtered dataset (raw - filtered) from four of the study taxa.
+Total species number in the raw data sets: Dipsadidae: 637, \\textit{Harengula}: 4, \\textit{Thozetella}: 9, \\textit{Tillandsia} 464.
 Photo credits for C) by Tiago Andrade Borges Santos, otherwise as in Figure 1.'), echo=FALSE}
-knitr::include_graphics(path = "./figures_tables/figure_number_species_richness_difference.jpg")
+knitr::include_graphics(path = "./figures_tables/Fig4_number_species_richness_difference.png")
 ```

### 2020-08-04 manuscript_peerj.tex
2nd resubmission

@@
 \usepackage[export]{adjustbox}
+\usepackage{float}
+\floatplacement{figure}{H}
 \usepackage{booktabs}
@@
 In particular, the availability of digitized collections from museums and herbaria and citizen science observations has increased drastically over the last few years.
-As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.4 billion geo-referenced occurrence records for species from across the globe and the tree of life.
+As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.5 billion geo-referenced occurrence records for species from across the globe and the tree of life.
 
@@
 2016).
-Problems mostly arise because data aggregators such as GBIF integrate records collected with different methodologies in different places at different times--- often without centralized curation and only rudimentary meta-data.
+Problems mostly arise because data aggregators such as GBIF integrate records collected with different methodologies in different places at different times---often without centralized curation and only rudimentary meta-data.
 For instance, problematic coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common (Maldonado et al.
@@
 
-This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 which gathered students and researchers working with different taxonomic groups of animals, fungi, and plants across the Neotropics.
+This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 which gathered students and researchers working with different taxonomic groups of animals, fungi, and plants across the Neotropics (Fig.
+\ref{fig:species}).
 Each participant analysed geographic occurrence data from their taxonomic group of interest and commented on the results for their group.
@@
 2018).
-The variety of existing databases is also the reason why we downloaded data without accounting for GBIF-internal quality testing (``Geospatial issues''), since quality control mechanisms might differ among databases and sometimes quality control might not be carried out (for instance in \emph{de novo} databases compiled by groups of researchers).
+GBIF provides information on the internal consistency of records, among others including information on decimal rounding of coordinates, geographic projection and date validity and geospatial issues (including the zero coordinates test used in this study).
+Since we specifically aimed to test the effect of user-level filtering we included records flagged with issues by GBIF (this was also the default option).
+Geospatial issues flagged by GBIF only concerned 0.4\% of the records used in this study and including them had the added benefit to make our results directly comparable to other databases, which may use different internal consistency checks or none at all.
 
@@
 We followed the cleaning pipeline outlined by Zizka et al.
-(2019) and first filtered the raw data downloaded from GBIF (``raw'', hereafter) using meta-data for those records for which they were available (although meta-data were often missing, Peterson et al.
+(2019) and first filtered the data as downloaded from GBIF (``raw'', hereafter) using meta-data for those records for which they were available (although meta-data were often missing, Peterson et al.
 2018), removing: 1) records with a coordinate precision below 100 km (as this represents the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99.
@@
 Overall, we consider the effect of the automated filters as positive since they identified the above-mentioned issues and increased the data precision and reduced computational burden (Table 3, Appendix 2).
-However, in some cases filters failed to remove major issues, often due to incomplete metadata.
-For instance, for Diogenidae we found at least two records of an extinct species (\emph{Paguristes mexicanus}) which slipped the ``basis of record'' test because they were marked as ``preserved specimen'' rather than ``fossil specimen''.
+However, in some cases filters failed to remove major issues, often due to incomplete meta-data.
+For instance, for Diogenidae we found at least two records of an species known only from Eocene fossils (\emph{Paguristes mexicanus}) which slipped the ``basis of record'' test because they were marked as ``preserved specimen'' rather than ``fossil specimen''.
 Furthermore, for Entomobryidae we found that for 1,996 records the meta-data on taxonomic rank was ``UNRANKED'' despite all of them being identified to species level, leading to a high fraction of records removed by the ``Identification level'' filter.
@@
 The filters we included in this study address a set of important but relatively easy to identify problems.
-In fact, the internal quality control of GBIF does flag some of the problems we tested for (i.e., zero coordinates, equal lat/lon) while others might be implemented in the near future (country centroids, \url{https://data-blog.gbif.org/post/country-centroids/}, \url{https://data-blog.gbif.org/post/country-centroids/}).
+In fact, the internal quality control of GBIF does flag some of the problems we tested for (i.e., zero coordinates, equal lat/lon) while others might be implemented in the near future (country centroids, \url{https://data-blog.gbif.org/post/country-centroids/}).
 While this internal quality is very helpful, we see a huge potential to overcome issues with data quality in a user-feedback system that allows users to provide expert assessments, i.e.~a meta-annotation of records being challenged (and why).
@@
 
-As a next step for automated filtering, it is possible to detect additional problematic records with the help of external species distribution information, which is available for some taxa, for example from www.iucn.org (vertebrates) or \url{https://wcsp.science.kew.org/} (selected seed plant families).
+As next steps for automated filtering, tests for intrinsic consistency and support by external data (if available) can help to detect additional problematic records.
+For instance, testing if records' coordinates fall within the state or province of collection noted for a record (intrinsic) or agree with external species distribution information, for example from www.iucn.org (vertebrates) or \url{https://wcsp.science.kew.org/} (selected seed plant families; extrinsic) can further corroborate the accuracy of a record's geographic referencing.
+If such tests are included it is essential to account for the sampling year, in particular for older records, since the names of provinces may change and the ranges of species may shift.
 Furthermore, while in this study we focused on meta-data and geographic filtering, taxonomic cleaning---the resolution of synonymies and identification of accepted names---is another important part of data curation, but depends on taxon-specific taxonomic backbones and synonymy lists which are not readily available for many groups and often are contradictory within individual taxa.
@@
 
-Our results suggest that between one to half of the occurrence records obtained from GBIF might be unsuitable for downstream biodiversity analyses.
+Our results suggest that between one quarter to half of the occurrence records obtained from GBIF might be unsuitable for downstream biodiversity analyses.
 While the majority of these records might not be erroneous \emph{per se}, they are overly imprecise and thereby increase uncertainty of downstream results or add computational burden on big data analyses.
@@
 
+\newpage{}
+
 \hypertarget{tables}{%
@@
 \addcontentsline{toc}{section}{Tables}
+
+\newpage{}
 
@@
 The outlines illustrate the broad taxonomic group.
-Pictures from www.phylopic.org.
-Phylopic credits: Diogenidae (Ekaterina Kopeykina (vectorized by T.
-Michael Keesey), Creative Commons Attribution-NonCommercial-ShareAlike 3.0; Entomobryidae and Neanuridae: Birgit Lang, Creative Commons Attribution 3.0; Tityus: Gareth Monger, Creative Commons Attribution 3.0; \textit{Gaylussacia}, \textit{Lepismium}, \textit{Pilosocereus}: T.
-Michael Keesey (vectorization) and Nadiatalent (photography), Creative Commons Attribution-ShareAlike 3.0; \textit{Iridaceae} adapted from Goldblatt and Manning (2008); \textit{Oocephalus}: Alexander Schmidt-Lebhuhn, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; \textit{Tillandsia}: Maija Karala, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; all others in the public domain.}
+Pictures from www.phylopic.org if available under a Public Domain Dedication license, otherwise generated from photographs available in the public domain.}
 \centering
@@
 \rowcolor{gray!6}  Biodiversity institutions & Error & Gazetteer-based & Records may have coordinates at the location of biodiversity institutions, e.g.
-because they were erroneously entered with the physical location of the specimen or because they represent individuals from captivity or horticulture\\
+because they were erroneously entered with the physical location of the specimen or because they represent individuals from captivity or horticulture, which are not clearly labeled as such\\
 Equal lat/lon & Error & Gazetteer-based & Coordinates with equal latitude and longitude are usually indicative of data entry errors\\
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{./figures_tables/Figure_study_groups_reduced} \caption{Examples of taxa included in this study.
-\textbf{A)} \textit{Pilosocereus pusillibaccatus} (\textit{Pilosocereus}), \textbf{B)} \textit{Conchocarpus macrocarpus} (\textit{Conchocarpus}); \textbf{C)} \textit{Tillandsia recurva} (\textit{Tillandsia}); \textbf{D)} \textit{Oxyrhopus guibei} (Dipsadidae); \textbf{E)} \textit{Aethiopella ricardoi} (Neanuridae); \textbf{F)} \textit{Tocoyena formosa} (\textit{Tocoyena}); \textbf{G)} \textit{Harengula jaguana} (\textit{Harengula}); \textbf{H)} \textit{Gaylussacia decipiens} (\textit{Gaylussacia}); \textbf{I)} \textit{Oocephalus foliosus} (\textit{Oocephalus}); \textbf{J)} \textit{Tityus carvalhoi} (\textit{Tityus}); \textbf{K)} \textit{Prosthechea vespa} (\textit{Prosthechea}), Image credits: A) Pamela Lavor, B) Juliana El-Ottra, C) Eduardo Tomaz, D) Filipe C Serrano, E) Raiane Vital da Paz, F) Fernanda GL Moreira, G) Thais Ferreira-Araujo, H) Luiz Menini Neto, I) Arthur Soares, J) Renata C Santos-Costa, K) Tiago Vieira.}\label{fig:species}
+\includegraphics[width=0.9\linewidth]{./figures_tables/Fig1_study_groups} \caption{Examples of taxa included in this study.
+\textbf{A)} \textit{Pilosocereus pusillibaccatus} (\textit{Pilosocereus}), \textbf{B)} \textit{Conchocarpus macrocarpus} (\textit{Conchocarpus}); \textbf{C)} \textit{Tillandsia recurva} (\textit{Tillandsia}); \textbf{D)} \textit{Oxyrhopus guibei} (Dipsadidae); \textbf{E)} \textit{Aethiopella ricardoi} (Neanuridae); \textbf{F)} \textit{Tocoyena formosa} (\textit{Tocoyena}); \textbf{G)} \textit{Harengula jaguana} (\textit{Harengula}); \textbf{H)} \textit{Gaylussacia decipiens} (\textit{Gaylussacia}); \textbf{I)} \textit{Oocephalus foliosus} (\textit{Oocephalus}); \textbf{J)} \textit{Tityus carvalhoi} (\textit{Tityus}); \textbf{K)} \textit{Prosthechea vespa} (\textit{Prosthechea}), Image credits: A) Pamela Lavor, B) Juliana El-Ottra, C) Eduardo Calisto Tomaz, D) Filipe C Serrano, E) Raiane Vital da Paz, F) Fernanda GL Moreira, G) Thais Ferreira-Araujo, H) Luiz Menini Neto, I) Arthur de Souza Soares, J) Renata C Santos-Costa, K) Tiago Vieira.}\label{fig:species}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\linewidth]{./output/figure_number_of_records} \caption{The difference in records available between the raw and filtered dataset of 18 Neotropical taxa including animals, fungi, and plants, plotted in a 100 x 100 km grid across the Neotropics.}\label{fig:total}
+\includegraphics[width=\linewidth]{./figures_tables/Fig2_number_of_records} \caption{The absolute number of records flagged as erroneous or unfit by automated geographic filters in dataset of 18 Neotropical taxa including animals, fungi, and plants, plotted in a 100 x 100 km grid across the Neotropics (Behrmann projection).}\label{fig:total}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\linewidth]{./output/figure_number_of_records_split} \caption{Geographic location of the occurrence records flagged by the automated tests applied in this study.
+\includegraphics[width=\linewidth]{./figures_tables/Fig3_number_of_records_split} \caption{Geographic location of the occurrence records flagged by the automated tests applied in this study.
 Only filters that flagged at least 0.1\% of records in any taxon are shown.}\label{fig:split}
@@
 \begin{figure}
-\includegraphics[width=\linewidth]{./figures_tables/figure_number_species_richness_difference} \caption{Illustrative examples for the difference in species richness between the raw and filtered dataset (raw - filtered) from four of the study taxa.
+\includegraphics[width=\linewidth]{./figures_tables/Fig4_number_species_richness_difference} \caption{Illustrative examples for the difference in species richness between the raw and filtered dataset (raw - filtered) from four of the study taxa.
+Total species number in the raw data sets: Dipsadidae: 637, \textit{Harengula}: 4, \textit{Thozetella}: 9, \textit{Tillandsia} 464.
 Photo credits for C) by Tiago Andrade Borges Santos, otherwise as in Figure 1.}\label{fig:speciesrichness}

### 2020-08-04 02_automated_cleaning.R

@@ -15,16 +15,29 @@ for(i in 1:length(li)){
   sub <- read_delim(paste("input/gbif/", li[i], "/occurrence.txt", sep = ""), delim = "\t", quote = "") %>% 
     mutate_all(as.character)
   if(nrow(sub) > 0){
+    quant <- bind_rows(quant, sub[,c("organismQuantity", "organismQuantityType", "occurrenceStatus" )])
     sub$taxon <- str_split(li[i], pattern = "_")[[1]][1]
     sub <- dplyr::select(sub, taxon, decimalLongitude, decimalLatitude,
                   coordinateUncertaintyInMeters, 
                   countryCode, gbifID, species, class, family, genus, taxonRank,
-                  year, basisOfRecord, individualCount)
+                  year, basisOfRecord, individualCount, organismQuantity, organismQuantityType, occurrenceStatus, hasGeospatialIssues)
     dat <- bind_rows(dat,sub)
+    
   }
 }
 
+# get the number of records with the columns needed to disambiguate the IndividualCount = 0 issue
+sum(!is.na(dat$organismQuantity)) / nrow(dat)
+sum(!is.na(dat$organismQuantityType)) / nrow(dat)
+sum(!is.na(dat$occurrenceStatus)) / nrow(dat)
 
+test <- dat %>% filter((individualCount == 0))
+sum(!is.na(test$organismQuantity)) / nrow(test)
+sum(!is.na(test$organismQuantityType)) / nrow(test)
+sum(!is.na(test$occurrenceStatus)) / nrow(test)
+
+dat <- dat %>% 
+  select(-organismQuantity, -organismQuantityType, -occurrenceStatus, -hasGeospatialIssues)
 
 # meta data filtering
 dat <- dat %>% 

### 2020-08-04 04_figure_number_of_occurrences.R

@@ -65,18 +65,28 @@ world.countries <- rnaturalearth::ne_countries(type = 'countries',
                                                scale = 110)
 countries.behr <- spTransform(world.countries, CRS(behr)) %>% fortify()
 
-# plot
+# Axis labels
+
+ticks_x <- SpatialPoints(data.frame(c(-120,-100,-80, -60, -40, -20),0), proj4string = CRS(wgs1984))
+ticks_x <-  data.frame(x = coordinates(spTransform(ticks_x, behr))[,1],
+                       long = c(-120,-100,-80, -60, -40, -20))
+
+
+ticks_y <- SpatialPoints(data.frame(0, c(40, 20, 0, -20, -40, -60)), proj4string = CRS(wgs1984))
+ticks_y <-  data.frame(y = coordinates(spTransform(ticks_y, behr))[,2],
+                       long = c(40, 20, 0, -20, -40, -60))# plot
+
 ggplot()+
   geom_polygon(data = countries.behr,
                aes(x = long, y = lat, group = group), 
                fill = "transparent", 
                color = "grey20")+
   geom_tile(data = plo, aes(x = x, y = y, fill = log(layer)), alpha = 0.8)+
-  scale_fill_viridis(name = "Number of\nrecords", direction = -1, na.value = "transparent",
+  scale_fill_viridis(name = "Absolute number\nof flagged records", direction = -1, na.value = "transparent",
                      breaks = c(log(1), log(10), log(100), log(1000), log(5000)),
                      labels = c(1,10,100,1000,5000))+
-  xlim(-12000000, -3000000)+
-  ylim(-6500000, 4500000)+
+  scale_x_continuous(breaks = ticks_x[,1], labels = ticks_x[,2], limits = c(-12000000, -3000000))+
+  scale_y_continuous(breaks = ticks_y[,1], labels = ticks_y[,2], limits = c(-6500000, 4500000))+
   coord_fixed()+
   theme_bw()+
   theme(legend.position = "bottom",
@@ -84,7 +94,7 @@ ggplot()+
         axis.title = element_blank())#+
   # facet_wrap(.~ dataset, ncol = 2)
 
-ggsave("output/figure_number_of_records.jpg", height = 8, width=8)
+ggsave("output/figure_number_of_records.png", height = 8, width=8)
 
 # Figure 2
 plo <- dat %>% 
@@ -158,7 +168,7 @@ ggplot()+
                fill = "transparent", 
                color = "grey50")+
[…]

### 2020-08-04 06_species_richness_maps.R

@@ -82,12 +82,20 @@ plo <- full_join(plo_rw, plo_cl, by = c("taxon", "x", "y")) %>%
 write_csv(plo, "output/specis_richness.csv")
 
 # plots of difference for the main manuscript
-
 ## select illustrative taxa
 
 plo <- plo %>% 
   filter(taxon %in% c("Thozetella", "Tillandsia", "Dipsadidae", "Harengula")) %>% 
   filter(difference != 0)
+
+## get the number of initial species for these focus groups to show in the figure legend
+dat %>% 
+  filter(taxon %in% c("Thozetella", "Tillandsia", "Dipsadidae", "Harengula")) %>%
+  dplyr::select(taxon, species) %>% 
+  distinct() %>% 
+  group_by(taxon) %>% 
+    summarize(species = n())
+
 
 # plot
 ggplot()+
